---
title: "riskutility: Record Linkage Risk Assessment for Perturbative Disclosure Control"
author:
  - name: Matthias Templ
    affiliation: "Institute for Competitiveness and Communication,
      School of Business, FHNW University of Applied Sciences and Arts
      Northwestern Switzerland; Swiss Data Anonymization Center
      (www.swissanon.ch)"
    email: matthias.templ@gmail.com
    orcid: 0000-0002-8638-5276
abstract: >
  The **riskutility** package for R provides a unified framework for
  measuring re-identification risk in perturbed microdata. The
  `recordLinkage()` function implements four linkage methods under a
  single S3 interface --- deterministic (Gower distance), probabilistic
  (Fellegi-Sunter), PRAM (transition matrix), and predictive (propensity
  score) --- with consistent risk weighting, blocking, and
  forward/reverse direction support. While several R packages provide
  record linkage functionality (RecordLinkage, fastLink, reclin2) and
  sdcMicro offers distance-based risk measures, no existing package
  combines all four methods with flexible risk weighting and a focus
  on disclosure risk quantification for perturbative SDC. This paper
  presents the methodology with mathematical derivations, hand-traceable
  toy examples, and realistic demonstrations using sdcMicro for
  perturbation. Practical guidance on method selection, risk
  interpretation, and common pitfalls is provided.
keywords: [record linkage, re-identification risk, statistical disclosure control,
  perturbative methods, PRAM, Fellegi-Sunter, R]
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{riskutility: Record Linkage Risk Assessment for Perturbative Disclosure Control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5,
  warning = FALSE, message = FALSE
)
library(riskutility)
library(sdcMicro)
```

# Introduction

## Perturbative SDC and Record Linkage Risk

Statistical disclosure control (SDC) encompasses a range of techniques for
protecting individual privacy in released microdata
[@hundepool2012statistical]. *Perturbative* methods -- such as PRAM
(Post-Randomization Method), record swapping, noise addition, and
microaggregation -- modify the original records rather than generating
entirely new ones [@templ2015sdcmicro]. This makes perturbative data
fundamentally different from fully synthetic data: every released record
has a real individual behind it, and the attacker's task is to undo the
perturbation and re-identify that individual.

A *quasi-identifier* (QI) is a variable that is not a direct identifier
(such as a name or social security number) but can be used in
combination with other variables to narrow down the identity of a record
holder. Typical quasi-identifiers include age, sex, region, education,
and occupation. The set of quasi-identifiers available to an attacker
determines the scope of a record linkage attack
[@skinner2002measure; @elliot2018functional].

Most disclosure risk functions in **riskutility** (DCAP, TCAP, DCR)
target *synthetic* data, where the question is whether the generator
memorized training records. RAPID can be applied to both synthetic
and traditionally anonymized data (it trains on the released data and
evaluates predictions on the original), but was primarily designed for
the synthetic case. For perturbative data, the more natural question is:
*can an attacker link a released record back to the original
individual?* This is the classical problem of
identity disclosure [@dewolf2002framework; @reiter2005estimating],
and the standard approach to its empirical assessment is record linkage
[@domingoferrer2003disclosure].
The `recordLinkage()` function addresses exactly this question.

## The Attacker Model

The attacker scenario assumes:

- **Membership knowledge**: the attacker knows that the target individual
  has a record in the released dataset.
- **Exact QI knowledge**: the attacker knows the target's quasi-identifier
  values (e.g., age, sex, region) from an external source.
- **Linking rule**: the attacker applies a strategy (nearest match,
  threshold, top-$k$) to select candidates from the released data and
  attributes re-identification risk accordingly.

Ground truth for evaluation is taken from row alignment
(`truth = "row"`) or a shared identifier column (`truth = "id"`).
Note that `truth` affects only the *evaluation* of success; it is not
an attacker capability.

These assumptions represent a *worst-case* scenario. In practice,
attackers rarely have perfect membership knowledge (they may not know
whether the target is in the dataset) or exact QI values (external
sources may be noisy or outdated). The computed risk should therefore
be interpreted as an *upper bound* on the re-identification probability
under the assumed attacker model. For population-level risk assessment
that accounts for unknown membership, see @franconi2004individual and
the `population_uniqueness()` function in **riskutility**.

## Related Software

Several R packages address record linkage or disclosure risk:

- **sdcMicro** [@templ2015sdcmicro] provides `dRisk()` and `dRiskRMD()`
  for distance-based record linkage risk [@templ2008robust], along with
  comprehensive SDC methods. These functions compute risk based on the
  proportion of records closer to their match than to any other record
  (using Euclidean or robust Mahalanobis distances for numeric data),
  but support only the deterministic (distance-based) approach and do
  not offer flexible risk weighting or candidate-set strategies.
- **RecordLinkage** [@sariyar2010recordlinkage] implements
  Fellegi-Sunter probabilistic linkage with EM-based weight estimation,
  designed for deduplication and data integration rather than disclosure
  risk quantification.
- **fastLink** [@enamorado2019fastlink] provides fast probabilistic
  record linkage with blocking and EM estimation, optimized for merging
  large administrative datasets.
- **reclin2** [@reclin2] offers a modular record linkage pipeline with
  blocking, comparison, and scoring stages.

These packages focus primarily on *linking* records (classification into
match/non-match) rather than *quantifying re-identification risk* per
record. The `recordLinkage()` function in **riskutility** fills this gap
by combining four linkage methods with risk-focused output: per-record
risk scores, candidate set diagnostics, and configurable risk weighting.

| Feature | sdcMicro | RecordLinkage | fastLink | riskutility |
|:--------|:--------:|:-------------:|:--------:|:-----------:|
| Deterministic (distance) | Yes | -- | -- | Yes |
| Probabilistic (Fellegi-Sunter) | -- | Yes | Yes | Yes |
| PRAM (transition matrix) | -- | -- | -- | Yes |
| Predictive (propensity score) | -- | -- | -- | Yes |
| Per-record risk scores | Partial[^1] | -- | -- | Yes |
| Risk weighting schemes | -- | -- | -- | Yes |
| Blocking | -- | Yes | Yes | Yes |
| Forward/reverse direction | -- | -- | -- | Yes |

[^1]: sdcMicro's `dRisk()` returns a global proportion of re-identified records rather than per-record continuous risk scores.

## Scope of This Paper

The `recordLinkage()` function in **riskutility** unifies four linkage
methods -- deterministic, probabilistic, PRAM, and predictive -- under a
single S3 interface with consistent risk weighting, blocking, and
direction support.

This paper covers all four methods, each following the same pedagogical
structure: *mathematical formulation* $\to$ *hand-traceable toy example*
$\to$ *realistic demonstration*:

1. **Deterministic** (Gower distance)
2. **Probabilistic** (Fellegi-Sunter log-likelihood ratios)
3. **PRAM** (transition matrix products)
4. **Predictive** (propensity-score-based)

The remaining sections compare methods and weighting schemes
side by side, present computational details of the implementation,
discuss advanced features (blocking, direction, ground truth), and
provide practical guidance for method selection and risk interpretation.

For the theoretical background on record linkage see @fellegi1969theory,
@winkler2006overview, and @domingo2002validating; for empirical
disclosure risk assessment via record linkage see
@domingoferrer2003disclosure; for PRAM see @gouweleeuw1998post; for the
general framework of record linkage risk see @dewolf2002framework.


# Mathematical Framework

## Notation

Let $X = \{x_1, \ldots, x_n\}$ denote the original microdata and
$X^a = \{x^a_1, \ldots, x^a_m\}$ the anonymized (perturbed) data.
In the typical case $m = n$ and record $x^a_i$ is the perturbed
version of $x_i$.

We denote:

- $\mathcal{K} = \{v_1, \ldots, v_p\}$ -- the set of quasi-identifier
  variables (keys) available to the attacker.
- $w_v \ge 0$ -- the weight assigned to variable $v$.
- $d(i, j)$ -- the distance (or dissimilarity) between original record
  $x_i$ and anonymized record $x^a_j$ over the quasi-identifiers
  (QIs).
  For the probabilistic method this becomes a log-likelihood ratio
  $\Lambda_{ij}$.
- $C_i$ -- the candidate set of anonymized records that the attacker
  considers as potential matches for $x_i$.
- $j^*_i$ -- the index of the true match of $x_i$ in the anonymized
  data ($j^*_i = i$ when rows are aligned).

## Re-identification Risk

The re-identification risk quantifies how likely an attacker is to
correctly identify an individual's record in the anonymized data using
only the quasi-identifiers $\mathcal{K}$.

Given a candidate set $C_i$ with weights $\pi_j \ge 0$,
$\sum_{j \in C_i} \pi_j = 1$, the re-identification risk for record $i$
is:
$$r_i = \sum_{j \in C_i} \pi_j \cdot \mathbf{1}[j = j^*_i].$$
When the true match is not in the candidate set, $r_i = 0$. When there
is a single candidate and it is the true match, $r_i = 1$.

The overall risk is summarized by the mean $\bar{r} = n^{-1}\sum_i r_i$,
the maximum $r_{\max}$, and the fraction of records exceeding a threshold
(typically 0.1).

The formulation above takes the *forward* perspective: for each original
record, search the anonymized data for matches. The *reverse*
perspective iterates over anonymized records and searches the original
data, answering "how disclosive is each released record?" For symmetric
distance methods (deterministic, predictive), the two perspectives
yield similar but not identical results because the evaluation criterion
changes (different records serve as query vs. search). For PRAM, the
asymmetric transition matrices cause forward and reverse risks to differ
substantially (Section 5.3).

## Risk Weighting Schemes

The choice of candidate weights $\pi_j$ controls how risk is attributed.
Three schemes are available; their effect is demonstrated empirically in
the method comparison section below.

**Uniform** (`risk_weighting = "uniform"`): Each candidate receives
equal weight:
$$\pi_j = \frac{1}{|C_i|}.$$
This is conservative: if three candidates are equally close and one is
the true match, the risk is $1/3$ regardless of distance.

**Softmax** (`risk_weighting = "softmax"`): Closer candidates receive
exponentially higher weight:
$$\pi_j = \frac{\exp(-\kappa \cdot d_j)}{\sum_{k \in C_i} \exp(-\kappa \cdot d_k)},$$
where $d_j \equiv d(i,j)$ is the distance from query record $i$ to
candidate $j$, and $\kappa > 0$ is a temperature parameter. If not supplied, $\kappa$
is auto-calibrated as $2 / (\max(d) - \min(d))$. As $\kappa \to \infty$
this approaches winner-take-all; as $\kappa \to 0$ it converges to
uniform.

**Kernel** (`risk_weighting = "kernel"`): Weights are proportional to a
kernel function evaluated at the scaled distance:
$$\pi_j = \frac{K(d_j / h)}{\sum_{k \in C_i} K(d_k / h)},$$
where $h$ is the bandwidth and $K$ is a kernel function.
If $h$ is not supplied, it is auto-selected via Silverman's rule of
thumb applied to the candidate distances:
$$h = 0.9 \cdot \min\!\left(\hat\sigma,\, \frac{\text{IQR}}{1.34}\right) \cdot |C_i|^{-1/5}.$$
The available kernels are:

- *Gaussian*: $K(u) = \exp(-u^2/2)$
- *Epanechnikov*: $K(u) = \max(0, 1 - u^2) \cdot 3/4$
- *Tricube*: $K(u) = \max(0, (1 - |u|^3)^3) \cdot 70/81$

With compact kernels (Epanechnikov, tricube), candidates beyond the
bandwidth receive exactly zero weight and are effectively excluded.
The effect of the bandwidth and kernel choice is demonstrated
empirically in the weighting scheme comparison (Section 6.2); readers
unfamiliar with kernel density estimation may skip these details on
first reading.


# Deterministic Method (Gower Distance)

The simplest approach to record linkage treats it as a matching problem:
for each original record, find the most similar record in the anonymized
data and assess whether it is the true match
[@domingoferrer2003disclosure; @domingo2002validating]. This requires no
knowledge of the perturbation mechanism -- only a sensible distance
metric -- making it the natural starting point for risk assessment.

## Weighted Gower Distance

The deterministic method computes a weighted Gower-type distance over the
quasi-identifiers [@gower1971general]. For original record $x_i$ and
anonymized record $x^a_j$:

$$d(i,j) = \frac{\sum_{v \in \mathcal{K}} w_v \, \delta_v(i,j)}{\sum_{v \in \mathcal{K}} w_v},$$

where the variable-specific dissimilarities are:

- **Numeric / ordinal**: $\delta_v(i,j) = |x_{iv} - x^a_{jv}| / R_v$,
  where $R_v = \max(x_v \cup x^a_v) - \min(x_v \cup x^a_v)$ is the
  range computed over both datasets
- **Nominal**: $\delta_v(i,j) = \mathbf{1}[x_{iv} \ne x^a_{jv}]$

The variable weights $w_v$ control the relative importance of each
quasi-identifier in the distance computation. By default, all weights
are equal ($w_v = 1$ for all $v$). Setting a higher weight for a variable
increases its influence on the distance -- for example, setting
$w_{\text{age}} = 2$ would make age twice as influential as variables
with $w_v = 1$. Weights can be specified via the `weights` parameter in
`recordLinkage()`.

The resulting distance $d(i,j) \in [0, 1]$, where 0 indicates perfect
agreement and 1 maximal disagreement.

## Toy Example: 6 Records

Consider a small dataset with three quasi-identifiers -- `age` (numeric),
`sex` (nominal), and `city` (nominal) -- and a sensitive attribute
`salary` that is *not* used in the linkage.

```{r det-toy-data}
orig <- data.frame(
  age    = c(30, 40, 50, 60, 70, 80),
  sex    = factor(c("M", "M", "F", "F", "M", "F")),
  city   = factor(c("A", "B", "A", "B", "A", "B")),
  salary = c(35000, 48000, 52000, 67000, 71000, 45000)
)
anon <- data.frame(
  age    = c(30, 60, 50, 40, 70, 80),  # rows 2 and 4: ages swapped (40<->60)
  sex    = factor(c("M", "M", "F", "F", "M", "F")),
  city   = factor(c("A", "B", "A", "B", "A", "B")),
  salary = c(35000, 67000, 52000, 48000, 71000, 45000)
)
knitr::kable(cbind(Row = 1:6, orig), caption = "Original data")
knitr::kable(cbind(Row = 1:6, anon), caption = "Anonymized data (rows 2 and 4 swapped)")
```

The perturbation is record swapping: rows 2 and 4 had their
ages swapped (40 $\leftrightarrow$ 60), and consequently their
salaries follow (48,000 $\leftrightarrow$ 67,000).
Rows 1, 3, 5, and 6 are unchanged.

The range of `age` is $80 - 30 = 50$. Each variable has equal weight
$w_v = 1$ (the default), so $\sum_v w_v = 3$. Only the three
quasi-identifiers (`age`, `sex`, `city`) enter the linkage; `salary`
is a sensitive attribute and is *not* used.

**Record 1** (age=30, sex=M, city=A) vs. anonymized record 1
(age=30, sex=M, city=A):

$$\delta_{\text{age}} = |30-30|/50 = 0, \quad
\delta_{\text{sex}} = 0, \quad
\delta_{\text{city}} = 0$$
$$d(1,1) = (0 + 0 + 0)/3 = 0.$$

An exact match -- record 1 was not perturbed, so it is the unique
nearest neighbor and $r_1 = 1.0$.

**Record 4** (age=60, sex=F, city=B) -- this record's age was swapped from
60 to 40. How does the attacker fare?

```{r det-toy-distances, echo=FALSE}
# Hand-computed distances for record 4
d4 <- data.frame(
  `Anon row` = 1:6,
  age_dissim = c(abs(60-30)/50, abs(60-60)/50, abs(60-50)/50,
                 abs(60-40)/50, abs(60-70)/50, abs(60-80)/50),
  sex_dissim = c(1, 1, 0, 0, 1, 0),
  city_dissim = c(1, 0, 1, 0, 1, 0)
)
d4$Gower_d <- round((d4$age_dissim + d4$sex_dissim + d4$city_dissim) / 3, 4)
names(d4) <- c("Anon row", "age", "sex", "city", "d(4,j)")
knitr::kable(d4, digits = 4,
             caption = "Variable dissimilarities and Gower distance for original record 4")
```

The two smallest distances are $d(4,4) = 0.1333$ and $d(4,6) = 0.1333$
(a tie). Why? Both anonymized records 4 and 6 are female from city B
(matching the original), and their ages (40 and 80) are equidistant
from 60 ($|60-40| = |60-80| = 20$). With the `nearest` strategy and
uniform weighting, the candidate set is $C_4 = \{4, 6\}$. Since the true
match $j^*_4 = 4$ is in the set, $r_4 = 1/2 = 0.5$. The age swap
halved the risk by creating an ambiguity the attacker cannot resolve.

**Pedagogical point**: nominal mismatches are expensive. A city
mismatch produces a dissimilarity of $\delta = 1.0$, equivalent to
spanning the entire age range ($50/50 = 1.0$). Even a moderate city
change contributes as much to the Gower distance as a 50-year age
difference. This makes nominal variables highly influential in the
linkage.

Now verify with `recordLinkage()`:

```{r det-toy-verify}
res_det <- recordLinkage(orig, anon, key = c("age", "sex", "city"))
knitr::kable(res_det$per_record[, c("risk", "cand_n", "true_in_set", "d_true")],
             digits = 4, caption = "Per-record risk (deterministic, nearest)")
```

Note that `salary` was *not* used as a quasi-identifier -- it
represents a sensitive attribute the attacker wants to learn. When
record 1 has $r_1 = 1.0$, the attacker uniquely identifies this person
*and* can read off their salary (35,000) from the anonymized data. This
is *attribute disclosure*: successful re-identification reveals all
non-QI attributes. By contrast, record 4's risk of 0.5 means the
attacker cannot distinguish between anonymized rows 4 and 6, whose
salaries are 48,000 and 45,000 -- the swapping provided some protection.
This illustrates why identity disclosure risk (measured by
`recordLinkage()`) is a precondition for attribute disclosure.

## Attacker Strategies

The `strategy` parameter controls how the attacker selects candidates.
We compare three strategies on the same toy data:

```{r det-toy-strategies}
res_nearest   <- recordLinkage(orig, anon, key = c("age", "sex", "city"),
                               strategy = "nearest")
res_threshold <- recordLinkage(orig, anon, key = c("age", "sex", "city"),
                               strategy = "threshold", threshold = 0.2)
res_topk      <- recordLinkage(orig, anon, key = c("age", "sex", "city"),
                               strategy = "topk", k = 3)
strat_df <- data.frame(
  Record = 1:6,
  nearest   = round(res_nearest$per_record$risk, 4),
  threshold = round(res_threshold$per_record$risk, 4),
  topk_3    = round(res_topk$per_record$risk, 4)
)
knitr::kable(strat_df, caption = "Risk under different attacker strategies")
```

With `threshold = 0.2`, only exact or near-exact matches qualify, so
records with moderate perturbation get zero risk. With `topk` and
$k = 3$, the candidate set always includes at least 3 records, diluting
risk for unique matches.

## Real-World Example (n = 500)

With 6 records, every record is a candidate for every other record and
ties are common. At scale, ties become rare and the nearest-neighbor
match is typically unique, so the `nearest` strategy often gives
$r_i = 1$ (exact match found) or $r_i = 0$ (true match not the
nearest). This motivates the weighting schemes introduced above.

We generate a realistic census-like dataset with 500 records and apply
perturbation using **sdcMicro** [@templ2015sdcmicro]: rank swapping for
the numeric variables (`age`, `income`) while the categorical variables
(`sex`, `region`, `educ`) remain unchanged. This is a typical workflow
in statistical disclosure control.

```{r det-realdata}
set.seed(123)  # ensures reproducibility for data generation
n <- 500
census <- data.frame(
  age    = sample(18:85, n, replace = TRUE),
  sex    = factor(sample(c("M", "F"), n, replace = TRUE)),
  region = factor(sample(paste0("R", 1:5), n, replace = TRUE)),
  educ   = factor(sample(c("primary", "secondary", "tertiary"), n,
                         replace = TRUE, prob = c(0.2, 0.5, 0.3))),
  income = round(rlnorm(n, meanlog = 10.5, sdlog = 0.6))
)

# Perturbation via sdcMicro: rank-swap numeric variables
census_anon <- census
census_num <- census[, c("age", "income")]
census_swapped <- rankSwap(census_num, variables = c("age", "income"),
                           P = 10, seed = 123)
census_anon$age    <- census_swapped$age
census_anon$income <- census_swapped$income

# Record linkage with 4 QIs
res_real <- recordLinkage(census, census_anon,
                          key = c("age", "sex", "region", "educ"))
print(res_real)
```

```{r det-realdata-summary}
summary(res_real)
```

Note the privacy assessment: the `privacy_pass` flag tests only the
*mean* risk against the default threshold of 0.1. A PASS does not
guarantee that individual records are protected -- `max_risk` may be
1.0, indicating that some records are uniquely re-identifiable. Always
examine `max_risk`, `n_high_risk`, and the risk band distribution
alongside the pass/fail label. A custom threshold can be set via
the `risk_threshold` parameter. For a full discussion, see
Section 8.2 (Interpreting Risk Bands).

```{r det-realdata-plot, fig.height=4, fig.cap="Distribution of per-record re-identification risk (deterministic method). The red dashed line indicates the risk threshold (0.1)."}
plot(res_real, which = 1)
```

The histogram reveals the risk distribution. Records whose rank-swapped
age is close to the original (small rank displacement) retain high risk,
while records swapped with a more distant value are harder to
re-identify.

```{r det-realdata-scatter, fig.height=4, fig.cap="Distance to true match vs. attributed risk. Records with zero distance (no perturbation effect on QIs) have maximum risk."}
plot(res_real, which = 2)
```

The scatter plot shows the relationship between the distance to the true
match and the attributed risk. Records with $d_{\text{true}} = 0$ (no
perturbation effect on these QIs) have maximum risk; as distance
increases, risk decreases because more competing candidates exist.


# Probabilistic Method (Fellegi-Sunter)

The deterministic method treats all variable agreements equally.
In practice, however, agreement on a rare variable value (e.g., an
uncommon occupation) provides much stronger evidence of a true match
than agreement on a common value (e.g., sex). The probabilistic method
addresses this by weighting agreements according to their informativeness.

## Log-Likelihood Ratios

The Fellegi-Sunter framework [@fellegi1969theory] models record pairs
as belonging to one of two classes: matches ($M$) or non-matches ($U$).
For each quasi-identifier $v$, the method estimates:

- $m_v = P(\text{agree on } v \mid \text{true match})$
- $u_v = P(\text{agree on } v \mid \text{non-match})$

The composite log-likelihood ratio for the pair $(i, j)$ is:
$$\Lambda_{ij} = \sum_{v \in \mathcal{K}} \left[\gamma_{ijv} \log\frac{m_v}{u_v} + (1 - \gamma_{ijv}) \log\frac{1-m_v}{1-u_v}\right],$$
where $\gamma_{ijv} = \mathbf{1}[\text{agree on } v]$. High $\Lambda$
indicates strong evidence of a match.

The standard Fellegi-Sunter model assumes *conditional independence* of
agreement indicators across variables, given the match/non-match status.
When quasi-identifiers are correlated (e.g., age and education), this
assumption is approximate, and the log-likelihood ratios may
overestimate or underestimate the evidence for a match.

The key insight is that *rare agreements are more informative*: if a
variable has many categories (low $u_v$), agreement provides stronger
evidence of a true match [@winkler2006overview].

## Estimation of m and u

When `m_probs` and `u_probs` are not supplied, `recordLinkage()` estimates
them internally:

- For **nominal** variables: $m_v$ is estimated from agreement rates among
  aligned pairs; $u_v$ from random pairings.
- For **numeric** variables: agreement is defined via a tolerance based
  on the Median Absolute Deviation (MAD). Two values agree if
  $|x_{iv} - x^a_{jv}| \le \text{MAD}(x_v)$, where
  $\text{MAD}(x_v) = 1.4826 \cdot \text{median}(|x_v - \text{median}(x_v)|)$
  is computed on the original data. MAD is preferred over the standard
  deviation for robustness to outliers.

User-supplied values via `m_probs` and `u_probs` override the estimation
and are recommended when the perturbation mechanism is well understood.

## Toy Example: 5 Records (All Categorical)

We use an all-categorical toy example to avoid the complication of
numeric tolerance.

```{r prob-toy-data}
orig_p <- data.frame(
  sex    = factor(c("M", "F", "M", "F", "M")),
  region = factor(c("N", "S", "E", "W", "N")),
  job    = factor(c("A", "B", "C", "A", "D"))
)
anon_p <- data.frame(
  sex    = factor(c("M", "F", "F", "F", "M")),
  region = factor(c("N", "S", "E", "N", "S")),
  job    = factor(c("A", "B", "C", "B", "D"))
)
knitr::kable(cbind(Row = 1:5, orig_p), caption = "Original data (probabilistic toy)")
knitr::kable(cbind(Row = 1:5, anon_p), caption = "Anonymized data (probabilistic toy)")
```

We supply explicit m/u probabilities for full transparency:

```{r prob-toy-mu}
m <- c(sex = 0.95, region = 0.80, job = 0.85)
u <- c(sex = 0.50, region = 0.25, job = 0.10)
```

The per-variable log-likelihood ratios are:

```{r prob-toy-logLR}
log_agree    <- round(log(m / u), 3)
log_disagree <- round(log((1 - m) / (1 - u)), 3)
lr_table <- data.frame(
  Variable = names(m),
  m = m, u = u,
  `log(m/u)` = log_agree,
  `log((1-m)/(1-u))` = log_disagree,
  check.names = FALSE
)
knitr::kable(lr_table, digits = 3,
             caption = "Per-variable log-likelihood ratios")
```

**Hand-tracing record 1** (M, N, A) against all 5 anonymized records:

```{r prob-toy-handtrace, echo=FALSE}
# Record 1: sex=M, region=N, job=A
# Anon records: (M,N,A), (F,S,B), (F,E,C), (F,N,B), (M,S,D)
agree_matrix <- matrix(c(
  1, 1, 1,  # anon 1: all agree
  0, 0, 0,  # anon 2: none agree
  0, 0, 0,  # anon 3: none agree
  0, 1, 0,  # anon 4: region agrees
  1, 0, 0   # anon 5: sex agrees
), nrow = 5, byrow = TRUE, dimnames = list(
  paste0("Anon ", 1:5),
  c("sex", "region", "job")
))

# Compute Lambda for each pair
lambda <- numeric(5)
for (j in 1:5) {
  lambda[j] <- sum(agree_matrix[j, ] * log(m / u) +
                   (1 - agree_matrix[j, ]) * log((1 - m) / (1 - u)))
}
ht <- data.frame(
  `Anon row` = 1:5,
  sex = ifelse(agree_matrix[, 1] == 1, "agree", "disagree"),
  region = ifelse(agree_matrix[, 2] == 1, "agree", "disagree"),
  job = ifelse(agree_matrix[, 3] == 1, "agree", "disagree"),
  Lambda = round(lambda, 3),
  check.names = FALSE
)
knitr::kable(ht, caption = "Log-LR scores for original record 1 vs all anonymized records")
```

For anonymized record 1 (all three variables agree):
$\Lambda = 0.642 + 1.163 + 2.140 = 3.945$. This is the highest score.
With `fs_threshold = 0` (the default -- include all pairs with positive
log-likelihood ratio), only anonymized record 1 qualifies, so
$r_1 = 1.0$.

**Key insight**: job agreement contributes $\log(0.85/0.10) = 2.140$ --
more than 3 times the contribution of sex agreement
($\log(0.95/0.50) = 0.642$) -- because rare agreements carry much
stronger evidence. The Fellegi-Sunter framework automatically accounts
for this.

**Hand-tracing record 4** (F, W, A) -- note that region "W" does not
appear in the anonymized data:

```{r prob-toy-rec4, echo=FALSE}
# Record 4: sex=F, region=W, job=A
agree4 <- matrix(c(
  0, 0, 1,  # anon 1: job agrees
  1, 0, 0,  # anon 2: sex agrees
  1, 0, 0,  # anon 3: sex agrees
  1, 0, 0,  # anon 4: sex agrees
  0, 0, 0   # anon 5: none
), nrow = 5, byrow = TRUE, dimnames = list(
  paste0("Anon ", 1:5),
  c("sex", "region", "job")
))
lambda4 <- numeric(5)
for (j in 1:5) {
  lambda4[j] <- sum(agree4[j, ] * log(m / u) +
                    (1 - agree4[j, ]) * log((1 - m) / (1 - u)))
}
ht4 <- data.frame(
  `Anon row` = 1:5,
  Lambda = round(lambda4, 3),
  check.names = FALSE
)
knitr::kable(ht4, caption = "Log-LR scores for original record 4 vs all anonymized records")
```

All scores are negative (evidence against a match), so with the default
threshold of 0, the candidate set is empty and $r_4 = 0$. The
perturbation successfully protected this record.

Now verify with `recordLinkage()`:

```{r prob-toy-verify}
res_prob <- recordLinkage(orig_p, anon_p,
                          key = c("sex", "region", "job"),
                          method = "probabilistic",
                          m_probs = m, u_probs = u)
knitr::kable(res_prob$per_record[, c("risk", "cand_n", "true_in_set", "d_true")],
             digits = 4, caption = "Per-record risk (probabilistic)")
```

The probabilistic method correctly identifies record 1 (all variables
agree, yielding the highest $\Lambda$, unique match, $r_1 = 1.0$)
and fully protects record 4 (region "W" absent in the anonymized
data, all scores negative, $r_4 = 0.0$). Compared to the deterministic
method, the probabilistic approach automatically gives more weight to
agreements on rare values: job agreement (with 4+ categories)
contributes 3.3 times more evidence than sex agreement (with only 2
categories), reflecting the discriminative power of each variable.

## Real-World Example (n = 500)

With the toy example, we supplied explicit m/u values. At scale, these
are typically estimated automatically from the data (see above). The
auto-estimation uses aligned pairs for m-probabilities and random
pairings for u-probabilities, which works well when the true match rate
is not too low.

Using the same 500-record census dataset, but now with the probabilistic
method and auto-estimated m/u:

```{r prob-real}
res_prob_real <- recordLinkage(census, census_anon,
                               key = c("age", "sex", "region", "educ"),
                               method = "probabilistic")
summary(res_prob_real)
```

The summary includes the estimated Fellegi-Sunter parameters. Note
how the m-probability for `sex` is typically high (sex is rarely
perturbed by within-region swapping) while `age` has a lower
m-probability (age is directly swapped).

```{r prob-real-plot, fig.height=4, fig.cap="Per-variable log-likelihood ratios for agreement and disagreement (probabilistic method). Variables with high log(m/u) are the most discriminating."}
plot(res_prob_real, which = 3)
```

The barplot shows per-variable log-likelihood ratios for agreement and
disagreement. Variables with high $\log(m/u)$ are the most discriminating
for identifying matches.


# PRAM Method

## Transition Matrices

For data protected via the Post-Randomization Method (PRAM)
[@gouweleeuw1998post], the perturbation mechanism is fully described by
transition matrices. For each categorical variable $v$, the matrix
$T_v$ has entry $T_v[a, b] = P(\text{output} = b \mid \text{input} = a)$.

The probability that original record $x_i$ produced anonymized record
$x^a_j$ is the product over all variables:
$$P(x^a_j \mid x_i) = \prod_{v \in \mathcal{K}} T_v[x_{iv}, x^a_{jv}].$$

The re-identification risk is then the normalized probability:
$$r_i = \frac{P(x^a_{j^*_i} \mid x_i)}{\sum_{j \in C_i} P(x^a_j \mid x_i)},$$
where $C_i = \{j : P(x^a_j \mid x_i) > 0\}$ is restricted to
records with nonzero transition probability (and within the blocking
window, if applicable). This assumes independence of
perturbation across variables, which is exact when PRAM is applied
independently to each variable.

## Toy Example: 5 Records

We use abstract categorical variables (`color`, `shape`) here to focus
attention on the PRAM transition mechanism itself, with $3 \times 3$
matrices that exhibit richer structure than the binary quasi-identifiers
used in the other sections.

```{r pram-toy-data}
orig_pram <- data.frame(
  color = factor(c("R", "G", "B", "R", "G")),
  shape = factor(c("C", "S", "T", "C", "T"))
)
anon_pram <- data.frame(
  color = factor(c("R", "B", "B", "G", "G")),
  shape = factor(c("C", "S", "T", "S", "T"))
)

tm_color <- matrix(
  c(0.8, 0.1, 0.1,
    0.1, 0.7, 0.2,
    0.1, 0.2, 0.7),
  3, 3, byrow = TRUE,
  dimnames = list(c("R", "G", "B"), c("R", "G", "B"))
)

tm_shape <- matrix(
  c(0.9, 0.05, 0.05,
    0.1, 0.8,  0.1,
    0.05, 0.05, 0.9),
  3, 3, byrow = TRUE,
  dimnames = list(c("C", "S", "T"), c("C", "S", "T"))
)

knitr::kable(tm_color, caption = "Transition matrix for 'color'")
knitr::kable(tm_shape, caption = "Transition matrix for 'shape'")
```

**Hand-tracing record 1** (R, C) $\to$ anonymized record 1 (R, C):

- $T_{\text{color}}[\text{R}, \text{R}] = 0.8$
- $T_{\text{shape}}[\text{C}, \text{C}] = 0.9$
- $P(x^a_1 \mid x_1) = 0.8 \times 0.9 = 0.720$

**Record 4** (R, C) $\to$ anonymized record 4 (G, S):

- $T_{\text{color}}[\text{R}, \text{G}] = 0.1$
- $T_{\text{shape}}[\text{C}, \text{S}] = 0.05$
- $P(x^a_4 \mid x_4) = 0.1 \times 0.05 = 0.005$

```{r pram-toy-allprobs, echo=FALSE}
# Compute all transition probabilities for records 1 and 4
probs <- matrix(0, nrow = 5, ncol = 2,
                dimnames = list(paste0("Anon ", 1:5),
                                c("P(anon_j | orig_1)", "P(anon_j | orig_4)")))
for (j in 1:5) {
  probs[j, 1] <- tm_color[as.character(orig_pram$color[1]),
                           as.character(anon_pram$color[j])] *
                 tm_shape[as.character(orig_pram$shape[1]),
                          as.character(anon_pram$shape[j])]
  probs[j, 2] <- tm_color[as.character(orig_pram$color[4]),
                           as.character(anon_pram$color[j])] *
                 tm_shape[as.character(orig_pram$shape[4]),
                          as.character(anon_pram$shape[j])]
}
probs_df <- as.data.frame(probs)
probs_df$`Normalized risk (rec 1)` <- round(probs[, 1] / sum(probs[, 1]), 4)
probs_df$`Normalized risk (rec 4)` <- round(probs[, 2] / sum(probs[, 2]), 4)
knitr::kable(probs_df, digits = 4,
             caption = "Transition probabilities and normalized risks for records 1 and 4")
```

Records 1 and 4 have identical original values (R, C) but completely
different risks. Record 1's output (R, C) has high probability under
the original (the diagonal is large), giving risk $\approx 0.97$.
Record 4's output (G, S) has very low probability, giving risk
$\approx 0.007$. **PRAM risk depends on the actual output, not just
the input.**

This is fundamentally different from the deterministic method, where risk
depends only on the distance between original and anonymized values.
Here, risk depends on the *transition probability* of the actual output
given the input, reflecting the perturbation mechanism directly. A
record that retains its original categories (record 1) is easy to
re-identify because the diagonal entries are large; a record that was
heavily perturbed (record 4) is protected because the off-diagonal
probability product is small relative to the total.

Now verify with `recordLinkage()`:

```{r pram-toy-verify}
res_pram <- recordLinkage(orig_pram, anon_pram,
                          key = c("color", "shape"),
                          method = "pram",
                          pram_matrix = list(color = tm_color,
                                            shape = tm_shape))
knitr::kable(res_pram$per_record[, c("risk", "d_true")],
             digits = 4, caption = "Per-record risk (PRAM)")
```

## Forward vs. Reverse Asymmetry

The PRAM method is unique among the four methods in that it is
*asymmetric*: the transition matrix $T_v[a, b]$ is not symmetric, so
forward risk (given the original, sum over anonymized candidates) and
reverse risk (given the anonymized record, sum over original candidates)
can differ.

```{r pram-direction-data}
orig_dir <- data.frame(grade = factor(c("A", "B", "C", "A", "B")))
anon_dir <- data.frame(grade = factor(c("A", "C", "C", "B", "A")))

tm_grade <- matrix(
  c(0.7, 0.2, 0.1,
    0.1, 0.6, 0.3,
    0.3, 0.1, 0.6),
  3, 3, byrow = TRUE,
  dimnames = list(c("A", "B", "C"), c("A", "B", "C"))
)
knitr::kable(tm_grade, caption = "Asymmetric transition matrix for 'grade'")
```

```{r pram-direction-compare}
res_fwd <- recordLinkage(orig_dir, anon_dir,
                         key = "grade", method = "pram",
                         pram_matrix = list(grade = tm_grade),
                         direction = "forward")
res_rev <- recordLinkage(orig_dir, anon_dir,
                         key = "grade", method = "pram",
                         pram_matrix = list(grade = tm_grade),
                         direction = "reverse")

dir_df <- data.frame(
  Record = 1:5,
  Orig_grade = as.character(orig_dir$grade),
  Anon_grade = as.character(anon_dir$grade),
  Forward_risk = round(res_fwd$per_record$risk, 4),
  Reverse_risk = round(res_rev$per_record$risk, 4)
)
knitr::kable(dir_df,
             caption = "Forward vs. reverse risk (PRAM with asymmetric matrix)")
```

In forward mode, for each original record's value (fixing a row of $T$),
the function evaluates $T[\text{original}, b]$ for each anonymized
candidate $b$. In reverse mode, for each anonymized record's value
(fixing a column of $T$), the function evaluates
$T[\text{original}_j, \text{anon}]$ for each original candidate $j$.
Both use $P(\text{output} \mid \text{input})$; only the roles of
query and search change. When $T$ is asymmetric, the normalization
denominators differ, yielding different risk profiles.

## Real-World Example (n = 500)

The toy examples above used hand-specified transition matrices.
In practice, transition matrices are generated by the perturbation
software. Here we use `sdcMicro::pram()` [@templ2015sdcmicro; @gouweleeuw1998post],
which generates the matrices internally and applies the perturbation.
We then extract the matrices and pass them directly to
`recordLinkage()`.

The `pd` parameter in `sdcMicro::pram()` sets the minimum diagonal
probability of the transition matrix -- the probability that a record
retains its original category. With `pd = 0.8`, each category is
preserved with at least 80% probability. The `alpha` parameter
controls the invariance property: `alpha = 1` produces a fully
invariant PRAM (marginal distributions are preserved in expectation),
while `alpha = 0` maximizes perturbation. The value `alpha = 0.5` is
a common compromise.

```{r pram-real}
set.seed(456)  # ensures reproducibility for PRAM data
n <- 500
pram_data <- data.frame(
  sex    = factor(sample(c("M", "F"), n, replace = TRUE)),
  region = factor(sample(c("N", "S", "E", "W"), n, replace = TRUE)),
  educ   = factor(sample(c("low", "mid", "high"), n, replace = TRUE,
                         prob = c(0.3, 0.5, 0.2)))
)

# Apply PRAM via sdcMicro
pram_vars <- c("sex", "region", "educ")
pram_res <- pram(pram_data, variables = pram_vars, pd = 0.8, alpha = 0.5)

# Build the anonymized dataset from _pram columns
pram_anon <- data.frame(
  sex    = pram_res$sex_pram,
  region = pram_res$region_pram,
  educ   = pram_res$educ_pram
)

# Extract transition matrices generated by sdcMicro
pram_params <- attr(pram_res, "pram_params")
tm_list <- setNames(
  lapply(pram_vars, function(v) pram_params[[v]]$Rs),
  pram_vars
)

# Inspect the perturbation summary
print(attr(pram_res, "summary"))

# Record linkage using the actual transition matrices
res_pram_real <- recordLinkage(
  pram_data, pram_anon,
  key = pram_vars,
  method = "pram",
  pram_matrix = tm_list
)
print(res_pram_real)
```

The PRAM risk distribution reflects the perturbation parameters
directly: with `pd = 0.8` (minimum diagonal probability), records
that retained their original categories appear with high probability
and thus receive high risk, while records that were perturbed (moved
off the diagonal) receive lower risk. The intermediate `alpha = 0.5`
balances perturbation strength against marginal preservation.

```{r pram-real-plot, fig.height=4, fig.cap="Risk distribution for the PRAM method on 500 records with sdcMicro-generated transition matrices."}
plot(res_pram_real, which = 1)
```

The histogram shows the characteristic bimodal pattern of PRAM risk:
a cluster of high-risk records (those retaining their original
categories) and a spread of lower-risk records (those that were
perturbed). Increasing `pd` (higher diagonal probability) would shift
mass toward the high-risk peak; decreasing it would spread mass
toward lower risk.


# Predictive Method (Propensity Score)

When quasi-identifiers are high-dimensional or exhibit nonlinear
interactions, pairwise distance computation may miss important
structure. The predictive method addresses this by using a propensity
score model to project the multivariate QI space into a single
informative dimension.

## Propensity Score Linkage

The concept of propensity scores originates in causal inference
[@rosenbaum1983propensity], where they are used to balance treatment
groups on observed covariates. Here we repurpose the idea for record
linkage: the propensity score measures how "original-like" a record
is, providing a natural 1D summary for matching.

The predictive method fits a logistic regression on the stacked
original + anonymized data to predict record origin (original = 1,
anonymized = 0) from the quasi-identifiers:
$$\log\frac{e(x)}{1 - e(x)} = \beta_0 + \sum_{v \in \mathcal{K}} \beta_v x_v,$$
where $e(x) = P(\text{original} \mid \text{QIs} = x)$ is the
propensity score. This projects the multivariate QI space into a single
dimension that captures the discriminability between the two datasets.
When the logistic model is correctly specified, this projection
maximally separates original from anonymized records; when
misspecified (e.g., important interactions are missing), the projection
may compress relevant variation. A random forest alternative
(`pred_model = "rf"`) is available for cases where nonlinearity is
suspected.

Records are then linked in the 1D propensity-score space. The distance
between original record $i$ and anonymized record $j$ is simply:
$$d(i,j) = |e(x_i) - e(x^a_j)|.$$

The standard strategies (nearest, threshold, top-$k$) and weighting
schemes apply to these propensity-score distances.

## Fisher SE Bandwidth

When `pred_se = TRUE` (default) and `pred_model = "logit"`, the
bandwidth for kernel weighting is derived from the Fisher information
matrix:
$$\text{SE}_i = \hat{e}_i(1 - \hat{e}_i) \sqrt{\mathbf{x}_i^\top \hat{V} \mathbf{x}_i},$$
where $\hat{V}$ is the variance-covariance matrix of the logistic
regression coefficients. Records with uncertain predictions (near 0.5)
get wider kernels, while confidently classified records get narrow
kernels -- and hence higher individual risk. Note that `pred_se` only
affects the bandwidth when `risk_weighting = "kernel"`; with uniform or
softmax weighting it has no effect.

## Toy Example: 6 Records

```{r pred-toy-data}
orig_pred <- data.frame(
  age = c(20, 40, 40, 60, 60, 80),
  sex = factor(c("F", "M", "F", "M", "F", "M"))
)
anon_pred <- data.frame(
  age = c(20, 40, 60, 40, 60, 80),  # rows 3 and 4 swapped ages
  sex = factor(c("F", "M", "F", "M", "F", "M"))
)
```

```{r pred-toy-run}
res_pred <- recordLinkage(orig_pred, anon_pred,
                          key = c("age", "sex"),
                          method = "predictive")
print(res_pred)
```

To understand the linkage, we inspect the propensity scores. The model
stacks original and anonymized records (12 rows total) and fits a
logistic regression predicting origin from `age` and `sex`.

```{r pred-toy-scores}
# Extract propensity scores from the result
p_orig <- res_pred$propensity_info$propensity_original
p_anon <- res_pred$propensity_info$propensity_synthetic

score_df <- data.frame(
  Record = 1:6,
  Age_orig = orig_pred$age,
  Sex_orig = as.character(orig_pred$sex),
  Score_orig = round(p_orig, 4),
  Age_anon = anon_pred$age,
  Score_anon = round(p_anon, 4)
)
knitr::kable(score_df, caption = "Propensity scores for original and anonymized records")
```

With only 12 records (6 + 6), the logistic model has limited ability to
separate the two groups, so all propensity scores cluster near 0.5.
The ordering is nonetheless informative: the model uses `age` and `sex`
jointly to distinguish original from anonymized records.

Rows 3 and 4 had their ages swapped (40 $\leftrightarrow$ 60).
Records that were not perturbed (rows 1, 2, 5, 6) share identical
covariates with their anonymized counterparts and thus receive the
same propensity score, giving $d(i,j) = 0$ and risk $r_i = 1.0$.
For swapped records 3 and 4, the perturbation shifts the propensity
score and the nearest anonymized record may no longer be the true match.

```{r pred-toy-risk}
knitr::kable(res_pred$per_record[, c("risk", "cand_n", "true_in_set", "d_true")],
             digits = 4, caption = "Per-record risk (predictive method)")
```

**Key insight**: The propensity score projects the multivariate QI
space into a single dimension that captures discriminability between
the two datasets. With this small toy example, the projection is
nearly flat. With larger samples and more QIs, the 1D projection may
capture multivariate structure that pairwise Gower distances miss,
though this advantage has not yet been empirically established in the
SDC literature (see Limitations).

```{r pred-toy-propensity, fig.height=4, fig.cap="Propensity score densities for original and anonymized records. Well-separated distributions indicate easy-to-link records."}
plot(res_pred, which = 4)
```

## Real-World Example (n = 500)

With 500 records and 4 QIs, the logistic model can meaningfully
separate original from anonymized records, producing a wider range
of propensity scores than the toy example. Using the same census
dataset:

```{r pred-real}
res_pred_real <- recordLinkage(census, census_anon,
                               key = c("age", "sex", "region", "educ"),
                               method = "predictive")
print(res_pred_real)
```

```{r pred-real-riskhist, fig.height=4, fig.cap="Per-record risk distribution (predictive method, nearest strategy)."}
plot(res_pred_real, which = 1)
```

To compare weighting schemes, we must use a strategy that produces
candidate sets with more than one record. With the default `nearest`
strategy, continuous propensity scores rarely produce ties -- nearly
every record has a unique nearest neighbor with `cand_n = 1`, giving
$r_i = 1$ regardless of the weighting scheme. Using `topk` with $k = 5$
ensures at least 5 candidates per record, making the weighting choice
meaningful:

```{r pred-real-compare}
res_pred_topk <- recordLinkage(census, census_anon,
                                key = c("age", "sex", "region", "educ"),
                                method = "predictive",
                                strategy = "topk", k = 5)
res_pred_kernel <- recordLinkage(census, census_anon,
                                 key = c("age", "sex", "region", "educ"),
                                 method = "predictive",
                                 strategy = "topk", k = 5,
                                 risk_weighting = "kernel")

comp_pred <- data.frame(
  Weighting = c("Uniform (top-5)", "Kernel (Fisher SE, top-5)"),
  Mean_risk = c(res_pred_topk$overall$mean_risk,
                res_pred_kernel$overall$mean_risk),
  Max_risk = c(res_pred_topk$overall$max_risk,
               res_pred_kernel$overall$max_risk),
  Pct_high = c(100 * res_pred_topk$overall$pct_high_risk,
               100 * res_pred_kernel$overall$pct_high_risk)
)
knitr::kable(comp_pred, digits = 4,
             caption = "Predictive method: uniform vs kernel weighting (top-5 strategy)")
```

The kernel weighting with Fisher SE bandwidth concentrates risk on records
where the propensity model is confident about the match, giving a more
nuanced risk profile than uniform weighting.  Note that the
propensity-score approach to record linkage risk is novel and its
properties are not yet fully established; see the Limitations section
for caveats.

```{r pred-real-plot, fig.height=4, fig.cap="Propensity score densities for the 500-record census dataset (predictive method)."}
plot(res_pred_real, which = 4)
```


# Comparing Methods and Weighting Schemes

## Side-by-Side Comparison

We run all four methods on the same 500-record census dataset and compare
results. For the PRAM method we use the categorical-only subset since
PRAM requires transition matrices.

```{r compare-methods}
# Deterministic (already computed)
# Probabilistic (already computed)
# Predictive (already computed)

# For fair comparison, also run deterministic on categorical-only keys
res_det_cat <- recordLinkage(census, census_anon,
                             key = c("sex", "region", "educ"))

# Apply sdcMicro::pram() to the census categorical variables
set.seed(789)
cat_vars <- c("sex", "region", "educ")
pram_census_res <- pram(census[, cat_vars], variables = cat_vars,
                        pd = 0.85, alpha = 0.5)
census_pram <- census
census_pram$sex    <- pram_census_res$sex_pram
census_pram$region <- pram_census_res$region_pram
census_pram$educ   <- pram_census_res$educ_pram

# Extract the actual transition matrices from sdcMicro
pp <- attr(pram_census_res, "pram_params")
tm_census <- setNames(lapply(cat_vars, function(v) pp[[v]]$Rs), cat_vars)

res_pram_census <- recordLinkage(
  census, census_pram,
  key = cat_vars,
  method = "pram",
  pram_matrix = tm_census
)

compare_df <- data.frame(
  Method = c("Deterministic (4 QIs)", "Probabilistic (4 QIs)",
             "Predictive (4 QIs)", "Deterministic (3 cat QIs)",
             "PRAM (3 cat QIs)"),
  Mean_risk = c(
    res_real$overall$mean_risk,
    res_prob_real$overall$mean_risk,
    res_pred_real$overall$mean_risk,
    res_det_cat$overall$mean_risk,
    res_pram_census$overall$mean_risk
  ),
  Max_risk = c(
    res_real$overall$max_risk,
    res_prob_real$overall$max_risk,
    res_pred_real$overall$max_risk,
    res_det_cat$overall$max_risk,
    res_pram_census$overall$max_risk
  ),
  Pct_high = c(
    100 * res_real$overall$pct_high_risk,
    100 * res_prob_real$overall$pct_high_risk,
    100 * res_pred_real$overall$pct_high_risk,
    100 * res_det_cat$overall$pct_high_risk,
    100 * res_pram_census$overall$pct_high_risk
  ),
  Pct_true_in_set = c(
    100 * res_real$overall$pct_true_in_set,
    100 * res_prob_real$overall$pct_true_in_set,
    100 * res_pred_real$overall$pct_true_in_set,
    100 * res_det_cat$overall$pct_true_in_set,
    100 * res_pram_census$overall$pct_true_in_set
  )
)
knitr::kable(compare_df, digits = 3,
             caption = "Method comparison on the census dataset")
```

**Important caveat**: This comparison is *illustrative*, not a
controlled experiment. The five configurations differ in three ways
simultaneously: (1) the linkage method, (2) the set of quasi-identifiers
(4 QIs including numeric `age` vs. 3 categorical QIs), and (3) the
perturbation mechanism (rank swapping vs. PRAM). In particular, the
PRAM row uses a *different anonymized dataset* (PRAM-perturbed) than
the other rows (rank-swapped), so differences in risk reflect the
combined effect of method, QI set, and perturbation strength. A fair
comparison of methods would require applying all methods to the same
anonymized data with the same QIs.

That said, two patterns are noteworthy: (1) the deterministic and
probabilistic methods with 4 QIs (including numeric `age`) tend to give
finer discrimination because continuous variables provide more
information for distinguishing records; (2) the PRAM method, which
uses the actual transition matrices, gives a risk profile that directly
reflects the perturbation mechanism rather than relying on distance
heuristics.

## Weighting Scheme Comparison

With the default `nearest` strategy, the candidate set for each record
contains only those anonymized records tied at the minimum distance.
When quasi-identifiers include continuous variables (such as `age`),
ties are extremely rare -- nearly every record has a unique nearest
neighbor. When $|C_i| = 1$, the risk is $1/1 = 1$ regardless of the
weighting scheme, because there is only one candidate to weight.

To demonstrate the effect of weighting, we use `strategy = "topk"` with
$k = 5$, which ensures each record has at least 5 candidates among
which the weighting can differentiate:

```{r compare-weights}
qi <- c("age", "sex", "region", "educ")
res_unif <- recordLinkage(census, census_anon, key = qi,
                          strategy = "topk", k = 5,
                          risk_weighting = "uniform")
res_soft <- recordLinkage(census, census_anon, key = qi,
                          strategy = "topk", k = 5,
                          risk_weighting = "softmax")
res_kern_gauss <- recordLinkage(census, census_anon, key = qi,
                                strategy = "topk", k = 5,
                                risk_weighting = "kernel",
                                kernel = "gaussian")
res_kern_epan <- recordLinkage(census, census_anon, key = qi,
                               strategy = "topk", k = 5,
                               risk_weighting = "kernel",
                               kernel = "epanechnikov")
res_kern_tri <- recordLinkage(census, census_anon, key = qi,
                              strategy = "topk", k = 5,
                              risk_weighting = "kernel",
                              kernel = "tricube")

wt_df <- data.frame(
  Weighting = c("Uniform", "Softmax", "Kernel (Gaussian)",
                "Kernel (Epanechnikov)", "Kernel (Tricube)"),
  Mean_risk = c(res_unif$overall$mean_risk,
                res_soft$overall$mean_risk,
                res_kern_gauss$overall$mean_risk,
                res_kern_epan$overall$mean_risk,
                res_kern_tri$overall$mean_risk),
  Max_risk = c(res_unif$overall$max_risk,
                res_soft$overall$max_risk,
                res_kern_gauss$overall$max_risk,
                res_kern_epan$overall$max_risk,
                res_kern_tri$overall$max_risk),
  Pct_high = c(100 * res_unif$overall$pct_high_risk,
               100 * res_soft$overall$pct_high_risk,
               100 * res_kern_gauss$overall$pct_high_risk,
               100 * res_kern_epan$overall$pct_high_risk,
               100 * res_kern_tri$overall$pct_high_risk)
)
knitr::kable(wt_df, digits = 4,
             caption = "Risk weighting schemes (deterministic method, top-5 strategy)")
```

Uniform weighting is most conservative -- it distributes risk equally
among all 5 candidates. Softmax and kernel weighting concentrate risk on
the closest candidate, giving higher risk to easy-to-link records and
lower risk to well-protected ones.

## Summary Table

| Method | Assumptions | Best when | Limitation |
|:-------|:------------|:----------|:-----------|
| Deterministic | Minimal (distance metric) | Default / first pass | Ignores agreement rarity |
| Probabilistic | m/u available or estimable | Audit documentation | Sensitive to m/u misspecification |
| PRAM | Transition matrix known | PRAM / swapping mechanism | Only categorical; requires exact matrices |
| Predictive | Logistic model adequate | Potentially useful for high-dimensional / mixed QIs | Model uncertainty; not yet established in SDC literature |


# Computational Details

## S3 Class Structure

The `recordLinkage()` function is an S3 generic with methods for
`data.frame` (the default) and `synth_pair` objects. It returns an
object of class `"recordLinkageRisk"` with the following components:

- `$per_record`: a `data.frame` with one row per query record,
  containing `risk` (re-identification risk), `cand_n` (candidate set
  size), `true_in_set` (logical, whether the true match is among
  candidates), and `d_true` (distance to true match).
- `$overall`: a list with summary statistics: `mean_risk`, `max_risk`,
  `median_risk`, `pct_high_risk` (fraction above threshold),
  `pct_true_in_set`, `n_unique_match` (records with $r = 1$), and
  `risk_bands` (tabulated risk distribution).
- `$privacy_pass`: logical, `TRUE` if `mean_risk` $\le 0.1$.
- `$params`: a list recording the method, direction, weighting,
  strategy, and other parameters used.
- For the probabilistic method: `$fs_params` with estimated or
  supplied m/u probabilities.
- For the predictive method: `$propensity_info` with fitted scores
  and, if `pred_se = TRUE`, Fisher standard errors.

Standard S3 methods are provided:

- `print()`: one-screen summary with method, risk statistics, and
  pass/fail status.
- `summary()`: detailed output including risk quantiles, risk band
  distribution, distance-to-true-match statistics, and per-variable
  Fellegi-Sunter parameters (for the probabilistic method). Returns
  an object of class `"summary.recordLinkageRisk"`.
- `plot(x, which = ...)`: up to 4 plot types: (1) risk histogram,
  (2) distance vs. risk scatter, (3) per-variable log-LR barplot
  (probabilistic only), (4) propensity score densities (predictive
  only).

## Computational Complexity

Let $n$ and $m$ denote the number of original and anonymized records,
and $p = |\mathcal{K}|$ the number of quasi-identifiers.

| Method | Complexity | Notes |
|:-------|:-----------|:------|
| Deterministic | $O(nmp)$ | Full Gower distance matrix |
| Probabilistic | $O(nmp)$ | Log-LR computation per pair per variable |
| PRAM | $O(nmp)$ | Transition probability lookup per pair per variable |
| Predictive | $O((n+m)p + nm)$ | $O((n+m)p)$ for model fitting + $O(nm)$ for score comparison |

Without blocking, all methods require $O(nm)$ pairwise comparisons.
For large datasets, blocking (described below) restricts comparisons
to records that agree on a blocking variable, reducing the effective
number of pairs. As a rough guideline:

| Dataset size | Recommendation |
|:-------------|:---------------|
| $n < 5{,}000$ | Any method, no blocking needed |
| $5{,}000 < n < 50{,}000$ | Use blocking; the predictive method may be faster (single model fit vs. full distance matrix) |
| $n > 50{,}000$ | Blocking required; consider the predictive method for its lower constant factor |

## Software Dependencies

The core `recordLinkage()` function requires only base R and the
**VIM** package (for Gower distance computation in the deterministic
method). The examples in this paper additionally use **sdcMicro**
[@templ2015sdcmicro] for perturbation (rank swapping and PRAM).
Both packages are listed in `Suggests` in the package `DESCRIPTION`.


# Advanced Features

## Blocking for Scalability

For large datasets, computing all $n \times m$ pairwise distances is
expensive. Blocking restricts comparisons to records that agree on one or
more blocking variables:

```{r blocking}
# Without blocking
system.time(
  res_noblock <- recordLinkage(census, census_anon,
                               key = c("age", "sex", "region", "educ"))
)

# With blocking on region
system.time(
  res_block <- recordLinkage(census, census_anon,
                             key = c("age", "sex", "region", "educ"),
                             block = "region")
)

block_df <- data.frame(
  Blocking = c("None", "region"),
  Mean_risk = c(res_noblock$overall$mean_risk,
                res_block$overall$mean_risk),
  Pct_true_in_set = c(100 * res_noblock$overall$pct_true_in_set,
                      100 * res_block$overall$pct_true_in_set)
)
knitr::kable(block_df, digits = 4,
             caption = "Effect of blocking on risk and true-match retention")
```

Blocking trades off completeness for speed. When the perturbation
preserves the blocking variable (as with within-region swapping), the
true match remains in the candidate set and `pct_true_in_set` stays
high. If the blocking variable itself is perturbed, some true matches
will be missed.

## Direction: Forward vs. Reverse

By default (`direction = "forward"`), the function iterates over original
records and searches in the anonymized data. This answers: *how easily
can each individual be re-identified?*

With `direction = "reverse"`, iteration is over anonymized records,
searching in the original data. This answers: *how disclosive is each
released record?*

```{r direction}
res_fwd2 <- recordLinkage(census, census_anon,
                          key = c("age", "sex", "region", "educ"),
                          direction = "forward")
res_rev2 <- recordLinkage(census, census_anon,
                          key = c("age", "sex", "region", "educ"),
                          direction = "reverse")

dir_comp <- data.frame(
  Direction = c("Forward", "Reverse"),
  Mean_risk = c(res_fwd2$overall$mean_risk,
                res_rev2$overall$mean_risk),
  Max_risk = c(res_fwd2$overall$max_risk,
               res_rev2$overall$max_risk),
  Pct_high = c(100 * res_fwd2$overall$pct_high_risk,
               100 * res_rev2$overall$pct_high_risk)
)
knitr::kable(dir_comp, digits = 4,
             caption = "Forward vs. reverse direction (deterministic)")
```

For the deterministic and predictive methods, distances are symmetric,
so only the evaluation perspective changes when switching direction.
For the probabilistic method, the log-likelihood ratios are also
symmetric, though auto-estimated m/u parameters are always fitted in the
forward direction. For PRAM, direction changes which side of the
transition matrix is used, so forward and reverse risks can differ
substantially (see the forward vs. reverse example in the PRAM section).

## Ground Truth Options

The `truth` parameter specifies how to determine the correct match for
scoring:

- `truth = "row"` (default): Row $i$ in the original corresponds to
  row $i$ in the anonymized data. Requires $n = m$.
- `truth = "id"`: A shared identifier column links records across
  datasets. Use when rows are shuffled or datasets have different sizes.

```{r truth-id}
# Add an ID column and shuffle the anonymized data
census_id <- census
census_id$id <- seq_len(n)
census_anon_id <- census_anon
census_anon_id$id <- seq_len(n)
set.seed(321)
census_anon_id <- census_anon_id[sample(n), ]  # shuffle

res_id <- recordLinkage(census_id, census_anon_id,
                        key = c("age", "sex", "region", "educ"),
                        truth = "id", id = "id")
cat("Mean risk (row-aligned):", round(res_real$overall$mean_risk, 4), "\n")
cat("Mean risk (id-based, shuffled):", round(res_id$overall$mean_risk, 4), "\n")
```

## The synth_pair Interface

For workflows that use the `synth_pair` container, `recordLinkage()`
has a dedicated method:

```{r synthpair}
pair <- synth_pair(census, census_anon,
                   key_vars = c("age", "sex", "region", "educ"))
res_pair <- recordLinkage(pair)
cat("Mean risk:", round(res_pair$overall$mean_risk, 4), "\n")
```

The `synth_pair` method extracts `key_vars` automatically, so you don't
need to specify `key` again.


# Practical Guidance

## Method Selection Guide

| What you know | Recommended method |
|:--------------|:-------------------|
| Nothing about the perturbation mechanism | `"deterministic"` |
| Transition matrices available (PRAM, swapping) | `"pram"` |
| m/u probabilities available or formal audit needed | `"probabilistic"` |
| Many QIs, nonlinear interactions suspected | `"predictive"` |
| Quick initial assessment | `"deterministic"` with `strategy = "nearest"` |
| Conservative upper bound | `"deterministic"` with `risk_weighting = "uniform"` |

When in doubt, start with the deterministic method. It requires no
assumptions beyond a sensible distance metric and gives a reasonable
baseline. Use the other methods when additional information about the
perturbation mechanism is available.

Two scenarios deserve special attention. First, when nonlinear
interactions among quasi-identifiers are suspected (e.g., age $\times$
education jointly identifying a record), the deterministic method may
underestimate risk because it computes variable-wise dissimilarities
and sums them -- it cannot capture that a *combination* of values is
unique even when individual values are common. The predictive method's
logistic model can include interaction terms and may detect such
patterns. Second, when the perturbation mechanism is known exactly
(transition matrices from PRAM, or m/u probabilities from a formal
audit), the mechanism-specific methods (PRAM or probabilistic) give
more accurate risk estimates than the distance-based default.

## Interpreting Risk Bands

The `summary()` output classifies records into risk bands:

| Band | Risk range | Interpretation |
|:-----|:-----------|:---------------|
| Unique match | $r = 1$ | Attacker can uniquely identify this record |
| Very high | $(0.5, 1)$ | More likely than not to be correctly identified |
| High | $(0.2, 0.5]$ | Substantial risk; consider additional protection |
| Moderate | $(0.1, 0.2]$ | Borderline; assess in context |
| Low | $(0.05, 0.1]$ | Acceptable for most use cases |
| Very low | $[0, 0.05]$ | Well protected |

Note that $r_i = 1$ means the attacker *could* uniquely identify this
record under the assumed model -- it does not mean re-identification has
occurred. The computed risk is an upper bound because the attacker model
assumes perfect membership and QI knowledge.

A common threshold for acceptable risk is $\bar{r} < 0.1$ (mean risk
below 10%), following the guidelines in @hundepool2012statistical
(Chapter 5). In practice, many statistical offices additionally require
that no record has $r > 0.5$ (no unique matches) and that the fraction
of high-risk records is small. The `privacy_pass` flag in
`recordLinkage()` tests only the *mean* risk against the threshold.
High tail risk (e.g., many records with $r > 0.5$) can coexist with a
passing mean -- always check `overall$max_risk` and the risk band
distribution alongside the pass/fail label.

These risk bands are also stored per record in `per_record$risk_band`
(a factor), enabling direct subsetting without recomputing thresholds.
The Gini coefficient (`overall$risk_gini`) measures how concentrated
risk is among a few records: values near 0 indicate uniform risk,
while values near 1 indicate that all risk is concentrated in a
small fraction of records.

## Identifying and Handling High-Risk Records

When the risk assessment returns a warning, practitioners need to
identify which records are at risk and decide on remedial action.
The `recordLinkageRisk` object provides several per-record diagnostics
and helper functions for this purpose.

### Per-record diagnostics

Each record in `per_record` now carries a categorical `risk_band`
(very_low, low, moderate, high, very_high, unique_match) and a
`d_rank` (the rank of the true match among all candidates, where
rank 1 is the closest/best match):

```{r per-record-diagnostics}
head(res_real$per_record)
table(res_real$per_record$risk_band)
```

### Top at-risk records

Use `top_at_risk()` to retrieve the highest-risk records with their
QI values:

```{r top-at-risk}
top <- top_at_risk(res_real, n = 10, data = census)
knitr::kable(top, digits = 4,
             caption = "Top 10 highest-risk records with QI values")
```

### Subpopulation analysis

Use `risk_by_group()` to identify which subgroups carry the most
risk:

```{r risk-by-group}
rg <- risk_by_group(res_real, group = "region", data = census)
knitr::kable(rg, digits = 4,
             caption = "Risk by region")
```

### Detailed record inspection

For forensic analysis of a single record, use `inspect_record()`
(requires `return_matches = TRUE`):

```{r inspect-record}
res_matches <- recordLinkage(census, census_anon,
                             key = c("age", "sex", "region", "educ"),
                             return_matches = TRUE)
ir <- inspect_record(res_matches, i = which.max(res_matches$per_record$risk),
                     data_orig = census, data_anon = census_anon)
print(ir)
```

### Variable importance

The `var_importance` field and `plot(x, which = 3)` show which
QIs contribute most to linkage success:

```{r var-importance}
res_real$var_importance
```
```{r var-importance-plot, fig.width=6, fig.height=4}
plot(res_real, which = 3)
```

### Risk band barplot

The risk band barplot (`which = 5`) gives a quick visual overview
of how records distribute across risk categories:

```{r risk-band-plot, fig.width=6, fig.height=4}
plot(res_real, which = 5)
```

### True-match rank distribution

The rank distribution (`which = 6`) shows where the true match
falls among candidates. A concentration at rank 1 means the
attacker's best guess is usually correct:

```{r rank-dist-plot, fig.width=6, fig.height=4}
plot(res_real, which = 6)
```

### Risk by subgroup

The risk-by-group barplot (`which = 7`) reveals which subpopulations
carry the most risk. Error bars show the per-group maximum:

```{r risk-by-group-plot, fig.width=6, fig.height=4}
plot(res_real, which = 7, group = "region", data = census)
```

### Lorenz curve and Gini coefficient

The Lorenz curve (`which = 8`) visualizes how concentrated risk is
among a few records. The Gini coefficient (0 = uniform, 1 = all
risk in one record) summarizes this:

```{r lorenz-plot, fig.width=5, fig.height=5}
plot(res_real, which = 8)
```
```{r risk-gini}
res_real$overall$risk_gini
```

### Merging risks back to data

Use `merge_per_record()` to join the risk diagnostics back to the
original data for further analysis or export:

```{r merge-per-record}
merged <- merge_per_record(res_real, data = census)
head(merged[order(merged$risk, decreasing = TRUE), ], 5)
```

Common remedial actions include:
(1) increasing perturbation strength for high-risk variables,
(2) applying local suppression to the most identifying QI values,
(3) adding noise to continuous variables, or
(4) microaggregating fine-grained variables.
In an iterative SDC workflow, the risk assessment is repeated after
each additional protection step until the risk profile is acceptable.

## Common Pitfalls

1. **All records flagged as high risk**: This often happens with
   fine-grained continuous QIs and the `nearest` strategy. With
   continuous variables, Gower distances are rarely tied, so each
   record has a unique nearest neighbor ($|C_i| = 1$), giving risk
   $r_i = 1$ whenever the true match is nearest (or $r_i = 0$ when
   it is not). This binary behavior masks the degree of protection.
   *Solution*: use `strategy = "topk"` with $k \ge 3$ to create
   candidate sets where weighting can differentiate, or use
   `strategy = "threshold"` to include all records within a distance,
   or bin continuous variables before linkage.

2. **No records flagged**: This happens with very few coarse categorical
   QIs (e.g., only sex and region). Many records share the same QI
   pattern, creating large equivalence classes that dilute risk.
   *Solution*: add more quasi-identifiers to increase discriminatory
   power, or use the probabilistic method (which weights by agreement
   rarity and can distinguish records within equivalence classes).

3. **Wrong variable types**: Factors treated as numeric produce
   meaningless scaled distances (factor levels are arbitrary integers);
   numeric variables treated as nominal ignore the ordering and distance
   structure. Always verify with `str()` and use the `type` parameter
   to override incorrect automatic detection.

4. **Blocking too aggressively**: Blocking restricts candidate search
   to records sharing the same blocking variable value. If the blocking
   variable is itself perturbed (e.g., region was swapped), the true
   match may end up in a different block and be missed entirely.
   Monitor `pct_true_in_set` -- if it drops below 80%, blocking may
   be too aggressive.

5. **Ignoring direction asymmetry in PRAM**: The transition matrix
   $T_v[a, b]$ is generally not symmetric, so the probability of
   observing output $b$ given input $a$ differs from the probability
   of input $a$ given observed output $b$. Forward and reverse risks
   can differ substantially. Always run both directions for a complete
   picture.

6. **PRAM with numeric variables**: The PRAM method requires transition
   matrices and is only applicable to categorical variables. For
   mixed-type data, use the deterministic or predictive method for
   numeric QIs, and optionally complement with PRAM for the categorical
   subset.

7. **Degenerate propensity scores**: When the predictive method
   produces propensity scores that all cluster near 0.5, the logistic
   model cannot separate original from anonymized records. This can
   occur with very few records, low-dimensional QIs, or when the
   perturbation preserves the covariate structure almost exactly.
   Inspect `res$propensity_info$propensity_original` -- if the
   standard deviation is less than 0.05, the risk estimates may be
   unreliable. Consider using the deterministic method instead.

8. **Overconfidence in `privacy_pass`**: A mean risk of 0.098 passes
   the default 0.1 threshold, but this does not guarantee adequate
   protection -- individual records may have $r_i = 1$ (unique match).
   Always examine the full risk distribution, including `max_risk`,
   the risk band table, and the number of unique matches.

## Relation to Other Risk Metrics

| Metric | Risk type | Data type | Package / function |
|:-------|:----------|:----------|:-------------------|
| `recordLinkage()` | Identity disclosure | Perturbative | riskutility |
| `dRisk()` / `dRiskRMD()` | Identity disclosure | Perturbative | sdcMicro |
| TCAP / DCAP | Attribute disclosure | Synthetic | riskutility |
| DCR / NNDR | Memorization | Generative models | riskutility |
| RAPID | Attribute prediction | Synthetic / anonymized | riskutility |
| `individual_risk()` | Pre-anonymization baseline | Any | riskutility |
| `population_uniqueness()` | Population-level risk | Any | riskutility |

These metrics address different disclosure scenarios and are
complementary rather than competing. Record linkage risk
(`recordLinkage()`) measures *identity disclosure*: can an attacker
match a released record to the original individual? This is the primary
concern for perturbative methods (PRAM, swapping, noise) where
records correspond to real individuals. TCAP and DCAP measure
*attribute disclosure*: can an attacker infer a sensitive attribute
from quasi-identifiers? These are most relevant for synthetic data,
where individual records may not correspond to real people but sensitive
patterns may still leak. DCR and NNDR detect *memorization* in
generative models by comparing distances between training and generated
records. RAPID bridges the gap: it uses machine learning to assess
attribute inference risk and can be applied to both synthetic and
perturbative data. For a comprehensive assessment, combine identity
disclosure metrics (such as `recordLinkage()`) with attribute disclosure
metrics (such as TCAP or RAPID) and utility measures to understand the
full risk-utility trade-off.


# Bijective Matching (GDBRL)

## Independent vs. Bijective Matching

By default, `recordLinkage()` scores each query record independently:
the attacker selects the best candidate for record $i$ without regard
to what candidate was chosen for record $j$. This "independent" mode
corresponds to the classical distance-based record linkage (DBRL)
framework of @domingoferrer2003disclosure.

@herranz2016gdbrl introduced *Global DBRL* (GDBRL), which models
a stronger attacker who enforces a one-to-one (bijective) assignment
between original and anonymized records. The attacker solves a global
optimization problem---minimizing total cost across all records
simultaneously---using the Hungarian algorithm (linear sum assignment).
Because the bijective constraint prevents multiple original records from
being matched to the same anonymized record, the resulting risk
estimates are typically higher than independent scoring: the attacker
cannot "waste" the best candidates on multiple queries.

Setting `matching = "bijective"` activates this mode. Risk becomes
binary: 1 if the Hungarian algorithm assigns a record to its true match,
0 otherwise. The `clue` package is required.

## Comparison Example

```{r bijective-comparison, cache=TRUE}
# Independent matching (default) -- reuses census / census_anon from above
res_ind <- recordLinkage(census, census_anon,
                         key = c("age", "sex", "region", "educ"),
                         matching = "independent")

# Bijective matching (GDBRL)
res_bij <- recordLinkage(census, census_anon,
                         key = c("age", "sex", "region", "educ"),
                         matching = "bijective")

cat("Independent:  mean risk =", round(res_ind$overall$mean_risk, 4),
    " | high-risk records:", res_ind$overall$n_high_risk, "\n")
cat("Bijective:    mean risk =", round(res_bij$overall$mean_risk, 4),
    " | high-risk records:", res_bij$overall$n_high_risk, "\n")
```

The bijective attacker typically achieves a higher re-identification
rate because the one-to-one constraint forces globally optimal
assignments. This is a worst-case scenario: in practice, an attacker
rarely has the full membership knowledge and computational resources
required for GDBRL. Nevertheless, reporting both independent and
bijective risk provides a useful bracket on the true risk level.

Bijective matching works with all four methods and with blocking. When
blocking is enabled, the Hungarian algorithm runs independently within
each block:

```{r bijective-blocking, cache=TRUE}
res_blk <- recordLinkage(census, census_anon,
                         key = c("age", "sex", "region", "educ"),
                         block = "region",
                         matching = "bijective")
cat("Bijective + blocking: mean risk =",
    round(res_blk$overall$mean_risk, 4), "\n")
```


# Discussion and Conclusion

The `recordLinkage()` function provides a unified framework for
assessing re-identification risk in perturbed microdata. The four methods
cover a range of scenarios: deterministic linkage for general-purpose
assessment, probabilistic linkage for formal audits, PRAM linkage when
the perturbation mechanism is known, and predictive linkage for
high-dimensional or mixed-type quasi-identifiers.

## Key Takeaways

The deterministic method provides a robust starting point for any
risk assessment -- it requires no knowledge of the perturbation
mechanism and gives a reasonable baseline. When additional information
is available (transition matrices for PRAM, m/u probabilities for
formal audits, or complex interaction patterns for the predictive
method), the specialized methods can refine the assessment. The choice
of risk weighting scheme matters: uniform weighting is conservative
(equal weight to all candidates), while softmax and kernel weighting
concentrate risk on the closest candidate, producing more nuanced risk
profiles. In particular, the `topk` or `threshold` strategies are
needed to make weighting differences visible, since the default
`nearest` strategy often yields singleton candidate sets. Both forward
and reverse directions should be considered: forward risk assesses how
well each individual is protected, while reverse risk assesses how
disclosive each released record is. Finally, always verify that
`pct_true_in_set` is high (ideally above 80%) to ensure that the
candidate sets are meaningful -- if the true match is rarely among the
candidates, risk estimates are uninformative.

## Limitations

Several limitations should be noted:

1. **Stylized attacker model**: The function assumes perfect membership
   knowledge and exact QI knowledge. Real attackers may have noisy or
   incomplete information, so the computed risk is an upper bound.
2. **Computational scaling**: Without blocking, all methods require
   $O(nm)$ pairwise comparisons, which limits applicability to datasets
   with more than approximately 50,000 records.
3. **Conditional independence**: The probabilistic method assumes
   conditional independence of agreement across variables, which is
   approximate when QIs are correlated.
4. **Predictive method**: The propensity-score approach to record
   linkage risk assessment is, to our knowledge, not yet established
   in the SDC literature. While it draws on well-understood methodology
   [@rosenbaum1983propensity], its properties for risk quantification
   (calibration, power) require further empirical investigation.
5. **Quasi-identifier selection**: Risk estimates are conditional on
   the assumed set of quasi-identifiers. Adding or removing a single
   QI can dramatically change risk levels. Practitioners must make
   judgment calls about which variables an attacker might know, and
   should consider running the assessment with multiple QI sets to
   gauge sensitivity.
6. **Sample vs. census**: The current implementation computes
   sample-level risk. For sample surveys where the attacker does not
   know whether the target is in the dataset, population-level
   adjustments [@franconi2004individual] are needed.

## Future Directions

Planned extensions include integration with Franconi-Polettini
population frequency estimation [@franconi2004individual] for
population-level risk, rank-based record linkage
[@muralidhar2016rankbased] as an alternative to distance-based matching
(which can outperform Gower-based linkage for certain masking methods
such as rank swapping), household-level matching for hierarchical
microdata, and attribute disclosure assessment (extending the function
to evaluate whether an attacker could infer sensitive attributes after
successful linkage). Performance improvements through parallel
computation and approximate nearest-neighbor search are also planned
for large-scale applications.


# References

<div id="refs"></div>


# Session Info

The results in this vignette were obtained with the following
configuration.

```{r sessioninfo}
sessionInfo()
```
