---
title: "riskutility: Record Linkage Risk Assessment for Perturbative Disclosure Control"
author:
  - name: Matthias Templ
    affiliation: "Institute for Competitiveness and Communication,
      School of Business, FHNW University of Applied Sciences and Arts
      Northwestern Switzerland; Swiss Data Anonymization Center
      (www.swissanon.ch)"
    email: matthias.templ@gmail.com
    orcid: 0000-0002-8638-5276
abstract: >
  The **riskutility** package for R provides a unified framework for
  measuring re-identification risk in perturbed microdata. The
  `recordLinkage()` function implements deterministic (Gower distance) and
  probabilistic (Fellegi-Sunter) linkage under a single S3 interface with
  consistent risk weighting, blocking, and forward/reverse direction support.
  While several R packages provide record linkage functionality
  (RecordLinkage, fastLink, reclin2) and sdcMicro offers distance-based
  risk measures, no existing package combines multiple linkage methods with
  flexible risk weighting and a focus on disclosure risk quantification
  for perturbative SDC. This paper
  presents the methodology with mathematical derivations, hand-traceable
  toy examples, and realistic demonstrations using sdcMicro for
  perturbation. Practical guidance on method selection, risk
  interpretation, and common pitfalls is provided.
keywords: [record linkage, re-identification risk, statistical disclosure control,
  perturbative methods, Fellegi-Sunter, R]
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{riskutility: Record Linkage Risk Assessment for Perturbative Disclosure Control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5,
  warning = FALSE, message = FALSE
)
if (!requireNamespace("sdcMicro", quietly = TRUE))
  stop("This vignette requires the sdcMicro package. ",
       "Install it with install.packages('sdcMicro').")
# Attach sdcMicro BEFORE riskutility. sdcMicro (>= 5.8.2) also exports a
# recordLinkage() (its own GDBRL implementation), so it would otherwise mask
# riskutility::recordLinkage() -- the subject of this vignette -- on the search
# path. Loading riskutility last makes its recordLinkage() win; sdcMicro's
# rankSwap() (used below) remains available.
library(sdcMicro)
library(riskutility, warn.conflicts = FALSE)

```

# Introduction

*Prerequisites*: This vignette assumes familiarity with microdata and
basic SDC concepts (quasi-identifiers, perturbation). Knowledge of R
at the level of running functions and reading data frames is sufficient.

## Perturbative SDC and Record Linkage Risk

Statistical disclosure control (SDC) encompasses a range of techniques for
protecting individual privacy in released microdata
[@hundepool2012statistical]. *Perturbative* methods -- such as record
swapping, noise addition, and microaggregation -- modify the original
records rather than generating entirely new ones [@templ2015sdcmicro]. This makes perturbative data
fundamentally different from fully synthetic data: every released record
has a real individual behind it, and the attacker's task is to undo the
perturbation and re-identify that individual.

A *quasi-identifier* (QI), also called a *key variable* (parameter `key`
in the R interface), is a variable that is not a direct identifier (such
as a name or social security number) but can be used in combination with
other variables to narrow down the identity of a record holder. Typical
quasi-identifiers include age, sex, region, education, and occupation. The set of quasi-identifiers available to an attacker
determines the scope of a record linkage attack
[@skinner2002measure; @elliot2018functional].

Most disclosure risk functions in **riskutility** (DCAP, TCAP, DCR)
target *synthetic* data, where the question is whether the generator
memorized training records. RAPID can be applied to both synthetic
and traditionally anonymized data (it trains on the released data and
evaluates predictions on the original), but was primarily designed for
the synthetic case. For perturbative data, the more natural question is:
*can an attacker link a released record back to the original
individual?* This is the classical problem of
identity disclosure [@dewolf2002framework; @reiter2005estimating],
and the standard approach to its empirical assessment is record linkage
[@domingoferrer2003disclosure].
The `recordLinkage()` function addresses exactly this question.

## The Attacker Model

The attacker scenario assumes:

- **Membership knowledge**: the attacker knows that the target individual
  has a record in the released dataset.
- **Exact QI knowledge**: the attacker knows the target's quasi-identifier
  values (e.g., age, sex, region) from an external source.
- **Linking rule**: the attacker applies a strategy (nearest match,
  threshold, top-$k$) to select candidates from the released data and
  attributes re-identification risk accordingly.

Ground truth for evaluation is taken from row alignment
(`truth = "row"`) or a shared identifier column (`truth = "id"`).
Note that `truth` affects only the *evaluation* of success; it is not
an attacker capability.

For example, an attacker with access to a voter registration list
(containing age, sex, and city) attempts to link those records to a
published health survey to learn about medical conditions.

These assumptions represent a *worst-case* scenario. In practice,
attackers rarely have perfect membership knowledge (they may not know
whether the target is in the dataset) or exact QI values (external
sources may be noisy or outdated). The computed risk should therefore
be interpreted as an *upper bound* on the re-identification probability
under the assumed attacker model. For population-level risk assessment
that accounts for unknown membership, see @franconi2004individual and
the `population_uniqueness()` function in **riskutility**.

## Related Software

Several R packages address record linkage or disclosure risk:

- **sdcMicro** [@templ2015sdcmicro] provides `dRisk()` and `dRiskRMD()`
  for distance-based record linkage risk [@templ2008robust], along with
  comprehensive SDC methods. These functions compute risk based on the
  proportion of records closer to their match than to any other record
  (using Euclidean or robust Mahalanobis distances for numeric data),
  but support only the deterministic (distance-based) approach and do
  not offer flexible risk weighting or candidate-set strategies.
- **RecordLinkage** [@sariyar2010recordlinkage] implements
  Fellegi-Sunter probabilistic linkage with EM-based weight estimation,
  designed for deduplication and data integration rather than disclosure
  risk quantification.
- **fastLink** [@enamorado2019fastlink] provides fast probabilistic
  record linkage with blocking and EM estimation, optimized for merging
  large administrative datasets.
- **reclin2** [@reclin2] offers a modular record linkage pipeline with
  blocking, comparison, and scoring stages.

These packages focus primarily on *linking* records (classification into
match/non-match) rather than *quantifying re-identification risk* per
record. The `recordLinkage()` function in **riskutility** fills this gap
by combining six linkage methods with risk-focused output: per-record
risk scores, candidate set diagnostics, and configurable risk weighting.

| Feature | sdcMicro | RecordLinkage | fastLink | riskutility |
|:--------|:--------:|:-------------:|:--------:|:-----------:|
| Deterministic (distance) | Yes | -- | -- | Yes |
| Probabilistic (Fellegi-Sunter) | -- | Yes | Yes | Yes |
| Embedding (autoencoder) | -- | -- | -- | Yes |
| RF proximity | -- | -- | -- | Yes |
| Rank-based (RBRL) | -- | -- | -- | Yes |
| Mahalanobis (robust) | Yes | -- | -- | Yes |
| Per-record risk scores | Partial[^1] | -- | -- | Yes |
| Risk weighting schemes | -- | -- | -- | Yes |
| Blocking | -- | Yes | Yes | Yes |
| Forward/reverse direction | -- | -- | -- | Yes |

[^1]: sdcMicro provides per-record risk via `individualRiskDT()` for frequency-based measures and `dRisk()` / `dRiskRMD()` for distance-based global proportions; however, it does not return continuous per-record record-linkage risk scores with candidate set diagnostics.

## Scope of This Paper

The `recordLinkage()` function in **riskutility** implements two
linkage methods under a single S3 interface with consistent risk
weighting, blocking, and direction support. This paper covers both
methods in detail, each following the same pedagogical structure:
*mathematical formulation* $\to$ *hand-traceable toy example* $\to$
*realistic demonstration*:

1. **Deterministic** (Gower distance)
2. **Probabilistic** (Fellegi-Sunter log-likelihood ratios)

The remaining sections compare methods and weighting schemes
side by side, present computational details of the implementation,
discuss advanced features (blocking, direction, ground truth), and
provide practical guidance for method selection and risk interpretation.

For the theoretical background on record linkage see @fellegi1969theory,
@winkler2006overview, and @domingo2002validating; for empirical
disclosure risk assessment via record linkage see
@domingoferrer2003disclosure; for the
general framework of record linkage risk see @dewolf2002framework; for a
textbook treatment of SDC risk estimation see @templ2017statistical
(Chapter 6).


# Mathematical Framework

## Notation

Let $X = \{x_1, \ldots, x_n\}$ denote the original microdata and
$X^a = \{x^a_1, \ldots, x^a_m\}$ the anonymized (perturbed) data.
In the typical case $m = n$ and record $x^a_i$ is the perturbed
version of $x_i$.

We denote:

- $\mathcal{K} = \{v_1, \ldots, v_p\}$ -- the set of quasi-identifier
  variables (keys) available to the attacker.
- $w_v \ge 0$ -- the weight assigned to variable $v$.
- $d(i, j)$ -- the distance (or dissimilarity) between original record
  $x_i$ and anonymized record $x^a_j$ over the quasi-identifiers
  (QIs).
  For the probabilistic method this becomes a log-likelihood ratio
  $\Lambda_{ij}$.
- $C_i$ -- the candidate set of anonymized records that the attacker
  considers as potential matches for $x_i$.
- $j^*_i$ -- the index of the true match of $x_i$ in the anonymized
  data ($j^*_i = i$ when rows are aligned).

## Re-identification Risk

The re-identification risk quantifies how likely an attacker is to
correctly identify an individual's record in the anonymized data using
only the quasi-identifiers $\mathcal{K}$.

Given a candidate set $C_i$ with weights $\pi_j \ge 0$,
$\sum_{j \in C_i} \pi_j = 1$, the re-identification risk for record $i$
is:
$$r_i = \sum_{j \in C_i} \pi_j \cdot \mathbf{1}[j = j^*_i].$$
Since the indicator is nonzero for at most one term, this simplifies to
$r_i = \pi_{j^*_i}$ if the true match $j^*_i \in C_i$, and $r_i = 0$
otherwise. When there is a single candidate and it is the true match,
$r_i = 1$.

The overall risk is summarized by the mean $\bar{r} = n^{-1}\sum_i r_i$,
the maximum $r_{\max}$, and the fraction of records exceeding a threshold
(typically 0.1).

The formulation above takes the *forward* perspective: for each original
record, search the anonymized data for matches. The *reverse*
perspective iterates over anonymized records and searches the original
data, answering "how easily can each released record be traced back
to an individual?" For symmetric
distance methods (deterministic), the two perspectives yield similar but
not identical results because the evaluation criterion changes (different
records serve as query vs. search).

## Risk Weighting Schemes

The choice of candidate weights $\pi_j$ controls how risk is attributed.
Three schemes are available; their effect is demonstrated empirically in
the method comparison section below.

**Uniform** (`risk_weighting = "uniform"`): Each candidate receives
equal weight:
$$\pi_j = \frac{1}{|C_i|}.$$
This is conservative: if three candidates are equally close and one is
the true match, the risk is $1/3$ regardless of distance.

**Softmax** (`risk_weighting = "softmax"`): Closer candidates receive
exponentially higher weight:
$$\pi_j = \frac{\exp(-\kappa \cdot d_j)}{\sum_{k \in C_i} \exp(-\kappa \cdot d_k)},$$
where $d_j \equiv d(i,j)$ is the distance from query record $i$ to
candidate $j$, and $\kappa > 0$ is a temperature parameter. If not
supplied, $\kappa$ is auto-calibrated as $2 / (\max(d) - \min(d))$,
so that the ratio of weights for the closest vs. farthest candidate is
approximately $e^2 \approx 7.4$ -- a moderate concentration that avoids
both winner-take-all and near-uniform behavior. As $\kappa \to \infty$
this approaches winner-take-all; as $\kappa \to 0$ it converges to
uniform.

**Kernel** (`risk_weighting = "kernel"`): Weights are proportional to a
kernel function evaluated at the scaled distance:
$$\pi_j = \frac{K(d_j / h)}{\sum_{k \in C_i} K(d_k / h)},$$
where $h$ is the bandwidth and $K$ is a kernel function.
If $h$ is not supplied, it is auto-selected via Silverman's rule of
thumb (a standard bandwidth selector from kernel density estimation),
applied heuristically to the candidate distances:
$$h = 0.9 \cdot \min\!\left(\hat\sigma,\, \frac{\widehat{\text{IQR}}}{1.34}\right) \cdot |C_i|^{-1/5}.$$
Note that this rule was designed for density estimation with larger
samples; for small candidate sets (e.g., $|C_i| \le 5$), the bandwidth
may be suboptimal. Users can supply a fixed $h$ via the `kernel_h`
parameter to override auto-selection.
The available kernels are:

- *Gaussian*: $K(u) = \exp(-u^2/2)$
- *Epanechnikov*: $K(u) = \max(0, 1 - u^2) \cdot 3/4$
- *Tricube*: $K(u) = \max(0, (1 - |u|^3)^3) \cdot 70/81$

With compact kernels (Epanechnikov, tricube), candidates beyond the
bandwidth receive exactly zero weight and are effectively excluded.
The effect of the bandwidth and kernel choice is demonstrated
empirically in the Weighting Scheme Comparison subsection below; readers
unfamiliar with kernel density estimation may skip these details on
first reading.


# Deterministic Method (Gower Distance)

The simplest approach to record linkage treats it as a matching problem:
for each original record, find the most similar record in the anonymized
data and assess whether it is the true match
[@domingoferrer2003disclosure; @domingo2002validating]. This requires no
knowledge of the perturbation mechanism -- only a sensible distance
metric -- making it the natural starting point for risk assessment.

## Weighted Gower Distance

The deterministic method computes a weighted Gower-type distance over the
quasi-identifiers [@gower1971general]. For original record $x_i$ and
anonymized record $x^a_j$:

$$d(i,j) = \frac{\sum_{v \in \mathcal{K}} w_v \, \delta_v(i,j)}{\sum_{v \in \mathcal{K}} w_v},$$

where the variable-specific dissimilarities are:

- **Numeric / ordinal**: $\delta_v(i,j) = |x_{iv} - x^a_{jv}| / R_v$,
  where $R_v = \max(x_v \cup x^a_v) - \min(x_v \cup x^a_v)$ is the
  range computed over both datasets. Ordinal variables (ordered factors
  in R) are treated identically to numeric variables: the integer codes
  of the ordered levels are used as numeric values, preserving the
  ordering. Use the `type` parameter to explicitly set a variable as
  `"ordinal"` if automatic detection is incorrect.
- **Nominal**: $\delta_v(i,j) = \mathbf{1}[x_{iv} \ne x^a_{jv}]$

The variable weights $w_v$ control the relative importance of each
quasi-identifier in the distance computation. By default, all weights
are equal ($w_v = 1$ for all $v$). Setting a higher weight for a variable
increases its influence on the distance -- for example, setting
$w_{\text{age}} = 2$ would make age twice as influential as variables
with $w_v = 1$. Weights can be specified via the `weights` parameter in
`recordLinkage()`. In practice, weights can reflect the attacker's
likely knowledge quality (higher for variables known with certainty,
lower for noisily known variables) or regulatory importance. In most
cases, equal weights are a reasonable default.

When missing values are present, the summation in both numerator and
denominator is restricted to variables with valid observations for both
records. The resulting distance $d(i,j) \in [0, 1]$, where 0 indicates perfect
agreement and 1 maximal disagreement.

## Toy Example: 6 Records

Consider a small dataset with three quasi-identifiers -- `age` (numeric),
`sex` (nominal), and `city` (nominal) -- and a sensitive attribute
`salary` that is *not* used in the linkage.

```{r det-toy-data}
orig <- data.frame(
  age    = c(30, 40, 50, 60, 70, 80),
  sex    = factor(c("M", "M", "F", "F", "M", "F")),
  city   = factor(c("A", "B", "A", "B", "A", "B")),
  salary = c(35000, 48000, 52000, 67000, 71000, 45000)
)
anon <- data.frame(
  age    = c(30, 60, 50, 40, 70, 80),  # rows 2 and 4: ages swapped (40<->60)
  sex    = factor(c("M", "M", "F", "F", "M", "F")),
  city   = factor(c("A", "B", "A", "B", "A", "B")),
  salary = c(35000, 67000, 52000, 48000, 71000, 45000)
)
knitr::kable(cbind(Row = 1:6, orig), caption = "Original data")
knitr::kable(cbind(Row = 1:6, anon), caption = "Anonymized data (rows 2 and 4 swapped)")
```

The perturbation is record swapping: rows 2 and 4 had their
ages swapped (40 $\leftrightarrow$ 60), and consequently their
salaries follow (48,000 $\leftrightarrow$ 67,000).
Rows 1, 3, 5, and 6 are unchanged.

The range of `age` is $80 - 30 = 50$. Each variable has equal weight
$w_v = 1$ (the default), so $\sum_v w_v = 3$. Only the three
quasi-identifiers (`age`, `sex`, `city`) enter the linkage; `salary`
is a sensitive attribute and is *not* used.

**Record 1** (age=30, sex=M, city=A) vs. anonymized record 1
(age=30, sex=M, city=A):

$$\delta_{\text{age}} = |30-30|/50 = 0, \quad
\delta_{\text{sex}} = 0, \quad
\delta_{\text{city}} = 0$$
$$d(1,1) = (0 + 0 + 0)/3 = 0.$$

An exact match -- record 1 was not perturbed, so it is the unique
nearest neighbor and $r_1 = 1.0$.

**Record 4** (age=60, sex=F, city=B) -- this record's age was swapped from
60 to 40. How does the attacker fare?

```{r det-toy-distances, echo=FALSE}
# Hand-computed distances for record 4
d4 <- data.frame(
  `Anon row` = 1:6,
  age_dissim = c(abs(60-30)/50, abs(60-60)/50, abs(60-50)/50,
                 abs(60-40)/50, abs(60-70)/50, abs(60-80)/50),
  sex_dissim = c(1, 1, 0, 0, 1, 0),
  city_dissim = c(1, 0, 1, 0, 1, 0)
)
d4$Gower_d <- round((d4$age_dissim + d4$sex_dissim + d4$city_dissim) / 3, 4)
names(d4) <- c("Anon row", "age", "sex", "city", "d(4,j)")
knitr::kable(d4, digits = 4,
             caption = "Variable dissimilarities and Gower distance for original record 4")
```

The two smallest distances are $d(4,4) = 0.1333$ and $d(4,6) = 0.1333$
(a tie). Why? Both anonymized records 4 and 6 are female from city B
(matching the original), and their ages (40 and 80) are equidistant
from 60 ($|60-40| = |60-80| = 20$). With the `nearest` strategy and
uniform weighting, the candidate set is $C_4 = \{4, 6\}$. Since the true
match $j^*_4 = 4$ is in the set, $r_4 = 1/2 = 0.5$. The age swap
halved the risk by creating an ambiguity the attacker cannot resolve.

**Pedagogical point**: nominal mismatches are expensive. A city
mismatch produces a dissimilarity of $\delta = 1.0$, equivalent to
spanning the entire age range ($50/50 = 1.0$). Even a moderate city
change contributes as much to the Gower distance as a 50-year age
difference. This makes nominal variables highly influential in the
linkage.

Now verify with `recordLinkage()`:

```{r det-toy-verify}
res_det <- recordLinkage(orig, anon, key = c("age", "sex", "city"))
knitr::kable(res_det$per_record[, c("risk", "cand_n", "true_in_set", "d_true")],
             digits = 4, caption = "Per-record risk (deterministic, nearest)")
```

Note that `salary` was *not* used as a quasi-identifier -- it
represents a sensitive attribute the attacker wants to learn. When
record 1 has $r_1 = 1.0$, the attacker uniquely identifies this person
*and* can read off their salary (35,000) from the anonymized data. This
is *attribute disclosure*: successful re-identification reveals all
non-QI attributes. By contrast, record 4's risk of 0.5 means the
attacker cannot distinguish between anonymized rows 4 and 6, whose
salaries are 48,000 and 45,000 -- the swapping provided some protection.
This illustrates why identity disclosure risk (measured by
`recordLinkage()`) is a precondition for attribute disclosure.

## Attacker Strategies

The `strategy` parameter controls how the attacker selects candidates.
We compare three strategies on the same toy data:

```{r det-toy-strategies}
res_nearest   <- recordLinkage(orig, anon, key = c("age", "sex", "city"),
                               strategy = "nearest")
res_threshold <- recordLinkage(orig, anon, key = c("age", "sex", "city"),
                               strategy = "threshold", threshold = 0.2)
res_topk      <- recordLinkage(orig, anon, key = c("age", "sex", "city"),
                               strategy = "topk", k = 3)
strat_df <- data.frame(
  Record = 1:6,
  nearest   = round(res_nearest$per_record$risk, 4),
  threshold = round(res_threshold$per_record$risk, 4),
  topk_3    = round(res_topk$per_record$risk, 4)
)
knitr::kable(strat_df, caption = "Risk under different attacker strategies")
```

With `threshold = 0.2`, only exact or near-exact matches qualify, so
records with moderate perturbation get zero risk. With `topk` and
$k = 3$, the candidate set always includes at least 3 records, diluting
risk for unique matches.

## Real-World Example (n = 500)

With 6 records, every record is a candidate for every other record and
ties are common. At scale, ties become rare and the nearest-neighbor
match is typically unique, so the `nearest` strategy often gives
$r_i = 1$ (exact match found) or $r_i = 0$ (true match not the
nearest). This motivates the weighting schemes introduced above.

We generate a realistic census-like dataset with 500 records and apply
perturbation using **sdcMicro** [@templ2015sdcmicro]: rank swapping for
the numeric variables (`age`, `income`) while the categorical variables
(`sex`, `region`, `educ`) remain unchanged. This is a typical workflow
in statistical disclosure control.

```{r det-realdata}
set.seed(123)  # ensures reproducibility for data generation
n <- 500
census <- data.frame(
  age    = sample(18:85, n, replace = TRUE),
  sex    = factor(sample(c("M", "F"), n, replace = TRUE)),
  region = factor(sample(paste0("R", 1:5), n, replace = TRUE)),
  educ   = factor(sample(c("primary", "secondary", "tertiary"), n,
                         replace = TRUE, prob = c(0.2, 0.5, 0.3))),
  income = round(rlnorm(n, meanlog = 10.5, sdlog = 0.6))
)

# Perturbation via sdcMicro: rank-swap numeric variables
census_anon <- census
census_num <- census[, c("age", "income")]
census_swapped <- rankSwap(census_num, variables = c("age", "income"),
                           P = 10, seed = 123)
census_anon$age    <- census_swapped$age
census_anon$income <- census_swapped$income

# Record linkage with 4 QIs
res_real <- recordLinkage(census, census_anon,
                          key = c("age", "sex", "region", "educ"))
print(res_real)
```

```{r det-realdata-summary}
summary(res_real)
```

Note the privacy assessment: the `privacy_pass` flag tests only the
*mean* risk against the default threshold of 0.1. A PASS does not
guarantee that individual records are protected -- `max_risk` may be
1.0, indicating that some records are uniquely re-identifiable. Always
examine `max_risk`, `n_high_risk`, and the risk band distribution
alongside the pass/fail label. A custom threshold can be set via
the `risk_threshold` parameter. For a full discussion, see
the Interpreting Risk Bands subsection in Practical Guidance.

```{r det-realdata-plot, fig.height=4, fig.cap="Distribution of per-record re-identification risk (deterministic method). The red dashed line indicates the risk threshold (0.1)."}
plot(res_real, which = 1)
```

The histogram reveals the risk distribution. Records whose rank-swapped
age is close to the original (small rank displacement) retain high risk,
while records swapped with a more distant value are harder to
re-identify.

```{r det-realdata-scatter, fig.height=4, fig.cap="Distance to true match vs. attributed risk. Records with zero distance (no perturbation effect on QIs) have maximum risk."}
plot(res_real, which = 2)
```

The scatter plot shows the relationship between the distance to the true
match and the attributed risk. Records with $d_{\text{true}} = 0$ (no
perturbation effect on these QIs) have maximum risk; as distance
increases, risk decreases because more competing candidates exist.


# Probabilistic Method (Fellegi-Sunter)

The deterministic method treats all variable agreements equally.
In practice, however, agreement on a rare variable value (e.g., an
uncommon occupation) provides much stronger evidence of a true match
than agreement on a common value (e.g., sex). The probabilistic method
addresses this by weighting agreements according to their informativeness.

## Log-Likelihood Ratios

The Fellegi-Sunter framework [@fellegi1969theory] models record pairs
as belonging to one of two classes: matches ($M$) or non-matches ($U$).
For each quasi-identifier $v$, the method estimates:

- $m_v = P(\text{agree on } v \mid \text{true match})$
- $u_v = P(\text{agree on } v \mid \text{non-match})$

The composite log-likelihood ratio for the pair $(i, j)$ is:
$$\Lambda_{ij} = \sum_{v \in \mathcal{K}} \left[\gamma_{ijv} \log\frac{m_v}{u_v} + (1 - \gamma_{ijv}) \log\frac{1-m_v}{1-u_v}\right],$$
where $\gamma_{ijv} = \mathbf{1}[\text{agree on } v]$. High $\Lambda$
indicates strong evidence of a match.

The standard Fellegi-Sunter model assumes *conditional independence* of
agreement indicators across variables, given the match/non-match status.
When quasi-identifiers are positively correlated (e.g., age and
education), agreement indicators become positively correlated given match
status, and the composite log-likelihood ratio typically *overestimates*
the evidence for a match -- the model double-counts shared information.
EM-based approaches [@winkler2006overview] can relax this assumption by
estimating latent match/non-match classes, but the current implementation
uses the standard model for simplicity and transparency.

The key insight is that *rare agreements are more informative*: if a
variable has many categories (low $u_v$), agreement provides stronger
evidence of a true match [@winkler2006overview].

## Estimation of m and u

When `m_probs` and `u_probs` are not supplied, `recordLinkage()` estimates
them internally:

- For **nominal** variables: $m_v$ is estimated from agreement rates among
  aligned pairs; $u_v$ from random pairings.
- For **numeric** variables: agreement is defined via a tolerance based
  on the Median Absolute Deviation (MAD). Two values agree if
  $|x_{iv} - x^a_{jv}| \le \text{MAD}(x_v)$, where
  $\text{MAD}(x_v) = 1.4826 \cdot \text{median}(|x_v - \text{median}(x_v)|)$
  is computed on the original data. MAD is preferred over the standard
  deviation for robustness to outliers.

User-supplied values via `m_probs` and `u_probs` override the estimation
and are recommended when the perturbation mechanism is well understood.

## Toy Example: 5 Records (All Categorical)

We use a new all-categorical toy example here to avoid the complication
of numeric tolerance thresholds (which the deterministic section's
6-record dataset with numeric `age` would require). The Fellegi-Sunter
model defines agreement/disagreement per variable, and for numeric
variables this requires a tolerance parameter that would obscure the
core mechanism.

```{r prob-toy-data}
orig_p <- data.frame(
  sex    = factor(c("M", "F", "M", "F", "M")),
  region = factor(c("N", "S", "E", "W", "N")),
  job    = factor(c("A", "B", "C", "A", "D"))
)
anon_p <- data.frame(
  sex    = factor(c("M", "F", "F", "F", "M")),
  region = factor(c("N", "S", "E", "N", "S")),
  job    = factor(c("A", "B", "C", "B", "D"))
)
knitr::kable(cbind(Row = 1:5, orig_p), caption = "Original data (probabilistic toy)")
knitr::kable(cbind(Row = 1:5, anon_p), caption = "Anonymized data (probabilistic toy)")
```

We supply explicit m/u probabilities for full transparency:

```{r prob-toy-mu}
m <- c(sex = 0.95, region = 0.80, job = 0.85)
u <- c(sex = 0.50, region = 0.25, job = 0.10)
```

The per-variable log-likelihood ratios are:

```{r prob-toy-logLR}
log_agree    <- round(log(m / u), 3)
log_disagree <- round(log((1 - m) / (1 - u)), 3)
lr_table <- data.frame(
  Variable = names(m),
  m = m, u = u,
  `log(m/u)` = log_agree,
  `log((1-m)/(1-u))` = log_disagree,
  check.names = FALSE
)
knitr::kable(lr_table, digits = 3,
             caption = "Per-variable log-likelihood ratios")
```

**Hand-tracing record 1** (M, N, A) against all 5 anonymized records:

```{r prob-toy-handtrace, echo=FALSE}
# Record 1: sex=M, region=N, job=A
# Anon records: (M,N,A), (F,S,B), (F,E,C), (F,N,B), (M,S,D)
agree_matrix <- matrix(c(
  1, 1, 1,  # anon 1: all agree
  0, 0, 0,  # anon 2: none agree
  0, 0, 0,  # anon 3: none agree
  0, 1, 0,  # anon 4: region agrees
  1, 0, 0   # anon 5: sex agrees
), nrow = 5, byrow = TRUE, dimnames = list(
  paste0("Anon ", 1:5),
  c("sex", "region", "job")
))

# Compute Lambda for each pair
lambda <- numeric(5)
for (j in 1:5) {
  lambda[j] <- sum(agree_matrix[j, ] * log(m / u) +
                   (1 - agree_matrix[j, ]) * log((1 - m) / (1 - u)))
}
ht <- data.frame(
  `Anon row` = 1:5,
  sex = ifelse(agree_matrix[, 1] == 1, "agree", "disagree"),
  region = ifelse(agree_matrix[, 2] == 1, "agree", "disagree"),
  job = ifelse(agree_matrix[, 3] == 1, "agree", "disagree"),
  Lambda = round(lambda, 3),
  check.names = FALSE
)
knitr::kable(ht, caption = "Log-LR scores for original record 1 vs all anonymized records")
```

For anonymized record 1 (all three variables agree):
$\Lambda = 0.642 + 1.163 + 2.140 = 3.945$. This is the highest score.
With `fs_threshold = 0` (the default -- include all pairs with non-negative
log-likelihood ratio), only anonymized record 1 qualifies, so
$r_1 = 1.0$.

**Key insight**: job agreement contributes $\log(0.85/0.10) = 2.140$ --
more than 3 times the contribution of sex agreement
($\log(0.95/0.50) = 0.642$) -- because rare agreements carry much
stronger evidence. The Fellegi-Sunter framework automatically accounts
for this.

**Hand-tracing record 4** (F, W, A) -- note that region "W" does not
appear in the anonymized data:

```{r prob-toy-rec4, echo=FALSE}
# Record 4: sex=F, region=W, job=A
agree4 <- matrix(c(
  0, 0, 1,  # anon 1: job agrees
  1, 0, 0,  # anon 2: sex agrees
  1, 0, 0,  # anon 3: sex agrees
  1, 0, 0,  # anon 4: sex agrees
  0, 0, 0   # anon 5: none
), nrow = 5, byrow = TRUE, dimnames = list(
  paste0("Anon ", 1:5),
  c("sex", "region", "job")
))
lambda4 <- numeric(5)
for (j in 1:5) {
  lambda4[j] <- sum(agree4[j, ] * log(m / u) +
                    (1 - agree4[j, ]) * log((1 - m) / (1 - u)))
}
ht4 <- data.frame(
  `Anon row` = 1:5,
  Lambda = round(lambda4, 3),
  check.names = FALSE
)
knitr::kable(ht4, caption = "Log-LR scores for original record 4 vs all anonymized records")
```

All scores are negative (evidence against a match), so with the default
threshold of 0, the candidate set is empty and $r_4 = 0$. The
perturbation successfully protected this record.

Now verify with `recordLinkage()`:

```{r prob-toy-verify}
res_prob <- recordLinkage(orig_p, anon_p,
                          key = c("sex", "region", "job"),
                          method = "probabilistic",
                          m_probs = m, u_probs = u)
knitr::kable(res_prob$per_record[, c("risk", "cand_n", "true_in_set", "d_true")],
             digits = 4, caption = "Per-record risk (probabilistic)")
```

The probabilistic method correctly identifies record 1 (all variables
agree, yielding the highest $\Lambda$, unique match, $r_1 = 1.0$)
and fully protects record 4 (region "W" absent in the anonymized
data, all scores negative, $r_4 = 0.0$). Compared to the deterministic
method, the probabilistic approach automatically gives more weight to
agreements on rare values: job agreement (with 4+ categories)
contributes more than three times as much evidence as sex agreement (with only 2
categories), reflecting the discriminative power of each variable.

## Real-World Example (n = 500)

With the toy example, we supplied explicit m/u values. At scale, these
are typically estimated automatically from the data (see above). The
auto-estimation uses aligned pairs for m-probabilities and random
pairings for u-probabilities, which works well when the true match rate
is not too low.

Using the same 500-record census dataset, but now with the probabilistic
method and auto-estimated m/u:

```{r prob-real}
res_prob_real <- recordLinkage(census, census_anon,
                               key = c("age", "sex", "region", "educ"),
                               method = "probabilistic")
summary(res_prob_real)
```

The summary includes the estimated Fellegi-Sunter parameters. Note
how the m-probability for `sex` is typically high (sex is rarely
perturbed by within-region swapping) while `age` has a lower
m-probability (age is directly swapped).

```{r prob-real-plot, fig.height=4, fig.cap="Per-variable log-likelihood ratios for agreement and disagreement (probabilistic method). Variables with high log(m/u) are the most discriminating."}
plot(res_prob_real, which = 3)
```

The barplot shows per-variable log-likelihood ratios for agreement and
disagreement. Variables with high $\log(m/u)$ are the most discriminating
for identifying matches.


# Comparing Methods and Weighting Schemes

## Side-by-Side Comparison

We run the deterministic and probabilistic methods on the same 500-record
census dataset and compare results.

```{r compare-methods}
# Deterministic (already computed)
# Probabilistic (already computed)

# For fair comparison, also run deterministic on categorical-only keys
res_det_cat <- recordLinkage(census, census_anon,
                             key = c("sex", "region", "educ"))

compare_df <- data.frame(
  Method = c("Deterministic (4 QIs)", "Probabilistic (4 QIs)",
             "Deterministic (3 cat QIs)"),
  Mean_risk = c(
    res_real$overall$mean_risk,
    res_prob_real$overall$mean_risk,
    res_det_cat$overall$mean_risk
  ),
  Max_risk = c(
    res_real$overall$max_risk,
    res_prob_real$overall$max_risk,
    res_det_cat$overall$max_risk
  ),
  Pct_high = c(
    100 * res_real$overall$pct_high_risk,
    100 * res_prob_real$overall$pct_high_risk,
    100 * res_det_cat$overall$pct_high_risk
  ),
  Pct_true_in_set = c(
    100 * res_real$overall$pct_true_in_set,
    100 * res_prob_real$overall$pct_true_in_set,
    100 * res_det_cat$overall$pct_true_in_set
  )
)

knitr::kable(compare_df, digits = 3,
             caption = "Method comparison on the census dataset")
```

**Important caveat**: This comparison is *illustrative*, not a
controlled experiment. The configurations differ in two ways
simultaneously: (1) the linkage method, and (2) the set of
quasi-identifiers (4 QIs including numeric `age` vs. 3 categorical QIs).
A fair comparison of methods would require applying all methods to the
same anonymized data with the same QIs.

The deterministic and probabilistic methods with 4 QIs (including numeric
`age`) tend to give finer discrimination because continuous variables
provide more information for distinguishing records.

## Weighting Scheme Comparison

With the default `nearest` strategy, the candidate set for each record
contains only those anonymized records tied at the minimum distance.
When quasi-identifiers include continuous variables (such as `age`),
ties are extremely rare -- nearly every record has a unique nearest
neighbor. When $|C_i| = 1$, the risk is $1/1 = 1$ regardless of the
weighting scheme, because there is only one candidate to weight.

To demonstrate the effect of weighting, we use `strategy = "topk"` with
$k = 5$, which ensures each record has at least 5 candidates among
which the weighting can differentiate:

```{r compare-weights}
qi <- c("age", "sex", "region", "educ")
res_unif <- recordLinkage(census, census_anon, key = qi,
                          strategy = "topk", k = 5,
                          risk_weighting = "uniform")
res_soft <- recordLinkage(census, census_anon, key = qi,
                          strategy = "topk", k = 5,
                          risk_weighting = "softmax")
res_kern_gauss <- recordLinkage(census, census_anon, key = qi,
                                strategy = "topk", k = 5,
                                risk_weighting = "kernel",
                                kernel = "gaussian")
res_kern_epan <- recordLinkage(census, census_anon, key = qi,
                               strategy = "topk", k = 5,
                               risk_weighting = "kernel",
                               kernel = "epanechnikov")
res_kern_tri <- recordLinkage(census, census_anon, key = qi,
                              strategy = "topk", k = 5,
                              risk_weighting = "kernel",
                              kernel = "tricube")

wt_df <- data.frame(
  Weighting = c("Uniform", "Softmax", "Kernel (Gaussian)",
                "Kernel (Epanechnikov)", "Kernel (Tricube)"),
  Mean_risk = c(res_unif$overall$mean_risk,
                res_soft$overall$mean_risk,
                res_kern_gauss$overall$mean_risk,
                res_kern_epan$overall$mean_risk,
                res_kern_tri$overall$mean_risk),
  Max_risk = c(res_unif$overall$max_risk,
                res_soft$overall$max_risk,
                res_kern_gauss$overall$max_risk,
                res_kern_epan$overall$max_risk,
                res_kern_tri$overall$max_risk),
  Pct_high = c(100 * res_unif$overall$pct_high_risk,
               100 * res_soft$overall$pct_high_risk,
               100 * res_kern_gauss$overall$pct_high_risk,
               100 * res_kern_epan$overall$pct_high_risk,
               100 * res_kern_tri$overall$pct_high_risk)
)
knitr::kable(wt_df, digits = 4,
             caption = "Risk weighting schemes (deterministic method, top-5 strategy)")
```

Uniform weighting is most conservative -- it distributes risk equally
among all 5 candidates. Softmax and kernel weighting concentrate risk on
the closest candidate, giving higher risk to easy-to-link records and
lower risk to well-protected ones.

## Summary Table

| Method | Assumptions | Best when | Limitation |
|:-------|:------------|:----------|:-----------|
| Deterministic | Minimal (distance metric) | Default / first pass | Ignores agreement rarity |
| Probabilistic | m/u available or estimable | Audit documentation | Sensitive to m/u misspecification |


# Computational Details

## S3 Class Structure

The `recordLinkage()` function is an S3 generic with methods for
`data.frame` (the default) and `synth_pair` objects. It returns an
object of class `"recordLinkageRisk"` with the following components:

- `$per_record`: a `data.frame` with one row per query record,
  containing `risk` (re-identification risk), `cand_n` (candidate set
  size), `true_in_set` (logical, whether the true match is among
  candidates), `d_true` (distance to true match), `d_min` (distance
  to nearest candidate), `d_rank` (rank of the true match among
  candidates by distance), and `risk_band` (factor categorizing risk
  into bands from "very_low" to "unique_match").
- `$overall`: a list with summary statistics: `mean_risk`, `max_risk`,
  `median_risk`, `pct_high_risk` (fraction above threshold),
  `pct_true_in_set`, `n_unique_match` (records with $r = 1$), and
  `risk_bands` (tabulated risk distribution).
- `$privacy_pass`: logical, `TRUE` if `mean_risk` $\le 0.1$.
- `$settings`: a list recording the method, direction, weighting,
  strategy, and other parameters used.
- For the probabilistic method: `$fs_params` with estimated or
  supplied m/u probabilities.
- `$var_importance`: per-variable importance scores (available for all
  methods; the computation varies: mean distance contributions for
  deterministic, log-LR weights for probabilistic).

Standard S3 methods are provided:

- `print()`: one-screen summary with method, risk statistics, and
  pass/fail status.
- `summary()`: detailed output including risk quantiles, risk band
  distribution, distance-to-true-match statistics, and per-variable
  Fellegi-Sunter parameters (for the probabilistic method). Returns
  an object of class `"summary.recordLinkageRisk"`.
- `plot(x, which = ...)`: up to 8 plot types: (1) risk histogram,
  (2) distance vs. risk scatter, (3) per-variable importance / log-LR
  barplot, (4) risk band barplot, (5) true-match rank distribution,
  (6) risk by subgroup, (7) Lorenz curve of risk concentration.

## Computational Complexity

Let $n$ and $m$ denote the number of original and anonymized records,
and $p = |\mathcal{K}|$ the number of quasi-identifiers.

| Method | Complexity | Notes |
|:-------|:-----------|:------|
| Deterministic | $O(nmp)$ | Full Gower distance matrix |
| Probabilistic | $O(nmp)$ | Log-LR computation per pair per variable |

Without blocking, all methods require $O(nm)$ pairwise comparisons.
For large datasets, blocking (described below) restricts comparisons
to records that agree on a blocking variable, reducing the effective
number of pairs. As a rough guideline:

| Dataset size | Recommendation |
|:-------------|:---------------|
| $n < 5{,}000$ | Any method, no blocking needed |
| $5{,}000 < n < 50{,}000$ | Use blocking |
| $n > 50{,}000$ | Blocking required |

As a rough benchmark, the deterministic method on 500 records with
4 QIs takes well under one second; at $n = 5{,}000$ without blocking,
expect a few seconds.

## Software Dependencies

The core `recordLinkage()` function requires only base R and the
**VIM** package (for Gower distance computation in the deterministic
method). An optional dependency is **clue** (bijective matching via
the Hungarian algorithm).
The examples in this paper additionally use **sdcMicro**
[@templ2015sdcmicro] for perturbation (rank swapping).
All optional packages are listed in `Suggests` in the package `DESCRIPTION`.


# Advanced Features

## Blocking for Scalability

For large datasets, computing all $n \times m$ pairwise distances is
expensive. Blocking restricts comparisons to records that agree on one or
more blocking variables:

```{r blocking}
# Without blocking
system.time(
  res_noblock <- recordLinkage(census, census_anon,
                               key = c("age", "sex", "region", "educ"))
)

# With blocking on region
system.time(
  res_block <- recordLinkage(census, census_anon,
                             key = c("age", "sex", "region", "educ"),
                             block = "region")
)

block_df <- data.frame(
  Blocking = c("None", "region"),
  Mean_risk = c(res_noblock$overall$mean_risk,
                res_block$overall$mean_risk),
  Pct_true_in_set = c(100 * res_noblock$overall$pct_true_in_set,
                      100 * res_block$overall$pct_true_in_set)
)
knitr::kable(block_df, digits = 4,
             caption = "Effect of blocking on risk and true-match retention")
```

Blocking trades off completeness for speed. When the perturbation
preserves the blocking variable (as with within-region swapping), the
true match remains in the candidate set and `pct_true_in_set` stays
high. If the blocking variable itself is perturbed, some true matches
will be missed.

## Direction: Forward vs. Reverse

By default (`direction = "forward"`), the function iterates over original
records and searches in the anonymized data. This answers: *how easily
can each individual be re-identified?*

With `direction = "reverse"`, iteration is over anonymized records,
searching in the original data. This answers: *how disclosive is each
released record?*

```{r direction}
res_fwd2 <- recordLinkage(census, census_anon,
                          key = c("age", "sex", "region", "educ"),
                          direction = "forward")
res_rev2 <- recordLinkage(census, census_anon,
                          key = c("age", "sex", "region", "educ"),
                          direction = "reverse")

dir_comp <- data.frame(
  Direction = c("Forward", "Reverse"),
  Mean_risk = c(res_fwd2$overall$mean_risk,
                res_rev2$overall$mean_risk),
  Max_risk = c(res_fwd2$overall$max_risk,
               res_rev2$overall$max_risk),
  Pct_high = c(100 * res_fwd2$overall$pct_high_risk,
               100 * res_rev2$overall$pct_high_risk)
)
knitr::kable(dir_comp, digits = 4,
             caption = "Forward vs. reverse direction (deterministic)")
```

For the deterministic method, distances are symmetric, so only the
evaluation perspective changes when switching direction. For the
probabilistic method, the log-likelihood ratios are also symmetric,
though auto-estimated m/u parameters are always fitted in the forward
direction.

## Ground Truth Options

The `truth` parameter specifies how to determine the correct match for
scoring:

- `truth = "row"` (default): Row $i$ in the original corresponds to
  row $i$ in the anonymized data. Requires $n = m$.
- `truth = "id"`: A shared identifier column links records across
  datasets. Use when rows are shuffled or datasets have different sizes.

```{r truth-id}
# Add an ID column and shuffle the anonymized data
census_id <- census
census_id$id <- seq_len(n)
census_anon_id <- census_anon
census_anon_id$id <- seq_len(n)
set.seed(321)
census_anon_id <- census_anon_id[sample(n), ]  # shuffle

res_id <- recordLinkage(census_id, census_anon_id,
                        key = c("age", "sex", "region", "educ"),
                        truth = "id", id = "id")
cat("Mean risk (row-aligned):", round(res_real$overall$mean_risk, 4), "\n")
cat("Mean risk (id-based, shuffled):", round(res_id$overall$mean_risk, 4), "\n")
```

## The synth_pair Interface

If you are already using **riskutility**'s `synth_pair` workflow for
other metrics (e.g., DCAP, RAPID, pMSE), you can pass the same object
directly to `recordLinkage()`:

```{r synthpair}
pair <- synth_pair(census, census_anon,
                   key_vars = c("age", "sex", "region", "educ"))
res_pair <- recordLinkage(pair)
cat("Mean risk:", round(res_pair$overall$mean_risk, 4), "\n")
```

The `synth_pair` method extracts `key_vars` automatically, so you don't
need to specify `key` again.


# Practical Guidance

## Method Selection Guide

| What you know | Recommended method |
|:--------------|:-------------------|
| Nothing about the perturbation mechanism | `"deterministic"` |
| m/u probabilities available or formal audit needed | `"probabilistic"` |
| Quick initial assessment | `"deterministic"` with `strategy = "nearest"` |
| Conservative upper bound | `"deterministic"` with `risk_weighting = "uniform"` |

When in doubt, start with the deterministic method. It requires no
assumptions beyond a sensible distance metric and gives a reasonable
baseline. Use the probabilistic method when m/u probabilities from a
formal audit are available, as it gives more accurate risk estimates.

When multiple perturbation methods have been applied to different
variables (e.g., local suppression for some, microaggregation for
income, rank swapping for age), use the deterministic method as a
default -- it makes no assumptions about the perturbation mechanism.

## Interpreting Risk Bands

The `summary()` output classifies records into risk bands. These
thresholds are package conventions for interpretive convenience, not
established standards:

| Band | Risk range | Interpretation |
|:-----|:-----------|:---------------|
| Unique match | $r = 1$ | Attacker can uniquely identify this record |
| Very high | $(0.5, 1)$ | More likely than not to be correctly identified |
| High | $(0.2, 0.5]$ | Substantial risk; consider additional protection |
| Moderate | $(0.1, 0.2]$ | Borderline; assess in context |
| Low | $(0.05, 0.1]$ | Acceptable for most use cases |
| Very low | $[0, 0.05]$ | Well protected |

Note that $r_i = 1$ means the attacker *could* uniquely identify this
record under the assumed model -- it does not mean re-identification has
occurred. The computed risk is an upper bound because the attacker model
assumes perfect membership and QI knowledge.

A common rule of thumb for acceptable risk is $\bar{r} < 0.1$ (mean risk
below 10%), loosely inspired by individual risk thresholds discussed in
@hundepool2012statistical (Chapter 5). This is a package convention, not
a universal standard -- practitioners should consult their organization's
data release policy for applicable thresholds. In practice, many
statistical offices additionally require
that no record has $r > 0.5$ (no unique matches) and that the fraction
of high-risk records is small. The `privacy_pass` flag in
`recordLinkage()` tests only the *mean* risk against the threshold.
High tail risk (e.g., many records with $r > 0.5$) can coexist with a
passing mean -- always check `overall$max_risk` and the risk band
distribution alongside the pass/fail label.

These risk bands are also stored per record in `per_record$risk_band`
(a factor), enabling direct subsetting without recomputing thresholds.
The Gini coefficient (`overall$risk_gini`) measures how concentrated
risk is among a few records: values near 0 indicate uniform risk,
while values near 1 indicate that all risk is concentrated in a
small fraction of records.

## Identifying and Handling High-Risk Records

When the risk assessment returns a warning, practitioners need to
identify which records are at risk and decide on remedial action.
The `recordLinkageRisk` object provides several per-record diagnostics
and helper functions for this purpose, following a natural workflow:
(1) *which records are at risk?* (diagnostics, `top_at_risk`),
(2) *why are they at risk?* (`inspect_record`, `var_importance`),
(3) *which groups need attention?* (`risk_by_group`),
(4) *how concentrated is the risk?* (Lorenz/Gini).

### Per-record diagnostics

Each record in `per_record` now carries a categorical `risk_band`
(very_low, low, moderate, high, very_high, unique_match) and a
`d_rank` (the rank of the true match among all candidates, where
rank 1 is the closest/best match):

```{r per-record-diagnostics}
head(res_real$per_record)
table(res_real$per_record$risk_band)
```

### Top at-risk records

Use `top_at_risk()` to retrieve the highest-risk records with their
QI values:

```{r top-at-risk}
top <- top_at_risk(res_real, n = 10, data = census)
knitr::kable(top, digits = 4,
             caption = "Top 10 highest-risk records with QI values")
```

### Subpopulation analysis

Use `risk_by_group()` to identify which subgroups carry the most
risk:

```{r risk-by-group}
rg <- risk_by_group(res_real, group = "region", data = census)
knitr::kable(rg, digits = 4,
             caption = "Risk by region")
```

### Detailed record inspection

For forensic analysis of a single record, use `inspect_record()`
(requires `return_matches = TRUE`):

```{r inspect-record}
res_matches <- recordLinkage(census, census_anon,
                             key = c("age", "sex", "region", "educ"),
                             return_matches = TRUE)
ir <- inspect_record(res_matches, i = which.max(res_matches$per_record$risk),
                     data_orig = census, data_anon = census_anon)
print(ir)
```

### Variable importance

The `var_importance` field and `plot(x, which = 3)` show which
QIs contribute most to linkage success:

```{r var-importance}
res_real$var_importance
```
```{r var-importance-plot, fig.width=6, fig.height=4}
plot(res_real, which = 3)
```

### Risk band barplot

The risk band barplot (`which = 5`) gives a quick visual overview
of how records distribute across risk categories:

```{r risk-band-plot, fig.width=6, fig.height=4}
plot(res_real, which = 5)
```

### True-match rank distribution

The rank distribution (`which = 6`) shows where the true match
falls among candidates. A concentration at rank 1 means the
attacker's best guess is usually correct:

```{r rank-dist-plot, fig.width=6, fig.height=4}
plot(res_real, which = 6)
```

### Risk by subgroup

The risk-by-group barplot (`which = 7`) reveals which subpopulations
carry the most risk. Error bars show the per-group maximum:

```{r risk-by-group-plot, fig.width=6, fig.height=4}
plot(res_real, which = 7, group = "region", data = census)
```

### Lorenz curve and Gini coefficient

The Lorenz curve (`which = 8`) visualizes how concentrated risk is
among a few records. The Gini coefficient (0 = uniform, 1 = all
risk in one record) summarizes this:

```{r lorenz-plot, fig.width=5, fig.height=5}
plot(res_real, which = 8)
```
```{r risk-gini}
res_real$overall$risk_gini
```

### Merging risks back to data

Use `merge_per_record()` to join the risk diagnostics back to the
original data for further analysis or export:

```{r merge-per-record}
merged <- merge_per_record(res_real, data = census)
head(merged[order(merged$risk, decreasing = TRUE), ], 5)
```

Common remedial actions include:
(1) increasing perturbation strength for high-risk variables,
(2) applying local suppression to the most identifying QI values,
(3) adding noise to continuous variables, or
(4) microaggregating fine-grained variables.
In an iterative SDC workflow, the risk assessment is repeated after
each additional protection step until the risk profile is acceptable.

### QI sensitivity

Risk estimates are conditional on the assumed set of quasi-identifiers.
Running the assessment with different QI sets provides a sensitivity
check:

```{r qi-sensitivity}
qi_sets <- list(
  "age + sex"         = c("age", "sex"),
  "age + sex + region" = c("age", "sex", "region"),
  "all 4 QIs"          = c("age", "sex", "region", "educ")
)
qi_sens <- data.frame(
  QI_set = names(qi_sets),
  Mean_risk = vapply(qi_sets, function(qi)
    recordLinkage(census, census_anon, key = qi)$overall$mean_risk,
    numeric(1))
)
knitr::kable(qi_sens, digits = 4, row.names = FALSE,
             caption = "Effect of QI set on mean risk")
```

Adding `region` and `educ` progressively increases the attacker's
discriminatory power. Practitioners should consider the broadest
plausible set of QIs that an attacker might access.

### Sample surveys vs. census data

`recordLinkage()` assumes the attacker has *membership knowledge* --
they know the target individual is in the dataset. For census or
administrative data, this is a reasonable assumption. For sample surveys,
the attacker may not know whether the target was sampled, and the
computed risk is an upper bound. In such cases, combine
`recordLinkage()` with `population_uniqueness()` to estimate
population-level risk. As a rule of thumb, for a sampling fraction $f$,
the actual re-identification probability is approximately $f \cdot r_i$
[@franconi2004individual].

## Common Pitfalls

1. **All records flagged as high risk**: This often happens with
   fine-grained continuous QIs and the `nearest` strategy. With
   continuous variables, Gower distances are rarely tied, so each
   record has a unique nearest neighbor ($|C_i| = 1$), giving risk
   $r_i = 1$ whenever the true match is nearest (or $r_i = 0$ when
   it is not). This binary behavior masks the degree of protection.
   *Solution*: use `strategy = "topk"` with $k \ge 3$ to create
   candidate sets where weighting can differentiate, or use
   `strategy = "threshold"` to include all records within a distance,
   or bin continuous variables before linkage.

2. **No records flagged**: This happens with very few coarse categorical
   QIs (e.g., only sex and region). Many records share the same QI
   pattern, creating large equivalence classes that dilute risk.
   *Solution*: add more quasi-identifiers to increase discriminatory
   power, or use the probabilistic method (which weights by agreement
   rarity and can distinguish records within equivalence classes).

3. **Wrong variable types**: Factors treated as numeric produce
   meaningless scaled distances (factor levels are arbitrary integers);
   numeric variables treated as nominal ignore the ordering and distance
   structure. Always verify with `str()` and use the `type` parameter
   to override incorrect automatic detection, e.g.,
   `recordLinkage(..., type = c(zip = "nominal", educ = "ordinal"))`.

4. **Blocking too aggressively**: Blocking restricts candidate search
   to records sharing the same blocking variable value. If the blocking
   variable is itself perturbed (e.g., region was swapped), the true
   match may end up in a different block and be missed entirely.
   Monitor `pct_true_in_set` -- if it drops below 80%, blocking may
   be too aggressive.

5. **Direction asymmetry**: Forward and reverse directions answer
   different attacker questions -- always run both for a complete
   picture when the direction of the attack is uncertain.

6. **Overconfidence in `privacy_pass`**: A mean risk of 0.098 passes
   the default 0.1 threshold, but this does not guarantee adequate
   protection -- individual records may have $r_i = 1$ (unique match).
   Always examine the full risk distribution, including `max_risk`,
   the risk band table, and the number of unique matches.

## Relation to Other Risk Metrics

| Metric | Risk type | Data type | Package / function |
|:-------|:----------|:----------|:-------------------|
| `recordLinkage()` | Identity disclosure | Perturbative | riskutility |
| `dRisk()` / `dRiskRMD()` | Identity disclosure | Perturbative | sdcMicro |
| TCAP / DCAP | Attribute disclosure | Synthetic | riskutility |
| DCR / NNDR | Memorization | Generative models | riskutility |
| RAPID | Attribute prediction | Synthetic / anonymized | riskutility |
| `individual_risk()` | Pre-anonymization baseline | Any | riskutility |
| `population_uniqueness()` | Population-level risk | Any | riskutility |

These metrics address different disclosure scenarios and are
complementary rather than competing. Record linkage risk
(`recordLinkage()`) measures *identity disclosure*: can an attacker
match a released record to the original individual? This is the primary
concern for perturbative methods (swapping, noise addition) where
records correspond to real individuals. TCAP and DCAP measure
*attribute disclosure*: can an attacker infer a sensitive attribute
from quasi-identifiers? These are most relevant for synthetic data,
where individual records may not correspond to real people but sensitive
patterns may still leak. DCR and NNDR detect *memorization* in
generative models by comparing distances between training and generated
records. RAPID bridges the gap: it uses machine learning to assess
attribute inference risk and can be applied to both synthetic and
perturbative data. For a comprehensive assessment, combine identity
disclosure metrics (such as `recordLinkage()`) with attribute disclosure
metrics (such as TCAP or RAPID) and utility measures to understand the
full risk-utility trade-off.


# Bijective Matching (GDBRL)

## Independent vs. Bijective Matching

By default, `recordLinkage()` scores each query record independently:
the attacker selects the best candidate for record $i$ without regard
to what candidate was chosen for record $j$. This "independent" mode
corresponds to the classical distance-based record linkage (DBRL)
framework of @domingoferrer2003disclosure.

@herranz2016gdbrl introduced *Global DBRL* (GDBRL), which models
a stronger attacker who enforces a one-to-one (bijective) assignment
between original and anonymized records. The attacker solves a global
optimization problem -- minimizing total cost across all records
simultaneously -- using the Hungarian algorithm (linear sum assignment).
Because the bijective constraint prevents multiple original records from
being matched to the same anonymized record, the resulting risk
estimates are typically higher than independent scoring: the attacker
cannot "waste" the best candidates on multiple queries.

Setting `matching = "bijective"` activates this mode. This implements
the GDBRL-min variant (minimizing total assignment cost); Herranz et al.
also define GDBRL-max (maximizing correct matches subject to a distance
threshold), which is not currently supported. Risk becomes binary: 1 if
the Hungarian algorithm assigns a record to its true match, 0 otherwise
-- a simplification of the graded risk discussed in the original paper.
The `clue` package is required.

A third option, `matching = "ot"`, uses entropy-regularized optimal
transport (Sinkhorn-Knopp algorithm) to produce continuous risk in
$[0,1]$ that interpolates between independent (smooth) and
bijective (hard) assignments, controlled by the `ot_epsilon` parameter.

## Comparison Example

```{r bijective-comparison}
# Independent matching (default) -- reuses census / census_anon from above
res_ind <- recordLinkage(census, census_anon,
                         key = c("age", "sex", "region", "educ"),
                         matching = "independent")

# Bijective matching (GDBRL)
res_bij <- recordLinkage(census, census_anon,
                         key = c("age", "sex", "region", "educ"),
                         matching = "bijective")

cat("Independent:  mean risk =", round(res_ind$overall$mean_risk, 4),
    " | high-risk records:", res_ind$overall$n_high_risk, "\n")
cat("Bijective:    mean risk =", round(res_bij$overall$mean_risk, 4),
    " | high-risk records:", res_bij$overall$n_high_risk, "\n")
```

The bijective attacker typically achieves a higher re-identification
rate because the one-to-one constraint forces globally optimal
assignments. This is a worst-case scenario: in practice, an attacker
rarely has the full membership knowledge and computational resources
required for GDBRL. Nevertheless, reporting both independent and
bijective risk provides a useful bracket on the true risk level.

Bijective (and OT) matching works with all six methods and with blocking. When
blocking is enabled, the Hungarian algorithm runs independently within
each block:

```{r bijective-blocking}
res_blk <- recordLinkage(census, census_anon,
                         key = c("age", "sex", "region", "educ"),
                         block = "region",
                         matching = "bijective")
cat("Bijective + blocking: mean risk =",
    round(res_blk$overall$mean_risk, 4), "\n")
```


# Discussion and Conclusion

The `recordLinkage()` function provides a unified framework for
assessing re-identification risk in perturbed microdata. The three
methods covered in this paper span a range of scenarios: deterministic
linkage for general-purpose assessment, probabilistic linkage for formal
audits, and embedding-based linkage for capturing complex non-linear
interactions via learned latent representations.

## Key Takeaways

The deterministic method provides a robust starting point for any
risk assessment -- it requires no knowledge of the perturbation
mechanism and gives a reasonable baseline. When additional information
is available (m/u probabilities for formal audits, or complex
interaction patterns with high-cardinality categoricals), the
specialized methods can refine the assessment. The choice
of risk weighting scheme matters: uniform weighting is conservative
(equal weight to all candidates), while softmax and kernel weighting
concentrate risk on the closest candidate, producing more nuanced risk
profiles. In particular, the `topk` or `threshold` strategies are
needed to make weighting differences visible, since the default
`nearest` strategy often yields singleton candidate sets. Both forward
and reverse directions should be considered: forward risk assesses how
well each individual is protected, while reverse risk assesses how
disclosive each released record is. Finally, always verify that
`pct_true_in_set` is high (ideally above 80%) to ensure that the
candidate sets are meaningful -- if the true match is rarely among the
candidates, risk estimates are uninformative.

## Limitations

Several limitations should be noted:

1. **Stylized attacker model**: The function assumes perfect membership
   knowledge and exact QI knowledge. Real attackers may have noisy or
   incomplete information, so the computed risk is an upper bound. The
   `compute_baseline = TRUE` option reports the risk *without* perturbation
   (linking the data against itself) as a reference point, so the reduction
   achieved by anonymization can be read off directly.
2. **Computational scaling**: Without blocking, all methods require
   $O(nm)$ pairwise comparisons, which limits applicability to datasets
   with more than approximately 50,000 records.
3. **Conditional independence**: The probabilistic method assumes
   conditional independence of agreement across variables, which is
   approximate when QIs are correlated.
4. **Quasi-identifier selection**: Risk estimates are conditional on
   the assumed set of quasi-identifiers. Adding or removing a single
   QI can dramatically change risk levels. Practitioners must make
   judgment calls about which variables an attacker might know, and
   should consider running the assessment with multiple QI sets to
   gauge sensitivity.
5. **Embedding method**: The autoencoder-based embedding method requires
   **torch** and sufficient training data (blocks with fewer than
   $\max(30, 5d)$ rows fall back to deterministic matching). Training
   is non-deterministic across platforms even with fixed seeds, and the
   method adds computational overhead compared to distance-based methods.
6. **Sample vs. census**: The current implementation computes
   sample-level risk. For sample surveys where the attacker does not
   know whether the target is in the dataset, population-level
   adjustments [@franconi2004individual] are needed.
7. **Realized vs. expected risk**: For perturbative mechanisms the reported
   risk reflects the single *realized* output. A mechanism-level expectation
   integrated over the perturbation distribution for arbitrary swapping
   schemes remains future work.
8. **Naive vs. informed attacker**: The function models an attacker who matches
   on all quasi-identifiers. An *informed* attacker who knows which variables
   were perturbed and matches only on the unchanged ones is not yet supported.
9. **Missing-value handling**: `na_anon` controls NA treatment uniformly across
   all methods (`"ignore"`, `"match"`, `"mismatch"`). For the numeric part of
   the Mahalanobis distance, `"ignore"` uses a zero-contribution approximation
   rather than the exact conditional sub-Mahalanobis distance.

## Future Directions

Planned extensions include:

- Integration with Franconi-Polettini population frequency estimation
  [@franconi2004individual] for population-level risk.
- Household-level matching for hierarchical microdata.
- Attribute disclosure assessment: extending the function to evaluate
  whether an attacker could infer sensitive attributes after successful
  linkage.
- Performance improvements through parallel computation and approximate
  nearest-neighbor search for large-scale applications.


# References

<div id="refs"></div>


# Session Info

The results in this vignette were obtained with the following
configuration.

```{r sessioninfo}
sessionInfo()
```
