---
title: "Record Linkage Risk Assessment with recordLinkage()"
author:
  - name: Matthias Templ
    affiliation: "Institute for Competitiveness and Communication,
      School of Business, FHNW University of Applied Sciences and Arts
      Northwestern Switzerland; Swiss Data Anonymization Center
      (www.swissanon.ch)"
    email: matthias.templ@gmail.com
    orcid: 0000-0002-8638-5276
abstract: >
  This vignette provides an in-depth guide to record linkage risk assessment
  using the `recordLinkage()` function in the **riskutility** package. Four linkage
  methods are covered --- deterministic (Gower distance), probabilistic
  (Fellegi-Sunter), PRAM (transition matrix), and predictive (propensity score)
  --- with mathematical derivations, hand-traceable toy examples, and realistic
  demonstrations. Practical guidance on risk weighting, blocking, direction,
  and method selection is provided.
keywords: [record linkage, re-identification risk, statistical disclosure control,
  perturbative methods, PRAM, Fellegi-Sunter]
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Record Linkage Risk Assessment with recordLinkage()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5,
  warning = FALSE, message = FALSE
)
library(riskutility)
library(sdcMicro)
```

# Introduction

## Perturbative SDC and Record Linkage Risk

Statistical disclosure control (SDC) encompasses a range of techniques for
protecting individual privacy in released microdata
[@hundepool2012statistical]. *Perturbative* methods -- such as PRAM
(Post-Randomization Method), record swapping, noise addition, and
microaggregation -- modify the original records rather than generating
entirely new ones. This makes perturbative data fundamentally different
from fully synthetic data: every released record has a real individual
behind it, and the attacker's task is to undo the perturbation and
re-identify that individual.

Most disclosure risk functions in **riskutility** (DCAP, TCAP, DCR,
RAPID) target *synthetic* data, where the question is whether the
generator memorized training records. For perturbative data the relevant
question is different: *can an attacker link a released record back to
the original individual?* The `recordLinkage()` function addresses
exactly this question.

## The Attacker Model

The attacker scenario assumes:

- **Membership knowledge**: the attacker knows that the target individual
  has a record in the released dataset.
- **Exact QI knowledge**: the attacker knows the target's quasi-identifier
  values (e.g., age, sex, region) from an external source.
- **Linking rule**: the attacker applies a strategy (nearest match,
  threshold, top-$k$) to select candidates from the released data and
  attributes re-identification risk accordingly.

Ground truth for evaluation is taken from row alignment
(`truth = "row"`) or a shared identifier column (`truth = "id"`).
Note that `truth` affects only the *evaluation* of success; it is not
an attacker capability.

## Scope of This Vignette

The `recordLinkage()` function in **riskutility** unifies four linkage
methods -- deterministic, probabilistic, PRAM, and predictive -- under a
single S3 interface with consistent risk weighting, blocking, and
direction support. To our knowledge, no existing R package provides this
combination.

This vignette covers all four methods:

1. **Deterministic** (Gower distance) -- Section 3
2. **Probabilistic** (Fellegi-Sunter log-likelihood ratios) -- Section 4
3. **PRAM** (transition matrix products) -- Section 5
4. **Predictive** (propensity-score-based) -- Section 6

Each section follows the same pedagogical funnel: *mathematical
formulation* $\to$ *hand-traceable toy example* $\to$ *realistic
demonstration*. Section 7 compares methods side by side, Section 8
covers advanced features, and Section 9 provides practical guidance.

For the theoretical background on record linkage see @fellegi1969theory
and @domingo2002validating; for PRAM see @gouweleeuw1998post.


# Mathematical Framework

## Notation

Let $X = \{x_1, \ldots, x_n\}$ denote the original microdata and
$X^a = \{x^a_1, \ldots, x^a_m\}$ the anonymized (perturbed) data.
In the typical case $m = n$ and record $x^a_i$ is the perturbed
version of $x_i$.

We denote:

- $\mathcal{K} = \{v_1, \ldots, v_p\}$ -- the set of quasi-identifier
  variables (keys) available to the attacker.
- $w_v \ge 0$ -- the weight assigned to variable $v$.
- $d(i, j)$ -- the distance (or dissimilarity) between original record
  $x_i$ and anonymized record $x^a_j$ over the quasi-identifiers
  (QIs).
  For the probabilistic method this becomes a log-likelihood ratio
  $\Lambda_{ij}$.
- $C_i$ -- the candidate set of anonymized records that the attacker
  considers as potential matches for $x_i$.
- $j^*_i$ -- the index of the true match of $x_i$ in the anonymized
  data ($j^*_i = i$ when rows are aligned).

## Re-identification Risk

Given a candidate set $C_i$ with weights $\pi_j \ge 0$,
$\sum_{j \in C_i} \pi_j = 1$, the re-identification risk for record $i$
is:
$$r_i = \sum_{j \in C_i} \pi_j \cdot \mathbf{1}[j = j^*_i].$$
When the true match is not in the candidate set, $r_i = 0$. When there
is a single candidate and it is the true match, $r_i = 1$.

The overall risk is summarized by the mean $\bar{r} = n^{-1}\sum_i r_i$,
the maximum $r_{\max}$, and the fraction of records exceeding a threshold
(typically 0.1).

## Risk Weighting Schemes

The choice of weights $w_j$ within the candidate set controls how risk
is attributed. Three schemes are available.

**Uniform** (`risk_weighting = "uniform"`): Each candidate receives
equal weight:
$$\pi_j = \frac{1}{|C_i|}.$$
This is conservative: if three candidates are equally close and one is
the true match, the risk is $1/3$ regardless of distance.

**Softmax** (`risk_weighting = "softmax"`): Closer candidates receive
exponentially higher weight:
$$\pi_j = \frac{\exp(-\kappa \cdot d_j)}{\sum_{k \in C_i} \exp(-\kappa \cdot d_k)},$$
where $\kappa > 0$ is a temperature parameter. If not supplied, $\kappa$
is auto-calibrated as $2 / (\max(d) - \min(d))$. As $\kappa \to \infty$
this approaches winner-take-all; as $\kappa \to 0$ it converges to
uniform.

**Kernel** (`risk_weighting = "kernel"`): Weights are proportional to a
kernel function evaluated at the scaled distance:
$$\pi_j = \frac{K(d_j / h)}{\sum_{k \in C_i} K(d_k / h)},$$
where $h$ is the bandwidth (auto-selected via Silverman's rule if not
supplied) and $K$ is one of:

- *Gaussian*: $K(u) = \exp(-u^2/2)$
- *Epanechnikov*: $K(u) = \max(0, 1 - u^2) \cdot 3/4$
- *Tricube*: $K(u) = \max(0, (1 - |u|^3)^3) \cdot 70/81$

With compact kernels (Epanechnikov, tricube), candidates beyond the
bandwidth receive exactly zero weight and are effectively excluded.


# Deterministic Method (Gower Distance)

## Weighted Gower Distance

The deterministic method computes a weighted Gower-type distance over the
quasi-identifiers [@gower1971general]. For original record $x_i$ and
anonymized record $x^a_j$:

$$d(i,j) = \frac{\sum_{v \in \mathcal{K}} w_v \, \delta_v(i,j)}{\sum_{v \in \mathcal{K}} w_v},$$

where the variable-specific dissimilarities are:

- **Numeric / ordinal**: $\delta_v(i,j) = |x_{iv} - x^a_{jv}| / R_v$,
  where $R_v = \max(x_v \cup x^a_v) - \min(x_v \cup x^a_v)$ is the
  range computed over both datasets
- **Nominal**: $\delta_v(i,j) = \mathbf{1}[x_{iv} \ne x^a_{jv}]$

The resulting distance $d(i,j) \in [0, 1]$, where 0 indicates perfect
agreement and 1 maximal disagreement.

## Toy Example: 6 Records

Consider a small dataset with three quasi-identifiers: `age` (numeric),
`sex` (nominal), and `city` (nominal).

```{r det-toy-data}
orig <- data.frame(
  age  = c(30, 40, 50, 60, 70, 80),
  sex  = factor(c("M", "M", "F", "F", "M", "F")),
  city = factor(c("A", "B", "A", "B", "A", "B"))
)
anon <- data.frame(
  age  = c(30, 45, 50, 50, 70, 80),
  sex  = factor(c("M", "M", "F", "F", "M", "F")),
  city = factor(c("A", "B", "B", "B", "A", "B"))
)
knitr::kable(cbind(Row = 1:6, orig), caption = "Original data")
knitr::kable(cbind(Row = 1:6, anon), caption = "Anonymized data (rows 2, 3, 4 perturbed)")
```

Three records were perturbed: row 2 (age changed from 40 to 45),
row 3 (city changed from A to B), and row 4 (age changed from 60 to 50).
Rows 1, 5, and 6 are unchanged.

The range of `age` is $80 - 30 = 50$. Each variable has equal weight
$w_v = 1/3$.

**Record 1** (age=30, sex=M, city=A) vs. anonymized record 1
(age=30, sex=M, city=A):

$$\delta_{\text{age}} = |30-30|/50 = 0, \quad
\delta_{\text{sex}} = 0, \quad
\delta_{\text{city}} = 0$$
$$d(1,1) = (0 + 0 + 0)/3 = 0.$$

An exact match -- so record 1 is the unique nearest neighbor and
$r_1 = 1.0$.

**Record 3** (age=50, sex=F, city=A) vs. the six anonymized records:

```{r det-toy-distances, echo=FALSE}
# Hand-computed distances for record 3
d3 <- data.frame(
  `Anon row` = 1:6,
  age_dissim = c(abs(50-30)/50, abs(50-45)/50, abs(50-50)/50,
                 abs(50-50)/50, abs(50-70)/50, abs(50-80)/50),
  sex_dissim = c(1, 1, 0, 0, 1, 0),
  city_dissim = c(0, 1, 1, 1, 0, 1)
)
d3$Gower_d <- round((d3$age_dissim + d3$sex_dissim + d3$city_dissim) / 3, 4)
names(d3) <- c("Anon row", "age", "sex", "city", "d(3,j)")
knitr::kable(d3, digits = 4,
             caption = "Variable dissimilarities and Gower distance for original record 3")
```

The two smallest distances are $d(3,3) = 0.3333$ and $d(3,4) = 0.3333$
(a tie). With the `nearest` strategy and uniform weighting, the candidate
set is $C_3 = \{3, 4\}$. Since the true match $j^*_3 = 3$ is in the
set, $r_3 = 1/2 = 0.5$.

**Pedagogical point**: nominal mismatches are expensive. A city
mismatch produces a dissimilarity of $\delta = 1.0$, equivalent to
spanning the entire age range ($50/50 = 1.0$). Even a moderate city
change contributes as much to the Gower distance as a 50-year age
difference. This makes nominal variables highly influential in the
linkage.

Now verify with `recordLinkage()`:

```{r det-toy-verify}
res_det <- recordLinkage(orig, anon, key = c("age", "sex", "city"))
knitr::kable(res_det$per_record[, c("risk", "cand_n", "true_in_set", "d_true")],
             digits = 4, caption = "Per-record risk (deterministic, nearest)")
```

## Attacker Strategies

The `strategy` parameter controls how the attacker selects candidates.
We compare three strategies on the same toy data:

```{r det-toy-strategies}
res_nearest   <- recordLinkage(orig, anon, key = c("age", "sex", "city"),
                               strategy = "nearest")
res_threshold <- recordLinkage(orig, anon, key = c("age", "sex", "city"),
                               strategy = "threshold", threshold = 0.2)
res_topk      <- recordLinkage(orig, anon, key = c("age", "sex", "city"),
                               strategy = "topk", k = 3)
strat_df <- data.frame(
  Record = 1:6,
  nearest   = round(res_nearest$per_record$risk, 4),
  threshold = round(res_threshold$per_record$risk, 4),
  topk_3    = round(res_topk$per_record$risk, 4)
)
knitr::kable(strat_df, caption = "Risk under different attacker strategies")
```

With `threshold = 0.2`, only exact or near-exact matches qualify, so
records with moderate perturbation get zero risk. With `topk` and
$k = 3$, the candidate set always includes at least 3 records, diluting
risk for unique matches.

## Real-World Example (n = 500)

We generate a realistic census-like dataset with 500 records and apply
perturbation using **sdcMicro**: rank swapping for the numeric variables
(`age`, `income`) and the original categorical variables remain
unchanged. This is a typical workflow in statistical disclosure control.

```{r det-realdata}
set.seed(123)  # ensures reproducibility for data generation
n <- 500
census <- data.frame(
  age    = sample(18:85, n, replace = TRUE),
  sex    = factor(sample(c("M", "F"), n, replace = TRUE)),
  region = factor(sample(paste0("R", 1:5), n, replace = TRUE)),
  educ   = factor(sample(c("primary", "secondary", "tertiary"), n,
                         replace = TRUE, prob = c(0.2, 0.5, 0.3))),
  income = round(rlnorm(n, meanlog = 10.5, sdlog = 0.6))
)

# Perturbation via sdcMicro: rank-swap numeric variables
census_anon <- census
census_num <- census[, c("age", "income")]
census_num$age <- as.numeric(census_num$age)
census_num$income <- as.numeric(census_num$income)
census_swapped <- rankSwap(census_num, variables = c("age", "income"),
                           P = 10, seed = 123)
census_anon$age    <- census_swapped$age
census_anon$income <- census_swapped$income

# Record linkage with 4 QIs
res_real <- recordLinkage(census, census_anon,
                          key = c("age", "sex", "region", "educ"))
print(res_real)
```

```{r det-realdata-summary}
summary(res_real)
```

```{r det-realdata-plot, fig.height=4}
plot(res_real, which = 1)
```

The histogram reveals the risk distribution. Records whose rank-swapped
age is close to the original (small rank displacement) retain high risk,
while records swapped with a more distant value are harder to
re-identify.

```{r det-realdata-scatter, fig.height=4}
plot(res_real, which = 2)
```

The scatter plot shows the relationship between the distance to the true
match and the attributed risk. Records with $d_{\text{true}} = 0$ (no
perturbation effect on these QIs) have maximum risk; as distance
increases, risk decreases because more competing candidates exist.


# Probabilistic Method (Fellegi-Sunter)

## Log-Likelihood Ratios

The Fellegi-Sunter framework [@fellegi1969theory] models record pairs
as belonging to one of two classes: matches ($M$) or non-matches ($U$).
For each quasi-identifier $v$, the method estimates:

- $m_v = P(\text{agree on } v \mid \text{true match})$
- $u_v = P(\text{agree on } v \mid \text{non-match})$

The composite log-likelihood ratio for the pair $(i, j)$ is:
$$\Lambda_{ij} = \sum_{v \in \mathcal{K}} \gamma_{ijv} \log\frac{m_v}{u_v} + (1 - \gamma_{ijv}) \log\frac{1-m_v}{1-u_v},$$
where $\gamma_{ijv} = \mathbf{1}[\text{agree on } v]$. High $\Lambda$
indicates strong evidence of a match.

The key insight is that *rare agreements are more informative*: if a
variable has many categories (low $u_v$), agreement provides stronger
evidence of a true match.

## Estimation of m and u

When `m_probs` and `u_probs` are not supplied, `recordLinkage()` estimates
them internally:

- For **nominal** variables: $m_v$ is estimated from agreement rates among
  aligned pairs; $u_v$ from random pairings.
- For **numeric** variables: agreement is defined via a MAD-based
  tolerance. Two values agree if $|x_{iv} - x^a_{jv}| \le
  \text{MAD}(x_v)$, where MAD is computed on the original data.

User-supplied values via `m_probs` and `u_probs` override the estimation
and are recommended when the perturbation mechanism is well understood.

## Toy Example: 5 Records (All Categorical)

We use an all-categorical toy example to avoid the complication of
numeric tolerance.

```{r prob-toy-data}
orig_p <- data.frame(
  sex    = factor(c("M", "F", "M", "F", "M")),
  region = factor(c("N", "S", "E", "W", "N")),
  job    = factor(c("A", "B", "C", "A", "D"))
)
anon_p <- data.frame(
  sex    = factor(c("M", "F", "F", "F", "M")),
  region = factor(c("N", "S", "E", "N", "S")),
  job    = factor(c("A", "B", "C", "B", "D"))
)
knitr::kable(cbind(Row = 1:5, orig_p), caption = "Original data (probabilistic toy)")
knitr::kable(cbind(Row = 1:5, anon_p), caption = "Anonymized data (probabilistic toy)")
```

We supply explicit m/u probabilities for full transparency:

```{r prob-toy-mu}
m <- c(sex = 0.95, region = 0.80, job = 0.85)
u <- c(sex = 0.50, region = 0.25, job = 0.10)
```

The per-variable log-likelihood ratios are:

```{r prob-toy-logLR}
log_agree    <- round(log(m / u), 3)
log_disagree <- round(log((1 - m) / (1 - u)), 3)
lr_table <- data.frame(
  Variable = names(m),
  m = m, u = u,
  `log(m/u)` = log_agree,
  `log((1-m)/(1-u))` = log_disagree,
  check.names = FALSE
)
knitr::kable(lr_table, digits = 3,
             caption = "Per-variable log-likelihood ratios")
```

**Hand-tracing record 1** (M, N, A) against all 5 anonymized records:

```{r prob-toy-handtrace, echo=FALSE}
# Record 1: sex=M, region=N, job=A
# Anon records: (M,N,A), (F,S,B), (F,E,C), (F,N,B), (M,S,D)
agree_matrix <- matrix(c(
  1, 1, 1,  # anon 1: all agree
  0, 0, 0,  # anon 2: none agree
  0, 0, 0,  # anon 3: none agree
  0, 1, 0,  # anon 4: region agrees
  1, 0, 0   # anon 5: sex agrees
), nrow = 5, byrow = TRUE, dimnames = list(
  paste0("Anon ", 1:5),
  c("sex", "region", "job")
))

# Compute Lambda for each pair
lambda <- numeric(5)
for (j in 1:5) {
  lambda[j] <- sum(agree_matrix[j, ] * log(m / u) +
                   (1 - agree_matrix[j, ]) * log((1 - m) / (1 - u)))
}
ht <- data.frame(
  `Anon row` = 1:5,
  sex = ifelse(agree_matrix[, 1] == 1, "agree", "disagree"),
  region = ifelse(agree_matrix[, 2] == 1, "agree", "disagree"),
  job = ifelse(agree_matrix[, 3] == 1, "agree", "disagree"),
  Lambda = round(lambda, 3),
  check.names = FALSE
)
knitr::kable(ht, caption = "Log-LR scores for original record 1 vs all anonymized records")
```

For anonymized record 1 (all three variables agree):
$\Lambda = 0.642 + 1.163 + 2.140 = 3.945$. This is the highest score.
With `fs_threshold = 0` (the default -- include all pairs with positive
log-likelihood ratio), only anonymized record 1 qualifies, so
$r_1 = 1.0$.

**Key insight**: job agreement contributes $\log(0.85/0.10) = 2.140$ --
more than 3 times the contribution of sex agreement
($\log(0.95/0.50) = 0.642$) -- because rare agreements carry much
stronger evidence. The Fellegi-Sunter framework automatically accounts
for this.

**Hand-tracing record 4** (F, W, A) -- note that region "W" does not
appear in the anonymized data:

```{r prob-toy-rec4, echo=FALSE}
# Record 4: sex=F, region=W, job=A
agree4 <- matrix(c(
  0, 0, 1,  # anon 1: job agrees
  1, 0, 0,  # anon 2: sex agrees
  1, 0, 0,  # anon 3: sex agrees
  1, 0, 0,  # anon 4: sex agrees
  0, 0, 0   # anon 5: none
), nrow = 5, byrow = TRUE, dimnames = list(
  paste0("Anon ", 1:5),
  c("sex", "region", "job")
))
lambda4 <- numeric(5)
for (j in 1:5) {
  lambda4[j] <- sum(agree4[j, ] * log(m / u) +
                    (1 - agree4[j, ]) * log((1 - m) / (1 - u)))
}
ht4 <- data.frame(
  `Anon row` = 1:5,
  Lambda = round(lambda4, 3),
  check.names = FALSE
)
knitr::kable(ht4, caption = "Log-LR scores for original record 4 vs all anonymized records")
```

All scores are negative (evidence against a match), so with the default
threshold of 0, the candidate set is empty and $r_4 = 0$. The
perturbation successfully protected this record.

Now verify with `recordLinkage()`:

```{r prob-toy-verify}
res_prob <- recordLinkage(orig_p, anon_p,
                          key = c("sex", "region", "job"),
                          method = "probabilistic",
                          m_probs = m, u_probs = u)
knitr::kable(res_prob$per_record[, c("risk", "cand_n", "true_in_set", "d_true")],
             digits = 4, caption = "Per-record risk (probabilistic)")
```

## Real-World Example (n = 500)

Using the same 500-record census dataset from Section 3.4, but now with
the probabilistic method and auto-estimated m/u:

```{r prob-real}
res_prob_real <- recordLinkage(census, census_anon,
                               key = c("age", "sex", "region", "educ"),
                               method = "probabilistic")
summary(res_prob_real)
```

The summary includes the estimated Fellegi-Sunter parameters. Note
how the m-probability for `sex` is typically high (sex is rarely
perturbed by within-region swapping) while `age` has a lower
m-probability (age is directly swapped).

```{r prob-real-plot, fig.height=4}
plot(res_prob_real, which = 3)
```

The barplot shows per-variable log-likelihood ratios for agreement and
disagreement. Variables with high $\log(m/u)$ are the most discriminating
for identifying matches.


# PRAM Method

## Transition Matrices

For data protected via the Post-Randomization Method (PRAM)
[@gouweleeuw1998post], the perturbation mechanism is fully described by
transition matrices. For each categorical variable $v$, the matrix
$T_v$ has entry $T_v[a, b] = P(\text{output} = b \mid \text{input} = a)$.

The probability that original record $x_i$ produced anonymized record
$x^a_j$ is the product over all variables:
$$P(x^a_j \mid x_i) = \prod_{v \in \mathcal{K}} T_v[x_{iv}, x^a_{jv}].$$

The re-identification risk is then the normalized probability:
$$r_i = \frac{P(x^a_{j^*_i} \mid x_i)}{\sum_{j \in C_i} P(x^a_j \mid x_i)},$$
where $C_i = \{j : P(x^a_j \mid x_i) > 0\}$ is restricted to
records with nonzero transition probability (and within the blocking
window, if applicable). This assumes independence of
perturbation across variables, which is exact when PRAM is applied
independently to each variable.

## Toy Example: 5 Records

```{r pram-toy-data}
orig_pram <- data.frame(
  color = factor(c("R", "G", "B", "R", "G")),
  shape = factor(c("C", "S", "T", "C", "T"))
)
anon_pram <- data.frame(
  color = factor(c("R", "B", "B", "G", "G")),
  shape = factor(c("C", "S", "T", "S", "T"))
)

tm_color <- matrix(
  c(0.8, 0.1, 0.1,
    0.1, 0.7, 0.2,
    0.1, 0.2, 0.7),
  3, 3, byrow = TRUE,
  dimnames = list(c("R", "G", "B"), c("R", "G", "B"))
)

tm_shape <- matrix(
  c(0.9, 0.05, 0.05,
    0.1, 0.8,  0.1,
    0.05, 0.05, 0.9),
  3, 3, byrow = TRUE,
  dimnames = list(c("C", "S", "T"), c("C", "S", "T"))
)

knitr::kable(tm_color, caption = "Transition matrix for 'color'")
knitr::kable(tm_shape, caption = "Transition matrix for 'shape'")
```

**Hand-tracing record 1** (R, C) $\to$ anonymized record 1 (R, C):

- $T_{\text{color}}[\text{R}, \text{R}] = 0.8$
- $T_{\text{shape}}[\text{C}, \text{C}] = 0.9$
- $P(x^a_1 \mid x_1) = 0.8 \times 0.9 = 0.720$

**Record 4** (R, C) $\to$ anonymized record 4 (G, S):

- $T_{\text{color}}[\text{R}, \text{G}] = 0.1$
- $T_{\text{shape}}[\text{C}, \text{S}] = 0.05$
- $P(x^a_4 \mid x_4) = 0.1 \times 0.05 = 0.005$

```{r pram-toy-allprobs, echo=FALSE}
# Compute all transition probabilities for records 1 and 4
probs <- matrix(0, nrow = 5, ncol = 2,
                dimnames = list(paste0("Anon ", 1:5),
                                c("P(anon_j | orig_1)", "P(anon_j | orig_4)")))
for (j in 1:5) {
  probs[j, 1] <- tm_color[as.character(orig_pram$color[1]),
                           as.character(anon_pram$color[j])] *
                 tm_shape[as.character(orig_pram$shape[1]),
                          as.character(anon_pram$shape[j])]
  probs[j, 2] <- tm_color[as.character(orig_pram$color[4]),
                           as.character(anon_pram$color[j])] *
                 tm_shape[as.character(orig_pram$shape[4]),
                          as.character(anon_pram$shape[j])]
}
probs_df <- as.data.frame(probs)
probs_df$`Normalized risk (rec 1)` <- round(probs[, 1] / sum(probs[, 1]), 4)
probs_df$`Normalized risk (rec 4)` <- round(probs[, 2] / sum(probs[, 2]), 4)
knitr::kable(probs_df, digits = 4,
             caption = "Transition probabilities and normalized risks for records 1 and 4")
```

Records 1 and 4 have identical original values (R, C) but completely
different risks. Record 1's output (R, C) has high probability under
the original (the diagonal is large), giving risk $\approx 0.97$.
Record 4's output (G, S) has very low probability, giving risk
$\approx 0.01$. **PRAM risk depends on the actual output, not just the
input.**

Now verify with `recordLinkage()`:

```{r pram-toy-verify}
res_pram <- recordLinkage(orig_pram, anon_pram,
                          key = c("color", "shape"),
                          method = "pram",
                          pram_matrix = list(color = tm_color,
                                            shape = tm_shape))
knitr::kable(res_pram$per_record[, c("risk", "d_true")],
             digits = 4, caption = "Per-record risk (PRAM)")
```

## Forward vs. Reverse Asymmetry

The PRAM method is unique among the four methods in that it is
*asymmetric*: the transition matrix $T_v[a, b]$ is not symmetric, so
forward risk (given the original, sum over anonymized candidates) and
reverse risk (given the anonymized record, sum over original candidates)
can differ.

```{r pram-direction-data}
orig_dir <- data.frame(grade = factor(c("A", "B", "C", "A", "B")))
anon_dir <- data.frame(grade = factor(c("A", "C", "C", "B", "A")))

tm_grade <- matrix(
  c(0.7, 0.2, 0.1,
    0.1, 0.6, 0.3,
    0.3, 0.1, 0.6),
  3, 3, byrow = TRUE,
  dimnames = list(c("A", "B", "C"), c("A", "B", "C"))
)
knitr::kable(tm_grade, caption = "Asymmetric transition matrix for 'grade'")
```

```{r pram-direction-compare}
res_fwd <- recordLinkage(orig_dir, anon_dir,
                         key = "grade", method = "pram",
                         pram_matrix = list(grade = tm_grade),
                         direction = "forward")
res_rev <- recordLinkage(orig_dir, anon_dir,
                         key = "grade", method = "pram",
                         pram_matrix = list(grade = tm_grade),
                         direction = "reverse")

dir_df <- data.frame(
  Record = 1:5,
  Orig_grade = as.character(orig_dir$grade),
  Anon_grade = as.character(anon_dir$grade),
  Forward_risk = round(res_fwd$per_record$risk, 4),
  Reverse_risk = round(res_rev$per_record$risk, 4)
)
knitr::kable(dir_df,
             caption = "Forward vs. reverse risk (PRAM with asymmetric matrix)")
```

In forward mode, for each original record's value (fixing a row of $T$),
the function evaluates $T[\text{original}, b]$ for each anonymized
candidate $b$. In reverse mode, for each anonymized record's value
(fixing a column of $T$), the function evaluates
$T[\text{original}_j, \text{anon}]$ for each original candidate $j$.
Both use $P(\text{output} \mid \text{input})$; only the roles of
query and search change. When $T$ is asymmetric, the normalization
denominators differ, yielding different risk profiles.

## Real-World Example (n = 500)

We create a categorical dataset and apply PRAM perturbation using
`sdcMicro::pram()`. The key advantage is that `pram()` generates the
transition matrices internally, which we can then pass directly to
`recordLinkage()`.

```{r pram-real}
set.seed(456)  # ensures reproducibility for PRAM data
n <- 500
pram_data <- data.frame(
  sex    = factor(sample(c("M", "F"), n, replace = TRUE)),
  region = factor(sample(c("N", "S", "E", "W"), n, replace = TRUE)),
  educ   = factor(sample(c("low", "mid", "high"), n, replace = TRUE,
                         prob = c(0.3, 0.5, 0.2)))
)

# Apply PRAM via sdcMicro
pram_vars <- c("sex", "region", "educ")
pram_res <- pram(pram_data, variables = pram_vars, pd = 0.8, alpha = 0.5)

# Build the anonymized dataset from _pram columns
pram_anon <- data.frame(
  sex    = pram_res$sex_pram,
  region = pram_res$region_pram,
  educ   = pram_res$educ_pram
)

# Extract transition matrices generated by sdcMicro
pram_params <- attr(pram_res, "pram_params")
tm_list <- setNames(
  lapply(pram_vars, function(v) pram_params[[v]]$Rs),
  pram_vars
)

# Inspect the perturbation summary
print(attr(pram_res, "summary"))

# Record linkage using the actual transition matrices
res_pram_real <- recordLinkage(
  pram_data, pram_anon,
  key = pram_vars,
  method = "pram",
  pram_matrix = tm_list
)
print(res_pram_real)
```

```{r pram-real-plot, fig.height=4}
plot(res_pram_real, which = 1)
```


# Predictive Method (Propensity Score)

## Propensity Score Linkage

The predictive method fits a propensity model on
the stacked original + anonymized data to predict record origin
(original = 1, anonymized = 0) from the quasi-identifiers. The
propensity score
$$e(x) = P(\text{original} \mid \text{QIs} = x)$$
projects the multivariate QI space into a single dimension that
maximally separates the two datasets.

Records are then linked in the 1D propensity-score space. The distance
between original record $i$ and anonymized record $j$ is simply:
$$d(i,j) = |e(x_i) - e(x^a_j)|.$$

The standard strategies (nearest, threshold, top-$k$) and weighting
schemes apply to these propensity-score distances.

## Fisher SE Bandwidth

When `pred_se = TRUE` (default) and `pred_model = "logit"`, the
bandwidth for kernel weighting is derived from the Fisher information
matrix:
$$\text{SE}_i = \hat{e}_i(1 - \hat{e}_i) \sqrt{\mathbf{x}_i^\top \hat{V} \mathbf{x}_i},$$
where $\hat{V}$ is the variance-covariance matrix of the logistic
regression coefficients. Records with uncertain predictions (near 0.5)
get wider kernels, while confidently classified records get narrow
kernels -- and hence higher individual risk. Note that `pred_se` only
affects the bandwidth when `risk_weighting = "kernel"`; with uniform or
softmax weighting it has no effect.

## Toy Example: 6 Records

```{r pred-toy-data}
orig_pred <- data.frame(
  age = c(20, 40, 40, 60, 60, 80),
  sex = factor(c("F", "M", "F", "M", "F", "M"))
)
anon_pred <- data.frame(
  age = c(25, 40, 45, 55, 60, 70),
  sex = factor(c("F", "M", "F", "M", "F", "M"))
)
```

```{r pred-toy-run}
res_pred <- recordLinkage(orig_pred, anon_pred,
                          key = c("age", "sex"),
                          method = "predictive")
print(res_pred)
```

```{r pred-toy-risk}
knitr::kable(res_pred$per_record[, c("risk", "cand_n", "true_in_set", "d_true")],
             digits = 4, caption = "Per-record risk (predictive method)")
```

The propensity score acts as a 1D summary of how "original-like" a record
is. Records at extreme ages (20, 80) get distinctive propensity scores
because the age distribution of the anonymized data does not extend as
far, making them easier to re-identify. Records at ages 40 and 60 --
where original and anonymized records overlap -- cluster near 0.5 and
are harder to link.

```{r pred-toy-propensity, fig.height=4}
plot(res_pred, which = 4)
```

## Real-World Example (n = 500)

Using the same 500-record census dataset:

```{r pred-real}
res_pred_real <- recordLinkage(census, census_anon,
                               key = c("age", "sex", "region", "educ"),
                               method = "predictive")
print(res_pred_real)
```

Compare default uniform weighting with kernel weighting (Fisher SE
adaptive bandwidth):

```{r pred-real-compare}
res_pred_kernel <- recordLinkage(census, census_anon,
                                 key = c("age", "sex", "region", "educ"),
                                 method = "predictive",
                                 risk_weighting = "kernel")

comp_pred <- data.frame(
  Weighting = c("Uniform", "Kernel (Fisher SE)"),
  Mean_risk = c(res_pred_real$overall$mean_risk,
                res_pred_kernel$overall$mean_risk),
  Max_risk = c(res_pred_real$overall$max_risk,
               res_pred_kernel$overall$max_risk),
  Pct_high = c(100 * res_pred_real$overall$pct_high_risk,
               100 * res_pred_kernel$overall$pct_high_risk)
)
knitr::kable(comp_pred, digits = 4,
             caption = "Predictive method: uniform vs kernel weighting")
```

The kernel weighting with Fisher SE bandwidth concentrates risk on records
where the propensity model is confident about the match, giving a more
nuanced risk profile than uniform weighting.

```{r pred-real-plot, fig.height=4}
plot(res_pred_real, which = 4)
```


# Comparing Methods and Weighting Schemes

## Side-by-Side Comparison

We run all four methods on the same 500-record census dataset and compare
results. For the PRAM method we use the categorical-only subset since
PRAM requires transition matrices.

```{r compare-methods}
# Deterministic (already computed)
# Probabilistic (already computed)
# Predictive (already computed)

# For fair comparison, also run deterministic on categorical-only keys
res_det_cat <- recordLinkage(census, census_anon,
                             key = c("sex", "region", "educ"))

# Apply sdcMicro::pram() to the census categorical variables
cat_vars <- c("sex", "region", "educ")
pram_census_res <- pram(census[, cat_vars], variables = cat_vars,
                        pd = 0.85, alpha = 0.5)
census_pram <- census
census_pram$sex    <- pram_census_res$sex_pram
census_pram$region <- pram_census_res$region_pram
census_pram$educ   <- pram_census_res$educ_pram

# Extract the actual transition matrices from sdcMicro
pp <- attr(pram_census_res, "pram_params")
tm_census <- setNames(lapply(cat_vars, function(v) pp[[v]]$Rs), cat_vars)

res_pram_census <- recordLinkage(
  census, census_pram,
  key = cat_vars,
  method = "pram",
  pram_matrix = tm_census
)

compare_df <- data.frame(
  Method = c("Deterministic (4 QIs)", "Probabilistic (4 QIs)",
             "Predictive (4 QIs)", "Deterministic (3 cat QIs)",
             "PRAM (3 cat QIs)"),
  Mean_risk = c(
    res_real$overall$mean_risk,
    res_prob_real$overall$mean_risk,
    res_pred_real$overall$mean_risk,
    res_det_cat$overall$mean_risk,
    res_pram_census$overall$mean_risk
  ),
  Max_risk = c(
    res_real$overall$max_risk,
    res_prob_real$overall$max_risk,
    res_pred_real$overall$max_risk,
    res_det_cat$overall$max_risk,
    res_pram_census$overall$max_risk
  ),
  Pct_high = c(
    100 * res_real$overall$pct_high_risk,
    100 * res_prob_real$overall$pct_high_risk,
    100 * res_pred_real$overall$pct_high_risk,
    100 * res_det_cat$overall$pct_high_risk,
    100 * res_pram_census$overall$pct_high_risk
  ),
  Pct_true_in_set = c(
    100 * res_real$overall$pct_true_in_set,
    100 * res_prob_real$overall$pct_true_in_set,
    100 * res_pred_real$overall$pct_true_in_set,
    100 * res_det_cat$overall$pct_true_in_set,
    100 * res_pram_census$overall$pct_true_in_set
  )
)
knitr::kable(compare_df, digits = 3,
             caption = "Method comparison on the census dataset")
```

In this particular example, the deterministic and probabilistic methods
with 4 QIs (including numeric `age`) tend to give lower risk because the
continuous variable provides fine-grained discrimination. The PRAM row
uses actual transition matrices from `sdcMicro::pram()`, applied to the
categorical variables of the census data. Note that the PRAM and
categorical-deterministic results use different anonymized datasets
(PRAM-perturbed vs. rank-swapped), so the comparison reflects both the
method and the perturbation mechanism.

## Weighting Scheme Comparison

We compare the three weighting schemes on the deterministic method:

```{r compare-weights}
res_unif <- recordLinkage(census, census_anon,
                          key = c("age", "sex", "region", "educ"),
                          risk_weighting = "uniform")
res_soft <- recordLinkage(census, census_anon,
                          key = c("age", "sex", "region", "educ"),
                          risk_weighting = "softmax")
res_kern_gauss <- recordLinkage(census, census_anon,
                                key = c("age", "sex", "region", "educ"),
                                risk_weighting = "kernel",
                                kernel = "gaussian")
res_kern_epan <- recordLinkage(census, census_anon,
                               key = c("age", "sex", "region", "educ"),
                               risk_weighting = "kernel",
                               kernel = "epanechnikov")
res_kern_tri <- recordLinkage(census, census_anon,
                              key = c("age", "sex", "region", "educ"),
                              risk_weighting = "kernel",
                              kernel = "tricube")

wt_df <- data.frame(
  Weighting = c("Uniform", "Softmax", "Kernel (Gaussian)",
                "Kernel (Epanechnikov)", "Kernel (Tricube)"),
  Mean_risk = c(res_unif$overall$mean_risk,
                res_soft$overall$mean_risk,
                res_kern_gauss$overall$mean_risk,
                res_kern_epan$overall$mean_risk,
                res_kern_tri$overall$mean_risk),
  Max_risk = c(res_unif$overall$max_risk,
                res_soft$overall$max_risk,
                res_kern_gauss$overall$max_risk,
                res_kern_epan$overall$max_risk,
                res_kern_tri$overall$max_risk),
  Pct_high = c(100 * res_unif$overall$pct_high_risk,
               100 * res_soft$overall$pct_high_risk,
               100 * res_kern_gauss$overall$pct_high_risk,
               100 * res_kern_epan$overall$pct_high_risk,
               100 * res_kern_tri$overall$pct_high_risk)
)
knitr::kable(wt_df, digits = 4,
             caption = "Risk weighting schemes (deterministic method, 4 QIs)")
```

Uniform weighting is most conservative -- it distributes risk equally
among all candidates. Softmax and kernel weighting concentrate risk on
the closest candidate, giving higher risk to easy-to-link records and
lower risk to well-protected ones.

## Summary Table

| Method | Assumptions | Best when | Limitation |
|:-------|:------------|:----------|:-----------|
| Deterministic | None | Default / first pass | Ignores agreement rarity |
| Probabilistic | m/u available or estimable | Audit documentation | Sensitive to m/u misspecification |
| PRAM | Transition matrix known | PRAM / swapping mechanism | Only categorical; requires exact matrices |
| Predictive | Logistic model adequate | High-dimensional / mixed QIs | Model uncertainty |


# Advanced Features

## Blocking for Scalability

For large datasets, computing all $n \times m$ pairwise distances is
expensive. Blocking restricts comparisons to records that agree on one or
more blocking variables:

```{r blocking}
# Without blocking
system.time(
  res_noblock <- recordLinkage(census, census_anon,
                               key = c("age", "sex", "region", "educ"))
)

# With blocking on region
system.time(
  res_block <- recordLinkage(census, census_anon,
                             key = c("age", "sex", "region", "educ"),
                             block = "region")
)

block_df <- data.frame(
  Blocking = c("None", "region"),
  Mean_risk = c(res_noblock$overall$mean_risk,
                res_block$overall$mean_risk),
  Pct_true_in_set = c(100 * res_noblock$overall$pct_true_in_set,
                      100 * res_block$overall$pct_true_in_set)
)
knitr::kable(block_df, digits = 4,
             caption = "Effect of blocking on risk and true-match retention")
```

Blocking trades off completeness for speed. When the perturbation
preserves the blocking variable (as with within-region swapping), the
true match remains in the candidate set and `pct_true_in_set` stays
high. If the blocking variable itself is perturbed, some true matches
will be missed.

## Direction: Forward vs. Reverse

By default (`direction = "forward"`), the function iterates over original
records and searches in the anonymized data. This answers: *how easily
can each individual be re-identified?*

With `direction = "reverse"`, iteration is over anonymized records,
searching in the original data. This answers: *how disclosive is each
released record?*

```{r direction}
res_fwd2 <- recordLinkage(census, census_anon,
                          key = c("age", "sex", "region", "educ"),
                          direction = "forward")
res_rev2 <- recordLinkage(census, census_anon,
                          key = c("age", "sex", "region", "educ"),
                          direction = "reverse")

dir_comp <- data.frame(
  Direction = c("Forward", "Reverse"),
  Mean_risk = c(res_fwd2$overall$mean_risk,
                res_rev2$overall$mean_risk),
  Max_risk = c(res_fwd2$overall$max_risk,
               res_rev2$overall$max_risk),
  Pct_high = c(100 * res_fwd2$overall$pct_high_risk,
               100 * res_rev2$overall$pct_high_risk)
)
knitr::kable(dir_comp, digits = 4,
             caption = "Forward vs. reverse direction (deterministic)")
```

For the deterministic and predictive methods, distances are symmetric,
so only the evaluation perspective changes when switching direction.
For the probabilistic method, the log-likelihood ratios are also
symmetric, though auto-estimated m/u parameters are always fitted in the
forward direction. For PRAM, direction changes which side of the
transition matrix is used, so forward and reverse risks can differ
substantially (see the forward vs. reverse example in the PRAM section).

## Ground Truth Options

The `truth` parameter specifies how to determine the correct match for
scoring:

- `truth = "row"` (default): Row $i$ in the original corresponds to
  row $i$ in the anonymized data. Requires $n = m$.
- `truth = "id"`: A shared identifier column links records across
  datasets. Use when rows are shuffled or datasets have different sizes.

```{r truth-id}
# Add an ID column and shuffle the anonymized data
census_id <- census
census_id$id <- seq_len(n)
census_anon_id <- census_anon
census_anon_id$id <- seq_len(n)
census_anon_id <- census_anon_id[sample(n), ]  # shuffle

res_id <- recordLinkage(census_id, census_anon_id,
                        key = c("age", "sex", "region", "educ"),
                        truth = "id", id = "id")
cat("Mean risk (row-aligned):", round(res_real$overall$mean_risk, 4), "\n")
cat("Mean risk (id-based, shuffled):", round(res_id$overall$mean_risk, 4), "\n")
```

## The synth_pair Interface

For workflows that use the `synth_pair` container, `recordLinkage()`
has a dedicated method:

```{r synthpair}
pair <- synth_pair(census, census_anon,
                   key_vars = c("age", "sex", "region", "educ"))
res_pair <- recordLinkage(pair)
cat("Mean risk:", round(res_pair$overall$mean_risk, 4), "\n")
```

The `synth_pair` method extracts `key_vars` automatically, so you don't
need to specify `key` again.


# Practical Guidance

## Method Selection Guide

| What you know | Recommended method |
|:--------------|:-------------------|
| Nothing about the perturbation mechanism | `"deterministic"` |
| Transition matrices available (PRAM, swapping) | `"pram"` |
| m/u probabilities available or formal audit needed | `"probabilistic"` |
| Many QIs, nonlinear interactions suspected | `"predictive"` |
| Quick initial assessment | `"deterministic"` with `strategy = "nearest"` |
| Conservative upper bound | `"deterministic"` with `risk_weighting = "uniform"` |

When in doubt, start with the deterministic method. It requires no
assumptions beyond a sensible distance metric and gives a reasonable
baseline. Use the other methods when additional information about the
perturbation mechanism is available.

## Interpreting Risk Bands

The `summary()` output classifies records into risk bands:

| Band | Risk range | Interpretation |
|:-----|:-----------|:---------------|
| Unique match | $r = 1$ | Attacker can uniquely identify this record |
| Very high | $(0.5, 1)$ | More likely than not to be correctly identified |
| High | $(0.2, 0.5]$ | Substantial risk; consider additional protection |
| Moderate | $(0.1, 0.2]$ | Borderline; assess in context |
| Low | $(0.05, 0.1]$ | Acceptable for most use cases |
| Very low | $[0, 0.05]$ | Well protected |

A common threshold for acceptable risk is $\bar{r} < 0.1$ (mean risk
below 10%). The `privacy_pass` flag tests only the *mean* risk against
this threshold. High tail risk (e.g., many records with $r > 0.5$) can
coexist with a passing mean -- always check `overall$max_risk` and
the risk band distribution alongside the pass/fail label.

## Common Pitfalls

1. **All records flagged as high risk**: This often happens with
   fine-grained continuous QIs and the `nearest` strategy. Every record
   has a unique nearest neighbor, giving risk $\ge 0.5$.
   *Solution*: use `strategy = "threshold"` or bin continuous variables.

2. **No records flagged**: This happens with very few coarse categorical
   QIs. Many records share the same QI pattern, diluting risk.
   *Solution*: add more quasi-identifiers or use the probabilistic method
   (which weights by agreement rarity).

3. **Wrong variable types**: Factors treated as numeric or vice versa
   produce incorrect distances. Always check `str()` and use the `type`
   parameter if needed.

4. **Blocking too aggressively**: If the blocking variable is itself
   perturbed, true matches may fall into different blocks. Monitor
   `pct_true_in_set` -- if it drops below 80%, blocking may be too
   aggressive.

5. **Ignoring direction asymmetry in PRAM**: Forward and reverse risks
   can differ substantially with asymmetric transition matrices. Always
   run both directions for a complete picture.

## Relation to Other Risk Metrics

| Metric | Risk type | Data type | Package function |
|:-------|:----------|:----------|:-----------------|
| `recordLinkage()` | Identity disclosure | Perturbative | `recordLinkage()` |
| TCAP / DCAP | Attribute disclosure | Synthetic | `tcap()`, `dcap()` |
| DCR / NNDR | Memorization | Generative models | `dcr()`, `nndr()` |
| RAPID | Attribute prediction | Synthetic (ML) | `rapid()` |
| `individual_risk()` | Pre-anonymization baseline | Any | `individual_risk()` |

Record linkage risk is most relevant for *perturbative* SDC methods
where the released data contains modified versions of real records. For
fully synthetic data, prefer the CAP family or distance-based metrics.


# Conclusion

The `recordLinkage()` function provides a flexible framework for
assessing re-identification risk in perturbed microdata. The four methods
cover a range of scenarios: deterministic linkage for general-purpose
assessment, probabilistic linkage for formal audits, PRAM linkage when
the perturbation mechanism is known, and predictive linkage for
high-dimensional or mixed-type quasi-identifiers.

Key takeaways:

- **Start with deterministic** as a baseline, then refine with
  method-specific information.
- **Weighting matters**: uniform is conservative; softmax/kernel give
  more nuanced risk profiles.
- **Direction matters**: forward risk assesses individual protection;
  reverse risk assesses the disclosiveness of released records.
- **Check `pct_true_in_set`** to verify that the candidate set is
  meaningful.

Future directions include integration with Franconi-Polettini population
frequency estimation, household-level matching, and attribute disclosure
assessment (extending the function to evaluate whether an attacker could
infer sensitive attributes after successful linkage).


# References

<div id="refs"></div>


# Session Info

The results in this vignette were obtained with the following
configuration.

```{r sessioninfo}
sessionInfo()
```
