---
title: "riskutility: Comprehensive Disclosure Risk and Data Utility Assessment for Anonymized and Synthetic Data in R"
author:
  - name: Matthias Templ
    affiliation: "Institute of Data Analysis and Process Design, Zurich University of Applied Sciences"
  - name: Oscar Thees
    affiliation: "Institute of Data Analysis and Process Design, Zurich University of Applied Sciences"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{riskutility: Comprehensive Disclosure Risk and Data Utility Assessment for Anonymized and Synthetic Data in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(riskutility)
```

# Introduction {#sec-intro}

Statistical agencies and data custodians face a fundamental challenge: releasing
data that is useful for analysis while protecting individual privacy. Two main
approaches exist: (i) traditional anonymization methods such as perturbation,
suppression, and generalization [@hundepool2012statistical;
@templ2017statistical], and (ii) synthetic data generation via statistical
models [@nowok2016synthpop; @templ2017simulation]. Both require rigorous
evaluation of *disclosure risk* and *data utility* --- yet existing tools
assess these two dimensions in isolation or cover only a narrow subset of the
relevant metrics.

The R package sdcMicro [@templ2015sdcmicro] provides frequency-based risk
estimation but limited utility assessment. synthpop [@nowok2016synthpop]
offers the CAP disclosure metric and propensity-score utility but not
distance-based or membership inference methods. In Python, SDMetrics covers
distance-based metrics and Anonymeter implements GDPR failure criteria, but
neither addresses attribution risk. No existing tool provides a unified
framework that spans all risk families and supports multivariate comparison
of multiple synthesis approaches.

The **riskutility** package for R [@R2025] fills this gap. It implements
over 30 risk measures spanning six paradigms --- frequency-based privacy
models, attribution-based (CAP), ML-based (RAPID), distance-based, record
linkage, and membership inference --- alongside 12 utility functions covering
global, distributional, structural, and predictive assessment. All functions
share a consistent S3 API with `print()`, `summary()`, and `plot()` methods
and accept a `synth_pair` container that bundles data and metadata.

**Quick start.** A comprehensive risk assessment requires just a few lines:

```{r intro-hook, eval=FALSE}
library(riskutility)
pair <- synth_pair(original, synthetic,
                   key_vars = c("age", "sex", "region"),
                   target_var = "income")
report <- disclosure_report(pair)
print(report)
```

This single call computes attribution-based risk (DCAP, TCAP, WEAP, DiSCO,
RAPID), distance-based risk (DCR, NNDR, IMS), privacy models ($k$-anonymity,
$l$-diversity, $t$-closeness), and membership inference measures (singling
out, linkability, NNAA) --- producing a pass/fail summary across all families.
For comparing multiple synthesis approaches, `rumap()` computes risk and
utility measures simultaneously, normalizes them to a common scale, and
identifies Pareto-optimal methods (Section \@ref(sec-case-rumap)).

**Paper overview.** Section \@ref(sec-background) introduces the disclosure
threat taxonomy and reviews related software. Section \@ref(sec-design)
describes the package architecture and the `synth_pair` container.
Sections \@ref(sec-risk) and \@ref(sec-utility) present the risk and utility
measures with mathematical formulations and worked examples.
Section \@ref(sec-comprehensive) demonstrates the complete practitioner
workflow on a realistic case study, comparing three synthesis approaches
with `disclosure_report()` and `rumap()`. Section \@ref(sec-discussion)
discusses limitations and future work. Section \@ref(sec-computational)
provides computational details.

```{r table1-scope, echo=FALSE}
scope <- data.frame(
  Category = c("Privacy models", "Attribution (CAP)",
                "ML-based (RAPID)", "Distance-based", "Record linkage",
                "Membership inference", "Utility measures", "Frameworks"),
  Functions = c(8, 4, 4, 6, "1 (4 methods)", 6, 12, 3),
  Paradigm = c("Equivalence class", "Matching", "Prediction",
                "Nearest neighbor", "Linkage", "Attack simulation",
                "Various", "Composite"),
  `Applicable to` = c(rep("Both", 8)),
  check.names = FALSE
)
knitr::kable(scope, caption = "Package scope at a glance.
             Both = applicable to traditionally anonymized and synthetic data.")
```


# Background: Threat Taxonomy and Related Software {#sec-background}

## Disclosure Threat Taxonomy {#sec-threats}

Every risk measure in riskutility addresses one or more of four disclosure
threats. The classical SDC taxonomy [@hundepool2012statistical;
@templ2017statistical] recognizes three types: identity, attribute, and
inferential disclosure. We extend this to four threats by separating
*membership disclosure* (from the ML privacy literature,
@shokri2017membership) and *memorization* (relevant to generative models)
as distinct categories, reflecting the broader scope of modern synthetic
data evaluation:

- **Identity disclosure**: An attacker links a released record to a specific
  individual. Measured by: `recordLinkage()`, `kanonymity()`,
  `individual_risk()`, `hitting_rate()`.
- **Attribute disclosure**: An attacker learns a sensitive attribute value
  through quasi-identifier matching. Measured by: `dcap()`, `tcap()`,
  `weap()`, `disco()`, `rapid()`.
- **Membership disclosure**: An attacker determines whether an individual's
  data was used to create the released dataset. Measured by:
  `mia_classifier()`, `domias()`, `nnaa()`, `singling_out()`,
  `delta_presence()`.
- **Memorization**: A generative model reproduces training records verbatim
  or near-verbatim. Measured by: `ims()`, `dcr()`, `nndr()`.

```{r table2-threats, echo=FALSE}
threats <- data.frame(
  Threat = c("Identity", "Attribute", "Membership", "Memorization"),
  Definition = c(
    "Attacker links record to individual",
    "Attacker learns sensitive value via linkage",
    "Attacker determines if individual is in dataset",
    "Generator reproduces training records"
  ),
  `Key measures` = c(
    "recordLinkage, kanonymity, individual_risk",
    "dcap, tcap, weap, disco, rapid",
    "mia_classifier, domias, nnaa, singling_out",
    "ims, dcr, nndr"
  ),
  check.names = FALSE
)
knitr::kable(threats, caption = "Disclosure threat taxonomy.")
```

## Risk Assessment Paradigms

riskutility implements six risk assessment paradigms, each addressing different
threats from the taxonomy above. We summarize each here; full definitions and
worked examples are in Section \@ref(sec-risk).

**Frequency-based (privacy models).** These methods assess privacy properties
of a single dataset based on quasi-identifier frequencies.
$k$-Anonymity [@samarati1998protecting] requires each combination of
quasi-identifiers to appear at least $k$ times. $l$-Diversity
[@machanavajjhala2007ldiversity] and $t$-closeness [@li2007tcloseness]
progressively strengthen the protection of sensitive attributes within
equivalence classes. See Section \@ref(sec-privacy-models).

**Attribution-based (CAP family).** The Correct Attribution Probability
framework [@taub2018differential] measures whether an attacker can infer
sensitive attribute values by matching quasi-identifiers between original and
released data. DCAP provides an aggregate measure; TCAP gives per-record
risk scores. See Section \@ref(sec-cap).

**ML-based (RAPID).** Risk of Attribute Prediction-Induced Disclosure
[@thees2026beyond] trains predictive models on the released data and evaluates
them on the original. Accurate predictions indicate information leakage. RAPID
captures complex non-linear relationships and variable interactions that the
CAP matching approach may miss. See Section \@ref(sec-rapid).

**Distance-based.** The holdout method compares distances from synthetic records
to training data vs. holdout data. If synthetic records are systematically
closer to training records, the generator has memorized rather than generalized.
However, @yao2025dcr demonstrate that passing distance-based tests does
not guarantee privacy (the "DCR Delusion"). See Section \@ref(sec-distance).

**Record linkage.** Directly simulates a re-identification attack using
deterministic (Gower distance), probabilistic (Fellegi-Sunter), PRAM-aware,
or propensity-score-based linkage [@fellegi1969theory;
@domingoferrer2003disclosure]. See Section \@ref(sec-recordlinkage) and the
companion vignette `vignette("recordLinkage")`.

**Membership inference.** Shadow model attacks [@shokri2017membership],
density-based detection [@hu2023domias], and GDPR failure criteria
[@stadler2022synthetic] assess whether an attacker can determine if a specific
individual's data was used during synthesis. See Section \@ref(sec-membership).

## Utility Assessment Paradigms

Utility measures quantify how well the released data preserves the
statistical properties of the original. We organize them into four groups
(details in Section \@ref(sec-utility)):

**Global utility** measures use propensity scores to quantify how
distinguishable the released data is from the original [@woo2009global;
@snoke2018general]. A single number summarizes overall data quality.

**Distributional utility** measures (Wasserstein, Hellinger, KS test) compare
marginal distributions variable by variable, identifying which specific
variables are poorly reproduced.

**Structural utility** measures (energy distance, PCA comparison, correlation
matrix comparison) assess whether multivariate relationships --- the
correlations and interactions that analysts depend on --- are preserved.

**Predictive utility** measures (feature importance stability, model
performance comparison) test whether models trained on released data
generalize to real-world predictions.

## Related Software

**In R.** sdcMicro [@templ2015sdcmicro] provides the most comprehensive
SDC toolkit, including frequency-based risk estimation, but focuses
primarily on *applying* anonymization methods. synthpop [@nowok2016synthpop]
provides the CAP disclosure metric and propensity-score utility, but does
not cover distance-based or membership inference methods. Neither package
provides a unified framework for comparing multiple risk and utility
measures simultaneously.

**In Python.** SDMetrics (part of the Synthetic Data Vault) provides
distance-based and statistical metrics for synthetic data quality. Anonymeter
[@giomi2023anonymeter] implements the three GDPR failure criteria from
@stadler2022synthetic (singling out, linkability, inference). Neither covers
attribution-based metrics (CAP, RAPID) or traditional privacy models
($k$-anonymity, $l$-diversity).

```{r table3-comparison, echo=FALSE}
comp <- data.frame(
  Measure = c("k-Anonymity", "l-Diversity", "t-Closeness",
              "DCAP/TCAP", "RAPID", "DCR/NNDR", "Record linkage",
              "MIA / GDPR criteria", "pMSE / SPECKS",
              "R-U map"),
  sdcMicro = c("freq()", "ldiversity()", "--",
               "--", "--", "dRisk()", "--",
               "--", "--", "--"),
  synthpop = c("--", "--", "--",
               "disclosure()", "--", "--", "--",
               "--", "utility.gen()", "--"),
  `SDMetrics (Py)` = c("--", "--", "--",
                        "--", "--", "Yes", "--",
                        "--", "--", "--"),
  `Anonymeter (Py)` = c("--", "--", "--",
                         "--", "--", "--", "--",
                         "Yes", "--", "--"),
  riskutility = c("kanonymity()", "ldiversity()", "tcloseness()",
                   "dcap(), tcap()", "rapid()", "dcr(), nndr()",
                   "recordLinkage()",
                   "mia_classifier(), singling_out()", "propscore(), specks()",
                   "rumap()"),
  check.names = FALSE
)
knitr::kable(comp,
             caption = "Risk/utility measures across R and Python packages.")
```

Py = Python.


# Software Design and Architecture {#sec-design}

## Design Philosophy

Existing R packages for statistical disclosure control --- sdcMicro
[@templ2015sdcmicro], synthpop [@nowok2016synthpop], and simPop
[@templ2017simulation] --- focus primarily on *applying* protection methods,
with risk and utility assessment provided as secondary features. riskutility
inverts this emphasis: it is a dedicated evaluation package, designed to be
used *after* protection has been applied, regardless of which tool generated
the released data.

The package follows four design principles:

1. **Consistent S3 architecture.** Every major function returns a typed S3
   object with `print()`, `summary()`, and `plot()` methods. We chose S3
   over S4 classes for three reasons: lighter memory footprint, simpler
   method dispatch for the typical R user, and easier interoperability with
   data.table and ggplot2 objects. All 24 risk/utility classes follow
   the same pattern, making the API predictable once one class is learned.

2. **Direction conventions.** Risk measures are oriented so that higher
   values always indicate higher disclosure risk. Utility measures are
   oriented so that higher values indicate higher utility (better
   preservation of statistical properties). Some measures (e.g., pMSE,
   Wasserstein distance) naturally use a "lower is better" scale; their
   interpretation is noted in the documentation, and `rumap()` applies
   the necessary direction transformation when normalizing to $[0, 1]$.

3. **Minimal core dependencies.** Frequency-based privacy models, CAP
   metrics, and distance-based measures require only base R, data.table, and
   ggplot2. ML-based methods require optional packages --- ranger for random
   forests in RAPID, xgboost for gradient boosting, and rpart for
   classification trees. These are loaded conditionally via
   `requireNamespace()` and produce informative error messages when absent.

4. **Integration over competition.** Rather than reimplementing synthesis or
   anonymization, the package provides `from_synthpop()`, `from_simPop()`,
   and `from_sdcMicro()` constructors that extract original and released
   data from objects created by these packages, wrapping them in the
   `synth_pair` container (Section \@ref(sec-synth-pair)). This ensures
   users can evaluate any protection method with a single consistent
   interface.

## The synth_pair Container {#sec-synth-pair}

The central data structure in riskutility is the `synth_pair` object, which
bundles original data, released data, variable roles, and metadata into a
single container:

```{r synth-pair-demo, eval=FALSE}
pair <- synth_pair(original, synthetic,
                   key_vars = c("age", "gender", "region"),
                   target_var = "income",
                   holdout = holdout_data)
```

The constructor stores the original and synthetic data frames alongside
their dimensions, automatically detects categorical (`cat_vars`) and numeric
(`num_vars`) columns, and retains user-specified quasi-identifiers
(`key_vars`), sensitive attribute (`target_var`), and optional holdout data.
This metadata elimination avoids a common source of error: specifying
different `key_vars` for different risk measures on the same dataset.

Once constructed, every risk and utility function in the package accepts a
`synth_pair` object as its first argument via S3 dispatch:

```{r synth-pair-dispatch, eval=FALSE}
# All functions accept synth_pair --- no parameter repetition:
dcap(pair)                    # Attribution risk
rapid(pair, model_type = "rf") # ML-based risk
propscore(pair)                # Propensity score utility
disclosure_report(pair)        # Full risk report
rumap(pair)                    # Risk-Utility map
```

This design means a practitioner sets up the data once and evaluates it
with as many measures as needed, confident that all analyses use the same
variable roles and data splits.

## S3 Method Pattern {#sec-s3-pattern}

Every exported risk/utility class follows an identical three-part pattern:

```{r s3-pattern, eval=FALSE}
# 1. Two equivalent calling conventions:
result <- dcap(pair)                                      # synth_pair method
result <- dcap(X, Y, key_vars = ..., target_var = ...)    # default method

# 2. Inspection:
print(result)                # One-screen summary with key statistic
s <- summary(result)         # Detailed statistics (returns summary.dcap)
print(s)                     # Formatted multi-line output

# 3. Visualization:
plot(result, which = 1)      # Plot type 1
plot(result, which = 1:2)    # Multiple plot types
```

The generic function dispatches via `UseMethod()`. The `synth_pair` method
extracts `original`, `synthetic`, `key_vars`, and `target_var` from the
container and delegates to the default method, which performs the actual
computation. The return object is a list with a class attribute (e.g.,
`"dcap"`). The `summary()` method returns a typed summary object (e.g.,
`"summary.dcap"`) with its own `print()` method, separating computation from
display.

Plot methods use an integer `which` parameter to select among multiple
visualization types. The number of available plot types varies by class,
from one (simple measures) to seven (`rumap`):

```{r table4-s3, echo=FALSE}
s3 <- data.frame(
  Class = c("dcap", "tcap", "weap", "disco", "rapid", "dcr", "nndr",
            "recordLinkageRisk", "kanonymity", "ldiversity",
            "tcloseness", "singling_out", "linkability",
            "propscore", "disclosure_report", "rumap"),
  `print` = rep("Yes", 16),
  `summary` = rep("Yes", 16),
  `plot (which)` = c("1--2", "1--3", "1--3", "1--3", "1--6", "1--3", "1--3",
                      "1--4", "1--2", "1--2",
                      "1--2", "1--2", "1--2",
                      "1--2", "1--2", "1--7"),
  check.names = FALSE
)
knitr::kable(s3, caption = "S3 classes and available methods (selected).")
```

This consistent API means that the code for evaluating one metric
generalizes directly to any other. A practitioner who has used `dcap()` can
immediately use `tcap()`, `rapid()`, or any other measure without consulting
the documentation for basic usage.

## Integration with the R Ecosystem {#sec-integration}

The `from_*` family of constructors bridges riskutility with the three main
R packages for statistical disclosure control and synthetic data generation:

```{r integration, eval=FALSE}
# From synthpop: pass synds object + original data
pair <- from_synthpop(synds_object, original_data,
                      key_vars = c("age", "sex"),
                      target_var = "income")

# From simPop: original data extracted automatically from simPopObj
pair <- from_simPop(simPopObj,
                    key_vars = c("age", "sex"),
                    target_var = "income")

# From sdcMicro: variable roles extracted from sdcMicroObj
pair <- from_sdcMicro(sdcMicroObj)
```

Each constructor returns a standard `synth_pair` object. `from_sdcMicro()`
additionally extracts variable roles (quasi-identifiers, sensitive
attributes, sample weights) from the sdcMicro S4 object, so that
`key_vars` and `target_var` need not be specified manually.
`from_synthpop()` supports multiple syntheses via the `m` parameter,
selecting a specific synthetic dataset from a `synds` object.
`from_simPop()` extracts sample weights when available, enabling
weighted risk calculations.


# Risk Measures {#sec-risk}

This section presents the six risk measure families, each illustrated with a
worked example using the same running dataset.

```{r running-data}
set.seed(42)
n <- 500
original <- data.frame(
  age = sample(18:85, n, replace = TRUE),
  sex = factor(sample(c("M", "F"), n, replace = TRUE)),
  education = factor(sample(c("Primary", "Secondary", "Tertiary"), n,
                            replace = TRUE, prob = c(0.3, 0.5, 0.2))),
  region = factor(sample(paste0("R", 1:5), n, replace = TRUE)),
  income = round(rlnorm(n, log(40000), 0.5))
)

# Synthetic: independent draws (low risk expected)
synthetic <- data.frame(
  age = sample(18:85, n, replace = TRUE),
  sex = factor(sample(c("M", "F"), n, replace = TRUE)),
  education = factor(sample(c("Primary", "Secondary", "Tertiary"), n,
                            replace = TRUE, prob = c(0.3, 0.5, 0.2))),
  region = factor(sample(paste0("R", 1:5), n, replace = TRUE)),
  income = round(rlnorm(n, log(40000), 0.5))
)

key_vars <- c("age", "sex", "education", "region")
target_var <- "income"

pair <- synth_pair(original, synthetic,
                   key_vars = key_vars, target_var = target_var)
```


## Privacy Models and Frequency-Based Risk {#sec-privacy-models}

These methods assess privacy properties of a *single* dataset based on
quasi-identifier frequencies. Originally developed for traditionally
anonymized data, they apply equally to synthetic data. They do not compare
original and released data; instead, they evaluate structural properties of
the released data alone.

**k-Anonymity** [@samarati1998protecting] partitions records into
equivalence classes (ECs) based on quasi-identifier values. The dataset
satisfies $k$-anonymity if every EC contains at least $k$ records:
$k = \min_i |\text{EC}(\mathbf{q}_i)|$, where $\mathbf{q}_i$ is the
quasi-identifier vector of record $i$. Small ECs are vulnerable to identity
disclosure because an attacker who knows a target's quasi-identifiers can
narrow them to fewer than $k$ candidates.

**l-Diversity** [@machanavajjhala2007ldiversity] strengthens $k$-anonymity by
requiring that each EC contains at least $l$ distinct values of the sensitive
attribute. This prevents attribute disclosure even when all records in an EC
share the same sensitive value (homogeneity attack).

**t-Closeness** [@li2007tcloseness] further requires that the distribution of
the sensitive attribute within each EC is close to its overall distribution.
The distance is measured by the Earth Mover's Distance (EMD), and $t$ is the
maximum EMD across all ECs.

```{r privacy-models}
# k-Anonymity: minimum equivalence class size
k_res <- kanonymity(synthetic, key_vars = key_vars)
print(k_res)

# l-Diversity: sensitive attribute diversity per EC
l_res <- ldiversity(synthetic, key_vars = key_vars,
                    sensitive_var = target_var)
print(l_res)

# t-Closeness: EMD between EC and overall distribution
t_res <- tcloseness(synthetic, key_vars = key_vars,
                    sensitive_var = target_var)
print(t_res)
```

With 68 unique age values, 2 sex levels, 3 education levels, and 5 regions,
there are up to $68 \times 2 \times 3 \times 5 = 2040$ possible QI
combinations for only $n = 500$ records. Most equivalence classes are
singletons, yielding $k = 1$ --- this is expected for fine-grained
quasi-identifiers and does not by itself indicate a problem with the
synthetic data. The $l$-diversity and $t$-closeness results should be
interpreted similarly.

The three models form a hierarchy of increasing protection: $k$-anonymity
guards against identity disclosure, $l$-diversity against homogeneity
attacks on sensitive attributes, and $t$-closeness against skewness attacks
where the EC distribution differs substantially from the population. In
practice, $k \geq 5$, $l \geq 2$, and $t \leq 0.15$ are common
thresholds, though appropriate values depend on the application context.

```{r table5-privacy, echo=FALSE}
privacy <- data.frame(
  Function = c("kanonymity()", "ldiversity()", "tcloseness()",
               "suda()", "individual_risk()", "population_uniqueness()",
               "attacker_risk()", "systemAnonymityLevel()"),
  Input = rep("Single dataset", 8),
  `Key output` = c("Min EC size", "Min distinct values per EC",
                    "Max EMD across ECs", "SUDA scores",
                    "Per-record frequency risk", "Estimated pop. uniques",
                    "Scenario-based risk", "Composite assessment"),
  `Threats` = c("Identity", "Attribute", "Attribute", "Identity",
                "Identity", "Identity", "Identity", "Identity"),
  check.names = FALSE
)
knitr::kable(privacy,
             caption = "Privacy models overview.")
```


## Attribution-Based Risk: The CAP Family {#sec-cap}

The Correct Attribution Probability framework [@taub2018differential] measures
attribute disclosure: can an attacker infer a sensitive value by matching
quasi-identifiers between original and released data?

For each original record $i$ with quasi-identifier values $\mathbf{q}_i$,
the attacker finds all records in the released data whose quasi-identifiers
match $\mathbf{q}_i$ (the *equivalence class*). If the sensitive attribute
is homogeneous within this class, the attacker learns the true value. The
**Targeted CAP** (TCAP) gives each record a risk score between 0 and 1:
$\text{TCAP}_i = \Pr(\text{correct attribution} \mid \mathbf{q}_i)$. The
**Differential CAP** (DCAP) is the mean CAP across all records:
$\text{DCAP} = n^{-1} \sum_i \text{CAP}_i$. The `summary()` method
computes a risk ratio $\text{DCAP} / \text{baseline}$ to contextualize
the result against random guessing.

The **WEAP** (Within-EC Attribution Probability) evaluates risk from the
released data alone, without access to the original, making it suitable
for data custodians who cannot share the original data with an auditor.
**DiSCO** (Disclosive in Synthetic, Correct in Original) identifies records
that are both confidently attributed in the released data and correctly
attributed in the original.

```{r cap-demo}
# TCAP: per-record risk (most informative member of CAP family)
tcap_res <- tcap(pair)
summary(tcap_res)
plot(tcap_res)
```

Since our running example uses independently generated synthetic data (no
relationship to the original), TCAP values should be close to the baseline
attribution probability. Records with TCAP above 0.1 warrant closer
inspection --- they may represent rare quasi-identifier combinations where
even random data can produce spurious matches.

```{r cap-table, echo=FALSE}
cap <- data.frame(
  Metric = c("DCAP", "TCAP", "WEAP", "DiSCO"),
  `Requires original?` = c("Yes", "Yes", "No", "Yes"),
  `Per-record?` = c("No", "Yes", "Yes", "Yes"),
  `Measures` = c("Mean attribution probability",
                 "Individual attribution risk",
                 "Within-EC homogeneity",
                 "Correct + confident attribution"),
  `Low risk` = c("ratio < 1.5", "< 0.1 per record",
                 "< 0.1", "< 5%"),
  check.names = FALSE
)
knitr::kable(cap,
             caption = "CAP family comparison with interpretation thresholds.")
```


## ML-Based Risk: RAPID {#sec-rapid}

Risk of Attribute Prediction-Induced Disclosure [@thees2026beyond] takes a
fundamentally different approach to attribute disclosure. Instead of matching
quasi-identifiers, RAPID trains a predictive model $\hat{f}$ on the released
data $(Y_{\mathcal{K}}, Y_s)$ and evaluates its predictions on the original
data: $\hat{s}_i = \hat{f}(X_{\mathcal{K},i})$. For numeric targets, a
record is *at risk* when the prediction error falls below a threshold
$\epsilon$: $e(s_i, \hat{s}_i) < \epsilon$, where $e(\cdot, \cdot)$ is
a configurable error metric (symmetric percentage error by default). The
RAPID score is the fraction of at-risk records:
$\text{RAPID} = n^{-1} \sum_i \mathbf{1}(e(s_i, \hat{s}_i) < \epsilon)$.
For categorical targets, a different evaluation applies: a record is at risk
when a gain or ratio score exceeds a threshold.

```{r rapid-demo}
rapid_res <- rapid(pair, model_type = "lm")
summary(rapid_res)
plot(rapid_res, which = c(1, 3))
```

With independently generated synthetic data, we expect the RAPID score to
be close to the baseline (random prediction accuracy). The threshold
sensitivity plot (`which = 3`) shows how the at-risk fraction changes
across threshold values, helping practitioners choose a threshold
appropriate for their context.

RAPID complements the CAP family in two ways. First, it captures non-linear
relationships and variable interactions --- a tree-based model can detect
that the combination of age and education jointly predicts income, even when
neither variable alone is informative. Second, it provides inferential tools
that the CAP family lacks: `rapid_test()` computes a permutation-based
$p$-value testing whether the RAPID score exceeds the null expectation,
`confint()` provides bootstrap confidence intervals, and
`rapid_threshold_select()` optimizes the threshold $\tau$ in a data-driven
manner.

Five model backends are available:

```{r rapid-models, echo=FALSE}
models <- data.frame(
  Model = c("lm", "rf", "cart", "gbm", "logit"),
  Package = c("stats", "ranger", "rpart", "xgboost", "stats"),
  Numeric = c("Yes", "Yes", "Yes", "Yes", "No"),
  Categorical = c("No", "Yes", "Yes", "Yes", "Yes"),
  Interactions = c("Manual", "Automatic", "Automatic", "Automatic", "Manual"),
  check.names = FALSE
)
knitr::kable(models, caption = "RAPID model backends.")
```


## Distance-Based Risk {#sec-distance}

Distance-based methods detect *memorization*: the failure mode where a
generative model reproduces training records verbatim or near-verbatim. The
key idea is the **holdout method** --- split the original data into a
training set $T$ (used for synthesis) and a holdout set $H$ (unseen by the
generator). For each synthetic record $y_j$, compute the Distance to
Closest Record in $T$ ($d_T$) and in $H$ ($d_H$). If the generator has
generalized, $d_T$ and $d_H$ should be comparable. If it has memorized,
$d_T$ will be systematically smaller:

$$\text{DCR\_share} = n^{-1} \sum_j \mathbf{1}\bigl(d_T(y_j) < d_H(y_j)\bigr), \qquad \text{DCR\_ratio} = \frac{\bar{d}_T}{\bar{d}_H}$$

A DCR share above 0.5 (or ratio below 1) indicates that synthetic records
are systematically closer to training data than holdout data, suggesting
memorization. The **NNDR** (Nearest
Neighbor Distance Ratio) provides a complementary view, and **IMS**
(Identical Match Share) counts exact copies. When an explicit holdout is
unavailable, `holdout_fraction` automatically splits the original data.

```{r distance-demo}
dcr_res <- dcr(pair, holdout_fraction = 0.2)
summary(dcr_res)
plot(dcr_res, which = 1)
```

**The DCR Delusion.** @yao2025dcr demonstrate that DCR can fail to
detect privacy leakage. Datasets deemed "private" by DCR may still be
vulnerable to membership inference attacks. The core problem is that DCR
measures distance in feature space, not in information space --- a record
can be far from any training record yet still leak information about
training individuals through learned correlations. Distance-based metrics
should always be complemented with other risk families, particularly
attribution and membership inference measures.

```{r distance-table, echo=FALSE}
dist <- data.frame(
  Metric = c("DCR", "NNDR", "IMS", "dRisk", "Hitting rate", "Epsilon ID"),
  Holdout = c("Yes", "Yes", "No", "No", "No", "No"),
  Detects = c("Memorization", "Memorization", "Exact copies",
              "Close records", "Close records", "Identifiability"),
  Limitation = c("DCR Delusion", "Assumes metric", "Misses near-copies",
                 "Continuous only", "Threshold-dependent", "Entropy estimation"),
  check.names = FALSE
)
knitr::kable(dist,
             caption = "Distance-based risk measures.")
```


## Record Linkage Risk {#sec-recordlinkage}

The `recordLinkage()` function directly simulates a re-identification attack by
linking each original record to the most similar record(s) in the anonymized
dataset. Four methods are implemented:

1. **Deterministic** (Gower distance): mixed-type distance with variable
   weighting and three candidate selection strategies.
2. **Probabilistic** (Fellegi-Sunter): agreement/disagreement weights via
   m/u-probabilities [@fellegi1969theory].
3. **PRAM-aware**: uses known transition matrices from Post-Randomization;
   supports forward and reverse risk perspectives.
4. **Predictive** (propensity score): logistic regression captures variable
   interactions that distance-based methods cannot detect. Note: this method
   is novel to riskutility; propensity scores are standard in causal
   inference but their use for disclosure risk assessment is not yet
   established in the SDC literature.

```{r recordlinkage-demo}
rl_res <- recordLinkage(pair, method = "deterministic")
print(rl_res)
```

For full details, see the companion vignette `vignette("recordLinkage")`.

```{r recordlinkage-table, echo=FALSE}
rl <- data.frame(
  Method = c("Deterministic", "Probabilistic", "PRAM", "Predictive"),
  Distance = c("Gower", "Fellegi-Sunter", "Transition prob.", "Propensity"),
  `Mixed types` = c("Yes", "Yes", "Categorical", "Yes"),
  Interactions = c("No", "No", "No", "Yes"),
  Requires = c("--", "--", "Transition matrix", "--"),
  check.names = FALSE
)
knitr::kable(rl,
             caption = "Record linkage methods.")
```


## Membership Inference and Anonymization Failure Criteria {#sec-membership}

This section groups two related but distinct concerns. The first three
measures (MIA, DOMIAS, NNAA) assess *membership disclosure* --- whether an
attacker can determine if a specific individual's data was used during
synthesis. The singling out and linkability attacks implement GDPR
anonymization failure criteria [@stadler2022synthetic]; singling out is
more precisely an *identity disclosure* test (does a predicate uniquely
identify an individual?), while linkability tests cross-dataset record
matching. We group them here because both originate from the same
regulatory framework.

```{r membership-demo, eval=FALSE}
# Membership inference attack (requires holdout + ranger)
mia_res <- mia_classifier(pair, holdout = holdout)

# GDPR singling out criterion (requires holdout)
so_res <- singling_out(pair, holdout = holdout, mode = "multivariate")
```

These examples are not evaluated because membership inference attacks require
a holdout dataset disjoint from both the training and synthetic data, and the
running example does not include one.

```{r membership-table, echo=FALSE}
mia <- data.frame(
  Metric = c("MIA classifier", "DOMIAS", "NNAA",
             "Singling out", "Linkability", "delta-Presence"),
  `Attack type` = c("Shadow model", "Density overfitting",
                     "Nearest neighbor", "Predicate-based",
                     "Record linkage", "Membership bounds"),
  Holdout = c("Yes", "Yes", "Yes", "Yes", "Yes", "No"),
  `GDPR criterion` = c("--", "--", "--",
                        "Art. 29 WP", "Art. 29 WP", "--"),
  `Low risk` = c("< 0.55", "< 0.6", "< 0.05",
                 "< 0.1", "< 0.1", "> 0.5"),
  check.names = FALSE
)
knitr::kable(mia,
             caption = "Membership inference and GDPR measures.")
```


## Cross-Family Comparison {#sec-rosetta}

No single metric tells the full story. Applying all families to the same
dataset reveals complementary and sometimes contradictory information.
Consider the contrast between our independently generated synthetic data
(which should be safe) and a near-copy:

```{r rosetta}
# Near-copy: original + small noise (high risk expected)
set.seed(99)
near_copy <- original
near_copy$age <- near_copy$age + sample(-1:1, n, replace = TRUE)
near_copy$income <- near_copy$income + round(rnorm(n, 0, 500))
pair_risky <- synth_pair(original, near_copy,
                         key_vars = key_vars, target_var = target_var)

# Compare key metrics across the two datasets
comparison <- data.frame(
  Metric = c("DCAP", "RAPID (lm)", "IMS"),
  Safe = c(
    dcap(pair)$dcap,
    rapid(pair, model_type = "lm", verbose = FALSE)$rapid,
    ims(pair)$ims
  ),
  Risky = c(
    dcap(pair_risky)$dcap,
    rapid(pair_risky, model_type = "lm", verbose = FALSE)$rapid,
    ims(pair_risky)$ims
  )
)
comparison$Safe <- round(comparison$Safe, 4)
comparison$Risky <- round(comparison$Risky, 4)
knitr::kable(comparison,
             caption = "Cross-family comparison: safe vs. risky synthetic data.")
```

The near-copy shows elevated risk across all families, but the magnitude
and interpretation differ. Attribution measures (DCAP) and ML-based measures
(RAPID) quantify *information leakage*: how much can an attacker learn about
sensitive attributes? Distance-based measures (IMS) quantify *memorization*:
are training records reproduced? These complementary perspectives mean
that a dataset can pass one family's tests while failing another's ---
a thorough evaluation uses at least one measure from each family.


# Data Utility Measures {#sec-utility}

A dataset that passes all risk checks but destroys the analytical value of
the data is useless. Utility measures quantify how well the released data
preserves the statistical properties of the original, completing the other
half of the risk-utility equation. We organize them by the practitioner
question they answer.

## "Is my synthetic data any good?" --- Quick Assessment {#sec-utility-quick}

Global utility measures give a single-number verdict by asking: *can a
classifier tell original and synthetic records apart?*

The propensity score method [@woo2009global; @snoke2018general] pools
original ($X$, $n_X$ records) and synthetic ($Y$, $n_Y$ records) data,
labels them (0/1), and fits a classifier (random forest by default;
logistic regression optionally). The **pMSE** (propensity score Mean
Squared Error) measures how well the model discriminates:

$$\text{pMSE} = \frac{1}{N} \sum_{i=1}^{N} \left(\hat{p}_i - c\right)^2$$

where $N = n_X + n_Y$ and $c = n_Y / N$ is the class ratio (equal to
0.5 when datasets are the same size). If original and synthetic records
are indistinguishable, $\hat{p}_i \approx c$ for all records and
pMSE $\approx 0$. The **SPECKS** variant applies the Kolmogorov-Smirnov
test statistic to the propensity score distributions, providing a
nonparametric alternative to pMSE.

```{r utility-quick}
# Propensity score utility
prop_res <- propscore(pair)
summary(prop_res)
plot(prop_res)
```

A pMSE below 0.1 generally indicates good utility. The propensity score
plot reveals *where* the synthetic data differs: if the density of
propensity scores is shifted rightward for original records, the classifier
has found systematic differences.

## "Which variables are poorly reproduced?" --- Univariate Diagnostics {#sec-utility-univariate}

When global utility is poor, per-variable measures identify which variables
are responsible. For **numeric** variables, the Wasserstein distance
(earth mover's distance) measures the cost of transforming one distribution
into another. For **categorical** variables, the Hellinger distance measures
distributional overlap:

$$H(p, q) = \frac{1}{\sqrt{2}} \sqrt{\sum_{k=1}^{K} \left(\sqrt{p_k} - \sqrt{q_k}\right)^2}$$

where $p$ and $q$ are the probability vectors over $K$ categories. $H = 0$
indicates identical distributions; $H = 1$ indicates disjoint support.

```{r utility-univariate}
# Wasserstein distance for numeric variables
w_res <- compare_wasserstein(original, synthetic, num_var = "income",
                             var_type = "continuous")
print(w_res)

# Hellinger distance for categorical variables
h_res <- hellinger(original, synthetic, vars = c("sex", "education"))
print(h_res)
```

Since both datasets are drawn from the same distributions, we expect small
Wasserstein distances and Hellinger distances close to 0. In practice, a
Hellinger distance above 0.1 for a categorical variable signals that the
synthesis has distorted that variable's marginal distribution.

## "Are multivariate relationships preserved?" --- Structural Utility {#sec-utility-structural}

Marginal distributions can match perfectly while joint distributions diverge.
The **energy distance** provides a multivariate two-sample test that is
sensitive to differences in both location and scale:

```{r utility-structural}
# Energy distance: multivariate two-sample test
e_res <- energy_distance(original[, c("age", "income")],
                         synthetic[, c("age", "income")])
print(e_res)
```

A non-significant $p$-value ($p > 0.05$, obtained via permutation test)
indicates that the multivariate distributions are not distinguishable. Additional structural measures test
specific aspects: `compare_pca()` compares the principal component structure,
`compare_correlation_matrices()` tests whether bivariate associations are
preserved, `compare_embedding()` visualizes distributional overlap via
t-SNE or UMAP, and `compare_multivariate_distribution()` provides a joint
assessment based on density ratios.

## "Can I still fit models on the synthetic data?" --- Predictive Utility {#sec-utility-predictive}

For many users, the ultimate test of synthetic data is whether models trained
on it generalize to real-world predictions. `compare_feature_importance()`
fits a random forest on both original and synthetic data and compares the
variable importance rankings --- if the same variables are important in both,
the synthetic data preserves the predictive structure.
`compare_model_performance()` goes further, comparing actual predictive
accuracy (RMSE, AUC) between models trained on original vs. synthetic data.

```{r utility-predictive, eval=FALSE}
# Feature importance stability
fi_res <- compare_feature_importance(original, synthetic,
                                     formula = income ~ age + sex + education + region,
                                     method = "rpart")
print(fi_res)
```

```{r table7-utility, echo=FALSE}
util <- data.frame(
  `Use case` = c(rep("Quick assessment", 3),
                 rep("Univariate", 4),
                 rep("Multivariate", 3),
                 rep("Predictive", 2)),
  Function = c("propscore()", "specks()", "mqs()",
               "compare_wasserstein()", "hellinger()",
               "ci_proximity()", "compare_ks_test()",
               "energy_distance()", "compare_pca()",
               "compare_correlation_matrices()",
               "compare_feature_importance()",
               "compare_model_performance()"),
  `Data type` = c("Mixed", "Mixed", "Mixed",
                   "Numeric", "Categorical", "Numeric", "Numeric",
                   "Numeric", "Numeric", "Numeric",
                   "Mixed", "Mixed"),
  Interpretation = c("< 0.1: good", "< 0.05: good", "Higher = better",
                     "Lower = better", "< 0.1: good",
                     "> 0.8: good", "p > 0.05: good",
                     "p > 0.05: good", "High overlap: good",
                     "High corr: good",
                     "High corr: good", "Small gap: good"),
  check.names = FALSE
)
knitr::kable(util,
             caption = "Utility measures by use case.")
```


# Comprehensive Assessment: A Case Study {#sec-comprehensive}

This section demonstrates the complete practitioner workflow on a realistic
dataset, comparing three synthesis approaches with different privacy-utility
trade-offs.

## Scenario and Data {#sec-case-data}

Consider a statistical agency that wants to release a survey dataset ($n = 1000$)
containing demographic variables (age, sex, education, region) and a sensitive
income variable. The agency must evaluate three candidate release strategies
before choosing one.

```{r case-data}
# Original data with realistic correlations
set.seed(123)
N <- 1000
edu_levels <- c("Primary", "Secondary", "Tertiary")
age_groups <- c("20-29", "30-39", "40-49", "50-59", "60-69")
orig <- data.frame(
  age_group = factor(sample(age_groups, N, replace = TRUE)),
  sex = factor(sample(c("M", "F"), N, replace = TRUE)),
  education = factor(sample(edu_levels, N, replace = TRUE,
                            prob = c(0.25, 0.50, 0.25))),
  region = factor(sample(paste0("R", 1:4), N, replace = TRUE))
)
# Income depends on age group and education (realistic structure)
edu_effect <- c(Primary = 0, Secondary = 0.3, Tertiary = 0.7)
age_effect <- c("20-29" = 0, "30-39" = 0.15, "40-49" = 0.3,
                "50-59" = 0.4, "60-69" = 0.35)
orig$income <- round(exp(
  10 + age_effect[as.character(orig$age_group)] +
    edu_effect[as.character(orig$education)] + rnorm(N, 0, 0.4)
))

qi <- c("age_group", "sex", "education", "region")
sens <- "income"
```

We create three synthetic datasets that span the privacy-utility spectrum:

```{r case-synthesis}
set.seed(456)

# Method A: Independent marginals (safest, but destroys correlations)
synA <- data.frame(
  age_group = factor(sample(age_groups, N, replace = TRUE)),
  sex = factor(sample(c("M", "F"), N, replace = TRUE)),
  education = factor(sample(edu_levels, N, replace = TRUE,
                            prob = c(0.25, 0.50, 0.25))),
  region = factor(sample(paste0("R", 1:4), N, replace = TRUE)),
  income = sample(orig$income, N, replace = TRUE)
)

# Method B: Category-preserving bootstrap with income noise
idx_B <- sample(N, N, replace = TRUE)
synB <- orig[idx_B, ]
rownames(synB) <- NULL
synB$income <- round(synB$income * exp(rnorm(N, 0, 0.15)))
# Swap some age groups to add noise to QIs
swap_idx <- sample(N, round(0.2 * N))
synB$age_group[swap_idx] <- factor(sample(age_groups,
                                          length(swap_idx), replace = TRUE))

# Method C: Near-copy with minimal perturbation (risky)
synC <- orig
synC$income <- round(synC$income * exp(rnorm(N, 0, 0.03)))
```

Method A draws each variable independently, destroying the income-education
relationship. Method B resamples records with bootstrap, adds income noise,
and perturbs 20% of age groups. Method C perturbs only income minimally,
preserving the full QI structure.

## Step 1: Quick Risk Screening with disclosure_report() {#sec-case-report}

The `disclosure_report()` function automates the cross-family risk comparison
from Section \@ref(sec-rosetta). It computes multiple risk measures, evaluates
each against a threshold, and produces an overall pass/fail assessment. The
`compute` parameter selects which families to include: `"attribution"` (DCAP,
TCAP, WEAP, DiSCO, RAPID), `"distance"` (DCR, NNDR, IMS, dRisk, hitting
rate), `"privacy"` ($k$-anonymity, $l$-diversity, $t$-closeness, individual risk),
`"membership"` (singling out, linkability, NNAA, DOMIAS), or `"all"` for
every available metric.

```{r case-report}
# Screen each method with attribution + privacy measures
pair_A <- synth_pair(orig, synA, key_vars = qi, target_var = sens)
pair_B <- synth_pair(orig, synB, key_vars = qi, target_var = sens)
pair_C <- synth_pair(orig, synC, key_vars = qi, target_var = sens)

rep_A <- disclosure_report(pair_A, compute = c("attribution", "privacy"),
                           seed = 42, verbose = FALSE)
rep_B <- disclosure_report(pair_B, compute = c("attribution", "privacy"),
                           seed = 42, verbose = FALSE)
rep_C <- disclosure_report(pair_C, compute = c("attribution", "privacy"),
                           seed = 42, verbose = FALSE)

# Compare overall verdicts
verdicts <- data.frame(
  Method = c("A: Independent", "B: Bootstrap+noise", "C: Near-copy"),
  Overall = c(rep_A$overall_risk, rep_B$overall_risk, rep_C$overall_risk),
  Pass = c(rep_A$n_pass, rep_B$n_pass, rep_C$n_pass),
  Warn = c(rep_A$n_warn, rep_B$n_warn, rep_C$n_warn)
)
knitr::kable(verdicts, caption = "Quick risk screening across three methods.")
```

Two patterns emerge. First, the attribution metrics (DCAP, RAPID)
differentiate the three methods in the expected order: Method A (independent)
has the lowest attribution risk, Method C (near-copy) the highest. Second,
the privacy models ($k$-anonymity, $l$-diversity) flag all three methods
because they evaluate the *released data's structure alone* --- with 120
possible QI combinations ($5 \times 2 \times 3 \times 4$) and $n = 1000$
records, some equivalence classes will inevitably be small regardless of how
the data was generated. This illustrates a key lesson: privacy models and
attribution metrics answer different questions and should be interpreted
together, not in isolation.

## Step 2: Comparative Assessment with rumap() {#sec-case-rumap}

The `rumap()` function implements the multivariate Risk-Utility framework of
@thees2026beyond. Traditional R-U analysis plots a single risk measure against
a single utility measure, producing a two-dimensional trade-off curve. This
can be misleading: a method may appear optimal on one pair of measures while
performing poorly on another.

`rumap()` computes multiple risk and utility measures simultaneously,
normalizes all measures to $[0, 1]$ (with consistent direction: higher =
more risk or more utility), and identifies the **Pareto-optimal** methods ---
those not dominated on any dimension.

```{r case-rumap}
ru <- rumap(orig,
            list("A: Independent" = synA,
                 "B: Bootstrap+noise" = synB,
                 "C: Near-copy" = synC),
            risk_measures = c("dcap", "tcap", "ims"),
            utility_measures = c("pmse", "wasserstein"),
            key_vars = qi, target_var = sens,
            seed = 42)
print(ru)
```

```{r case-rumap-scatter, fig.width=8, fig.height=6}
plot(ru, which = 1)  # R-U scatterplot with Pareto front
```

The R-U scatterplot places each method in the composite risk-utility plane.
Methods in the lower-right corner (low risk, high utility) are preferred.
Method B occupies the middle ground --- moderate risk and moderate utility ---
making it the natural compromise candidate.

```{r case-rumap-heatmap, fig.width=8, fig.height=5}
plot(ru, which = 2)  # Heatmap of individual measures
```

The heatmap reveals *why* the methods differ. Method A achieves low risk
across all measures but has poor utility (high pMSE, large Wasserstein
distance). Method C has excellent utility but elevated attribution risk.
Method B balances the two, though its exact acceptability depends on the
agency's risk tolerance.

## Step 3: Decision {#sec-case-decision}

The analysis supports a structured decision:

- **Method A** is appropriate when the data is released to the general
  public and any re-identification would be unacceptable. The utility loss
  is the price of maximum safety.
- **Method B** is appropriate for controlled-access research environments
  where moderate risk is acceptable in exchange for preserving analytical
  relationships.
- **Method C** should be rejected --- its risk profile is too close to the
  original data to provide meaningful protection.

This iterative workflow --- screen with `disclosure_report()`, compare with
`rumap()`, and refine the synthesis parameters --- is the core use case that
riskutility is designed to support.


# Summary and Discussion {#sec-discussion}

## Contributions

The riskutility package makes five contributions to the R ecosystem for
statistical disclosure control:

1. **Comprehensive coverage.** It is the first R package to unify all six
   risk assessment paradigms --- frequency-based privacy models, attribution
   (CAP), ML-based (RAPID), distance-based, record linkage, and membership
   inference --- under a single API. No existing R or Python package provides
   this breadth.

2. **Novel implementations.** It provides the first R implementations of
   RAPID (Risk of Attribute Prediction-Induced Disclosure), two of the three
   GDPR failure criteria from @stadler2022synthetic (singling out and
   linkability; the third criterion, inference, is addressed by RAPID via a
   complementary ML-based approach), $t$-closeness, DOMIAS density-based
   membership inference, and four-method record linkage with predictive
   (propensity-score) matching.

3. **Unified API.** The `synth_pair` container and consistent S3 class
   pattern (Section \@ref(sec-design)) eliminate parameter repetition and
   ensure that practitioners can switch between risk measures without
   learning new interfaces.

4. **Multivariate R-U mapping.** The `rumap()` function
   (Section \@ref(sec-case-rumap)) implements the framework of
   @thees2026beyond for comparing multiple synthesis approaches on multiple
   risk and utility dimensions simultaneously, with Pareto frontier
   identification and seven visualization types.

5. **Ecosystem integration.** The `from_sdcMicro()`, `from_synthpop()`, and
   `from_simPop()` constructors allow practitioners to evaluate data
   produced by any of the three main R packages for SDC and synthesis
   without manual data extraction.

## Limitations and Recommendations

**riskutility evaluates but does not generate protected data.** It is
designed to work alongside synthesis and anonymization tools, not to replace
them. The package cannot tell users *how* to improve a failing synthesis ---
only *that* it fails and *where*.

**Computational scalability.** Most measures scale linearly or quadratically
with $n$. The main bottlenecks are RAPID with random forests ($O(n \log n)$
per tree), energy distance ($O(n^2)$), and membership inference attacks
(requiring model training). For datasets exceeding $n = 50{,}000$ records,
we recommend using the faster model backends (RAPID with `model_type = "lm"`)
and sampling-based approximations.

**No formal privacy guarantees.** All measures in riskutility provide
*empirical* risk assessment, not *formal* guarantees such as differential
privacy (DP). A low DCAP score does not prove that no attacker can
succeed --- it only shows that the specific attack simulated by DCAP did
not succeed. Empirical and formal approaches are complementary: DP
provides worst-case guarantees independent of the attacker's capabilities,
while empirical measures reveal the *actual* risk profile for specific
attack scenarios and help practitioners interpret whether formal guarantees
translate into meaningful protection in their context.

**Key variable selection.** Results depend heavily on the choice of
quasi-identifiers. Including too few QIs underestimates risk (the attacker
may know more than assumed). Including too many QIs overestimates risk
(most equivalence classes become singletons). Practitioners should base QI
selection on a realistic threat model, not on convenience.

**Threshold interpretation.** The pass/fail thresholds used by
`disclosure_report()` are pragmatic defaults, not universal standards.
Different application contexts (public release vs. controlled access,
health data vs. economic surveys) require different thresholds.
Practitioners should calibrate thresholds to their specific risk tolerance.

## Practitioner Guide

```{r table9-guide, echo=FALSE}
guide <- data.frame(
  Question = c(
    "Can an attacker infer sensitive values?",
    "Did the generator memorize training records?",
    "Can an attacker determine membership?",
    "Are small groups exposed?",
    "How distinguishable is the released data?",
    "Are variable distributions preserved?",
    "Are multivariate relationships preserved?",
    "Which synthesis method is best overall?"
  ),
  Family = c(
    "Attribution (CAP, RAPID)",
    "Distance (DCR, IMS)",
    "Membership (MIA, NNAA)",
    "Privacy models (k-anon, l-div)",
    "Global utility (pMSE)",
    "Univariate (Wasserstein, Hellinger)",
    "Structural (energy, PCA)",
    "Framework (rumap)"
  ),
  `Key functions` = c(
    "dcap(), rapid()",
    "dcr(), ims()",
    "mia_classifier(), nnaa()",
    "kanonymity(), ldiversity()",
    "propscore()",
    "compare_wasserstein(), hellinger()",
    "energy_distance(), compare_pca()",
    "rumap(), disclosure_report()"
  ),
  check.names = FALSE
)
knitr::kable(guide,
             caption = "Practitioner guide: matching questions to metric families.")
```

We recommend the following minimal evaluation protocol:

1. **Always run** `disclosure_report()` with `compute = "all"` as a first
   screening. This takes seconds and covers all metric families.
2. **For publication-quality assessment**, use `rumap()` to compare multiple
   synthesis approaches and identify Pareto-optimal methods.
3. **For regulatory compliance** (e.g., GDPR), include singling out and
   linkability tests, which implement the Article 29 Working Party criteria.
4. **Interpret distance-based metrics cautiously.** Following @yao2025dcr,
   do not rely on DCR/NNDR alone for privacy assessment.

## Future Work

Four extensions are planned: (i) a Shiny dashboard for interactive
evaluation, making the package accessible to non-programmers;
(ii) integration with differential privacy frameworks, enabling comparison
of empirical risk scores against formal privacy budgets;
(iii) computational optimizations for large datasets ($n > 50{,}000$),
including sampling-based approximations for quadratic-cost measures; and
(iv) population-level risk estimation from sample data, connecting to the
super-population models used in survey statistics.


# Computational Details {#sec-computational}

All computations in this paper were performed using R `r getRversion()`
[@R2025] with the riskutility package version `r packageVersion("riskutility")`.
The package is available from GitHub and will be submitted to CRAN upon
publication.

Core dependencies include data.table for efficient data manipulation and
ggplot2 for all visualizations. ML-based methods require optional packages:
ranger (random forests for RAPID), rpart (CART), xgboost (gradient boosting),
and caret (model training for propensity scores and MIA). These are loaded
conditionally via `requireNamespace()` and produce informative error messages
when absent.

The running examples in Sections \@ref(sec-risk) and \@ref(sec-utility) use
a synthetic dataset with $n = 500$ records and five variables. The case study
in Section \@ref(sec-comprehensive) uses $n = 1000$ records. On a standard
laptop (Apple M1, 16 GB RAM), the full `disclosure_report()` with all metric
families completes in under 30 seconds for $n = 1000$.

```{r session-info}
sessionInfo()
```


# References {-}
