---
title: "riskutility: Measuring Disclosure Risk and Data Utility for Synthetic Data"
author:
  - name: Matthias Templ
    affiliation: Institute of Statistics and Mathematical Methods in Economics, Vienna University of Technology
    address: Wiedner Hauptstraße 8-10, 1040 Vienna, Austria
    email: matthias.templ@gmail.com
    orcid: 0000-0002-8638-5276
abstract: >
  The R package **riskutility** provides a comprehensive framework for measuring
  disclosure risk and data utility in synthetic and anonymized datasets. It implements
  attribution-based privacy metrics (DCAP, TCAP, WEAP, DiSCO), distance-based metrics
  (DCR, NNDR, IMS), RF proximity-based memorization detection, and various utility measures. This vignette demonstrates the
  complete workflow from synthetic data generation to risk assessment, explains metric
  interpretation, and shows integration with existing R packages for statistical
  disclosure control.
keywords:
  - synthetic data

  - disclosure risk
  - data utility
  - privacy
  - statistical disclosure control
  - R
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{riskutility: Measuring Disclosure Risk and Data Utility for Synthetic Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

Synthetic data generation has emerged as a key technique for protecting privacy while
preserving data utility for statistical analysis. However, evaluating the quality of
synthetic data requires careful assessment of both **disclosure risk** (how much
private information might be revealed) and **data utility** (how well the synthetic
data preserves analytical properties of the original).
The **riskutility** package provides a unified framework for these assessments,
implementing state-of-the-art metrics from the statistical disclosure control
literature. This vignette demonstrates:

1. Core concepts of disclosure risk and data utility
2. Attribution-based privacy metrics (CAP family)
3. Distance-based privacy metrics (holdout method)
4. Utility assessment methods
5. Comprehensive reporting with `disclosure_report()`
6. Integration with synthpop and simPop packages

# Installation and Setup

```{r install, eval=FALSE}
# Install from CRAN (when available)
install.packages("riskutility")

# Or install development version from GitHub
# devtools::install_github("mtempl/riskutility")
```

```{r load}
library(riskutility)
library(ggplot2)
```
# Example Data

Throughout this vignette, we use a simulated dataset representing survey data with
demographic and income information. We create both an "original" dataset and a
"synthetic" version with intentionally introduced differences.

```{r example-data}
set.seed(42)
n <- 500

# Original data
original <- data.frame(
  age = as.integer(sample(18:80, n, replace = TRUE)),
  gender = factor(sample(c("Male", "Female"), n, replace = TRUE)),
  education = factor(sample(c("Primary", "Secondary", "Tertiary"), n,
                           replace = TRUE, prob = c(0.2, 0.5, 0.3))),
  region = factor(sample(paste0("Region_", 1:5), n, replace = TRUE)),
  income = round(rgamma(n, shape = 2, scale = 15000), -2),
  health_status = factor(sample(c("Good", "Fair", "Poor"), n,
                                replace = TRUE, prob = c(0.6, 0.3, 0.1)))
)

# Synthetic data (with some noise and distributional shifts)
synthetic <- data.frame(
  age = as.integer(pmin(80, pmax(18, original$age + sample(-3:3, n, replace = TRUE)))),
  gender = factor(sample(c("Male", "Female"), n, replace = TRUE)),
  education = factor(sample(c("Primary", "Secondary", "Tertiary"), n,
                           replace = TRUE, prob = c(0.22, 0.48, 0.30))),
  region = factor(sample(paste0("Region_", 1:5), n, replace = TRUE)),
  income = round(rgamma(n, shape = 2.1, scale = 14500), -2),
  health_status = factor(sample(c("Good", "Fair", "Poor"), n,
                                replace = TRUE, prob = c(0.58, 0.32, 0.10)))
)

head(original)
```

# Disclosure Risk Metrics

## Conceptual Framework

Disclosure risk in synthetic data can be understood through two main paradigms:

1. **Attribution disclosure**: Can an intruder learn sensitive information about
   an individual by matching quasi-identifiers between original and synthetic data?

2. **Identity disclosure**: Can an intruder determine that a specific synthetic
   record corresponds to a real individual (memorization)?

The **riskutility** package implements metrics for both paradigms.

## Attribution-Based Metrics (CAP Family)

### Correct Attribution Probability (DCAP)

The Differential Correct Attribution Probability (DCAP) measures how much better
an intruder can guess a sensitive target variable when they have access to the
synthetic data, compared to random guessing.

```{r dcap}
# Define quasi-identifiers (key variables) and sensitive attribute (target)
key_vars <- c("age", "gender", "education", "region")
target_var <- "health_status"

# Compute DCAP
dcap_result <- dcap(
  X = original,
  Y = synthetic,
  key_vars = key_vars,
  target_var = target_var
)

print(dcap_result)
```

```{r dcap-summary}
summary(dcap_result)
```

**Interpretation Guidelines:**

| DCAP Value | Risk Level | Interpretation |
|------------|------------|----------------|
| < 0.05 | Low | Minimal additional disclosure risk |
| 0.05 - 0.15 | Moderate | Some risk; review key variable selection |
| 0.15 - 0.30 | Elevated | Consider additional protection |
| > 0.30 | High | Significant privacy concern |

### Targeted CAP (TCAP)

While DCAP provides an aggregate measure, TCAP computes per-record risk scores,
identifying which individuals face the highest disclosure risk.

```{r tcap}
tcap_result <- tcap(
  X = original,
  Y = synthetic,
  key_vars = key_vars,
  target_var = target_var
)

print(tcap_result)
```

```{r tcap-plot, fig.cap="Distribution of per-record TCAP scores"}
plot(tcap_result)
```

High TCAP values for specific records indicate that those individuals' sensitive
attributes are more easily inferred from the synthetic data.

### Within Equivalence Class Attribution Probability (WEAP)

WEAP focuses on the risk within equivalence classes (groups of records sharing
the same quasi-identifier values).

```{r weap}
# WEAP only uses synthetic data (Y) - it measures disclosure risk
# within the synthetic dataset itself
weap_result <- weap(
  synthetic,
  key_vars = key_vars,
  target_var = target_var
)

print(weap_result)
summary(weap_result)
```

### DiSCO (Disclosive in Synthetic Correct Original)
DiSCO identifies records where the synthetic data correctly reveals the original
sensitive value with high confidence.

```{r disco}
disco_result <- disco(
  X = original,
  Y = synthetic,
  key_vars = key_vars,
  target_var = target_var
)

print(disco_result)
```

### RAPID (Risk of Attribute Prediction-Induced Disclosure)

RAPID uses machine learning models to assess disclosure risk. Unlike CAP-based methods
that use exact or fuzzy matching, RAPID trains a predictive model on synthetic data
and evaluates how well it can predict sensitive attributes in the original data.

```{r rapid, eval=requireNamespace("ranger", quietly = TRUE)}
rapid_result <- rapid(
  X = original,
  Y = synthetic,
  key_vars = key_vars,
  target_var = target_var,
  model_type = "rf",
  cat_tau = 0.3,
  cat_eval_method = "RCS_marginal"
)

print(rapid_result)
summary(rapid_result)
```

**Interpretation**: A low RAPID score (< 0.05) indicates that the ML model cannot
predict the sensitive attribute much better than baseline, suggesting good privacy
protection. Higher scores indicate potential disclosure risk.

## Distance-Based Metrics (Holdout Method)

Distance-based metrics detect **memorization** by comparing how close synthetic
records are to training data versus held-out data. If synthetic records are
systematically closer to training data, this suggests the model has memorized
training examples.

### Distance to Closest Record (DCR)

```{r dcr}
# Split original data into training and holdout
set.seed(123)
n_orig <- nrow(original)
train_idx <- sample(n_orig, size = floor(0.7 * n_orig))

train_data <- original[train_idx, ]
holdout_data <- original[-train_idx, ]

# For this example, synthetic was generated from full original
# In practice, synthetic should be generated only from train_data

dcr_result <- dcr(
  X = train_data,
  Y = synthetic,
  holdout = holdout_data,
  progress = FALSE
)

print(dcr_result)
```

```{r dcr-plot, fig.cap="DCR distribution comparison"}
plot(dcr_result)
```

**Interpretation**: If the median ratio of distances (synthetic-to-training vs
synthetic-to-holdout) is close to 1, there is no evidence of memorization.
Values significantly below 1 suggest memorization.

### Nearest Neighbor Distance Ratio (NNDR)

NNDR refines DCR by computing the ratio of the distance to the nearest neighbor
versus the second-nearest neighbor.

```{r nndr}
nndr_result <- nndr(
  X = train_data,
  Y = synthetic,
  holdout = holdout_data,
  progress = FALSE
)

print(nndr_result)
summary(nndr_result)
```

### Identical Match Share (IMS)

IMS measures the proportion of synthetic records that exactly match a training record.

```{r ims}
ims_result <- ims(
  X = train_data,
  Y = synthetic
)

print(ims_result)
```

An IMS close to 0 indicates good privacy protection (no exact copies of training data).

### RF Proximity Privacy (`rf_privacy`)

RF proximity offers a data-adaptive alternative to DCR for memorization detection.
Instead of computing Gower or Euclidean distances, it trains a Random Forest to
discriminate original from synthetic records and measures how often synthetic
records share terminal nodes with training vs. holdout records. This approach
naturally handles mixed data types and captures complex variable interactions.

The key metric is `max_prox_share`: the fraction of synthetic records whose
nearest real-data neighbor (by terminal node co-occurrence) belongs to the
training set rather than the holdout. Values near 0.5 indicate no memorization;
values above 0.5 suggest the synthetic data is systematically closer to training
records.

```{r rf-privacy, eval=requireNamespace("ranger", quietly=TRUE)}
rfp_result <- rf_privacy(
  X = train_data,
  Y = synthetic,
  holdout = holdout_data,
  n_trees = 200,
  null_test = FALSE,  # skip permutation test for speed
  seed = 42
)

print(rfp_result)
```

```{r rf-privacy-plot, fig.cap="RF proximity: training vs holdout", eval=requireNamespace("ranger", quietly=TRUE)}
plot(rfp_result, which = 1)
```

**When to use RF proximity vs DCR**: Use `rf_privacy()` when you have 20+
variables with complex interactions or mixed data types. Use `dcr()` with Gower
distance for simpler settings where interpretability matters more, or when speed
is a priority (DCR is faster for n < 5,000).

# Data Utility Metrics

Protecting privacy is only valuable if the synthetic data remains useful for
analysis. The **riskutility** package provides several utility metrics.

## Propensity Score Analysis

The propensity score method trains a classifier to distinguish original from
synthetic records. If the classifier performs poorly (accuracy near 0.5), the
distributions are similar.

```{r propscore}
ps_result <- propscore(
  X = original,
  Y = synthetic,
  form = ~ age + income + education + gender
)

print(ps_result)
```

```{r propscore-summary}
summary(ps_result)
```

**Interpretation**: A propensity score close to 0 indicates good utility
preservation (the classifier cannot distinguish original from synthetic).
The closer the score distributions for original and synthetic data, the
better the utility.

### Ranger-Based Propensity Scores

The `method = "ranger"` option uses a probability forest (via the **ranger**
package) instead of `randomForest`, providing OOB propensity scores and
proximity-based structural metrics that quantify within-class vs. cross-class
similarity.

```{r propscore-ranger, eval=requireNamespace("ranger", quietly=TRUE)}
ps_ranger <- propscore(
  X = original,
  Y = synthetic,
  method = "ranger",
  proximity = "summary"
)

summary(ps_ranger)
```

The `structure_ratio` compares cross-class proximity to within-class proximity.
Values near 1 indicate the forest cannot distinguish the two datasets (good
utility). Values well below 1 suggest the datasets occupy distinct regions of
feature space.

## Distribution Comparisons

### Continuous Variables

```{r compare-dist}
dist_result <- compare_distributions_cont(
  X = original,
  Y = synthetic,
  variables = c("age", "income")
)

print(dist_result)
```

```{r compare-dist-plot, fig.cap="Distribution comparison for continuous variables"}
plot(dist_result)
```

### Categorical Variables

```{r compare-cat}
chisq_result <- compare_chisq_gof(
  X = original,
  Y = synthetic,
  cat_vars = "gender"
)

print(chisq_result)
```

## Summary Statistics Comparison

```{r summary-stats}
stats_result <- compare_multivariate_summary_statistics(
  X = original,
  Y = synthetic,
  cont_vars = c("age", "income"),
  cat_vars = c("gender", "education")
)

print(stats_result$continuous)
```

# Comprehensive Disclosure Report

The `disclosure_report()` function provides a unified assessment combining multiple
risk and utility metrics.

```{r disclosure-report}
report <- disclosure_report(
  X = original,
  Y = synthetic,
  key_vars = key_vars,
  target_var = target_var
)

print(report)
```

```{r report-summary}
summary(report)
```

```{r report-plot, fig.cap="Comprehensive disclosure risk visualization"}
plot(report)
```

# Integration with synthpop

The **riskutility** package integrates seamlessly with the **synthpop** package
for synthetic data generation.

```{r synthpop-integration, eval=FALSE}
library(synthpop)

# Generate synthetic data with synthpop
synth_obj <- syn(original, seed = 123)

# Extract synthetic data for riskutility
synthetic_sp <- synth_obj$syn

# Run disclosure assessment
dcap_sp <- dcap(
  X = original,
  Y = synthetic_sp,
  key_vars = key_vars,
  target_var = target_var
)

# Or use the from_synthpop helper (if available)
# pair <- from_synthpop(synth_obj, original)
```

# Integration with simPop

For users of the **simPop** package, **riskutility** provides S4 method support.

```{r simPop-integration, eval=FALSE}
library(simPop)

# Assuming you have a simPopObj object
# synthetic_data <- simPop::pop(simPopObj)

# Use with riskutility
# dcap_result <- dcap(X = original, Y = synthetic_data, ...)
```

# Metric Comparison

Different metrics capture different aspects of disclosure risk. Here we compare
results across metrics for the same dataset.

```{r metric-comparison}
# Extract scalar values from results
dcap_val <- dcap_result$dcap[1]
weap_val <- weap_result$weap_mean[1]
disco_val <- disco_result$pct_disco[1]
ims_val <- ims_result$ims[1]
ps_score_val <- ps_result$ps_score[1]

# Include RAPID if available
if (exists("rapid_result")) {
  rapid_val <- rapid_result$rapid[1]
  metrics <- c("DCAP", "Mean WEAP", "DiSCO %", "RAPID", "IMS", "PS Score")
  values <- round(c(dcap_val, weap_val, disco_val, rapid_val, ims_val, ps_score_val), 4)
  interp <- c(
    ifelse(dcap_val < 0.05, "Low risk",
           ifelse(dcap_val < 0.15, "Moderate", "Elevated")),
    ifelse(weap_val < 0.5, "Low risk",
           ifelse(weap_val < 0.8, "Moderate", "Elevated")),
    ifelse(disco_val < 1, "Low risk",
           ifelse(disco_val < 5, "Moderate", "Elevated")),
    ifelse(rapid_val < 0.05, "Low risk",
           ifelse(rapid_val < 0.15, "Moderate", "Elevated")),
    ifelse(ims_val < 0.01, "No memorization", "Potential memorization"),
    ifelse(ps_score_val < 0.1, "Good utility",
           ifelse(ps_score_val < 0.3, "Acceptable", "Poor utility"))
  )
} else {
  metrics <- c("DCAP", "Mean WEAP", "DiSCO %", "IMS", "PS Score")
  values <- round(c(dcap_val, weap_val, disco_val, ims_val, ps_score_val), 4)
  interp <- c(
    ifelse(dcap_val < 0.05, "Low risk",
           ifelse(dcap_val < 0.15, "Moderate", "Elevated")),
    ifelse(weap_val < 0.5, "Low risk",
           ifelse(weap_val < 0.8, "Moderate", "Elevated")),
    ifelse(disco_val < 1, "Low risk",
           ifelse(disco_val < 5, "Moderate", "Elevated")),
    ifelse(ims_val < 0.01, "No memorization", "Potential memorization"),
    ifelse(ps_score_val < 0.1, "Good utility",
           ifelse(ps_score_val < 0.3, "Acceptable", "Poor utility"))
  )
}

# Add RF Privacy if available
if (exists("rfp_result")) {
  rfp_val <- rfp_result$max_prox_share
  metrics <- c(metrics, "RF Privacy")
  values <- c(values, round(rfp_val, 4))
  interp <- c(interp,
    ifelse(rfp_val <= 0.55, "No memorization signal",
           "Memorization signal"))
}

comparison <- data.frame(
  Metric = metrics,
  Value = values,
  Interpretation = interp
)

knitr::kable(comparison, caption = "Summary of disclosure risk and utility metrics")
```

# Best Practices

## Choosing Key Variables

Key variables (quasi-identifiers) should be selected based on:

1. **Availability to potential intruders**: Variables an attacker might know
2. **Uniqueness**: Combinations that identify individuals
3. **Linkage potential**: Variables present in external databases

Common quasi-identifiers include age, gender, geographic location, and occupation.

## Interpreting Results

1. **No single metric is sufficient**: Use multiple metrics to get a complete picture
2. **Context matters**: Acceptable risk levels depend on data sensitivity and use case
3. **Compare to baselines**: DCAP values should be compared to the baseline (random guessing)
4. **Consider record-level risk**: Aggregate metrics may hide high-risk individuals (use TCAP)

## Iterative Refinement

If disclosure risk is too high:

1. Add more noise to the synthesis process
2. Reduce the granularity of quasi-identifiers (e.g., age groups instead of exact age)
3. Apply additional anonymization techniques (suppression, generalization)
4. Re-evaluate with **riskutility** metrics

# Computational Considerations

For large datasets, some functions support parallel processing and progress reporting:

```{r computational, eval=FALSE}
# Enable progress bars for long computations
dcr_result <- dcr(
  X = train_data,
  Y = synthetic,
  holdout = holdout_data,
  progress = TRUE  # Show progress bar
)
```

The DCAP and TCAP functions have been optimized using **data.table** for efficient
computation on datasets with hundreds of thousands of records.

# Conclusion

The **riskutility** package provides a comprehensive toolkit for evaluating synthetic
data quality. By combining attribution-based metrics (DCAP, TCAP, WEAP, DiSCO, RAPID)
with distance-based metrics (DCR, NNDR, IMS), RF proximity-based memorization
detection, and utility measures, practitioners can make informed decisions about the
privacy-utility trade-off in their synthetic data products.

Key takeaways:

- Use **DCAP/TCAP** to assess attribute disclosure risk via matching
- Use **RAPID** for ML-based attribute inference risk (captures complex relationships)
- Use **DCR/NNDR/IMS** to detect memorization via distance comparisons
- Use **rf_privacy()** as a data-adaptive alternative to DCR for mixed-type data
- Use **propensity scores** (including ranger-based) and **distribution comparisons** for utility assessment
- Use **disclosure_report()** for comprehensive assessment
- Always evaluate multiple metrics and consider the specific use case

# Session Info

```{r session-info}
sessionInfo()
```

# References
