---
title: "riskutility: Measuring Disclosure Risk and Data Utility for Synthetic Data"
author:
  - name: Matthias Templ
    affiliation: Institute of Statistics and Mathematical Methods in Economics, Vienna University of Technology
    address: Wiedner Hauptstraße 8-10, 1040 Vienna, Austria
    email: matthias.templ@gmail.com
    orcid: 0000-0002-8638-5276
abstract: >
  The R package **riskutility** provides a comprehensive framework for measuring
  disclosure risk and data utility in synthetic and anonymized datasets. It implements
  over 30 risk and utility metrics spanning attribution-based privacy metrics
  (DCAP, TCAP, WEAP, DiSCO, RAPID), distance-based memorization detection
  (DCR, NNDR, IMS, RF proximity), equivalence class privacy models
  (k-anonymity, l-diversity, t-closeness), attacker risk models, membership
  inference attacks (DOMIAS, NNAA), re-identification risk measures (singling out,
  linkability, record linkage), and a broad suite of utility measures including
  distributional distances (Hellinger, energy, MMD), dependence structure fidelity
  (copula, contingency), regression fidelity, and subgroup-stratified utility.
  The multivariate Risk-Utility map (`rumap()`) provides a unified framework
  for jointly evaluating multiple risk and utility dimensions following
  @thees2026beyond. This vignette demonstrates the complete workflow from
  synthetic data generation to comprehensive risk-utility assessment.
keywords:
  - synthetic data
  - disclosure risk
  - data utility
  - privacy
  - statistical disclosure control
  - R
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{riskutility: Measuring Disclosure Risk and Data Utility for Synthetic Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

Synthetic data generation has emerged as a key technique for protecting privacy while
preserving data utility for statistical analysis. However, evaluating the quality of
synthetic data requires careful assessment of both **disclosure risk** (how much
private information might be revealed) and **data utility** (how well the synthetic
data preserves analytical properties of the original).

The **riskutility** package provides a unified framework for these assessments,
implementing state-of-the-art metrics from the statistical disclosure control
literature [@hundepool2012statistical; @templ2017statistical]. The package covers
five major risk paradigms and a comprehensive utility assessment suite:

**Disclosure risk metrics:**

1. Attribution-based metrics (CAP family): DCAP, TCAP, WEAP, DiSCO, RAPID
2. Distance-based memorization detection: DCR, NNDR, IMS, RF proximity
3. Equivalence class privacy models: k-anonymity, l-diversity, t-closeness
4. Attacker risk models, record-level risk (SUDA, individual risk, dRisk)
5. Membership inference and re-identification: DOMIAS, NNAA, singling out, linkability, record linkage

**Data utility metrics:**

1. Propensity score analysis (pMSE, SPECKS)
2. Distributional distances: Hellinger, energy distance, MMD, Wasserstein, KS
3. Dependence structure: copula fidelity, contingency fidelity
4. Model-based utility: regression fidelity, TSTR
5. Tail preservation, CI proximity, subgroup-stratified utility

**Joint assessment:**

- The `rumap()` function provides multivariate Risk-Utility maps following the
  framework of @thees2026beyond
- The `disclosure_report()` function generates comprehensive risk reports

# Installation and Setup

```{r install, eval=FALSE}
# Install from CRAN (when available)
install.packages("riskutility")

# Or install development version from GitHub
# devtools::install_github("mtempl/riskutility")
```

```{r load}
library(riskutility)
library(ggplot2)
```

# Example Data

Throughout this vignette, we use a simulated dataset representing survey data with
demographic and income information. We create both an "original" dataset and a
"synthetic" version with intentionally introduced differences.

```{r example-data}
set.seed(42)
n <- 500

# Original data
original <- data.frame(
  age = as.integer(sample(18:80, n, replace = TRUE)),
  gender = factor(sample(c("Male", "Female"), n, replace = TRUE)),
  education = factor(sample(c("Primary", "Secondary", "Tertiary"), n,
                           replace = TRUE, prob = c(0.2, 0.5, 0.3))),
  region = factor(sample(paste0("Region_", 1:5), n, replace = TRUE)),
  income = round(rgamma(n, shape = 2, scale = 15000), -2),
  health_status = factor(sample(c("Good", "Fair", "Poor"), n,
                                replace = TRUE, prob = c(0.6, 0.3, 0.1)))
)

# Synthetic data (with some noise and distributional shifts)
synthetic <- data.frame(
  age = as.integer(pmin(80, pmax(18, original$age + sample(-3:3, n, replace = TRUE)))),
  gender = factor(sample(c("Male", "Female"), n, replace = TRUE)),
  education = factor(sample(c("Primary", "Secondary", "Tertiary"), n,
                           replace = TRUE, prob = c(0.22, 0.48, 0.30))),
  region = factor(sample(paste0("Region_", 1:5), n, replace = TRUE)),
  income = round(rgamma(n, shape = 2.1, scale = 14500), -2),
  health_status = factor(sample(c("Good", "Fair", "Poor"), n,
                                replace = TRUE, prob = c(0.58, 0.32, 0.10)))
)

head(original)
```

We also prepare a training/holdout split for distance-based metrics:

```{r train-holdout}
set.seed(123)
n_orig <- nrow(original)
train_idx <- sample(n_orig, size = floor(0.7 * n_orig))
train_data <- original[train_idx, ]
holdout_data <- original[-train_idx, ]

# Define quasi-identifiers and sensitive attribute
key_vars <- c("age", "gender", "education", "region")
target_var <- "health_status"
```

# Disclosure Risk Metrics

## Conceptual Framework

Disclosure risk in synthetic data can be understood through several paradigms:

1. **Attribution disclosure**: Can an intruder learn sensitive information about
   an individual by matching quasi-identifiers between original and synthetic data?
2. **Identity disclosure**: Can an intruder determine that a specific synthetic
   record corresponds to a real individual (memorization)?
3. **Membership inference**: Can an intruder determine whether a specific individual
   was in the training data?
4. **Equivalence class violations**: Does the data satisfy formal privacy guarantees
   such as k-anonymity?

The **riskutility** package implements metrics for all four paradigms.

## Attribution-Based Metrics (CAP Family)

### Correct Attribution Probability (DCAP)

The Differential Correct Attribution Probability [DCAP; @taub2018differential]
measures how much better an intruder can guess a sensitive target variable when
they have access to the synthetic data, compared to random guessing.

```{r dcap}
dcap_result <- dcap(
  X = original,
  Y = synthetic,
  key_vars = key_vars,
  target_var = target_var
)

print(dcap_result)
```

```{r dcap-summary}
summary(dcap_result)
```

**Interpretation Guidelines:**

| DCAP Value | Risk Level | Interpretation |
|------------|------------|----------------|
| < 0.05 | Low | Minimal additional disclosure risk |
| 0.05 - 0.15 | Moderate | Some risk; review key variable selection |
| 0.15 - 0.30 | Elevated | Consider additional protection |
| > 0.30 | High | Significant privacy concern |

### Targeted CAP (TCAP)

While DCAP provides an aggregate measure, TCAP computes per-record risk scores,
identifying which individuals face the highest disclosure risk.

```{r tcap}
tcap_result <- tcap(
  X = original,
  Y = synthetic,
  key_vars = key_vars,
  target_var = target_var
)

print(tcap_result)
```

```{r tcap-plot, fig.cap="Distribution of per-record TCAP scores"}
plot(tcap_result)
```

High TCAP values for specific records indicate that those individuals' sensitive
attributes are more easily inferred from the synthetic data.

### Within Equivalence Class Attribution Probability (WEAP)

WEAP focuses on the risk within equivalence classes (groups of records sharing
the same quasi-identifier values).

```{r weap}
weap_result <- weap(
  synthetic,
  key_vars = key_vars,
  target_var = target_var
)

print(weap_result)
summary(weap_result)
```

### DiSCO (Disclosive in Synthetic Correct Original)

DiSCO identifies records where the synthetic data correctly reveals the original
sensitive value with high confidence.

```{r disco}
disco_result <- disco(
  X = original,
  Y = synthetic,
  key_vars = key_vars,
  target_var = target_var
)

print(disco_result)
```

### RAPID (Risk of Attribute Prediction-Induced Disclosure)

RAPID uses machine learning models to assess disclosure risk. Unlike CAP-based methods
that use exact or fuzzy matching, RAPID trains a predictive model on synthetic data
and evaluates how well it can predict sensitive attributes in the original data.

```{r rapid, eval=requireNamespace("ranger", quietly = TRUE)}
rapid_result <- rapid(
  X = original,
  Y = synthetic,
  key_vars = key_vars,
  target_var = target_var,
  model_type = "rf",
  cat_tau = 0.3,
  cat_eval_method = "RCS_marginal"
)

print(rapid_result)
summary(rapid_result)
```

A low RAPID score (< 0.05) indicates that the ML model cannot predict the
sensitive attribute much better than baseline, suggesting good privacy protection.

## Distance-Based Metrics (Holdout Method)

Distance-based metrics detect **memorization** by comparing how close synthetic
records are to training data versus held-out data. If synthetic records are
systematically closer to training data, this suggests the model has memorized
training examples. Note that @yao2025dcr demonstrate limitations of these metrics
(the "DCR delusion"), so they should always be used alongside other risk measures.

### Distance to Closest Record (DCR)

```{r dcr}
dcr_result <- dcr(
  X = train_data,
  Y = synthetic,
  holdout = holdout_data,
  progress = FALSE
)

print(dcr_result)
```

```{r dcr-plot, fig.cap="DCR distribution comparison"}
plot(dcr_result)
```

If the median ratio of distances (synthetic-to-training vs synthetic-to-holdout)
is close to 1, there is no evidence of memorization. Values significantly below 1
suggest memorization.

### Nearest Neighbor Distance Ratio (NNDR)

NNDR refines DCR by computing the ratio of the distance to the nearest neighbor
versus the second-nearest neighbor.

```{r nndr}
nndr_result <- nndr(
  X = train_data,
  Y = synthetic,
  holdout = holdout_data,
  progress = FALSE
)

print(nndr_result)
summary(nndr_result)
```

### Identical Match Share (IMS)

IMS measures the proportion of synthetic records that exactly match a training record.

```{r ims}
ims_result <- ims(
  X = train_data,
  Y = synthetic
)

print(ims_result)
```

An IMS close to 0 indicates good privacy protection (no exact copies of training data).

### RF Proximity Privacy

RF proximity offers a data-adaptive alternative to DCR for memorization detection.
Instead of computing Gower or Euclidean distances, it trains a Random Forest to
discriminate original from synthetic records and measures how often synthetic
records share terminal nodes with training vs. holdout records.

```{r rf-privacy, eval=requireNamespace("ranger", quietly=TRUE)}
rfp_result <- rf_privacy(
  X = train_data,
  Y = synthetic,
  holdout = holdout_data,
  n_trees = 200,
  null_test = FALSE,
  seed = 42
)

print(rfp_result)
```

```{r rf-privacy-plot, fig.cap="RF proximity: training vs holdout", eval=requireNamespace("ranger", quietly=TRUE)}
plot(rfp_result, which = 1)
```

Use `rf_privacy()` when you have 20+ variables with complex interactions or mixed
data types. Use `dcr()` for simpler settings where interpretability matters more.

## Equivalence Class Privacy Models {#equiv-class}

These metrics evaluate formal privacy guarantees based on equivalence classes---groups
of records sharing the same quasi-identifier values. They originate from the database
privacy literature and are useful for assessing whether anonymized data provides
structural protection against re-identification.

### k-Anonymity

A dataset satisfies k-anonymity [@samarati1998protecting] if every combination of
quasi-identifier values appears at least $k$ times. Records in smaller equivalence
classes are more vulnerable to re-identification.

```{r kanonymity}
kanon_result <- kanonymity(
  X = original,
  key_vars = key_vars,
  k = 5
)

print(kanon_result)
summary(kanon_result)
```

The result shows the achieved k-level (minimum equivalence class size), the
percentage of records violating the k-threshold, and the distribution of
equivalence class sizes.

### l-Diversity

While k-anonymity protects against identity disclosure, it does not prevent
attribute disclosure within an equivalence class. l-Diversity
[@machanavajjhala2007ldiversity] requires that each equivalence class contains
at least $l$ "well-represented" values of the sensitive attribute. The function
computes distinct, entropy, and recursive (c,l)-diversity:

```{r ldiversity}
ldiv_result <- ldiversity(
  X = original,
  key_vars = key_vars,
  sensitive_var = target_var,
  l = 2
)

print(ldiv_result)
summary(ldiv_result)
```

### t-Closeness

t-Closeness [@li2007tcloseness] strengthens l-diversity by requiring that the
distribution of the sensitive attribute within each equivalence class is "close"
to the overall distribution, measured by Earth Mover's Distance (EMD):

```{r tcloseness}
tclose_result <- tcloseness(
  X = original,
  key_vars = key_vars,
  sensitive_var = target_var,
  t = 0.2
)

print(tclose_result)
summary(tclose_result)
```

Values of `t_achieved` near zero indicate that the sensitive attribute is
distributed similarly within every equivalence class, providing strong protection
against attribute inference.

## Attacker Risk Models

The `attacker_risk()` function implements three classical attacker scenarios from
the SDC literature [@hundepool2012statistical; @templ2017statistical]:

- **Prosecutor**: knows a target is in the data and tries to identify them
- **Journalist**: does not know whether the target is in the data
- **Marketer**: tries to identify as many records as possible

```{r attacker-risk}
ar_result <- attacker_risk(
  X = original,
  key_vars = key_vars,
  sampling_fraction = 0.01
)

print(ar_result)
summary(ar_result)
```

The `privacy_pass` field indicates whether the prosecutor risk is below the
conventional 0.1 threshold.

## Record-Level Risk Measures

### Individual Risk

Individual risk [@franconi2004individual; @skinner2002measure] estimates the
re-identification probability for each record based on equivalence class
frequencies and an optional super-population model:

```{r individual-risk}
ir_result <- individual_risk(
  X = original,
  key_vars = key_vars,
  method = "both",
  threshold = 0.1
)

print(ir_result)
```

### SUDA (Special Uniques Detection Algorithm)

SUDA identifies records at high risk by detecting Minimal Sample Uniques (MSUs)---the
smallest subsets of quasi-identifiers that make a record unique:

```{r suda}
suda_result <- suda(
  X = original,
  key_vars = key_vars
)

print(suda_result)
summary(suda_result)
```

Higher SUDA scores indicate records that are unique on many small subsets
of quasi-identifiers and therefore more vulnerable.

### dRisk (Distance-Based Risk for Continuous Variables)

dRisk [@templ2008robust] assesses disclosure risk for continuous variables by
measuring how many synthetic records fall within a narrow interval around each
original record. It includes a robust variant (dRiskRMD) based on Mahalanobis distances:

```{r drisk}
drisk_result <- drisk(
  X = original,
  Y = synthetic,
  vars = c("age", "income"),
  method = "both"
)

print(drisk_result)
```

## Population-Level Risk

### Population Uniqueness

Population uniqueness estimates how many records are unique not just in the sample
but in the population, using super-population models (Pitman, Zayatz, SNB):

```{r pop-unique}
pu_result <- population_uniqueness(
  X = original,
  key_vars = key_vars,
  method = "all",
  sampling_fraction = 0.01
)

print(pu_result)
```

### Epsilon-Identifiability

Epsilon-identifiability flags synthetic records that are within distance $\epsilon$
of an original record, weighted by variable entropy:

```{r epsilon-id}
eid_result <- epsilon_identifiability(
  X = original,
  Y = synthetic,
  epsilon = 0.05
)

print(eid_result)
```

### Delta-Presence

Delta-presence measures the membership probability of each quasi-identifier
combination and checks whether all probabilities fall within acceptable bounds
$[\delta_{\min}, \delta_{\max}]$:

```{r delta-presence}
dp_result <- delta_presence(
  X = original,
  Y = synthetic,
  key_vars = key_vars,
  delta_min = 0.0,
  delta_max = 0.5
)

print(dp_result)
```

## Membership Inference Attacks

Membership inference attacks [@shokri2017membership] test whether an adversary can
determine if a specific individual was in the training data used to generate the
synthetic dataset.

### NNAA (Nearest Neighbor Adversarial Accuracy)

NNAA compares the distance from synthetic records to training data versus holdout data.
If synthetic records are systematically closer to training data (`privacy_loss > 0`),
there is evidence of information leakage:

```{r nnaa}
nnaa_result <- nnaa(
  X = train_data,
  Y = synthetic,
  holdout = holdout_data,
  method = "gower",
  seed = 42
)

print(nnaa_result)
summary(nnaa_result)
```

### DOMIAS

DOMIAS [@hu2023domias] detects membership through overfitting by comparing local
density ratios of training vs. holdout records in the neighborhood of synthetic
records:

```{r domias}
domias_result <- domias(
  X = train_data,
  Y = synthetic,
  holdout = holdout_data,
  radius = 0.1,
  seed = 42
)

print(domias_result)
```

An AUC near 0.5 indicates the attacker cannot distinguish training from holdout
records (good privacy). AUC above 0.6 suggests potential memorization.

## Re-identification Attacks (Anonymeter Framework)

The singling out, linkability, and hitting rate metrics follow the framework of
@giomi2023anonymeter and evaluate whether an adversary can isolate, link, or
re-identify individuals in the synthetic data.

### Singling Out

Singling out measures the risk that randomly generated predicates (e.g., "age > 50
AND region = Region_3") identify a single individual:

```{r singling-out}
so_result <- singling_out(
  X = original,
  Y = synthetic,
  n_attacks = 500,
  n_cols = 3,
  mode = "multivariate",
  seed = 42
)

print(so_result)
```

### Linkability

Linkability evaluates whether an adversary who knows some attributes of a target
can link to the correct record via the synthetic data:

```{r linkability}
link_result <- linkability(
  X = original,
  Y = synthetic,
  n_attacks = 500,
  n_neighbors = 1,
  seed = 42
)

print(link_result)
```

### Hitting Rate

Hitting rate measures the fraction of original records that have a synthetic
record within a distance threshold:

```{r hitting-rate}
hr_result <- hitting_rate(
  X = original,
  Y = synthetic,
  threshold = 0.05,
  method = "gower"
)

print(hr_result)
```

## Record Linkage Risk

For a comprehensive treatment of record linkage risk, see the dedicated vignette
`vignette("recordLinkage", package = "riskutility")`. The `recordLinkage()` function
supports 8 linkage methods (deterministic, probabilistic, PRAM, predictive, RF, RBRL,
Mahalanobis, embedding) and 3 matching strategies (independent, bijective via
Hungarian algorithm, optimal transport via Sinkhorn):

```{r record-linkage, eval=FALSE}
# Deterministic record linkage with bijective matching (GDBRL)
rl_result <- recordLinkage(
  X = original,
  Y = synthetic,
  key_vars = c("age", "gender", "education", "region", "income"),
  method = "deterministic",
  matching = "bijective"
)

print(rl_result)
summary(rl_result)
```

See @domingoferrer2003disclosure for the original distance-based record linkage
framework and @herranz2016gdbrl for the bijective (GDBRL) extension.

# Data Utility Metrics

Protecting privacy is only valuable if the synthetic data remains useful for
analysis. The **riskutility** package provides a comprehensive suite of utility
measures covering marginal distributions, dependence structure, tail behavior,
model-based fidelity, and subgroup fairness.

## Propensity Score Analysis

The propensity score method [@rosenbaum1983propensity; @woo2009global] trains a
classifier to distinguish original from synthetic records. If the classifier
performs poorly (accuracy near 0.5), the distributions are similar. The pMSE
statistic measures the mean squared deviation of propensity scores from 0.5.

```{r propscore}
ps_result <- propscore(
  X = original,
  Y = synthetic,
  form = ~ age + income + education + gender
)

print(ps_result)
```

```{r propscore-summary}
summary(ps_result)
```

### Ranger-Based Propensity Scores

The `method = "ranger"` option uses a probability forest providing OOB propensity
scores and proximity-based structural metrics:

```{r propscore-ranger, eval=requireNamespace("ranger", quietly=TRUE)}
ps_ranger <- propscore(
  X = original,
  Y = synthetic,
  method = "ranger",
  proximity = "summary"
)

summary(ps_ranger)
```

The `structure_ratio` compares cross-class proximity to within-class proximity.
Values near 1 indicate the forest cannot distinguish the two datasets.

## Distributional Distance Measures

### Hellinger Distance

The Hellinger distance measures the similarity between categorical distributions.
It ranges from 0 (identical) to 1 (completely disjoint):

```{r hellinger}
hell_result <- hellinger(
  X = original,
  Y = synthetic,
  vars = c("gender", "education", "health_status")
)

print(hell_result)
summary(hell_result)
```

### Energy Distance

Energy distance [@szekely2013energy] is a multivariate two-sample statistic that
equals zero if and only if the two distributions are identical. It handles
arbitrary dimensions and does not require density estimation:

```{r energy-distance}
ed_result <- energy_distance(
  X = original,
  Y = synthetic,
  vars = c("age", "income"),
  seed = 42
)

print(ed_result)
```

The `utility_score` transforms energy distance to a 0--1 scale where 1 is perfect fidelity.

### Maximum Mean Discrepancy (MMD)

MMD [@gretton2012kernel] is a kernel-based two-sample test statistic. The package
supports exact computation (for moderate $n$) and random Fourier features (RFF)
for large datasets:

```{r mmd}
mmd_result <- mmd(
  X = original,
  Y = synthetic,
  vars = c("age", "income"),
  kernel = "gaussian",
  method = "rff",
  n_features = 500,
  seed = 42
)

print(mmd_result)
```

Use `n_perm = 200` to add a permutation $p$-value for the two-sample test.

## Continuous Distribution Comparisons

### Wasserstein and KS Tests

```{r compare-dist}
dist_result <- compare_distributions_cont(
  X = original,
  Y = synthetic,
  variables = c("age", "income")
)

print(dist_result)
```

```{r compare-dist-plot, fig.cap="Distribution comparison for continuous variables"}
plot(dist_result)
```

### Categorical Variables (Chi-Squared)

```{r compare-cat}
chisq_result <- compare_chisq_gof(
  X = original,
  Y = synthetic,
  cat_vars = "gender"
)

print(chisq_result)
```

## Confidence Interval Proximity

CI proximity [@karr2006framework] compares confidence intervals of summary
statistics (means) between original and synthetic data. It provides three
complementary views: overlap, proximity, and relative error:

```{r ci-proximity}
cip_result <- ci_proximity(
  X = original,
  Y = synthetic,
  vars = c("age", "income")
)

print(cip_result)
summary(cip_result)
```

An overlap near 1 means the confidence intervals for the means coincide;
a relative error near 0 means the point estimates are close.

## Model-Based Utility

### Train on Synthetic, Test on Real (TSTR)

TSTR [@snoke2018general; @zhao2021ctgan] trains a predictive model on the synthetic
data and evaluates its performance on held-out real data. The ratio of
TSTR-to-TRTR performance quantifies how well the synthetic data preserves
predictive relationships:

```{r tstr, eval=requireNamespace("ranger", quietly=TRUE)}
tstr_result <- tstr(
  X = original,
  Y = synthetic,
  target_var = "income",
  model = "rf",
  test_fraction = 0.3,
  seed = 42
)

print(tstr_result)
summary(tstr_result)
```

A TSTR ratio close to 1 indicates that models trained on synthetic data perform
nearly as well as models trained on real data.

### Regression Fidelity

Regression fidelity [@karr2006framework] fits the same regression model on both
original and synthetic data and compares coefficient estimates via CI overlap
(using average-width denominator), standardized bias, and significance agreement:

```{r regression-fidelity}
rf_result <- regression_fidelity(
  X = original,
  Y = synthetic,
  formula = income ~ age + gender + education
)

print(rf_result)
summary(rf_result)
```

```{r regression-fidelity-plot, fig.cap="Regression coefficient comparison"}
plot(rf_result, which = 1)
```

The `utility_score` is the mean CI overlap across all coefficients (excluding the
intercept). Values above 0.8 indicate excellent agreement; below 0.5 suggests
the synthetic data distorts regression relationships.

## Dependence Structure Fidelity

### Copula Fidelity

Copula fidelity compares the empirical copula (rank dependence structure) of
numeric variables between original and synthetic data using the Cramer-von Mises
statistic on pairwise copula CDFs:

```{r copula-fidelity}
cop_result <- copula_fidelity(
  X = original,
  Y = synthetic,
  vars = c("age", "income")
)

print(cop_result)
```

### Contingency Fidelity

Contingency fidelity [@snoke2018general] complements copula fidelity for
categorical variables. It computes the total variation distance between bivariate
contingency tables for all pairs of categorical variables:

```{r contingency-fidelity}
ctf_result <- contingency_fidelity(
  X = original,
  Y = synthetic,
  vars = c("gender", "education", "health_status", "region")
)

print(ctf_result)
summary(ctf_result)
```

```{r contingency-fidelity-plot, fig.cap="TV distance heatmap for categorical variable pairs"}
plot(ctf_result, which = 1)
```

Low TV distances (below 0.05) indicate that bivariate categorical relationships
are well preserved.

## Tail Fidelity

Tail fidelity assesses how well the synthetic data preserves extreme values---critical
for applications where tail behavior matters (financial risk, rare diseases):

```{r tail-fidelity}
tf_result <- tail_fidelity(
  X = original,
  Y = synthetic,
  vars = c("age", "income"),
  percentile = 95,
  tails = "both"
)

print(tf_result)
```

The QQ tail divergence measures the integrated squared difference between empirical
quantile functions in the tails. Values near zero indicate good tail preservation.

## Subgroup Utility

Synthetic data can look good overall but perform poorly for minority subgroups.
`subgroup_utility()` [@snoke2018general] applies any utility measure to each
subgroup defined by a grouping variable and identifies groups with low utility:

```{r subgroup-utility}
su_result <- subgroup_utility(
  X = original,
  Y = synthetic,
  group_var = "region",
  utility_fun = energy_distance,
  threshold = 0.5,
  seed = 42
)

print(su_result)
summary(su_result)
```

```{r subgroup-utility-plot, fig.cap="Utility scores by region subgroup"}
plot(su_result)
```

The conservative `utility_score` is the worst subgroup score. A `ratio` near 1
indicates homogeneous utility; below 0.5 indicates substantial disparity.

## Other Utility Measures

### Summary Statistics Comparison

```{r summary-stats}
stats_result <- compare_multivariate_summary_statistics(
  X = original,
  Y = synthetic,
  cont_vars = c("age", "income"),
  cat_vars = c("gender", "education")
)

print(stats_result$continuous)
```

### Gower Distance

The average Gower distance [@gower1971general] provides a simple overall measure
of dissimilarity between original and synthetic datasets:

```{r gower}
gow_result <- gower(
  X = original,
  Y = synthetic
)

print(gow_result)
```

# Multivariate Risk-Utility Framework {#rumap}

## The rumap() Function

Traditional risk-utility evaluation relies on a single risk metric plotted against
a single utility metric. This is insufficient: different metrics capture different
aspects, and a method that scores well on one dimension may fail on another.

The `rumap()` function [@thees2026beyond] computes multiple risk and utility measures
simultaneously, normalizes them to a common [0,1] scale, and provides composite
scores, Pareto frontier identification, and seven visualization types:

```{r rumap, eval=requireNamespace("ranger", quietly=TRUE)}
ru_result <- rumap(
  X = original,
  synthetic = synthetic,
  risk_measures = c("dcap", "ims", "repu"),
  utility_measures = c("pmse", "hellinger", "energy", "contingency"),
  key_vars = key_vars,
  target_var = target_var,
  seed = 42
)

print(ru_result)
```

```{r rumap-summary, eval=requireNamespace("ranger", quietly=TRUE)}
summary(ru_result)
```

## Visualization

```{r rumap-plot-1, fig.cap="Risk-Utility scatterplot with Pareto front", eval=requireNamespace("ranger", quietly=TRUE)}
plot(ru_result, which = 1)
```

```{r rumap-plot-2, fig.cap="Risk-Utility heatmap", fig.height=6, eval=requireNamespace("ranger", quietly=TRUE)}
plot(ru_result, which = 2)
```

Additional plot types include dot plot (`which = 3`), parallel coordinates
(`which = 4`), radial/origami plot (`which = 5`), PCA biplot (`which = 6`),
and blockwise PCA (`which = 7`).

## Comparing Multiple Generators

When evaluating multiple synthetic data generators, pass a named list to the
`synthetic` parameter:

```{r rumap-multi, eval=FALSE}
# Assume synth_A and synth_B are synthetic datasets from different generators
ru_multi <- rumap(
  X = original,
  synthetic = list(Generator_A = synth_A, Generator_B = synth_B),
  risk_measures = c("dcap", "ims"),
  utility_measures = c("pmse", "hellinger", "energy"),
  key_vars = key_vars,
  target_var = target_var,
  seed = 42
)

# Heatmap comparing all generators
plot(ru_multi, which = 2)

# Pareto front with labeled generators
plot(ru_multi, which = 1)
```

# Comprehensive Disclosure Report

The `disclosure_report()` function provides a unified assessment combining multiple
risk and utility metrics in a single call:

```{r disclosure-report}
report <- disclosure_report(
  X = original,
  Y = synthetic,
  key_vars = key_vars,
  target_var = target_var
)

print(report)
```

```{r report-summary}
summary(report)
```

```{r report-plot, fig.cap="Comprehensive disclosure risk visualization"}
plot(report)
```

# Integration with synthpop and simPop

The **riskutility** package integrates with the major synthetic data generation
packages in R. Helper functions `from_synthpop()` and `from_simPop()` extract
the original and synthetic datasets into a `synth_pair` object that can be passed
directly to any risk or utility function:

```{r synthpop-integration, eval=FALSE}
library(synthpop)

synth_obj <- syn(original, seed = 123)

# Create synth_pair from synthpop object
pair <- from_synthpop(synth_obj, original)

# Use with any riskutility function
dcap(pair, key_vars = key_vars, target_var = target_var)
energy_distance(pair, vars = c("age", "income"))
subgroup_utility(pair, group_var = "region")
```

```{r simPop-integration, eval=FALSE}
library(simPop)

# Create synth_pair from simPop object
# pair <- from_simPop(simPopObj)
```

# Metric Comparison

Different metrics capture different aspects of disclosure risk. Here we compare
results across metrics for the same dataset:

```{r metric-comparison}
comparison <- data.frame(
  Category = c("Attribution", "Attribution", "Attribution",
               "Distance", "Privacy Model", "Privacy Model",
               "Record-Level", "Membership Inference",
               "Re-identification"),
  Metric = c("DCAP", "Mean WEAP", "DiSCO %",
             "IMS", "k-Anonymity (k)", "t-Closeness (t)",
             "dRisk (interval)", "NNAA (privacy loss)",
             "Singling out"),
  Value = round(c(
    dcap_result$dcap[1],
    weap_result$weap_mean[1],
    disco_result$pct_disco[1],
    ims_result$ims[1],
    kanon_result$k_level,
    tclose_result$t_achieved,
    drisk_result$drisk_interval,
    nnaa_result$privacy_loss,
    so_result$risk
  ), 4),
  Pass = c(
    dcap_result$dcap[1] < 0.15,
    weap_result$weap_mean[1] < 0.8,
    disco_result$pct_disco[1] < 5,
    ims_result$ims[1] < 0.01,
    kanon_result$satisfies_k,
    tclose_result$satisfies_t,
    drisk_result$privacy_pass,
    nnaa_result$privacy_pass,
    so_result$privacy_pass
  ),
  stringsAsFactors = FALSE
)

knitr::kable(comparison, caption = "Summary of disclosure risk metrics across paradigms")
```

# Best Practices

## Choosing Key Variables

Key variables (quasi-identifiers) should be selected based on:

1. **Availability to potential intruders**: Variables an attacker might know
2. **Uniqueness**: Combinations that identify individuals
3. **Linkage potential**: Variables present in external databases

Common quasi-identifiers include age, gender, geographic location, and occupation.

## Recommended Evaluation Workflow

A thorough evaluation should cover multiple risk paradigms:

1. **Attribution risk** (DCAP/TCAP or RAPID): Can sensitive attributes be inferred?
2. **Memorization** (DCR or RF proximity + IMS): Are records copied?
3. **Equivalence class** (k-anonymity, l-diversity): Are structural guarantees met?
4. **Membership inference** (NNAA or DOMIAS): Can training set membership be detected?
5. **Utility** (multiple measures via rumap): Is the data useful?
6. **Subgroup fairness** (subgroup_utility): Is utility preserved for all subgroups?

## Interpreting Results

1. **No single metric is sufficient**: Use multiple metrics to get a complete picture
2. **Context matters**: Acceptable risk levels depend on data sensitivity and use case
3. **Compare to baselines**: DCAP values should be compared to the baseline (random guessing)
4. **Consider record-level risk**: Aggregate metrics may hide high-risk individuals (use TCAP)
5. **Check subgroups**: Overall utility can mask poor performance for minorities

## Iterative Refinement

If disclosure risk is too high:

1. Add more noise to the synthesis process
2. Reduce the granularity of quasi-identifiers (e.g., age groups instead of exact age)
3. Apply additional anonymization techniques (suppression, generalization)
4. Re-evaluate with **riskutility** metrics

# Computational Considerations

For large datasets, some functions support sampling and progress reporting:

```{r computational, eval=FALSE}
# Enable progress bars for long computations
dcr_result <- dcr(
  X = train_data,
  Y = synthetic,
  holdout = holdout_data,
  progress = TRUE
)

# MMD with random Fourier features scales to large n
mmd_result <- mmd(X = original, Y = synthetic, method = "rff", n_features = 500)

# Energy distance with subsampling
ed_result <- energy_distance(X = original, Y = synthetic, n_sample = 1000)
```

The DCAP and TCAP functions have been optimized using **data.table** for efficient
computation on datasets with hundreds of thousands of records.

# Conclusion

The **riskutility** package provides a comprehensive toolkit for evaluating
synthetic data quality across multiple dimensions of risk and utility. With
over 30 metrics spanning attribution risk, memorization detection, equivalence
class privacy, membership inference, and a broad suite of utility measures,
practitioners can make informed decisions about the privacy-utility trade-off
in their synthetic data products.

Key recommendations:

- Use **DCAP/TCAP** for attribute disclosure risk via matching
- Use **RAPID** for ML-based attribute inference (captures complex relationships)
- Use **DCR/NNDR/IMS** or **rf_privacy()** for memorization detection
- Use **k-anonymity/l-diversity/t-closeness** for structural privacy guarantees
- Use **NNAA/DOMIAS** for membership inference risk
- Use **singling out/linkability** for re-identification risk
- Use **energy distance, Hellinger, MMD** for distributional fidelity
- Use **copula/contingency fidelity** for dependence structure preservation
- Use **regression fidelity** for model-based analytical validity
- Use **subgroup_utility()** to check fairness across subpopulations
- Use **rumap()** for joint multivariate risk-utility evaluation
- Use **disclosure_report()** for comprehensive automated assessment
- Always evaluate multiple metrics and consider the specific use case

# Session Info

```{r session-info}
sessionInfo()
```

# References
