---
title: "riskutility: Comprehensive Disclosure Risk and Data Utility Assessment for Anonymized and Synthetic Data in R"
author:
  - name: Matthias Templ
    affiliation: "School of Business, FHNW University of Applied Sciences and Arts Northwestern Switzerland"
    email: matthias.templ@gmail.com
    orcid: 0000-0002-8638-5276
  - name: Oscar Thees
    affiliation: "School of Business, FHNW University of Applied Sciences and Arts Northwestern Switzerland"
    orcid: 0009-0001-9378-4988
abstract: >
  The R package **riskutility** provides a comprehensive framework for measuring
  disclosure risk and data utility in synthetic and anonymized datasets. It
  implements over 30 risk measures spanning six paradigms --- frequency-based
  privacy models ($k$-anonymity, $l$-diversity, $t$-closeness), attribution-based
  (DCAP, TCAP, WEAP, DiSCO), ML-based (RAPID), distance-based (DCR, NNDR, IMS,
  RF proximity), record linkage, membership inference (DOMIAS, NNAA), and GDPR
  anonymization criteria (singling out, linkability) --- alongside a broad suite of
  utility measures including
  propensity scores, distributional distances (Hellinger, energy, MMD),
  dependence structure fidelity (copula, contingency), regression fidelity,
  subgroup-stratified utility, and Train on Synthetic, Test on Real (TSTR).
  The multivariate Risk-Utility map (`rumap()`) provides a unified framework for
  jointly evaluating multiple risk and utility dimensions. All functions share a
  consistent S3 API with `print()`, `summary()`, and `plot()` methods and accept
  a `synth_pair` container. This paper describes the package architecture,
  demonstrates the complete workflow from synthetic data generation to
  comprehensive risk-utility assessment, and presents a case study comparing
  three synthesis approaches.
keywords:
  - synthetic data
  - disclosure risk
  - data utility
  - privacy
  - statistical disclosure control
  - R
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{riskutility: Comprehensive Disclosure Risk and Data Utility Assessment for Anonymized and Synthetic Data in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = TRUE,
  message = FALSE
)
library(riskutility)
```

# Introduction {#sec-intro}

Statistical agencies and data custodians face a fundamental challenge: releasing
data that are useful for analysis while protecting individual privacy. Two main
approaches exist: (i) traditional anonymization methods such as perturbation,
suppression, and generalization [@hundepool2012statistical;
@templ2017statistical], and (ii) synthetic data generation via statistical
models [@nowok2016synthpop; @templ2017simulation]. Both require rigorous
evaluation of *disclosure risk* and *data utility* --- yet existing tools
assess these two dimensions in isolation or cover only a narrow subset of the
relevant metrics.

The R package sdcMicro [@templ2015sdcmicro] provides frequency-based risk
estimation but limited utility assessment. synthpop [@nowok2016synthpop]
offers the CAP disclosure metric and propensity-score utility but not
distance-based or membership inference methods. In Python, SDMetrics covers
distance-based metrics and Anonymeter implements GDPR failure criteria, but
neither addresses attribution risk. No existing *R* package provides a unified
framework that spans all risk families and supports multivariate comparison of
multiple synthesis approaches; the most comprehensive evaluation suites (the
Python libraries synthcity and SynthEval) sit outside the R ecosystem.

The **riskutility** package for R [@R2025] fills this gap. It implements
over 30 risk measures spanning six paradigms --- frequency-based privacy
models, attribution-based (CAP), ML-based (RAPID), distance-based, record
linkage, and membership inference --- alongside more than a dozen utility
functions covering global, distributional, structural, and predictive assessment. All functions
share a consistent S3 API with `print()`, `summary()`, and `plot()` methods
and accept a `synth_pair` container that bundles data and metadata.

The need for rigorous evaluation is well established in the statistical
disclosure control literature and has gained urgency as synthetic data enters
regulatory frameworks. The European Article 29 Working Party
[@wp29anonymisation] identifies three criteria for anonymization: singling out,
linkability, and inference. **riskutility** measures the first two directly
(`singling_out()`, `linkability()`); the inference criterion is approached
through the attribution-based CAP and RAPID measures rather than a dedicated
inference-attack function.
Importantly, synthetic data does not automatically satisfy these criteria
[@stadler2022synthetic], making empirical assessment essential.

The distinction between *general* and *specific* utility [@snoke2018general]
is important: general utility measures (propensity scores, distributional
distances) assess overall distributional fidelity, while specific utility
measures (regression fidelity, TSTR) assess whether particular analyses
yield similar results on original and synthetic data. A thorough evaluation
should include both.

**riskutility** provides empirical privacy metrics rather than formal
mathematical guarantees. Unlike differential privacy, which provides
worst-case bounds, our metrics quantify observed risk on concrete datasets.
This approach is appropriate for practical SDC workflows where the protection
mechanism is not differentially private. For differentially private
synthesizers, formal epsilon budgets should be used alongside empirical
evaluation.

**Quick start.** A comprehensive risk assessment requires just a few lines:

```{r intro-hook, eval=FALSE}
library(riskutility)
pair <- synth_pair(original, synthetic,
                   key_vars = c("age", "sex", "region"),
                   target_var = "income")
report <- disclosure_report(pair)
print(report)
```

This single call computes attribution-based risk (DCAP, TCAP, WEAP, DiSCO,
RAPID), distance-based risk (DCR, NNDR, IMS, dRisk, hitting rate), privacy
models ($k$-anonymity, $l$-diversity, $t$-closeness), and membership
inference measures (singling out, linkability, NNAA) --- producing a
pass/fail summary across all families. For comparing multiple synthesis
approaches, `rumap()` computes risk and utility measures simultaneously,
normalizes them to a common scale, and identifies Pareto-optimal methods
(Section \@ref(sec-case-rumap)).

**Paper overview.** Section \@ref(sec-background) introduces the disclosure
threat taxonomy and reviews related software. Section \@ref(sec-design)
describes the package architecture and the `synth_pair` container.
Sections \@ref(sec-risk) and \@ref(sec-utility) present the risk and utility
measures with mathematical formulations and worked examples.
Section \@ref(sec-comprehensive) demonstrates the complete practitioner
workflow on a realistic case study, comparing three synthesis approaches
with `disclosure_report()` and `rumap()`. Section \@ref(sec-discussion)
discusses limitations, remediation strategies, and future work.
Section \@ref(sec-computational) provides computational details.

```{r table1-scope, echo=FALSE}
scope <- data.frame(
  Category = c("Privacy models", "Attribution (CAP)",
                "ML-based (RAPID)", "Distance-based", "Record linkage",
                "Membership inference", "Utility measures", "Frameworks"),
  Functions = c(8, 4, 4, 6, "1 (8 methods)", 6, 15, 3),
  Paradigm = c("Equivalence class", "Matching", "Prediction",
                "Nearest neighbor", "Linkage", "Attack simulation",
                "Various", "Composite"),
  `Applicable to` = c(rep("Both", 8)),
  check.names = FALSE
)
knitr::kable(scope, caption = "Package scope at a glance.
             Both = applicable to traditionally anonymized and synthetic data.")
```

All analyses in this paper were conducted in R [@R2025].


# Background: Threat Taxonomy and Related Software {#sec-background}

## Disclosure Threat Taxonomy {#sec-threats}

Every risk measure in riskutility addresses one or more of four disclosure
threats. The classical SDC taxonomy [@hundepool2012statistical;
@templ2017statistical] recognizes three types: identity, attribute, and
inferential disclosure. We extend this to four threats by separating
*membership disclosure* (from the ML privacy literature,
@shokri2017membership) and *memorization* (relevant to generative models)
as distinct categories, reflecting the broader scope of modern synthetic
data evaluation:

- **Identity disclosure**: An attacker links a released record to a specific
  individual. Measured by: `recordLinkage()`, `kanonymity()`,
  `individual_risk()`, `hitting_rate()`.
- **Attribute disclosure**: An attacker learns a sensitive attribute value
  through quasi-identifier (QI) matching. Measured by: `dcap()`, `tcap()`,
  `weap()`, `disco()`, `rapid()`.
- **Membership disclosure**: An attacker determines whether an individual's
  data was used to create the released dataset. Measured by:
  `mia_classifier()`, `domias()`, `nnaa()`, `singling_out()`,
  `delta_presence()`.
- **Memorization**: A generative model reproduces training records verbatim
  or near-verbatim. Measured by: `ims()`, `dcr()`, `nndr()`.

```{r table2-threats, echo=FALSE}
threats <- data.frame(
  Threat = c("Identity", "Attribute", "Membership", "Memorization"),
  Definition = c(
    "Attacker links record to individual",
    "Attacker learns sensitive value via linkage",
    "Attacker determines if individual is in dataset",
    "Generator reproduces training records"
  ),
  `Key measures` = c(
    "recordLinkage, kanonymity, individual_risk",
    "dcap, tcap, weap, disco, rapid",
    "mia_classifier, domias, nnaa, singling_out",
    "ims, dcr, nndr"
  ),
  check.names = FALSE
)
knitr::kable(threats, caption = "Disclosure threat taxonomy.")
```

## Risk Assessment Paradigms

riskutility implements six risk assessment paradigms, each addressing different
threats from the taxonomy above. We summarize each here; full definitions and
worked examples are in Section \@ref(sec-risk).

**Frequency-based (privacy models).** These methods assess privacy properties
of a single dataset based on quasi-identifier frequencies.
$k$-Anonymity [@samarati1998protecting] requires each combination of
quasi-identifiers to appear at least $k$ times. $l$-Diversity
[@machanavajjhala2007ldiversity] and $t$-closeness [@li2007tcloseness]
progressively strengthen the protection of sensitive attributes within
equivalence classes. See Section \@ref(sec-privacy-models).

**Attribution-based (CAP family).** The Correct Attribution Probability
framework [@taub2018differential] measures whether an attacker can infer
sensitive attribute values by matching quasi-identifiers between original and
released data. DCAP provides an aggregate measure; TCAP gives per-record
risk scores. See Section \@ref(sec-cap).

**ML-based (RAPID).** Risk of Attribute Prediction-Induced Disclosure
[@thees2026beyond] trains predictive models on the released data and evaluates
them on the original. Accurate predictions indicate information leakage. RAPID
captures complex non-linear relationships and variable interactions that the
CAP matching approach may miss. See Section \@ref(sec-rapid).

**Distance-based.** The holdout method compares distances from synthetic records
to training data vs. holdout data. If synthetic records are systematically
closer to training records, the generator has memorized rather than generalized.
However, @yao2025dcr demonstrate that passing distance-based tests does
not guarantee privacy (the "DCR Delusion"). See Section \@ref(sec-distance).

**Record linkage.** Directly simulates a re-identification attack using
deterministic (Gower distance), probabilistic (Fellegi-Sunter), PRAM-aware,
predictive (propensity-score), or random forest linkage
[@fellegi1969theory; @domingoferrer2003disclosure]. See
Section \@ref(sec-recordlinkage) and the companion vignette
`vignette("recordLinkage")`.

**Membership inference.** Shadow model attacks [@shokri2017membership],
density-based detection [@hu2023domias], and GDPR failure criteria
[@stadler2022synthetic] assess whether an attacker can determine if a specific
individual's data was used during synthesis. See Section \@ref(sec-membership).

## Utility Assessment Paradigms

Utility measures quantify how well the released data preserves the
statistical properties of the original. We organize them into four groups
(details in Section \@ref(sec-utility)):

**Global utility** measures use propensity scores to quantify how
distinguishable the released data is from the original [@woo2009global;
@snoke2018general]. A single number summarizes overall data quality.

**Distributional utility** measures (Wasserstein, Hellinger, KS test,
energy distance, MMD) compare marginal or joint distributions, identifying
which specific variables or relationships are poorly reproduced.

**Structural utility** measures (copula fidelity, contingency fidelity,
PCA comparison, correlation matrix comparison) assess whether multivariate
relationships --- the correlations and interactions that analysts depend
on --- are preserved.

**Predictive utility** measures (TSTR, regression fidelity, feature
importance stability) test whether models trained on released data
generalize to real-world predictions.

## Related Software

**In R.** sdcMicro [@templ2015sdcmicro] provides the most comprehensive
SDC toolkit, including frequency-based risk estimation, but focuses
primarily on *applying* anonymization methods. synthpop [@nowok2016synthpop]
provides the CAP disclosure metric and propensity-score utility, but does
not cover distance-based or membership inference methods. Neither package
provides a unified framework for comparing multiple risk and utility
measures simultaneously.

**In Python.** SDMetrics (part of the Synthetic Data Vault) provides
distance-based and statistical metrics for synthetic data quality. Anonymeter
[@giomi2023anonymeter] implements the three GDPR failure criteria from
@stadler2022synthetic (singling out, linkability, inference). Neither covers
attribution-based metrics (CAP, RAPID) or traditional privacy models
($k$-anonymity, $l$-diversity).

```{r table3-comparison, echo=FALSE}
comp <- data.frame(
  Measure = c("k-Anonymity", "l-Diversity", "t-Closeness",
              "DCAP/TCAP", "RAPID", "DCR/NNDR", "Record linkage",
              "MIA / GDPR criteria", "pMSE / SPECKS",
              "R-U map"),
  sdcMicro = c("freq()", "ldiversity()", "--",
               "--", "--", "dRisk()", "--",
               "--", "--", "--"),
  synthpop = c("--", "--", "--",
               "disclosure()", "--", "--", "--",
               "--", "utility.gen()", "--"),
  `SDMetrics (Py)` = c("--", "--", "--",
                        "--", "--", "Yes", "--",
                        "--", "--", "--"),
  `Anonymeter (Py)` = c("--", "--", "--",
                         "--", "--", "--", "--",
                         "Yes", "--", "--"),
  riskutility = c("kanonymity()", "ldiversity()", "tcloseness()",
                   "dcap(), tcap()", "rapid()", "dcr(), nndr()",
                   "recordLinkage()",
                   "mia_classifier(), singling_out()", "propscore(), specks()",
                   "rumap()"),
  check.names = FALSE
)
knitr::kable(comp,
             caption = "Risk/utility measures across R and Python packages.")
```

Py = Python.


# Software Design and Architecture {#sec-design}

## Design Philosophy

Existing R packages for statistical disclosure control --- sdcMicro
[@templ2015sdcmicro], synthpop [@nowok2016synthpop], and simPop
[@templ2017simulation] --- focus primarily on *applying* protection methods,
with risk and utility assessment provided as secondary features. riskutility
inverts this emphasis: it is a dedicated evaluation package, designed to be
used *after* protection has been applied, regardless of which tool generated
the released data.

The package follows four design principles:

1. **Consistent S3 architecture.** Every major function returns a typed S3
   object with `print()`, `summary()`, and `plot()` methods. We chose S3
   over S4 classes for three reasons: lighter memory footprint, simpler
   method dispatch for the typical R user, and easier interoperability with
   data.table and ggplot2 objects. All risk/utility classes follow
   the same pattern, making the API predictable once one class is learned.

2. **Direction conventions.** Risk measures are oriented so that higher
   values always indicate higher disclosure risk. Utility measures are
   oriented so that higher values indicate higher utility (better
   preservation of statistical properties). Some measures (e.g., pMSE,
   Wasserstein distance) naturally use a "lower is better" scale; their
   interpretation is noted in the documentation, and `rumap()` applies
   the necessary direction transformation when normalizing to $[0, 1]$.

3. **Minimal core dependencies.** Frequency-based privacy models, CAP
   metrics, and distance-based measures require only base R, data.table, and
   ggplot2. ML-based methods require optional packages --- ranger for random
   forests in RAPID, xgboost for gradient boosting, and rpart for
   classification trees. These are loaded conditionally via
   `requireNamespace()` and produce informative error messages when absent.

4. **Integration over competition.** Rather than reimplementing synthesis or
   anonymization, the package provides `from_synthpop()`, `from_simPop()`,
   and `from_sdcMicro()` constructors that extract original and released
   data from objects created by these packages, wrapping them in the
   `synth_pair` container (Section \@ref(sec-synth-pair)). This ensures
   users can evaluate any protection method with a single consistent
   interface.

## The synth_pair Container {#sec-synth-pair}

The central data structure in riskutility is the `synth_pair` object, which
bundles original data, released data, variable roles, and metadata into a
single container:

```{r synth-pair-demo, eval=FALSE}
pair <- synth_pair(original, synthetic,
                   key_vars = c("age", "gender", "region"),
                   target_var = "income",
                   holdout = holdout_data)
```

The constructor stores the original and synthetic data frames alongside
their dimensions, automatically detects categorical (`cat_vars`) and numeric
(`num_vars`) columns, and retains user-specified quasi-identifiers
(`key_vars`), sensitive attribute (`target_var`), and optional holdout data.
This metadata eliminates a common source of error: specifying
different `key_vars` for different risk measures on the same dataset.

Once constructed, every risk and utility function in the package accepts a
`synth_pair` object as its first argument via S3 dispatch:

```{r synth-pair-dispatch, eval=FALSE}
# All functions accept synth_pair --- no parameter repetition:
dcap(pair)                    # Attribution risk
rapid(pair, model_type = "rf") # ML-based risk
propscore(pair)                # Propensity score utility
disclosure_report(pair)        # Full risk report
rumap(pair)                    # Risk-Utility map
```

## S3 Method Pattern {#sec-s3-pattern}

Every exported risk/utility class follows an identical three-part pattern:

```{r s3-pattern, eval=FALSE}
# 1. Two equivalent calling conventions:
result <- dcap(pair)                                      # synth_pair method
result <- dcap(X, Y, key_vars = ..., target_var = ...)    # default method

# 2. Inspection:
print(result)                # One-screen summary with key statistic
s <- summary(result)         # Detailed statistics (returns summary.dcap)
print(s)                     # Formatted multi-line output

# 3. Visualization:
plot(result, which = 1)      # Plot type 1
plot(result, which = 1:2)    # Multiple plot types
```

The generic function dispatches via `UseMethod()`. The `synth_pair` method
extracts `original`, `synthetic`, `key_vars`, and `target_var` from the
container and delegates to the default method, which performs the actual
computation. The return object is a list with a class attribute (e.g.,
`"dcap"`). The `summary()` method returns a typed summary object (e.g.,
`"summary.dcap"`) with its own `print()` method, separating computation from
display.

Plot methods use an integer `which` parameter to select among multiple
visualization types. The number of available plot types varies by class,
from one (simple measures) to seven (`rumap`).

## Integration with the R Ecosystem {#sec-integration}

The `from_*` family of constructors bridges riskutility with the three main
R packages for statistical disclosure control and synthetic data generation:

```{r integration, eval=FALSE}
# From synthpop: pass synds object + original data
pair <- from_synthpop(synds_object, original_data,
                      key_vars = c("age", "sex"),
                      target_var = "income")

# From simPop: original data extracted automatically from simPopObj
pair <- from_simPop(simPopObj,
                    key_vars = c("age", "sex"),
                    target_var = "income")

# From sdcMicro: variable roles extracted from sdcMicroObj
pair <- from_sdcMicro(sdcMicroObj)
```

Each constructor returns a standard `synth_pair` object. `from_sdcMicro()`
additionally extracts variable roles (quasi-identifiers, sensitive
attributes, sample weights) from the sdcMicro S4 object, so that
`key_vars` and `target_var` need not be specified manually.
`from_synthpop()` supports multiple syntheses via the `m` parameter,
selecting a specific synthetic dataset from a `synds` object.
`from_simPop()` extracts sample weights when available, enabling
weighted risk calculations.


# Risk Measures {#sec-risk}

This section presents the six risk measure families, each illustrated with a
worked example using the same running dataset.

```{r running-data}
set.seed(42)
n <- 500
original <- data.frame(
  age = sample(18:85, n, replace = TRUE),
  sex = factor(sample(c("M", "F"), n, replace = TRUE)),
  education = factor(sample(c("Primary", "Secondary", "Tertiary"), n,
                            replace = TRUE, prob = c(0.3, 0.5, 0.2))),
  region = factor(sample(paste0("R", 1:5), n, replace = TRUE)),
  income = round(rlnorm(n, log(40000), 0.5))
)

# Synthetic: independent draws (low risk expected)
synthetic <- data.frame(
  age = sample(18:85, n, replace = TRUE),
  sex = factor(sample(c("M", "F"), n, replace = TRUE)),
  education = factor(sample(c("Primary", "Secondary", "Tertiary"), n,
                            replace = TRUE, prob = c(0.3, 0.5, 0.2))),
  region = factor(sample(paste0("R", 1:5), n, replace = TRUE)),
  income = round(rlnorm(n, log(40000), 0.5))
)

key_vars <- c("age", "sex", "education", "region")
target_var <- "income"

pair <- synth_pair(original, synthetic,
                   key_vars = key_vars, target_var = target_var)

# Train/holdout split for distance-based metrics
set.seed(123)
train_idx <- sample(n, size = floor(0.7 * n))
train_data <- original[train_idx, ]
holdout_data <- original[-train_idx, ]
```


## Privacy Models and Frequency-Based Risk {#sec-privacy-models}

These methods assess privacy properties of a *single* dataset based on
quasi-identifier frequencies. Originally developed for traditionally
anonymized data, they apply equally to synthetic data. They do not compare
original and released data; instead, they evaluate structural properties of
the released data alone.

**$k$-Anonymity** [@samarati1998protecting] partitions records into
equivalence classes (ECs) based on quasi-identifier values. The dataset
satisfies $k$-anonymity if every EC contains at least $k$ records:
$k = \min_i |\text{EC}(\mathbf{q}_i)|$, where $\mathbf{q}_i$ is the
quasi-identifier vector of record $i$. Small ECs are vulnerable to identity
disclosure because an attacker who knows a target's quasi-identifiers can
narrow them to fewer than $k$ candidates. $k$-Anonymity protects against
identity disclosure but not attribute disclosure: if all $k$ records in a
class share the same sensitive value, the attribute is trivially revealed.

**$l$-Diversity** [@machanavajjhala2007ldiversity] strengthens $k$-anonymity by
requiring that each EC contains at least $l$ distinct values of the sensitive
attribute. This prevents attribute disclosure even when all records in an EC
share the same sensitive value (homogeneity attack).

**$t$-Closeness** [@li2007tcloseness] further requires that the distribution of
the sensitive attribute within each EC is close to its overall distribution.
The distance is measured by the Earth Mover's Distance (EMD), and $t$ is the
maximum EMD across all ECs.

```{r privacy-models}
# k-Anonymity: minimum equivalence class size
k_res <- riskutility::kanonymity(synthetic, key_vars = key_vars)
k_res

# l-Diversity: sensitive attribute diversity per EC
l_res <- riskutility::ldiversity(synthetic, key_vars = key_vars,
                                 sensitive_var = target_var)
print(l_res)

# t-Closeness: EMD between EC and overall distribution
t_res <- riskutility::tcloseness(synthetic, key_vars = key_vars,
                                 sensitive_var = target_var)
t_res
```

With 68 unique age values, 2 sex levels, 3 education levels, and 5 regions,
there are up to $68 \times 2 \times 3 \times 5 = 2040$ possible QI
combinations for only $n = 500$ records. Most equivalence classes are
singletons, yielding $k = 1$ --- this is expected for fine-grained
quasi-identifiers and does not by itself indicate a problem with the
synthetic data. The three models form a hierarchy: $k$-anonymity guards
against identity disclosure, $l$-diversity against homogeneity attacks,
and $t$-closeness against skewness attacks.

Additional frequency-based measures include `individual_risk()` for
per-record re-identification probability based on EC frequencies
[@franconi2004individual; @skinner2002measure], `attacker_risk()` for
scenario-based assessment under prosecutor, journalist, and marketer
attacker models [@hundepool2012statistical], `suda()` for detecting records
unique on small QI subsets, and `population_uniqueness()` for estimating
population-level uniques via super-population models
[@reiter2005estimating].

```{r table5-privacy, echo=FALSE}
privacy <- data.frame(
  Function = c("kanonymity()", "ldiversity()", "tcloseness()",
               "suda()", "individual_risk()", "population_uniqueness()",
               "attacker_risk()", "epsilon_identifiability()"),
  Input = rep("Single dataset", 8),
  `Key output` = c("Min EC size", "Min distinct values per EC",
                    "Max EMD across ECs", "SUDA scores",
                    "Per-record frequency risk", "Estimated pop. uniques",
                    "Scenario-based risk", "Identifiability fraction"),
  `Threats` = c("Identity", "Attribute", "Attribute", "Identity",
                "Identity", "Identity", "Identity", "Identity"),
  check.names = FALSE
)
knitr::kable(privacy,
             caption = "Privacy models overview.")
```


## Attribution-Based Risk: The CAP Family {#sec-cap}

The Correct Attribution Probability framework [@taub2018differential] measures
attribute disclosure: can an attacker infer a sensitive value by matching
quasi-identifiers between original and released data?

For each original record $i$ with quasi-identifier values $\mathbf{q}_i$,
the attacker finds all records in the released data whose quasi-identifiers
match $\mathbf{q}_i$ (the *equivalence class*). If the sensitive attribute
is homogeneous within this class, the attacker learns the true value. The
**Targeted CAP** (TCAP) gives each record a risk score between 0 and 1:
$\text{TCAP}_i = \Pr(\text{correct attribution} \mid \mathbf{q}_i)$. The
mean CAP across all records is $\overline{\text{CAP}} = n^{-1} \sum_i \text{CAP}_i$
(returned as `cap`). The **Differential CAP** subtracts the baseline (modal-class)
attribution rate, $\text{DCAP} = \overline{\text{CAP}} - \text{baseline}$ (returned
as `dcap`; @taub2018differential), so a value near zero indicates no attribution
gain over random guessing. The `summary()` method also reports the risk ratio
$\overline{\text{CAP}} / \text{baseline}$ to contextualize the result.

The **WEAP** (Within-EC Attribution Probability) evaluates risk from the
released data alone, without access to the original, making it suitable
for data custodians who cannot share the original data with an auditor.
**DiSCO** (Disclosive in Synthetic, Correct in Original) identifies records
that are both confidently attributed in the released data and correctly
attributed in the original.

```{r cap-demo}
# TCAP: per-record risk (most informative member of CAP family)
tcap_res <- tcap(pair)
summary(tcap_res)
plot(tcap_res)
```

Since our running example uses independently generated synthetic data (no
relationship to the original), TCAP values should be close to the baseline
attribution probability. Records with TCAP above 0.1 warrant closer
inspection.

```{r cap-table, echo=FALSE}
cap <- data.frame(
  Metric = c("DCAP", "TCAP", "WEAP", "DiSCO"),
  `Requires original?` = c("Yes", "Yes", "No", "Yes"),
  `Per-record?` = c("No", "Yes", "Yes", "Yes"),
  `Measures` = c("Mean attribution probability",
                 "Individual attribution risk",
                 "Within-EC homogeneity",
                 "Correct + confident attribution"),
  `Low risk` = c("ratio < 1.5", "< 0.1 per record",
                 "< 0.1", "< 5%"),
  check.names = FALSE
)
knitr::kable(cap,
             caption = "CAP family comparison with interpretation thresholds.")
```


## ML-Based Risk: RAPID {#sec-rapid}

Risk of Attribute Prediction-Induced Disclosure [@thees2026beyond] takes a
fundamentally different approach to attribute disclosure. Instead of matching
quasi-identifiers, RAPID trains a predictive model $\hat{f}$ on the released
data $(Y_{\mathcal{K}}, Y_s)$ and evaluates its predictions on the original
data: $\hat{s}_i = \hat{f}(X_{\mathcal{K},i})$. For numeric targets, a
record is *at risk* when the prediction error falls below a threshold
$\epsilon$: $e(s_i, \hat{s}_i) < \epsilon$, where $e(\cdot, \cdot)$ is
a configurable error metric (symmetric percentage error by default). The
RAPID score is the fraction of at-risk records:
$\text{RAPID} = n^{-1} \sum_i \mathbf{1}(e(s_i, \hat{s}_i) < \epsilon)$.
For categorical targets, a different evaluation applies: a record is at risk
when a gain or ratio score exceeds a threshold.

```{r rapid-demo, warning=FALSE}
rapid_res <- rapid(pair, model_type = "lm")
summary(rapid_res)
plot(rapid_res, which = c(1, 3))
```

With independently generated synthetic data, we expect the RAPID score to
be close to the baseline. The threshold sensitivity plot (`which = 3`)
shows how the at-risk fraction changes across threshold values.

RAPID complements the CAP family in two ways. First, it captures non-linear
relationships and variable interactions. Second, it provides inferential tools:
`rapid_test()` computes a permutation-based $p$-value, `confint()` provides
bootstrap confidence intervals, and `rapid_threshold_select()` optimizes the
threshold in a data-driven manner.

```{r rapid-models, echo=FALSE}
models <- data.frame(
  Model = c("lm", "rf", "cart", "gbm", "logit"),
  Package = c("stats", "ranger", "rpart", "xgboost", "stats"),
  Numeric = c("Yes", "Yes", "Yes", "Yes", "No"),
  Categorical = c("No", "Yes", "Yes", "Yes", "Yes"),
  Interactions = c("Manual", "Automatic", "Automatic", "Automatic", "Manual"),
  check.names = FALSE
)
knitr::kable(models, caption = "RAPID model backends.")
```

| RAPID Score | Risk Level | Interpretation |
|-------------|------------|----------------|
| < 0.05 | Low | ML model cannot predict target much better than baseline |
| 0.05--0.15 | Moderate | Some predictive signal from synthetic data |
| 0.15--0.30 | Elevated | Significant predictive leakage |
| > 0.30 | High | Strong evidence of disclosure risk |


## Distance-Based Risk {#sec-distance}

Distance-based methods detect *memorization*: the failure mode where a
generative model reproduces training records verbatim or near-verbatim. The
key idea is the **holdout method** --- split the original data into a
training set $T$ (used for synthesis) and a holdout set $H$ (unseen by the
generator). For each synthetic record $y_j$, compute the Distance to
Closest Record in $T$ ($d_T$) and in $H$ ($d_H$). If the generator has
generalized, $d_T$ and $d_H$ should be comparable. If it has memorized,
$d_T$ will be systematically smaller:

$$\text{DCR\_share} = n^{-1} \sum_j \mathbf{1}\bigl(d_T(y_j) < d_H(y_j)\bigr), \qquad \text{DCR\_ratio} = \frac{\bar{d}_T}{\bar{d}_H}$$

A DCR share meaningfully above 0.5 (the package flags shares above 0.55), or a
ratio below about 1, suggests memorization. The **NNDR** (Nearest Neighbor
Distance Ratio) provides a complementary view: for each synthetic record, it is
the ratio of the distance to its nearest neighbor over the distance to its
second-nearest neighbor. A ratio near 0 indicates a single dominant match;
near 1 indicates no distinctive match. **IMS** (Identical
Match Share) counts exact copies. When an explicit holdout is unavailable,
`holdout_fraction` automatically splits the original data.

```{r distance-demo, warning=FALSE}
dcr_res <- dcr(pair, holdout_fraction = 0.2)
summary(dcr_res)
plot(dcr_res, which = 1)
```

**The DCR Delusion.** @yao2025dcr show that DCR can fail to
detect privacy leakage: datasets deemed "private" by DCR may still be
vulnerable to membership inference attacks. Their central recommendation is that
DCR be interpreted relative to a proper null distribution rather than in
absolute terms; `dcr()` implements exactly this, comparing the observed share
against a permutation null (`null_test`) and reporting a Wilcoxon test alongside
the point estimate. Even so, distance-based metrics should always be
complemented with other risk families.

```{r distance-table, echo=FALSE}
dist <- data.frame(
  Metric = c("DCR", "NNDR", "IMS", "RF proximity", "dRisk", "Hitting rate",
             "Epsilon ID", "Delta-presence"),
  Holdout = c("Yes", "Yes", "No", "Yes", "No", "No", "No", "No"),
  Detects = c("Memorization", "Memorization", "Exact copies",
              "Memorization (non-linear)", "Close records", "Close records",
              "Identifiability", "Membership bounds"),
  `Low risk` = c("share < 0.55", "share < 0.55", "< 0.01",
                 "ratio near 1", "< 0.05", "< 0.05",
                 "< 0.01", "> 0.5"),
  check.names = FALSE
)
knitr::kable(dist,
             caption = "Distance-based and proximity risk measures.")
```

**RF proximity** offers a data-adaptive alternative: it trains a random forest
to discriminate original from synthetic records and measures how often
synthetic records share terminal nodes with training vs. holdout records,
capturing non-linear proximity that fixed distance metrics may miss. Use
`rf_privacy()` when complex interactions are expected.


## Record Linkage Risk {#sec-recordlinkage}

The `recordLinkage()` function directly simulates a re-identification attack by
linking each original record to the most similar record(s) in the anonymized
dataset. Eight methods are implemented, spanning deterministic (Gower distance),
probabilistic (Fellegi-Sunter, @fellegi1969theory), PRAM-aware, predictive
(propensity score), random forest proximity, rank-based (RBRL,
@muralidhar2016rankbased), robust Mahalanobis [@templ2008robust], and
autoencoder embedding [@guo2016entity]. Three matching modes are available:
independent (many-to-one), bijective (one-to-one via Hungarian algorithm,
@herranz2016gdbrl), and optimal transport (Sinkhorn).

```{r recordlinkage-demo}
# Namespace-qualified: sdcMicro (>= 5.8.2) also exports recordLinkage(), and
# vignettes may be re-built in a shared session, so qualify to be unambiguous.
rl_res <- riskutility::recordLinkage(pair, method = "deterministic")
print(rl_res)
```

For full details, see the companion vignette `vignette("recordLinkage")`.

```{r recordlinkage-table, echo=FALSE}
rl <- data.frame(
  Method = c("Deterministic", "Probabilistic", "PRAM", "Predictive",
             "RF", "RBRL", "Mahalanobis", "Embedding"),
  Distance = c("Gower", "Fellegi-Sunter", "Transition prob.", "Propensity",
               "RF proximity", "Rank-based", "Mahalanobis", "Autoencoder"),
  `Mixed types` = c("Yes", "Yes", "Categorical", "Yes",
                     "Yes", "Yes", "Numeric", "Yes"),
  Matching = c("All 3", "All 3", "All 3", "All 3",
               "All 3", "Independent", "All 3", "All 3"),
  check.names = FALSE
)
knitr::kable(rl,
             caption = "Record linkage methods. All 3 = independent, bijective, OT.")
```


## Membership Inference and Anonymization Failure Criteria {#sec-membership}

This section groups two related but distinct concerns. The first three
measures (`mia_classifier()`, `domias()`, `nnaa()`) assess *membership
disclosure* --- whether a membership inference attack (MIA) can determine if a
specific individual's data was used during
synthesis. The singling out and linkability attacks operationalize two of the GDPR
anonymization criteria [@wp29anonymisation], following the attack-based
approach of @giomi2023anonymeter.

**NNAA** (Nearest Neighbor Adversarial Accuracy, @yale2020generation) is based on
the adversarial accuracy of a nearest-neighbour two-sample comparison,
$\text{AA}(A,S) = \tfrac{1}{2}\bigl[\Pr(d_{AS} > d_{AA}) + \Pr(d_{SA} > d_{SS})\bigr]$.
The reported privacy loss is $\text{AA}(\text{holdout}, S) - \text{AA}(\text{train}, S)$;
a positive value means synthetic records resemble training records more closely
than holdout records, indicating memorization:

```{r nnaa-demo}
nnaa_res <- nnaa(train_data, synthetic, holdout = holdout_data,
                 method = "gower", seed = 42)
print(nnaa_res)
```

| Privacy Loss | Interpretation |
|-------------|----------------|
| Near 0 | No detectable leakage (ideal) |
| 0.01--0.05 | Minor leakage, likely acceptable |
| > 0.10 | Significant memorization |

**Singling out** and **linkability** operationalize two of the Article 29
Working Party's three anonymization criteria (the third, inference, is
addressed by the attribution-based CAP and RAPID measures):

```{r membership-demo}
so_res <- singling_out(original, synthetic,
                       n_attacks = 500, n_cols = 3,
                       mode = "multivariate", seed = 42)
print(so_res)

link_res <- linkability(original, synthetic,
                        n_attacks = 500, n_neighbors = 1, seed = 42)
print(link_res)
```

```{r membership-table, echo=FALSE}
mia <- data.frame(
  Metric = c("MIA classifier", "DOMIAS", "NNAA",
             "Singling out", "Linkability", "delta-Presence"),
  `Attack type` = c("Shadow model", "Density overfitting",
                     "Nearest neighbor", "Predicate-based",
                     "Record linkage", "Membership bounds"),
  Holdout = c("Yes", "Yes", "Yes", "Yes", "Yes", "No"),
  `GDPR criterion` = c("--", "--", "--",
                        "Art. 29 WP", "Art. 29 WP", "--"),
  `Low risk` = c("< 0.55", "< 0.6", "< 0.05",
                 "< 0.1", "< 0.1", "> 0.5"),
  check.names = FALSE
)
knitr::kable(mia,
             caption = "Membership inference and GDPR measures.")
```


## Cross-Family Comparison {#sec-rosetta}

No single metric tells the full story. Applying all families to the same
dataset reveals complementary and sometimes contradictory information:

```{r rosetta}
# Near-copy: original + small noise (high risk expected)
set.seed(99)
near_copy <- original
near_copy$age <- near_copy$age + sample(-1:1, n, replace = TRUE)
near_copy$income <- near_copy$income + round(rnorm(n, 0, 500))
pair_risky <- synth_pair(original, near_copy,
                         key_vars = key_vars, target_var = target_var)

# Compare key metrics across the two datasets
comparison <- data.frame(
  Metric = c("DCAP", "RAPID (lm)", "IMS"),
  Safe = c(
    dcap(pair)$dcap,
    rapid(pair, model_type = "lm", verbose = FALSE)$rapid,
    ims(pair)$ims
  ),
  Risky = c(
    dcap(pair_risky)$dcap,
    rapid(pair_risky, model_type = "lm", verbose = FALSE)$rapid,
    ims(pair_risky)$ims
  )
)
comparison$Safe <- round(comparison$Safe, 4)
comparison$Risky <- round(comparison$Risky, 4)
knitr::kable(comparison,
             caption = "Cross-family comparison: safe vs. risky synthetic data.")
```

The near-copy shows elevated risk across all families, but the magnitude
and interpretation differ. Attribution measures quantify *information
leakage*; distance-based measures quantify *memorization*. These
complementary perspectives mean that a dataset can pass one family's tests
while failing another's --- a thorough evaluation uses at least one measure
from each family.


# Data Utility Measures {#sec-utility}

A dataset that passes all risk checks but destroys the analytical value of
the data is useless. Utility measures quantify how well the released data
preserves the statistical properties of the original.

## Global Utility: Propensity Scores {#sec-utility-quick}

Global utility measures give a single-number verdict by asking: *can a
classifier tell original and synthetic records apart?*

The propensity score method [@woo2009global; @snoke2018general] pools
original ($X$, $n_X$ records) and synthetic ($Y$, $n_Y$ records) data,
labels them (0/1), and fits a classifier. The **pMSE** (propensity score
Mean Squared Error) measures how well the model discriminates:

$$\text{pMSE} = \frac{1}{N} \sum_{i=1}^{N} \left(\hat{p}_i - c\right)^2$$

where $N = n_X + n_Y$ and $c = n_Y / N$. If original and synthetic records
are indistinguishable, pMSE $\approx 0$.

```{r utility-quick, warning=FALSE}
prop_res <- propscore(pair)
summary(prop_res)
```

| pMSE Value | Interpretation |
|------------|----------------|
| < 0.01 | Excellent fidelity |
| 0.01--0.05 | Good fidelity |
| 0.05--0.10 | Moderate differences |
| > 0.10 | Poor fidelity |

## Univariate Diagnostics {#sec-utility-univariate}

When global utility is poor, per-variable measures identify which variables
are responsible. For **numeric** variables, the Wasserstein distance
measures the cost of transforming one distribution into another. For
**categorical** variables, the Hellinger distance measures distributional
overlap:

$$H(p, q) = \frac{1}{\sqrt{2}} \sqrt{\sum_{k=1}^{K} \left(\sqrt{p_k} - \sqrt{q_k}\right)^2}$$

```{r utility-univariate}
# Hellinger distance for categorical variables
h_res <- hellinger(original, synthetic, vars = c("sex", "education"))
print(h_res)

# CI proximity: confidence interval overlap for means
cip_res <- ci_proximity(original, synthetic, vars = c("age", "income"))
print(cip_res)
```

The **CI proximity** measure [@karr2006framework] compares confidence intervals
of summary statistics (means) between original and synthetic data. An overlap
near 1 means the intervals coincide; a relative error near 0 means point
estimates are close.


## Multivariate and Structural Utility {#sec-utility-structural}

Marginal distributions can match perfectly while joint distributions diverge.
The **energy distance** [@szekely2013energy] is a multivariate two-sample
statistic sensitive to differences in both location and scale (lower values
indicate closer joint distributions):

```{r utility-structural}
e_res <- energy_distance(original[, c("age", "income")],
                         synthetic[, c("age", "income")],
                         seed = 42)
print(e_res)
```

The **MMD** (Maximum Mean Discrepancy, @gretton2012kernel) provides a
kernel-based alternative supporting exact computation and random Fourier
features (RFF) for large datasets:

```{r mmd-demo}
mmd_res <- mmd(original[, c("age", "income")],
               synthetic[, c("age", "income")],
               kernel = "gaussian", method = "rff",
               n_features = 500, seed = 42)
print(mmd_res)
```

**Copula fidelity** compares the empirical copula (rank dependence structure)
using the Cramér-von Mises statistic on pairwise copula CDFs. **Contingency
fidelity** [@snoke2018general] is its categorical complement, computing total
variation distance between bivariate contingency tables:

```{r fidelity-demo}
cop_res <- copula_fidelity(original, synthetic, vars = c("age", "income"))
print(cop_res)

ctf_res <- contingency_fidelity(original, synthetic,
                                vars = c("sex", "education", "region"))
print(ctf_res)
```


## Predictive Utility {#sec-utility-predictive}

**TSTR** (Train on Synthetic, Test on Real, @zhao2021ctgan) trains a
predictive model on the synthetic data and evaluates performance on held-out
real data. The ratio of TSTR-to-TRTR performance quantifies how well
predictive relationships are preserved:

```{r tstr-demo, warning=FALSE, eval=requireNamespace("ranger", quietly=TRUE)}
set.seed(42)
tstr_res <- tstr(pair, target_var = "income", model = "rf",
                 test_fraction = 0.3, seed = 42)
print(tstr_res)
```

**Regression fidelity** [@karr2006framework] fits the same regression model on
both datasets and compares coefficient estimates via CI overlap, standardized
bias, and significance agreement:

```{r regression-demo}
reg_res <- regression_fidelity(original, synthetic,
                               formula = income ~ age + sex + education)
summary(reg_res)
plot(reg_res, which = 1)
```

**Tail fidelity** assesses how well extreme values are preserved --- critical
for applications where tail behavior matters (financial risk, rare diseases):

```{r tail-demo}
tail_res <- tail_fidelity(original, synthetic, vars = c("age", "income"),
                          percentile = 95, tails = "both")
print(tail_res)
```

**Subgroup utility** [@snoke2018general] applies any utility measure to each
subgroup defined by a grouping variable, identifying groups with low utility:

```{r subgroup-demo}
su_res <- subgroup_utility(original, synthetic, group_var = "region",
                           utility_fun = energy_distance,
                           threshold = 0.5, seed = 42)
print(su_res)
```

The conservative `utility_score` is the worst subgroup score. A `ratio` near
1 indicates homogeneous utility; below 0.5 indicates substantial disparity.

```{r table7-utility, echo=FALSE}
util <- data.frame(
  `Use case` = c(rep("Quick assessment", 2),
                 rep("Univariate", 3),
                 rep("Multivariate", 4),
                 rep("Predictive", 3),
                 "Subgroup"),
  Function = c("propscore()", "specks()",
               "compare_wasserstein()", "hellinger()", "ci_proximity()",
               "energy_distance()", "mmd()",
               "copula_fidelity()", "contingency_fidelity()",
               "tstr()", "regression_fidelity()",
               "compare_feature_importance()",
               "subgroup_utility()"),
  `Data type` = c("Mixed", "Mixed",
                   "Numeric", "Categorical", "Numeric",
                   "Numeric", "Numeric",
                   "Numeric", "Categorical",
                   "Mixed", "Mixed", "Mixed",
                   "Mixed"),
  Interpretation = c("< 0.1: good", "< 0.05: good",
                     "Lower = better", "< 0.1: good", "> 0.8: good",
                     "Lower = better", "Lower = better",
                     "< 0.1: good", "< 0.05: good",
                     "ratio near 1: good", "overlap > 0.8: good",
                     "High corr: good",
                     "min > 0.5: good"),
  check.names = FALSE
)
knitr::kable(util,
             caption = "Utility measures by use case.")
```


# Comprehensive Assessment: A Case Study {#sec-comprehensive}

This section demonstrates the complete practitioner workflow on a realistic
dataset, comparing three synthesis approaches with different privacy-utility
trade-offs.

## Scenario and Data {#sec-case-data}

Consider a statistical agency that wants to release a survey dataset ($n = 1000$)
containing demographic variables (age, sex, education, region) and a sensitive
income variable.

```{r case-data}
set.seed(123)
N <- 1000
edu_levels <- c("Primary", "Secondary", "Tertiary")
age_groups <- c("20-29", "30-39", "40-49", "50-59", "60-69")
orig <- data.frame(
  age_group = factor(sample(age_groups, N, replace = TRUE)),
  sex = factor(sample(c("M", "F"), N, replace = TRUE)),
  education = factor(sample(edu_levels, N, replace = TRUE,
                            prob = c(0.25, 0.50, 0.25))),
  region = factor(sample(paste0("R", 1:4), N, replace = TRUE))
)
edu_effect <- c(Primary = 0, Secondary = 0.3, Tertiary = 0.7)
age_effect <- c("20-29" = 0, "30-39" = 0.15, "40-49" = 0.3,
                "50-59" = 0.4, "60-69" = 0.35)
orig$income <- round(exp(
  10 + age_effect[as.character(orig$age_group)] +
    edu_effect[as.character(orig$education)] + rnorm(N, 0, 0.4)
))

qi <- c("age_group", "sex", "education", "region")
sens <- "income"
```

We create three synthetic datasets spanning the privacy-utility spectrum:

```{r case-synthesis}
set.seed(456)

# Method A: Independent marginals (safest, but destroys correlations)
synA <- data.frame(
  age_group = factor(sample(age_groups, N, replace = TRUE)),
  sex = factor(sample(c("M", "F"), N, replace = TRUE)),
  education = factor(sample(edu_levels, N, replace = TRUE,
                            prob = c(0.25, 0.50, 0.25))),
  region = factor(sample(paste0("R", 1:4), N, replace = TRUE)),
  income = sample(orig$income, N, replace = TRUE)
)

# Method B: Category-preserving bootstrap with income noise
idx_B <- sample(N, N, replace = TRUE)
synB <- orig[idx_B, ]
rownames(synB) <- NULL
synB$income <- round(synB$income * exp(rnorm(N, 0, 0.15)))
swap_idx <- sample(N, round(0.2 * N))
synB$age_group[swap_idx] <- factor(sample(age_groups,
                                          length(swap_idx), replace = TRUE))

# Method C: Near-copy with minimal perturbation (risky)
synC <- orig
synC$income <- round(synC$income * exp(rnorm(N, 0, 0.03)))
```

## Step 1: Quick Risk Screening with disclosure_report() {#sec-case-report}

The `disclosure_report()` function computes multiple risk measures, evaluates
each against a threshold, and produces a pass/fail assessment:

```{r case-report, warning=FALSE}
pair_A <- synth_pair(orig, synA, key_vars = qi, target_var = sens)
pair_B <- synth_pair(orig, synB, key_vars = qi, target_var = sens)
pair_C <- synth_pair(orig, synC, key_vars = qi, target_var = sens)

rep_A <- disclosure_report(pair_A, compute = c("attribution", "privacy"),
                           seed = 42, verbose = FALSE)
rep_B <- disclosure_report(pair_B, compute = c("attribution", "privacy"),
                           seed = 42, verbose = FALSE)
rep_C <- disclosure_report(pair_C, compute = c("attribution", "privacy"),
                           seed = 42, verbose = FALSE)

verdicts <- data.frame(
  Method = c("A: Independent", "B: Bootstrap+noise", "C: Near-copy"),
  Overall = c(rep_A$overall_risk, rep_B$overall_risk, rep_C$overall_risk),
  Pass = c(rep_A$n_pass, rep_B$n_pass, rep_C$n_pass),
  Warn = c(rep_A$n_warn, rep_B$n_warn, rep_C$n_warn)
)
knitr::kable(verdicts, caption = "Quick risk screening across three methods.")
```

Two patterns emerge. First, attribution metrics differentiate the three methods
in the expected order: Method A (independent) has the lowest attribution risk,
Method C (near-copy) the highest. Second, privacy models flag all three methods
because they evaluate the released data's structure alone --- with 120 possible
QI combinations and $n = 1000$ records, some equivalence classes are small
regardless of synthesis method. This illustrates a key lesson: privacy models
and attribution metrics answer different questions and should be interpreted
together.

## Step 2: Comparative Assessment with rumap() {#sec-case-rumap}

The `rumap()` function implements the multivariate Risk-Utility framework of
@thees2026beyond. Traditional R-U analysis plots a single risk measure against
a single utility measure, producing a two-dimensional trade-off curve. This
can be misleading: a method may appear optimal on one pair of measures while
performing poorly on another. `rumap()` computes multiple risk and utility
measures simultaneously, normalizes to $[0, 1]$, and identifies
**Pareto-optimal** methods.

```{r case-rumap, warning=FALSE}
set.seed(42)
ru <- rumap(orig,
            list("A: Independent" = synA,
                 "B: Bootstrap+noise" = synB,
                 "C: Near-copy" = synC),
            risk_measures = c("dcap", "tcap", "ims"),
            utility_measures = c("pmse", "wasserstein"),
            key_vars = qi, target_var = sens,
            seed = 42)
print(ru)
```

```{r case-rumap-scatter, fig.width=8, fig.height=6}
plot(ru, which = 1)  # R-U scatterplot with Pareto front
```

The R-U scatterplot places each method in the composite risk-utility plane.
Methods in the lower-right corner (low risk, high utility) are preferred.

```{r case-rumap-heatmap, fig.width=8, fig.height=5}
plot(ru, which = 2)  # Heatmap of individual measures
```

The heatmap reveals *why* the methods differ. Method A achieves low risk
across all measures but has poor utility. Method C has excellent utility but
elevated attribution risk. Method B balances the two.

## Step 3: Decision {#sec-case-decision}

The analysis supports a structured decision:

- **Method A** is appropriate when data is released to the general public
  and any re-identification would be unacceptable.
- **Method B** is appropriate for controlled-access research environments
  where moderate risk is acceptable.
- **Method C** should be rejected --- its risk profile is too close to the
  original data.

This iterative workflow --- screen with `disclosure_report()`, compare with
`rumap()`, and refine synthesis parameters --- is the core use case that
riskutility is designed to support.


# Summary and Discussion {#sec-discussion}

## Contributions

The riskutility package makes five contributions to the R ecosystem for
statistical disclosure control:

1. **Comprehensive coverage.** It is the first R package to unify all six
   risk assessment paradigms --- frequency-based privacy models, attribution
   (CAP), ML-based (RAPID), distance-based, record linkage, and membership
   inference --- under a single API.

2. **Novel implementations.** It provides the first R implementations of
   RAPID, two of the three GDPR failure criteria from @stadler2022synthetic
   (singling out and linkability), $t$-closeness, DOMIAS density-based
   membership inference, and eight-method record linkage with bijective and
   optimal transport matching.

3. **Unified API.** The `synth_pair` container and consistent S3 class
   pattern eliminate parameter repetition and ensure practitioners can switch
   between risk measures without learning new interfaces.

4. **Multivariate R-U mapping.** The `rumap()` function implements the
   framework of @thees2026beyond for comparing multiple synthesis approaches
   on multiple risk and utility dimensions simultaneously.

5. **Ecosystem integration.** The `from_sdcMicro()`, `from_synthpop()`, and
   `from_simPop()` constructors allow practitioners to evaluate data produced
   by any of the three main R packages.

## Partially vs Fully Synthetic Data

The privacy-utility evaluation differs depending on the synthesis approach
[@drechsler2011synthetic]:

- **Fully synthetic data**: All records are synthetic; all risk metrics are
  directly applicable.
- **Partially synthetic data**: Only sensitive values are replaced.
  Attribution metrics (DCAP, RAPID) are particularly relevant because original
  quasi-identifiers provide a direct matching key. Distance-based metrics
  should focus on the synthesized variables.
- **Multiple synthetic datasets**: When $m > 1$ datasets are generated, evaluate
  each separately and report the worst-case risk across all $m$ releases.

## Remediation

If disclosure risk is too high:

1. Add more noise to the synthesis process
2. Reduce granularity of quasi-identifiers (e.g., age groups instead of exact age)
3. Apply additional anonymization techniques (suppression, generalization)
4. Re-synthesize with different model settings
5. Re-evaluate with **riskutility** --- iterate until risk-utility balance is acceptable

If utility is too low:

1. Use a more flexible synthesizer (CART, Bayesian network, GAN)
2. Reduce the amount of perturbation
3. Identify affected variables/subgroups (`subgroup_utility()`, per-variable Hellinger)
4. Consider partially synthetic data (synthesize only sensitive variables)

## Limitations and Recommendations

**No formal privacy guarantees.** All measures provide empirical risk
assessment. A low DCAP score does not prove that no attacker can succeed.
Empirical and formal approaches (differential privacy) are complementary.

**Key variable selection.** Results depend heavily on the choice of
quasi-identifiers. Practitioners should base QI selection on a realistic
threat model, not on convenience.

**Threshold interpretation.** The pass/fail thresholds used by
`disclosure_report()` are pragmatic defaults. Different contexts require
different thresholds.

We recommend the following minimal evaluation protocol:

1. **Always run** `disclosure_report()` with `compute = "all"` as a first
   screening.
2. **For publication-quality assessment**, use `rumap()` to compare multiple
   synthesis approaches and identify Pareto-optimal methods.
3. **For regulatory compliance** (GDPR), include singling out and
   linkability tests.
4. **Interpret distance-based metrics cautiously.** Following @yao2025dcr,
   do not rely on DCR/NNDR alone.

## Future Work

Four extensions are planned: (i) a Shiny dashboard for interactive evaluation;
(ii) integration with differential privacy frameworks;
(iii) computational optimizations for large datasets ($n > 50{,}000$); and
(iv) population-level risk estimation from sample data.


# Computational Details {#sec-computational}

All computations were performed using R `r getRversion()` [@R2025] with the
riskutility package version `r packageVersion("riskutility")`. Core
dependencies include data.table for efficient data manipulation and ggplot2
for all visualizations. ML-based methods require optional packages: ranger,
rpart, xgboost, and caret, loaded conditionally via `requireNamespace()`.

```{r scalability-table, echo=FALSE}
scale_df <- data.frame(
  Metric = c("dcap()", "dcr()", "kanonymity()", "energy_distance()",
             "mmd(method='rff')", "propscore()", "rumap()"),
  `n=1000` = c("< 1 s", "< 1 s", "< 1 s", "< 1 s", "< 1 s", "~1 s", "~10 s"),
  `n=10000` = c("~5 s", "~10 s", "~1 s", "~2 s", "~1 s", "~5 s", "~60 s"),
  `n=100000` = c("~60 s", "~5 min", "~5 s", "~30 s", "~5 s", "~30 s", "depends"),
  Scaling = c("O(n*k)", "O(n^2)", "O(n log n)", "O(n^2)", "O(n*D)",
              "O(n*p)", "Sum of components"),
  check.names = FALSE
)
knitr::kable(scale_df, caption = "Approximate runtimes for key metrics.")
```

```{r session-info}
sessionInfo()
```


# References {-}
