---
documentclass: jss
author:
  - name: Matthias Templ
    orcid: 0000-0002-8638-5276
    affiliation: University of Applied Sciences and Arts Northwestern Switzerland
    address: |
      | School of Business
      | University of Applied Sciences and Arts
      | Northwestern Switzerland (FHNW)
      | Riggenbachstrasse 16, 4600 Olten, Switzerland
    email: \email{matthias.templ@fhnw.ch}
  - name: Oscar Thees
    orcid: 0009-0001-9378-4988
    affiliation: University of Applied Sciences and Arts Northwestern Switzerland
    address: |
      | School of Business
      | University of Applied Sciences and Arts
      | Northwestern Switzerland (FHNW)
title:
  formatted: "\\pkg{riskutility}: Disclosure Risk and Data Utility Assessment for Anonymized and Synthetic Data in \\proglang{R}"
  plain:     "riskutility: Disclosure Risk and Data Utility Assessment for Anonymized and Synthetic Data in R"
  short:     "\\pkg{riskutility}: Risk and Utility Assessment"
abstract: >
  The \pkg{riskutility} package for \proglang{R} provides a comprehensive
  framework for measuring disclosure risk and data utility in synthetic and
  anonymized datasets. It implements over 30 risk measures spanning six
  paradigms---frequency-based privacy models ($k$-anonymity, $l$-diversity,
  $t$-closeness), attribution-based measures (CAP, TCAP, WEAP, DiSCO), an
  ML-based measure (RAPID), distance-based measures (DCR, NNDR, IMS), record
  linkage, and membership inference (DOMIAS, NNAA)---together with the GDPR
  anonymization criteria of singling out and linkability, and a broad suite of
  utility measures including propensity scores, distributional distances
  (Hellinger, energy, MMD), dependence-structure fidelity (copula, contingency),
  regression fidelity, subgroup-stratified utility, and Train on Synthetic, Test
  on Real (TSTR). The multivariate Risk-Utility map (\code{rumap()}) computes
  many measures on several candidate releases under one call and summarises
  them as composite risk and utility scores; we show that Pareto membership
  derived from such composites is itself a random quantity, and quantify its
  stability. All functions share a consistent S3 API with \code{print()},
  \code{summary()}, and \code{plot()} methods and accept a \code{synth\_pair}
  container. We describe the package architecture, demonstrate the complete
  workflow from synthetic-data generation to comprehensive risk-utility
  assessment, and present a case study comparing three synthesis approaches.
keywords:
  formatted: [synthetic data, disclosure risk, data utility, statistical disclosure control, "\\proglang{R}"]
  plain:     [synthetic data, disclosure risk, data utility, statistical disclosure control, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{booktabs}
  \usepackage{longtable}
output: rticles::jss_article
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  prompt = TRUE,
  comment = NA,
  collapse = FALSE,
  fig.width = 4.9,
  fig.height = 3.675,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
options(prompt = "R> ", continue = "+  ", width = 76, digits = 4,
        knitr.table.format = "latex")
library("riskutility")

## Helper: format a regression coefficient for inline reporting (no sci. notation)
fmt_coef <- function(x, term, which = c("orig", "synth")) {
  which <- match.arg(which)
  i <- grep(term, x$coefficients$term)[1]
  v <- x$coefficients[[paste0("estimate_", which)]][i]
  formatC(round(v), format = "d", big.mark = ",")
}

## Helper: render a Graphviz DOT specification to PDF for inclusion as a figure
## (requires DiagrammeR, DiagrammeRsvg, and rsvg).
render_dot <- function(dot) {
  f <- tempfile(fileext = ".pdf")
  svg <- DiagrammeRsvg::export_svg(DiagrammeR::grViz(dot))
  rsvg::rsvg_pdf(charToRaw(svg), f)
  knitr::include_graphics(f)
}
```

# Introduction {#sec-intro}

Statistical agencies and data custodians face a fundamental challenge: releasing
data that are useful for analysis while protecting individual privacy. Two main
approaches exist: (i) traditional anonymization methods such as perturbation,
suppression, and generalization [@hundepool2012statistical;
@templ2017statistical], and (ii) synthetic data generation via statistical
models [@nowok2016synthpop; @templ2017simulation]. Both require rigorous
evaluation of *disclosure risk* and *data utility* --- yet existing tools
assess these two dimensions in isolation or cover only a narrow subset of the
relevant metrics.

The \proglang{R} package \pkg{sdcMicro} [@templ2015sdcmicro] provides frequency-based risk
estimation but limited utility assessment. \pkg{synthpop} [@nowok2016synthpop]
offers the CAP disclosure metric and propensity-score utility but not
distance-based or membership inference methods. In \proglang{Python}, \pkg{SDMetrics} [@patki2016sdv] covers
distance-based metrics and \pkg{Anonymeter} [@giomi2023anonymeter] implements the GDPR
failure criteria, but neither addresses attribution risk; the broader evaluation
libraries \pkg{synthcity} [@qian2023synthcity] and \pkg{SynthEval} [@lautrup2025syntheval]
are also \proglang{Python}-only. No existing *R* package provides a unified
framework that spans all risk families and supports multivariate comparison of
multiple synthesis approaches; the most comprehensive evaluation suites (the
\proglang{Python} libraries \pkg{synthcity} and \pkg{SynthEval}) sit outside the R ecosystem.

The **\pkg{riskutility}** package for R [@R2025] fills this gap. It implements
over 30 risk measures spanning six paradigms --- frequency-based privacy
models, attribution-based (CAP), ML-based (\pkg{RAPID}), distance-based, record
linkage, and membership inference --- alongside more than a dozen utility
functions covering global, distributional, structural, and predictive assessment. All functions
share a consistent S3 API with `print()`, `summary()`, and `plot()` methods
and accept a `synth_pair` container that bundles data and metadata.

The multivariate Risk-Utility methodology underlying `rumap()` is developed,
with its statistical motivation and properties, in a companion paper
[@thees2026beyond, under review at the *Journal of Official Statistics*]. The
present paper is a *software* contribution: it describes the package design, the
unified API, and the implementation of more than 30 risk and utility measures,
of which `rumap()` is one component.

The need for rigorous evaluation is well established in the statistical
disclosure control literature and has gained urgency as synthetic data enters
regulatory frameworks. The European Article 29 Working Party
[@wp29anonymisation] identifies three criteria for anonymization: singling out,
linkability, and inference. **\pkg{riskutility}** measures the first two directly
(`singling_out()`, `linkability()`); the inference criterion is approached
through the attribution-based CAP and \pkg{RAPID} measures rather than a dedicated
inference-attack function.
Importantly, synthetic data does not automatically satisfy these criteria
[@stadler2022synthetic], making empirical assessment essential.

The distinction between *general* and *specific* utility [@snoke2018general]
is important: general utility measures (propensity scores, distributional
distances) assess overall distributional fidelity, while specific utility
measures (regression fidelity, TSTR) assess whether particular analyses
yield similar results on original and synthetic data. A thorough evaluation
should include both.

**\pkg{riskutility}** provides empirical privacy metrics rather than formal
mathematical guarantees. Unlike differential privacy, which provides
worst-case bounds, our metrics quantify observed risk on concrete datasets.
This approach is appropriate for practical SDC workflows where the protection
mechanism is not differentially private. For differentially private
synthesizers, formal epsilon budgets should be used alongside empirical
evaluation.

**Quick start.** A comprehensive risk assessment requires just a few lines:

```{r intro-hook, eval=FALSE}
library(riskutility)
pair <- synth_pair(original, synthetic,
                   key_vars = c("age", "sex", "region"),
                   target_var = "income")
report <- disclosure_report(pair)
print(report)
```

This single call computes attribution-based risk (DCAP, TCAP, WEAP, DiSCO,
\pkg{RAPID}), distance-based risk (DCR, NNDR, IMS, dRisk, hitting rate), privacy
models ($k$-anonymity, $l$-diversity, $t$-closeness), and membership
inference measures (singling out, linkability, NNAA) --- producing a
pass/fail summary across all families. For comparing multiple synthesis
approaches, `rumap()` computes risk and utility measures simultaneously,
normalizes them to a common scale, and identifies Pareto-optimal methods
(Section \ref{sec-case-rumap}).

**Paper overview.** Section \ref{sec-background} introduces the disclosure
threat taxonomy and reviews related software. Section \ref{sec-design}
describes the package architecture and the `synth_pair` container.
Sections \ref{sec-risk} and \ref{sec-utility} present the risk and utility
measures with mathematical formulations and worked examples.
Section \ref{sec-comprehensive} demonstrates the complete practitioner
workflow on a realistic case study, comparing three synthesis approaches
with `disclosure_report()` and `rumap()`. Section \ref{sec-discussion}
discusses limitations, remediation strategies, and future work.
Section \ref{sec-computational} provides computational details.
Table \ref{tab:scope} summarizes the package's scope at a glance.

```{r table1-scope, echo=FALSE}
scope <- data.frame(
  Category = c("Privacy models", "Attribution (CAP)",
                "ML-based (RAPID)", "Distance-based", "Record linkage",
                "Membership inference", "Utility measures", "Frameworks"),
  Functions = c(8, 4, 4, 6, "1 (8 methods)", 6, 15, 3),
  Paradigm = c("Equivalence class", "Matching", "Prediction",
                "Nearest neighbor", "Linkage", "Attack simulation",
                "Various", "Composite"),
  check.names = FALSE
)
knitr::kable(scope, caption = "Package scope at a glance. All families apply to both traditionally anonymized and synthetic data.\\label{tab:scope}")
```

All analyses in this paper were conducted in \proglang{R} [@R2025].
Figure \ref{fig:decision} is a decision guide to the package's main functions,
organized by the question the analyst is asking.

```{r fig-decision, echo=FALSE, out.width="95%", fig.cap="Decision guide: choosing a riskutility function by assessment goal and disclosure threat.\\label{fig:decision}"}
render_dot('
digraph decision {
  graph [rankdir=LR, fontsize=11];
  node [shape=box, style="rounded,filled", fillcolor="#EAF2F8", fontname="Helvetica", fontsize=9];
  edge [fontname="Helvetica", fontsize=8];
  Q  [label="What do you want\nto assess?", fillcolor="#D4E6F1"];
  R  [label="Disclosure risk"];
  U  [label="Data utility"];
  M  [label="Compare multiple\nsynthesizers?", fillcolor="#D4E6F1"];
  Q -> R; Q -> U; Q -> M;
  R -> Tid [label="identity"];
  R -> Tat [label="attribute"];
  R -> Tme [label="membership"];
  R -> Tmo [label="memorization"];
  Tid [label="recordLinkage()\nkanonymity()\nindividual_risk()", fillcolor="#FDEDEC"];
  Tat [label="dcap(), tcap()\nweap(), disco(), rapid()", fillcolor="#FDEDEC"];
  Tme [label="domias(), nnaa()\nsingling_out(), linkability()", fillcolor="#FDEDEC"];
  Tmo [label="dcr(), nndr(), ims()", fillcolor="#FDEDEC"];
  U -> Ug [label="global"];
  U -> Ud [label="distributional"];
  U -> Us [label="structural"];
  U -> Up [label="predictive"];
  Ug [label="propscore()\nspecks()", fillcolor="#E8F8F5"];
  Ud [label="hellinger()\nenergy_distance(), mmd()", fillcolor="#E8F8F5"];
  Us [label="copula_fidelity()\ncontingency_fidelity()", fillcolor="#E8F8F5"];
  Up [label="tstr()\nregression_fidelity()", fillcolor="#E8F8F5"];
  M -> RU [label="yes"];
  RU [label="rumap()", fillcolor="#FCF3CF"];
}
')
```


# Background: Threat Taxonomy and Related Software {#sec-background}

## Disclosure Threat Taxonomy {#sec-threats}

Every risk measure in \pkg{riskutility} addresses one or more disclosure threats.
The classical SDC taxonomy [@hundepool2012statistical; @templ2017statistical]
recognizes three types: identity, attribute, and inferential disclosure. Modern
synthetic-data evaluation adds *membership disclosure* (from the ML privacy
literature, @shokri2017membership) and *memorization* (relevant to generative
models). We organize the measures around the four operational threats below.
Inferential disclosure---learning a statistical property of a group rather than
an exact record value---is, at the record level, the quantity the
attribute-disclosure measures (CAP, \pkg{RAPID}) estimate; we therefore fold it into
attribute disclosure rather than treating it as a separate operational category:

- **Identity disclosure**: An attacker links a released record to a specific
  individual. Measured by: `recordLinkage()`, `kanonymity()`,
  `individual_risk()`, `hitting_rate()`.
- **Attribute disclosure**: An attacker learns a sensitive attribute value
  through quasi-identifier (QI) matching. Measured by: `dcap()`, `tcap()`,
  `weap()`, `disco()`, `rapid()`.
- **Membership disclosure**: An attacker determines whether an individual's
  data was used to create the released dataset. Measured by membership-inference
  attacks (`mia_classifier()`, `domias()`, `nnaa()`) and bounded by
  `delta_presence()`. The related GDPR criteria of *singling out*
  (`singling_out()`) and *linkability* (`linkability()`) are reported alongside
  these measures (Section \ref{sec-membership}).
- **Memorization**: A generative model reproduces training records verbatim
  or near-verbatim. Measured by: `ims()`, `dcr()`, `nndr()`.

Table \ref{tab:threats} summarizes the four threats and their principal measures.

```{r table2-threats, echo=FALSE}
threats <- data.frame(
  Threat = c("Identity", "Attribute", "Membership", "Memorization"),
  Definition = c(
    "Link a record to an individual",
    "Learn a sensitive value via linkage",
    "Determine whether an individual is in the data",
    "Generator reproduces training records"
  ),
  `Key measures` = c(
    "recordLinkage, kanonymity",
    "dcap, tcap, weap, disco, rapid",
    "mia_classifier, domias, nnaa",
    "ims, dcr, nndr"
  ),
  check.names = FALSE
)
knitr::kable(threats, caption = "Disclosure threat taxonomy.\\label{tab:threats}")
```

## Risk Assessment Paradigms

\pkg{riskutility} implements six risk assessment paradigms, each addressing different
threats from the taxonomy above. We summarize each here; full definitions and
worked examples are in Section \ref{sec-risk}.

**Frequency-based (privacy models).** These methods assess privacy properties
of a single dataset based on quasi-identifier frequencies.
$k$-Anonymity [@samarati1998protecting] requires each combination of
quasi-identifiers to appear at least $k$ times. $l$-Diversity
[@machanavajjhala2007ldiversity] and $t$-closeness [@li2007tcloseness]
progressively strengthen the protection of sensitive attributes within
equivalence classes. See Section \ref{sec-privacy-models}.

**Attribution-based (CAP family).** The Correct Attribution Probability
framework [@taub2018differential] measures whether an attacker can infer
sensitive attribute values by matching quasi-identifiers between original and
released data. DCAP provides an aggregate measure; TCAP gives per-record
risk scores. See Section \ref{sec-cap}.

**ML-based (\pkg{RAPID}).** Risk of Attribute Prediction-Induced Disclosure
[@templ2026rapid] trains predictive models on the released data and evaluates
them on the original. Accurate predictions indicate information leakage. \pkg{RAPID}
captures complex non-linear relationships and variable interactions that the
CAP matching approach may miss. See Section \ref{sec-rapid}.

**Distance-based.** The holdout method compares distances from synthetic records
to training data vs. holdout data. If synthetic records are systematically
closer to training records, the generator has memorized rather than generalized.
However, @yao2025dcr demonstrate that passing distance-based tests does
not guarantee privacy (the "DCR Delusion"). See Section \ref{sec-distance}.

**Record linkage.** Directly simulates a re-identification attack using
deterministic (Gower distance), probabilistic (Fellegi-Sunter), PRAM-aware,
predictive (propensity-score), or random forest linkage
[@fellegi1969theory; @domingoferrer2003disclosure]. See
Section \ref{sec-recordlinkage} and `?recordLinkage`.

**Membership inference.** Shadow model attacks [@shokri2017membership],
density-based detection [@hu2023domias], and the Article 29 Working Party
anonymization criteria [@wp29anonymisation] assess whether an attacker can
determine if a specific individual's data was used during synthesis. See Section \ref{sec-membership}.

## Utility Assessment Paradigms

Utility measures quantify how well the released data preserves the
statistical properties of the original. We organize them into four groups
(details in Section \ref{sec-utility}):

**Global utility** measures use propensity scores to quantify how
distinguishable the released data is from the original [@woo2009global;
@snoke2018general]. A single number summarizes overall data quality.

**Distributional utility** measures (Wasserstein, Hellinger, KS test,
energy distance, MMD) compare marginal or joint distributions, identifying
which specific variables or relationships are poorly reproduced.

**Structural utility** measures (copula fidelity, contingency fidelity,
PCA comparison, correlation matrix comparison) assess whether multivariate
relationships --- the correlations and interactions that analysts depend
on --- are preserved.

**Predictive utility** measures (TSTR, regression fidelity, feature
importance stability) test whether models trained on released data
generalize to real-world predictions.

## Related Software

**In R.** \pkg{sdcMicro} [@templ2015sdcmicro] is the established toolkit for
frequency-based risk estimation and, since version 5.8.2, distance-based record
linkage, but it focuses primarily on *applying* anonymization methods.
\pkg{synthpop} [@nowok2016synthpop] has grown well beyond the CAP metric: its
`disclosure()` function now returns identity measures (UiO, UiS, repU) and
attribute measures (DiSCO, DiSDiO) alongside CAP and TCAP
[@raab2024practical], together with propensity-score utility. It does not
cover distance-based memorization measures, ML-based attribute inference, or
membership inference. \pkg{syntheval} (Urban Institute, not on CRAN) targets
utility rather than risk. The authors' own \pkg{RAPID} package
[@templ2026rapid] implements the \pkg{RAPID} measure alone; \pkg{riskutility} supersedes
it and adds the inferential tools of Section \ref{sec-rapid}. What no \proglang{R} package offers is a single interface in
which measures from all six risk paradigms and the utility families can be
computed on one pair of datasets and compared.

**In \proglang{Python}.** \pkg{SDMetrics} (part of the Synthetic Data Vault
[@patki2016sdv]) provides distance-based and statistical quality metrics and
also implements CAP-style attribution measures. \pkg{Anonymeter}
[@giomi2023anonymeter] operationalizes the three Article 29 Working Party
criteria [@wp29anonymisation] (singling out, linkability, inference). Neither
covers the traditional privacy models ($k$-anonymity, $l$-diversity,
$t$-closeness), and neither is available to R users without leaving the
language.

Table \ref{tab:software} contrasts the coverage of the main R and \proglang{Python} tools.

```{r table3-comparison, echo=FALSE}
comp <- data.frame(
  Measure = c("k-Anonymity", "l-Diversity", "t-Closeness",
              "CAP family", "RAPID", "DCR / NNDR", "Record linkage",
              "WP29 criteria / MIA", "pMSE / SPECKS",
              "R-U map"),
  sdcMicro = c("kAnon()", "ldiversity()", "--",
               "--", "--", "dRisk()*", "recordLinkage()",
               "--", "dUtility()", "--"),
  synthpop = c("--", "--", "--",
               "disclosure()", "--", "--", "--",
               "--", "utility.gen()", "--"),
  Python = c("--", "--", "--",
             "SDMetrics", "--", "SDMetrics", "--",
             "Anonymeter", "--", "--"),
  riskutility = c("kanonymity()", "ldiversity()", "tcloseness()",
                  "dcap(), tcap()", "rapid()",
                  "dcr(), nndr()", "recordLinkage()",
                  "singling_out()",
                  "propscore(), specks()",
                  "rumap()"),
  check.names = FALSE
)
knitr::kable(comp,
             caption = "Risk and utility measures across the main R and Python tools. *sdcMicro's dRisk() is an interval/RMD risk for paired perturbed records, not a train-versus-holdout comparison.\\label{tab:software}")
```



# Software Design and Architecture {#sec-design}

## Design Philosophy

Existing \proglang{R} packages for statistical disclosure control --- \pkg{sdcMicro}
[@templ2015sdcmicro], \pkg{synthpop} [@nowok2016synthpop], and \pkg{simPop}
[@templ2017simulation] --- focus primarily on *applying* protection methods,
with risk and utility assessment provided as secondary features. \pkg{riskutility}
inverts this emphasis: it is a dedicated evaluation package, designed to be
used *after* protection has been applied, regardless of which tool generated
the released data.

The package follows four design principles:

1. **Consistent S3 architecture.** Every major function returns a typed S3
   object with `print()`, `summary()`, and `plot()` methods. We chose S3
   over S4 classes for three reasons: lighter memory footprint, simpler
   method dispatch for the typical R user, and easier interoperability with
   data.table and ggplot2 objects. All risk/utility classes follow
   the same pattern, making the API predictable once one class is learned.

2. **Direction conventions.** Risk measures are oriented so that higher
   values always indicate higher disclosure risk. Utility measures are
   oriented so that higher values indicate higher utility (better
   preservation of statistical properties). Some measures (e.g., pMSE,
   Wasserstein distance) naturally use a "lower is better" scale; their
   interpretation is noted in the documentation, and `rumap()` applies
   the necessary direction transformation when normalizing to $[0, 1]$.

3. **Small, stable dependency set.** The required dependencies are
   \pkg{data.table} [@pkg_datatable] for the frequency and matching
   operations, \pkg{ggplot2} [@pkg_ggplot2] for all graphics, \pkg{VIM}
   [@pkg_vim] for Gower distances, and \pkg{ranger} [@pkg_ranger] /
   \pkg{randomForest} for the random-forest backends used by `rapid()` and
   `propscore()`. Only the remaining model backends are optional ---
   \pkg{xgboost} [@pkg_xgboost] for gradient boosting, \pkg{rpart}
   [@pkg_rpart] for classification trees, \pkg{caret} [@pkg_caret] for the
   `tstr()` training interface, and \pkg{clue} [@pkg_clue] and \pkg{torch}
   for bijective and embedding-based record linkage. These are loaded
   conditionally via `requireNamespace()` and produce informative error
   messages when absent.

4. **Integration over competition.** Rather than reimplementing synthesis or
   anonymization, the package provides `from_synthpop()`, `from_simPop()`,
   and `from_sdcMicro()` constructors that extract original and released
   data from objects created by these packages, wrapping them in the
   `synth_pair` container (Section \ref{sec-synth-pair}). This ensures
   users can evaluate any protection method with a single consistent
   interface.

## The Paired-Data Container {#sec-synth-pair}

The central data structure in \pkg{riskutility} is the `synth_pair` object, which
bundles original data, released data, variable roles, and metadata into a
single container:

```{r synth-pair-demo, eval=FALSE}
pair <- synth_pair(original, synthetic,
                   key_vars = c("age", "gender", "region"),
                   target_var = "income",
                   holdout = holdout_data)
```

The constructor stores the original and synthetic data frames alongside
their dimensions, automatically detects categorical (`cat_vars`) and numeric
(`num_vars`) columns, and retains user-specified quasi-identifiers
(`key_vars`), sensitive attribute (`target_var`), and optional holdout data.
This metadata eliminates a common source of error: specifying
different `key_vars` for different risk measures on the same dataset.

Once constructed, every risk and utility function in the package accepts a
`synth_pair` object as its first argument via S3 dispatch:

```{r synth-pair-dispatch, eval=FALSE}
# All functions accept synth_pair --- no parameter repetition:
dcap(pair)                    # Attribution risk
rapid(pair, model_type = "rf") # ML-based risk
propscore(pair)                # Propensity score utility
disclosure_report(pair)        # Full risk report
rumap(pair)                    # Risk-Utility map
```

## S3 Method Pattern {#sec-s3-pattern}

Every exported risk/utility class follows an identical three-part pattern:

```{r s3-pattern, eval=FALSE}
# 1. Two equivalent calling conventions:
result <- dcap(pair)                                      # synth_pair method
result <- dcap(X, Y, key_vars = ..., target_var = ...)    # default method

# 2. Inspection:
print(result)                # One-screen summary with key statistic
s <- summary(result)         # Detailed statistics (returns summary.dcap)
print(s)                     # Formatted multi-line output

# 3. Visualization:
plot(result, which = 1)      # Plot type 1
plot(result, which = 1:2)    # Multiple plot types
```

The generic function dispatches via `UseMethod()`. The `synth_pair` method
extracts `original`, `synthetic`, `key_vars`, and `target_var` from the
container and delegates to the default method, which performs the actual
computation. The return object is a list with a class attribute (e.g.,
`"dcap"`). The `summary()` method returns a typed summary object (e.g.,
`"summary.dcap"`) with its own `print()` method, separating computation from
display.

Plot methods use an integer `which` parameter to select among multiple
visualization types. The number of available plot types varies by class,
from one (simple measures) to seven (`rumap`).

## Integration with the R Ecosystem {#sec-integration}

Figure \ref{fig:integration} shows how \pkg{riskutility} connects to the wider R
ecosystem.

```{r fig-integration, echo=FALSE, out.width="90%", fig.cap="Ecosystem integration: converters map synthpop, simPop, and sdcMicro objects (and plain data frames) into the paired-data container consumed by all measures.\\label{fig:integration}"}
render_dot('
digraph integration {
  graph [rankdir=LR, fontsize=10];
  node [shape=box, style="rounded,filled", fontname="Helvetica", fontsize=9];
  edge [fontname="Helvetica", fontsize=8];
  synthpop [label="synthpop\n(synds)", fillcolor="#FDEBD0"];
  simPop   [label="simPop\n(simPopObj)", fillcolor="#FDEBD0"];
  sdcMicro [label="sdcMicro\n(sdcMicroObj)", fillcolor="#FDEBD0"];
  raw      [label="data frames\n(original, synthetic)", fillcolor="#FDEBD0"];
  sp [label="synth_pair", fillcolor="#D4E6F1"];
  ru [label="riskutility\nrisk + utility measures\ndisclosure_report(), rumap()", fillcolor="#D5F5E3"];
  synthpop -> sp [label="from_synthpop()"];
  simPop   -> sp [label="from_simPop()"];
  sdcMicro -> sp [label="from_sdcMicro()"];
  raw      -> sp [label="synth_pair()"];
  sp -> ru;
}
')
```

The `from_*` family of constructors bridges \pkg{riskutility} with the three main
\proglang{R} packages for statistical disclosure control and synthetic data generation:

```{r integration, eval=FALSE}
# From synthpop: pass synds object + original data
pair <- from_synthpop(synds_object, original_data,
                      key_vars = c("age", "sex"),
                      target_var = "income")

# From simPop: original data extracted automatically from simPopObj
pair <- from_simPop(simPopObj,
                    key_vars = c("age", "sex"),
                    target_var = "income")

# From sdcMicro: variable roles extracted from sdcMicroObj
pair <- from_sdcMicro(sdcMicroObj)
```

Each constructor returns a standard `synth_pair` object. `from_sdcMicro()`
additionally extracts variable roles (quasi-identifiers, sensitive
attributes, sample weights) from the \pkg{sdcMicro} S4 object, so that
`key_vars` and `target_var` need not be specified manually.
`from_synthpop()` supports multiple syntheses via the `m` parameter,
selecting a specific synthetic dataset from a `synds` object.
`from_simPop()` extracts sample weights when available, enabling
weighted risk calculations.


# Risk Measures {#sec-risk}

This section presents the six risk measure families, each illustrated with a
worked example using the same running dataset.

```{r running-data}
set.seed(42)
n <- 500
edu_levels <- c("Primary", "Secondary", "Tertiary")
regions <- paste0("R", 1:5)

age <- sample(25:75, n, replace = TRUE)
sex <- factor(sample(c("M", "F"), n, replace = TRUE))

# Education depends on age: younger cohorts are better educated.
education <- factor(ifelse(runif(n) < 0.5 * plogis(-0.05 * (age - 45)),
                           "Tertiary",
                           sample(edu_levels[1:2], n, replace = TRUE,
                                  prob = c(0.4, 0.6))),
                    levels = edu_levels)

# Region depends on education: R1 and R2 are urban and attract graduates.
region <- factor(ifelse(education == "Tertiary",
                        sample(regions, n, TRUE,
                               prob = c(.35, .30, .15, .10, .10)),
                        sample(regions, n, TRUE,
                               prob = c(.10, .15, .25, .25, .25))),
                 levels = regions)

# Income depends on age, education and sex.
edu_effect <- c(Primary = 0, Secondary = 0.35, Tertiary = 0.75)
income <- round(exp(log(20000) + 0.020 * age +
                      edu_effect[as.character(education)] -
                      0.15 * (sex == "F") + rnorm(n, 0, 0.18)))

original <- data.frame(age, sex, education, region, income)

# Synthetic: every variable is resampled independently from its own marginal.
# Marginal distributions are preserved; the joint structure is destroyed.
synthetic <- as.data.frame(lapply(original, function(v)
  sample(v, n, replace = TRUE)))

key_vars <- c("age", "sex", "education", "region")
target_var <- "income"

pair <- synth_pair(original, synthetic,
                   key_vars = key_vars, target_var = target_var)

# Train/holdout split for distance-based metrics
set.seed(123)
train_idx <- sample(n, size = floor(0.7 * n))
train_data <- original[train_idx, ]
holdout_data <- original[-train_idx, ]
```


## Privacy Models and Frequency-Based Risk {#sec-privacy-models}

These methods assess privacy properties of a *single* dataset based on
quasi-identifier frequencies. Originally developed for traditionally
anonymized data, they apply equally to synthetic data. They do not compare
original and released data; instead, they evaluate structural properties of
the released data alone.

**$k$-Anonymity** [@samarati1998protecting] partitions records into
equivalence classes (ECs) based on quasi-identifier values. The dataset
satisfies $k$-anonymity if every EC contains at least $k$ records:
$k = \min_i |\text{EC}(\mathbf{q}_i)|$, where $\mathbf{q}_i$ is the
quasi-identifier vector of record $i$. Small ECs are vulnerable to identity
disclosure because an attacker who knows a target's quasi-identifiers can
narrow them to fewer than $k$ candidates. $k$-Anonymity protects against
identity disclosure but not attribute disclosure: if all $k$ records in a
class share the same sensitive value, the attribute is trivially revealed.

**$l$-Diversity** [@machanavajjhala2007ldiversity] strengthens $k$-anonymity by
requiring that each EC contains at least $l$ distinct values of the sensitive
attribute. This prevents attribute disclosure even when all records in an EC
share the same sensitive value (homogeneity attack).

**$t$-Closeness** [@li2007tcloseness] further requires that the distribution of
the sensitive attribute within each EC is close to its overall distribution.
The distance is measured by the Earth Mover's Distance (EMD), and $t$ is the
maximum EMD across all ECs.

```{r privacy-models}
# k-Anonymity: minimum equivalence class size
k_res <- riskutility::kanonymity(synthetic, key_vars = key_vars)
k_res

# l-Diversity: sensitive attribute diversity per EC
l_res <- riskutility::ldiversity(synthetic, key_vars = key_vars,
                                 sensitive_var = target_var)
print(l_res)

# t-Closeness: EMD between EC and overall distribution
t_res <- riskutility::tcloseness(synthetic, key_vars = key_vars,
                                 sensitive_var = target_var)
t_res
```

With 51 unique age values, two sex levels, three education levels and five
regions, there are up to $51 \times 2 \times 3 \times 5 = 1{,}530$ possible QI
combinations for only $n = 500$ records. Most equivalence classes are
singletons, yielding $k = 1$ --- this is expected for fine-grained
quasi-identifiers and does not by itself indicate a problem with the
synthetic data. The three models form a hierarchy: $k$-anonymity guards
against identity disclosure, $l$-diversity against homogeneity attacks,
and $t$-closeness against skewness attacks.

Additional frequency-based measures include `individual_risk()` for
per-record re-identification probability based on EC frequencies
[@franconi2004individual; @skinner2002measure], `attacker_risk()` for
scenario-based assessment under prosecutor, journalist, and marketer
attacker models [@hundepool2012statistical], `suda()` for detecting records
unique on small QI subsets, and `population_uniqueness()` for estimating
population-level uniques via super-population models
[@reiter2005estimating].

```{r table5-privacy, echo=FALSE}
privacy <- data.frame(
  Function = c("kanonymity()", "ldiversity()", "tcloseness()",
               "suda()", "individual_risk()", "population_uniqueness()",
               "attacker_risk()", "epsilon_identifiability()"),
  Input = c(rep("Single dataset", 7), "Both datasets"),
  `Key output` = c("Min EC size", "Min distinct values per EC",
                    "Max EMD across ECs", "SUDA scores",
                    "Per-record frequency risk", "Estimated pop. uniques",
                    "Scenario-based risk", "Identifiability fraction"),
  `Threats` = c("Identity", "Attribute", "Attribute", "Identity",
                "Identity", "Identity", "Identity", "Identity"),
  check.names = FALSE
)
knitr::kable(privacy,
             caption = "Privacy models overview.\\label{tab:privacy}")
```


## Attribution-Based Risk: The CAP Family {#sec-cap}

The Correct Attribution Probability framework [@taub2018differential] measures
attribute disclosure: can an attacker infer a sensitive value by matching
quasi-identifiers between original and released data?

For each original record $i$ with quasi-identifier values $\mathbf{q}_i$,
the attacker finds all records in the released data whose quasi-identifiers
match $\mathbf{q}_i$ (the *equivalence class*). If the sensitive attribute
is homogeneous within this class, the attacker learns the true value. The
**Targeted CAP** (TCAP) gives each record a risk score between 0 and 1:
$\text{TCAP}_i = \Pr(\text{correct attribution} \mid \mathbf{q}_i)$. The
mean CAP across all records is $\overline{\text{CAP}} = n^{-1} \sum_i \text{CAP}_i$
(returned as `cap`). The **Differential CAP** subtracts the baseline (modal-class)
attribution rate, $\text{DCAP} = \overline{\text{CAP}} - \text{baseline}$ (returned
as `dcap`; @taub2018differential), so a value near zero indicates no attribution
gain over random guessing. The `summary()` method also reports the risk ratio
$\overline{\text{CAP}} / \text{baseline}$ to contextualize the result.

The **WEAP** (Within-EC Attribution Probability) evaluates risk from the
released data alone, without access to the original, making it suitable
for data custodians who cannot share the original data with an auditor.
**DiSCO** [@raab2024practical] (Disclosive in Synthetic, Correct in Original) identifies records
that are both confidently attributed in the released data and correctly
attributed in the original.

```{r cap-demo, fig.cap="Per-record TCAP attribution-risk scores for the running example.\\label{fig:tcap}"}
# TCAP: per-record risk (most informative member of CAP family)
tcap_res <- tcap(pair)
summary(tcap_res)
plot(tcap_res)
```

Because the synthesiser resamples each variable independently, no synthetic
record retains the original association between a quasi-identifier vector and
an income band, and the mean TCAP (`r round(tcap_res$tcap_mean, 4)`) sits
*below* the modal-guess baseline (`r round(tcap_res$baseline, 4)`), giving a
risk ratio of `r round(tcap_res$tcap_mean / tcap_res$baseline, 2)`. Values at or below the
baseline are the signature of no attribution leakage. Note that `tcap()`, like
`dcap()`, discretises a continuous target into `cont_bins` quantile bins
before matching; without that step an exact match on a continuous income value
essentially never occurs and every TCAP score would collapse to zero.

```{r cap-table, echo=FALSE}
cap <- data.frame(
  Metric = c("DCAP", "TCAP", "WEAP", "DiSCO"),
  `Requires original?` = c("Yes", "Yes", "No", "Yes"),
  `Per-record?` = c("No", "Yes", "Yes", "Yes"),
  `Measures` = c("Mean attribution probability",
                 "Individual attribution risk",
                 "Within-EC homogeneity",
                 "Correct + confident attribution"),
  `Low risk` = c("ratio <= 1.5", "ratio <= 1.5",
                 "<= 5% disclosive", "ratio <= 1.5"),
  check.names = FALSE
)
knitr::kable(cap,
             caption = "CAP family comparison with interpretation thresholds.\\label{tab:cap}")
```


## Validation Against Existing Tools {#sec-validation}

Where definitions overlap, \pkg{riskutility} reproduces the values of established
tools exactly. \pkg{synthpop}'s `disclosure()` returns several attribution
measures under one heading. Its `DCAP` is the mean correct-attribution
probability of the *original* records under the *synthetic* conditional
distribution --- precisely the quantity `dcap()` reports as `cap`. Its `CAPd`
is a different estimand, computed within the original data alone, and is not
the correct comparison. Using `education` as a categorical target and
\pkg{synthpop} to generate the synthetic data:

```{r synthpop-comparison, warning=FALSE, eval=requireNamespace("synthpop", quietly = TRUE)}
keys_cmp <- c("sex", "region")
syn_sp   <- synthpop::syn(original, seed = 2024, print.flag = FALSE)
pair_sp  <- synth_pair(original, syn_sp$syn,
                       key_vars = keys_cmp, target_var = "education")
disc_sp  <- synthpop::disclosure(syn_sp, original, keys = keys_cmp,
                                 target = "education", print.flag = FALSE)

# from_synthpop() builds the paired container directly from a synds object,
# which is Contribution 5 of Section 7.1 in one line:
pair_conv <- from_synthpop(syn_sp, original,
                           key_vars = keys_cmp, target_var = "education")
stopifnot(identical(dcap(pair_conv)$cap, dcap(pair_sp)$cap))

cmp <- data.frame(
  Quantity = c("riskutility: dcap()$cap", "synthpop: DCAP", "synthpop: CAPd"),
  Value = round(c(100 * dcap(pair_sp)$cap,
                  as.numeric(disc_sp$allCAPs[["DCAP"]]),
                  as.numeric(disc_sp$allCAPs[["CAPd"]])), 4)
)
```

```{r synthpop-comparison-tab, echo=FALSE}
knitr::kable(cmp, row.names = FALSE, col.names = c("Quantity", "Value (%)"),
             caption = "Mean CAP from riskutility and synthpop on identical data. The first two rows are the same estimand and agree to all printed digits.\\label{tab:synthpop}")
```

The first two rows agree to every printed digit, as they must: the two packages
implement the same estimand. `CAPd` differs because it measures attribution
*within* the original data rather than attribution of original records from the
synthetic distribution; reporting it alongside `dcap()` would compare two
different quantities. The frequency-based measures `kanonymity()` and
`ldiversity()` are defined identically to their \pkg{sdcMicro} counterparts.


## ML-Based Risk: RAPID {#sec-rapid}

Risk of Attribute Prediction-Induced Disclosure (\pkg{RAPID}) [@templ2026rapid] takes a
fundamentally different approach to attribute disclosure. Instead of matching
quasi-identifiers, \pkg{RAPID} trains a predictive model $\hat{f}$ on the released
data $(Y_{\mathcal{K}}, Y_s)$ and evaluates its predictions on the original
data: $\hat{s}_i = \hat{f}(X_{\mathcal{K},i})$. For numeric targets, a
record is *at risk* when the prediction error falls below a threshold
$\epsilon$: $e(s_i, \hat{s}_i) < \epsilon$, where $e(\cdot, \cdot)$ is
a configurable error metric (symmetric percentage error by default). The
\pkg{RAPID} score is the fraction of at-risk records:
$\text{RAPID} = n^{-1} \sum_i \mathbf{1}(e(s_i, \hat{s}_i) < \epsilon)$.
For categorical targets, a different evaluation applies: a record is at risk
when a gain or ratio score exceeds a threshold.

```{r rapid-demo, warning=FALSE, fig.cap="RAPID threshold-sensitivity diagnostic for the running example.\\label{fig:rapid}"}
rapid_res <- rapid(pair, model_type = "lm")
summary(rapid_res)

# The raw score is not interpretable on its own: a fixed error band captures a
# non-zero share of records by chance. rapid_test() supplies the permutation
# null against which the score must be read.
rapid_null <- rapid_test(original, synthetic,
                         quasi_identifiers = key_vars,
                         sensitive_attribute = target_var,
                         model_type = "lm", n_permutations = 199, seed = 42)
print(rapid_null)

plot(rapid_res, which = 3)
```

The raw score is `r round(rapid_res$rapid, 4)`, which `summary()` labels
`r summary(rapid_res)$risk_level`. Read on its own that is misleading: a fixed
$\pm 10\%$ error band captures a non-trivial share of records by chance
whatever the synthesiser does, and `rapid()` subtracts no baseline. The
permutation test puts the score in context --- `rapid_test()` returns
$p = `r round(rapid_null$p_value, 3)`$, so the observed score is not
distinguishable from the null obtained by permuting the target. This is why we
report the two together, and why the absolute bands below should be treated as
rough conventions rather than decision rules. Figure \ref{fig:rapid} shows how
the at-risk fraction changes across threshold values.

\pkg{RAPID} complements the CAP family in two ways. First, it captures non-linear
relationships and variable interactions. Second, it provides inferential tools:
`rapid_test()` computes a permutation-based $p$-value, `confint()` provides
bootstrap confidence intervals, and `rapid_threshold_select()` optimizes the
threshold in a data-driven manner.

```{r rapid-models, echo=FALSE}
models <- data.frame(
  Model = c("lm", "rf", "cart", "gbm", "logit"),
  Package = c("stats", "ranger", "rpart", "xgboost", "stats"),
  Numeric = c("Yes", "Yes", "Yes", "Yes", "No"),
  Categorical = c("No", "Yes", "Yes", "Yes", "Binary only"),
  Interactions = c("Manual", "Automatic", "Automatic", "Automatic", "Manual"),
  check.names = FALSE
)
knitr::kable(models, caption = "RAPID model backends.")
```

| \pkg{RAPID} Score | Risk Level | Interpretation |
|-------------|------------|----------------|
| < 0.05 | Low | ML model cannot predict target much better than baseline |
| 0.05--0.15 | Moderate | Some predictive signal from synthetic data |
| 0.15--0.30 | Elevated | Significant predictive leakage |
| > 0.30 | High | Strong evidence of disclosure risk |


## Distance-Based Risk {#sec-distance}

Distance-based methods detect *memorization*: the failure mode where a
generative model reproduces training records verbatim or near-verbatim. The
key idea is the **holdout method** --- split the original data into a
training set $T$ (used for synthesis) and a holdout set $H$ (unseen by the
generator). For each synthetic record $y_j$, compute the Distance to
Closest Record in $T$ ($d_T$) and in $H$ ($d_H$). If the generator has
generalized, $d_T$ and $d_H$ should be comparable. If it has memorized,
$d_T$ will be systematically smaller:

$$\text{DCR\_share} = n^{-1} \sum_j \mathbf{1}\bigl(d_T(y_j) < d_H(y_j)\bigr), \qquad \text{DCR\_ratio} = \frac{\bar{d}_T}{\bar{d}_H}$$

A DCR share meaningfully above 0.5 (the package flags shares above 0.55), or a
ratio below about 1, suggests memorization. The **NNDR** (Nearest Neighbor
Distance Ratio) provides a complementary view: for each synthetic record, it is
the ratio of the distance to its nearest neighbor over the distance to its
second-nearest neighbor. A ratio near 0 indicates a single
dominant match, hence a distinctive and potentially re-identifiable record;
near 1 indicates no distinctive match. `nndr()` reports the ratio of the fifth
percentile of the NNDR distribution against training to that against holdout,
together with the number of records below 0.1; values near 1 and few such
records indicate no memorization. **IMS** (Identical
Match Share) counts exact copies. When an explicit holdout is unavailable,
`holdout_fraction` automatically splits the original data.

```{r distance-demo, warning=FALSE, fig.cap="Distance to closest record (DCR): synthetic-to-training versus synthetic-to-holdout distances.\\label{fig:dcr}"}
dcr_res <- dcr(pair, holdout_fraction = 0.5)
summary(dcr_res)
plot(dcr_res, which = 1)
```

The printed verdict is `WARNING`, and it is not a finding. The observed share
is `r sprintf("%.1f%%", 100*dcr_res$dcr_share)` against a permutation null whose
own mean is `r round(dcr_res$null_distribution$mean, 3)` with a 95% interval of
[`r round(dcr_res$null_distribution$quantiles[["2.5%"]], 2)`,
`r round(dcr_res$null_distribution$quantiles[["97.5%"]], 2)`]; the
null-comparison $p$-value is
`r round(dcr_res$null_share_pvalue, 2)`. What tripped `privacy_pass` is the
paired Wilcoxon test ($p =$ `r signif(dcr_res$wilcox_test$p.value, 2)`), which
at $n = 500$ resolves a difference in mean Gower distance of
`r round(dcr_res$mean_dcr_train, 4)` against
`r round(dcr_res$mean_dcr_holdout, 4)`. On a synthesiser that copies nothing,
the absolute criterion fires and the null-based comparison does not.

**The DCR Delusion.** @yao2025dcr show that DCR can fail to
detect privacy leakage: datasets deemed "private" by DCR may still be
vulnerable to membership inference attacks. Their central recommendation is that
DCR be interpreted relative to a proper null distribution rather than in
absolute terms; `dcr()` implements exactly this, comparing the observed share
against a permutation null (`null_test`) and reporting a Wilcoxon test alongside
the point estimate. Even so, distance-based metrics should always be
complemented with other risk families.

```{r distance-table, echo=FALSE}
dist <- data.frame(
  Metric = c("DCR", "NNDR", "IMS", "RF proximity", "dRisk", "Hitting rate",
             "Epsilon ID", "Delta-presence"),
  Holdout = c("Yes", "Yes", "No", "Yes", "No", "No", "No", "No"),
  Detects = c("Memorization", "Memorization", "Exact copies",
              "Memorization (non-linear)", "Close records", "Close records",
              "Identifiability", "Membership bounds"),
  `Low risk` = c("share < 0.55", "ratio >= 0.8", "< 0.01",
                 "ratio near 1", "< 0.05", "< 0.05",
                 "< 0.01", "delta_max well below 1"),
  check.names = FALSE
)
knitr::kable(dist,
             caption = "Distance-based and proximity risk measures.\\label{tab:distance}")
```

**RF proximity** offers a data-adaptive alternative: it trains a random forest
to discriminate original from synthetic records and measures how often
synthetic records share terminal nodes with training vs. holdout records,
capturing non-linear proximity that fixed distance metrics may miss. Use
`rf_privacy()` when complex interactions are expected.


## Record Linkage Risk {#sec-recordlinkage}

The `recordLinkage()` function directly simulates a re-identification attack by
linking each original record to the most similar record(s) in the anonymized
dataset. Eight methods are implemented, spanning deterministic (Gower distance),
probabilistic (Fellegi-Sunter, @fellegi1969theory), PRAM-aware, predictive
(propensity score), random forest proximity, rank-based (RBRL,
@muralidhar2016rankbased), robust Mahalanobis [@templ2008robust], and
autoencoder embedding [@guo2016entity]. Three matching modes are available:
independent (many-to-one), bijective (one-to-one via Hungarian algorithm,
@herranz2016gdbrl), and optimal transport (Sinkhorn).

```{r recordlinkage-demo}
# Namespace-qualified: sdcMicro (>= 5.8.2) exports an unrelated
# recordLinkage() with a different signature, which masks this one whenever
# sdcMicro is attached afterwards.
rl_res <- riskutility::recordLinkage(pair, method = "deterministic")
print(rl_res)
```

The mean risk is the average probability mass the attacker places on the *true*
match, so it is directly comparable with the no-information baseline
$1/n = 0.002$: a value at or below the baseline means the linkage attack does
no better than guessing. Here the mean is `r round(rl_res$overall$mean_risk, 4)`. The
maximum, though, is `r round(rl_res$overall$max_risk, 3)`: one record is matched to its
true counterpart with that probability. Custodians act on the per-record
maximum, not on the mean, which is why `recordLinkage()` returns both. For the full set of methods and matching modes, see
`?recordLinkage` and the record-linkage vignette in the package repository.

```{r recordlinkage-table, echo=FALSE}
rl <- data.frame(
  Method = c("Deterministic", "Probabilistic", "PRAM", "Predictive",
             "RF", "RBRL", "Mahalanobis", "Embedding"),
  Distance = c("Gower", "Fellegi-Sunter", "Transition prob.", "Propensity",
               "RF proximity", "Rank-based", "Mahalanobis", "Autoencoder"),
  `Mixed types` = c("Yes", "Yes", "Categorical", "Yes",
                     "Yes", "Yes", "Numeric", "Yes"),
  check.names = FALSE
)
knitr::kable(rl,
             caption = "Record linkage methods. All eight support all three matching modes (independent, bijective, optimal transport).\\label{tab:rl}")
```


## Membership Inference and Anonymization Failure Criteria {#sec-membership}

This section groups two related but distinct concerns. The first three
measures (`mia_classifier()`, `domias()`, `nnaa()`) assess *membership
disclosure* --- whether a membership inference attack (MIA) can determine if a
specific individual's data was used during
synthesis. The singling out and linkability attacks operationalize two of the GDPR
anonymization criteria [@wp29anonymisation], following the attack-based
approach of @giomi2023anonymeter.

**NNAA** (Nearest Neighbor Adversarial Accuracy, @yale2020generation) is based on
the adversarial accuracy of a nearest-neighbour two-sample comparison,
$\text{AA}(A,S) = \tfrac{1}{2}\bigl[\Pr(d_{AS} > d_{AA}) + \Pr(d_{SA} > d_{SS})\bigr]$.
The reported privacy loss is $\text{AA}(\text{holdout}, S) - \text{AA}(\text{train}, S)$;
a positive value means synthetic records resemble training records more closely
than holdout records, indicating memorization:

```{r nnaa-demo}
nnaa_res <- nnaa(train_data, synthetic, holdout = holdout_data,
                 method = "gower", seed = 42)
print(nnaa_res)
```

| Privacy Loss | Interpretation |
|-------------|----------------|
| Near 0 | No detectable leakage (ideal) |
| 0.01--0.05 | Minor leakage, likely acceptable |
| > 0.10 | Significant memorization |

**Singling out** and **linkability** operationalize two of the Article 29
Working Party's three anonymization criteria (the third, inference, is
addressed by the attribution-based CAP and \pkg{RAPID} measures):

```{r membership-demo}
so_res <- singling_out(original, synthetic,
                       n_attacks = 500, n_cols = 3,
                       mode = "multivariate", seed = 42)
print(so_res)

link_res <- linkability(original, synthetic,
                        n_attacks = 500, n_neighbors = 1, seed = 42)
print(link_res)
```

Both attacks pass. Residual singling-out risk is
`r round(so_res$risk, 4)` with an upper 95% bound of
`r round(so_res$risk_ci[2], 4)`, and residual linkability
`r round(link_res$risk, 4)` with an upper bound of
`r round(link_res$risk_ci[2], 4)`. "Residual" is the key word: both are
control-corrected, i.e. the attack success rate against the original data minus
the rate achievable against a holdout the synthesiser never saw. The two
rates nearly coincide here, which is the same no-leakage signature as
TCAP at its baseline and a DCR share at one half. A data-protection officer
should be shown the upper confidence bounds rather than the point estimates.

```{r membership-table, echo=FALSE}
mia <- data.frame(
  Metric = c("MIA classifier", "DOMIAS", "NNAA",
             "Singling out", "Linkability", "delta-Presence"),
  `Attack type` = c("Shadow model", "Density overfitting",
                     "Nearest neighbor", "Predicate-based",
                     "Record linkage", "Membership bounds"),
  Holdout = c("Yes", "Yes", "Yes", "Yes", "Yes", "No"),
  `GDPR criterion` = c("--", "--", "--",
                        "Art. 29 WP", "Art. 29 WP", "--"),
  `Low risk` = c("< 0.55", "< 0.6", "< 0.05",
                 "< 0.1", "< 0.1", "> 0.5"),
  check.names = FALSE
)
knitr::kable(mia,
             caption = "Membership inference and GDPR measures.\\label{tab:membership}")
```


## Cross-Family Comparison {#sec-rosetta}

No single metric tells the full story. Applying all families to the same
dataset reveals complementary and sometimes contradictory information:

```{r rosetta}
# Near-copy: original + small noise (high risk expected)
set.seed(99)
near_copy <- original
near_copy$age <- near_copy$age + sample(-1:1, n, replace = TRUE)
near_copy$income <- near_copy$income + round(rnorm(n, 0, 500))
pair_risky <- synth_pair(original, near_copy,
                         key_vars = key_vars, target_var = target_var)

# Compare key metrics across the two datasets
comparison <- data.frame(
  Metric = c("DCAP", "RAPID (lm)", "IMS"),
  Safe = c(
    dcap(pair)$dcap,
    rapid(pair, model_type = "lm", verbose = FALSE)$rapid,
    ims(pair)$ims
  ),
  Risky = c(
    dcap(pair_risky)$dcap,
    rapid(pair_risky, model_type = "lm", verbose = FALSE)$rapid,
    ims(pair_risky)$ims
  )
)
comparison$Range <- c("(-1, 1), differential", "[0, 1]", "[0, 1]")
comparison$Safe <- sprintf("%.3f", comparison$Safe)
comparison$Risky <- sprintf("%.3f", comparison$Risky)
comparison <- comparison[, c("Metric", "Range", "Safe", "Risky")]
knitr::kable(comparison,
             caption = "Cross-family comparison: safe versus risky synthetic data.")
```

The near-copy shows elevated risk across all families, but the magnitude
and interpretation differ. Attribution measures quantify *information
leakage*; distance-based measures quantify *memorization*. These
complementary perspectives mean that a dataset can pass one family's tests
while failing another's --- a thorough evaluation uses at least one measure
from each family.


# Data Utility Measures {#sec-utility}

A dataset that passes all risk checks but destroys the analytical value of
the data is useless. Utility measures quantify how well the released data
preserves the statistical properties of the original.

## Global Utility: Propensity Scores {#sec-utility-quick}

Global utility measures give a single-number verdict by asking: *can a
classifier tell original and synthetic records apart?*

The propensity score method [@woo2009global; @snoke2018general] pools
original ($X$, $n_X$ records) and synthetic ($Y$, $n_Y$ records) data,
labels them (0/1), and fits a classifier. The **pMSE** (propensity score
Mean Squared Error) measures how well the model discriminates:

$$\text{pMSE} = \frac{1}{N} \sum_{i=1}^{N} \left(\hat{p}_i - c\right)^2$$

where $N = n_X + n_Y$ and $c = n_Y / N$. If original and synthetic records
are indistinguishable, pMSE $\approx 0$.

```{r utility-quick, warning=FALSE}
prop_res <- propscore(pair)
summary(prop_res)
```

| pMSE Value | Interpretation |
|------------|----------------|
| < 0.01 | Excellent fidelity |
| 0.01--0.05 | Good fidelity |
| 0.05--0.10 | Moderate differences |
| > 0.10 | Poor fidelity |

For the running example the pMSE is
`r round(prop_res$ps_score, 4)`. The synthesiser reproduces every marginal
exactly, so a classifier has only the *joint* structure to work with; the
resulting value sits in the *moderate differences* band rather than at either
extreme. Two caveats matter when reading this number in absolute terms. It
depends on the classifier (`propscore()` defaults to a random forest, which
attains a non-zero pMSE even for two samples from the same distribution), and
on $c$; the bands above are therefore conventions, not calibrated cut-offs. A
value should always be compared against the pMSE obtained for a synthesiser
known to be faithful, or against the standardized pMSE of @snoke2018general. Like `propscore()`,
every utility function below also accepts a `synth_pair` (illustrated by the
`tstr()` and `regression_fidelity()` calls); we pass explicit data frames where
it is clearer to highlight a variable subset.

## Univariate Diagnostics {#sec-utility-univariate}

When global utility is poor, per-variable measures identify which variables
are responsible. For **numeric** variables, the Wasserstein distance
measures the cost of transforming one distribution into another. For
**categorical** variables, the Hellinger distance measures distributional
overlap:

$$H(p, q) = \frac{1}{\sqrt{2}} \sqrt{\sum_{k=1}^{K} \left(\sqrt{p_k} - \sqrt{q_k}\right)^2}$$

```{r utility-univariate}
# Hellinger distance for categorical variables
h_res <- hellinger(original, synthetic, vars = c("sex", "education"))
print(h_res)

# CI proximity: confidence interval overlap for means
cip_res <- ci_proximity(original, synthetic, vars = c("age", "income"))
print(cip_res)
```

The **CI proximity** measure [@karr2006framework] compares confidence intervals
of summary statistics (means) between original and synthetic data. An overlap
near 1 means the intervals coincide; a relative error near 0 means point
estimates are close. The mean Hellinger distance is
`r round(h_res$hellinger_mean, 4)`: the categorical marginals are reproduced
almost exactly, as they must be, since each variable was resampled from its
own marginal. The numeric means are close too --- the mean relative error is
`r round(cip_res$relative_error_mean, 4)`. Yet the mean confidence-interval
overlap is only `r round(cip_res$overlap_mean, 4)`. That gap is not evidence of
broken dependence: a sample mean is a functional of a single margin and cannot
see the joint structure at all. At $n = 500$ the intervals are narrow enough
that the ordinary resampling drift in the point estimate already pulls them
apart. CI overlap is a strict criterion and should be read next to the relative
error, not instead of it. Neither measure says anything about what this
synthesiser actually destroyed --- both look at one variable at a time.


## Multivariate and Structural Utility {#sec-utility-structural}

Marginal distributions can match perfectly while joint distributions diverge.
The **energy distance** [@szekely2013energy] is a multivariate two-sample
statistic sensitive to differences in both location and scale (lower values
indicate closer joint distributions):

```{r utility-structural}
e_res <- energy_distance(original[, c("age", "income")],
                         synthetic[, c("age", "income")],
                         seed = 42)
print(e_res)
```

The **MMD** (Maximum Mean Discrepancy, @gretton2012kernel) provides a
kernel-based alternative supporting exact computation and random Fourier
features (RFF) for large datasets:

```{r mmd-demo}
mmd_res <- mmd(original[, c("age", "income")],
               synthetic[, c("age", "income")],
               kernel = "gaussian", method = "rff",
               n_features = 500, seed = 42)
print(mmd_res)
```

Both statistics remain small (energy distance
`r round(e_res$energy_distance, 4)`, $\text{MMD}^2$
`r round(mmd_res$mmd2, 4)`), even though the dependence between `age` and
`income` has been removed entirely. This is not a failure of the
implementations but a property of the statistics: on two standardized
variables whose marginals are identical, the contribution of the dependence
term to an energy or kernel distance is small relative to the marginal spread.
Energy distance and MMD are powerful against shifts in location, scale or
shape; they are comparatively blunt against a pure change in the dependence
structure. The measures below are the ones that see it.

**Copula fidelity** compares the empirical copula (rank dependence structure)
using the Cramér-von Mises statistic on pairwise copula CDFs. **Contingency
fidelity** [@snoke2018general] is its categorical complement, computing total
variation distance between bivariate contingency tables:

```{r fidelity-demo}
cop_res <- copula_fidelity(original, synthetic, vars = c("age", "income"))
print(cop_res)

ctf_res <- contingency_fidelity(original, synthetic,
                                vars = c("sex", "education", "region"))
print(ctf_res)
```

One caution before reading these numbers together. Each function prints a line
called `Utility score`, but the transformations differ --- $1-\bar H$ for
Hellinger, $\exp(-E)$ for the energy distance, $1/(1+100\,\overline{\text{CvM}})$
for copula fidelity, $1-\overline{\text{TV}}$ for contingency fidelity. They are
comparable *within* a measure across datasets, not *across* measures. Putting
several measures on one scale is exactly what `rumap()` does in
Section \ref{sec-comprehensive}.

Copula fidelity registers the change: the mean Cramér--von Mises distance
rises to `r round(cop_res$mean_cvm, 5)`, giving a utility score of
`r round(cop_res$utility_score, 4)`. Contingency fidelity is affected less
(`r round(ctf_res$utility_score, 4)`), because total variation distance on a
3 $\times$ 5 table is a coarse instrument for a dependence that is real but
moderate. Reading the two together --- rank dependence clearly degraded,
categorical cross-tabulation only mildly so --- is more informative than
either number alone.


## Predictive Utility {#sec-utility-predictive}

**TSTR** (Train on Synthetic, Test on Real, @zhao2021ctgan) trains a
predictive model on the synthetic data and evaluates performance on held-out
real data. The ratio of TSTR-to-TRTR performance quantifies how well
predictive relationships are preserved:

```{r tstr-demo, warning=FALSE, eval=requireNamespace("ranger", quietly=TRUE)}
set.seed(42)
tstr_res <- tstr(pair, target_var = "income", model = "rf",
                 test_fraction = 0.3, seed = 42)
print(tstr_res)
```

The ratio is `r round(tstr_res$tstr_ratio, 4)`: a model trained on the
synthetic data carries essentially none of the original's predictive
structure across to real test data, which is exactly what independent
resampling of the marginals should produce. TSTR is the sharpest of the
utility measures here precisely because it is the only one that asks a
*conditional* question. Note that the ratio is interpretable only when the
train-on-real baseline itself has predictive power; `tstr()` returns `NA` with
a warning when it does not, rather than reporting the quotient of two
near-zero quantities.

**Regression fidelity** [@karr2006framework] fits the same regression model on
both datasets and compares coefficient estimates via CI overlap, standardized
bias, and significance agreement:

```{r regression-demo, fig.cap="Regression-coefficient fidelity: original versus synthetic estimates with confidence intervals.\\label{fig:reg}"}
reg_res <- regression_fidelity(pair,
                               formula = income ~ age + sex + education)
summary(reg_res)
plot(reg_res, which = 1)
```

The coefficients do not survive. On the original data the model puts
`r fmt_coef(reg_res, "age", "orig")`
on each year of age and `r fmt_coef(reg_res, "Tertiary", "orig")`
on a tertiary degree; fitted on the synthetic data the same two coefficients are
`r fmt_coef(reg_res, "age", "synth")` and
`r fmt_coef(reg_res, "Tertiary", "synth")`.
The mean confidence-interval overlap is
`r round(reg_res$mean_ci_overlap, 3)`, only
`r round(100*reg_res$sig_agreement_rate)`% of terms agree on significance, and
the mean absolute standardized bias is
`r round(reg_res$mean_abs_std_bias, 1)`. The synthetic regression
has collapsed to an intercept. This is the same dataset whose categorical
marginals scored `r round(h_res$utility_score, 3)` two subsections ago, which
is the entire argument of this section in one comparison.

**Tail fidelity** assesses how well extreme values are preserved --- critical
for applications where tail behavior matters (financial risk, rare diseases):

```{r tail-demo}
tail_res <- tail_fidelity(original, synthetic, vars = c("age", "income"),
                          percentile = 95, tails = "both")
print(tail_res)
```

**Subgroup utility** [@snoke2018general] applies any utility measure to each
subgroup defined by a grouping variable, identifying groups with low utility:

```{r subgroup-demo}
su_res <- subgroup_utility(original, synthetic, group_var = "region",
                           utility_fun = energy_distance,
                           threshold = 0.5, seed = 42)
print(su_res)
```

The conservative `utility_score` is the worst subgroup score. A `ratio` near
1 indicates homogeneous utility; below 0.5 indicates substantial disparity.

```{r table7-utility, echo=FALSE}
util <- data.frame(
  `Use case` = c(rep("Quick assessment", 2),
                 rep("Univariate", 3),
                 rep("Multivariate", 4),
                 rep("Predictive", 3),
                 "Subgroup"),
  Function = c("propscore()", "specks()",
               "compare_wasserstein()", "hellinger()", "ci_proximity()",
               "energy_distance()", "mmd()",
               "copula_fidelity()", "contingency_fidelity()",
               "tstr()", "regression_fidelity()",
               "compare_feature_importance()",
               "subgroup_utility()"),
  `Data type` = c("Mixed", "Mixed",
                   "Numeric", "Categorical", "Numeric",
                   "Numeric", "Numeric",
                   "Numeric", "Categorical",
                   "Mixed", "Mixed", "Mixed",
                   "Mixed"),
  Interpretation = c("< 0.1: good", "< 0.05: good",
                     "Lower = better", "< 0.1: good", "> 0.8: good",
                     "Lower = better", "Lower = better",
                     "< 0.1: good", "< 0.05: good",
                     "ratio near 1: good", "overlap > 0.8: good",
                     "High corr: good",
                     "min > 0.5: good"),
  check.names = FALSE
)
knitr::kable(util,
             caption = "Utility measures by use case.\\label{tab:utility}")
```


# Comprehensive Assessment: A Case Study {#sec-comprehensive}

This section demonstrates the complete practitioner workflow on a realistic
dataset, comparing three synthesis approaches with different privacy-utility
trade-offs. Figure \ref{fig:workflow} summarizes this iterative process.

```{r fig-workflow, echo=FALSE, out.width="95%", fig.cap="The iterative risk-utility evaluation workflow.\\label{fig:workflow}"}
render_dot('
digraph workflow {
  graph [rankdir=LR, fontsize=10];
  node [shape=box, style="rounded,filled", fillcolor="#EAF2F8", fontname="Helvetica", fontsize=9];
  edge [fontname="Helvetica", fontsize=8];
  gen  [label="Generate /\nanonymize data"];
  risk [label="Assess risk\n(disclosure_report())"];
  util [label="Assess utility"];
  dec  [label="Acceptable\ntrade-off?", shape=diamond, fillcolor="#D4E6F1"];
  rel  [label="Release", fillcolor="#D5F5E3"];
  ref  [label="Refine synthesis /\nprotection parameters", fillcolor="#FCF3CF"];
  cmp  [label="Compare candidates\n(rumap())", fillcolor="#FCF3CF"];
  gen -> risk -> util -> dec;
  dec -> rel [label="yes"];
  dec -> ref [label="no"];
  ref -> cmp -> gen;
}
')
```

## Scenario and Data {#sec-case-data}

Consider a statistical agency that wants to release a survey dataset ($n = 1000$)
containing demographic variables (age, sex, education, region) and a sensitive
income variable.

```{r case-data}
set.seed(123)
N <- 1000
edu_levels <- c("Primary", "Secondary", "Tertiary")
age_groups <- c("20-29", "30-39", "40-49", "50-59", "60-69")
orig <- data.frame(
  age_group = factor(sample(age_groups, N, replace = TRUE)),
  sex = factor(sample(c("M", "F"), N, replace = TRUE)),
  education = factor(sample(edu_levels, N, replace = TRUE,
                            prob = c(0.25, 0.50, 0.25))),
  region = factor(sample(paste0("R", 1:4), N, replace = TRUE))
)
edu_effect <- c(Primary = 0, Secondary = 0.3, Tertiary = 0.7)
age_effect <- c("20-29" = 0, "30-39" = 0.15, "40-49" = 0.3,
                "50-59" = 0.4, "60-69" = 0.35)
orig$income <- round(exp(
  10 + age_effect[as.character(orig$age_group)] +
    edu_effect[as.character(orig$education)] + rnorm(N, 0, 0.4)
))

qi <- c("age_group", "sex", "education", "region")
sens <- "income"
```

We create three synthetic datasets spanning the privacy-utility spectrum:

```{r case-synthesis}
set.seed(456)

# Method A: Independent marginals (safest, but destroys correlations)
synA <- data.frame(
  age_group = factor(sample(age_groups, N, replace = TRUE)),
  sex = factor(sample(c("M", "F"), N, replace = TRUE)),
  education = factor(sample(edu_levels, N, replace = TRUE,
                            prob = c(0.25, 0.50, 0.25))),
  region = factor(sample(paste0("R", 1:4), N, replace = TRUE)),
  income = sample(orig$income, N, replace = TRUE)
)

# Method B: Category-preserving bootstrap with income noise
idx_B <- sample(N, N, replace = TRUE)
synB <- orig[idx_B, ]
rownames(synB) <- NULL
synB$income <- round(synB$income * exp(rnorm(N, 0, 0.15)))
swap_idx <- sample(N, round(0.2 * N))
synB$age_group[swap_idx] <- factor(sample(age_groups,
                                          length(swap_idx), replace = TRUE))

# Method C: Near-copy with minimal perturbation (risky)
synC <- orig
synC$income <- round(synC$income * exp(rnorm(N, 0, 0.03)))
```

## Step 1: Quick Risk Screening {#sec-case-report}

The `disclosure_report()` function computes multiple risk measures, evaluates
each against a threshold, and produces a pass/fail assessment:

```{r case-report, warning=FALSE}
pair_A <- synth_pair(orig, synA, key_vars = qi, target_var = sens)
pair_B <- synth_pair(orig, synB, key_vars = qi, target_var = sens)
pair_C <- synth_pair(orig, synC, key_vars = qi, target_var = sens)

rep_A <- disclosure_report(pair_A, compute = c("attribution", "privacy"),
                           seed = 42, verbose = FALSE)
rep_B <- disclosure_report(pair_B, compute = c("attribution", "privacy"),
                           seed = 42, verbose = FALSE)
rep_C <- disclosure_report(pair_C, compute = c("attribution", "privacy"),
                           seed = 42, verbose = FALSE)

# The report's own print method is the practitioner-facing artefact:
print(rep_A)

verdicts <- data.frame(
  Method = c("A: Independent", "B: Bootstrap+noise", "C: Near-copy"),
  Overall = c(rep_A$overall_risk, rep_B$overall_risk, rep_C$overall_risk),
  Pass = c(rep_A$n_pass, rep_B$n_pass, rep_C$n_pass),
  Warn = c(rep_A$n_warn, rep_B$n_warn, rep_C$n_warn)
)
```

```{r case-report-tab, echo=FALSE}
knitr::kable(verdicts,
             caption = "Quick risk screening across the three methods.\\label{tab:verdicts}")
```

Table \ref{tab:verdicts} reveals two patterns. First, attribution metrics differentiate the three methods
in the expected order: Method A (independent) has the lowest attribution risk,
Method C (near-copy) the highest. Second, privacy models flag all three methods
because they evaluate the released data's structure alone --- with 120 possible
QI combinations and $n = 1000$ records, some equivalence classes are small
regardless of synthesis method. This illustrates a key lesson: privacy models
and attribution metrics answer different questions and should be interpreted
together.

## Step 2: Comparative Assessment with rumap() {#sec-case-rumap}

The `rumap()` function implements the multivariate Risk-Utility framework of
@thees2026beyond. Traditional R-U analysis plots a single risk measure against
a single utility measure, producing a two-dimensional trade-off curve. This
can be misleading: a method may appear optimal on one pair of measures while
performing poorly on another. `rumap()` computes multiple risk and utility
measures simultaneously, normalizes to $[0, 1]$, and identifies
**Pareto-optimal** methods.

```{r case-rumap, warning=FALSE}
set.seed(42)
ru <- rumap(orig,
            list("A: Independent" = synA,
                 "B: Bootstrap+noise" = synB,
                 "C: Near-copy" = synC),
            risk_measures = c("dcap", "tcap", "ims"),
            utility_measures = c("pmse", "wasserstein"),
            key_vars = qi, target_var = sens,
            seed = 42)
print(ru)
```

```{r case-rumap-scatter, fig.height=4.2, fig.cap="Risk-Utility map of the three synthesizers; the Pareto-optimal front is highlighted.\\label{fig:rumap-scatter}"}
plot(ru, which = 1)  # R-U scatterplot with Pareto front
```

Figure \ref{fig:rumap-scatter} places each method in the composite
risk-utility plane; methods towards the lower right (low risk, high utility)
are preferred. On this replicate the front consists of Methods A and C. The
error bars are $\pm 1$ standard deviation across the constituent measures, and
they overlap for all three methods --- a first indication that the composite
positions are less sharply separated than the point symbols suggest.

```{r case-rumap-heatmap, fig.height=3.6, fig.cap="Heatmap of normalized risk and utility measures across the three synthesizers.\\label{fig:rumap-heatmap}"}
plot(ru, which = 2)  # Heatmap of individual measures
```

Figure \ref{fig:rumap-heatmap} shows the component measures whose averages
produce those composites, and it repays careful reading. DCAP and TCAP order
the methods as one would expect --- A lowest, C highest. **IMS runs the other
way.** Its raw values are `r round(ru$risk$ims[1], 3)`, `r round(ru$risk$ims[2], 3)`
and `r round(ru$risk$ims[3], 3)` for A, B and C: the independent-marginals
synthesiser produces the *most* exact matches and the near-copy the *fewest*.
The reason is mechanical. Method A resamples genuine income values into only
`r length(unique(interaction(orig[, qi])))` occupied quasi-identifier cells, so
coincidental exact duplicates arise; Method C perturbs every income
continuously, so no record is reproduced exactly. IMS is blind to near-copies,
which is precisely the disclosure it looks designed to catch.

That single column drives the composite. After min-max normalization Method A
scores 1 on `ims` and 0 on both attribution measures, so its entire composite
risk of `r round(ru$composites$risk_mean[1], 3)` comes from the inverted
measure, while Method C's is pulled *down* by it. The differences being
stretched across the full $[0,1]$ range are `r round(1000*ru$risk$ims[1])`,
`r round(1000*ru$risk$ims[2])` and `r round(1000*ru$risk$ims[3])` matching
records out of 1000.

Three cautions therefore apply to the composites. The normalization is min-max
*across the compared methods*, so every column contains a 0 and a 1 by
construction: the composites rank the candidates, they do not measure them, and
adding or removing a candidate changes them. When a measure's raw range is
degenerate, as `ims` is here, that rescaling manufactures contrast out of
sampling noise. And with three candidates the internal-consistency
coefficients reported by `summary()` rest on three points and should not be
interpreted.

The practical consequence is that Pareto membership computed from composites is
itself a random quantity. Repeating the whole exercise over independent draws
of the data and the three synthesizers makes this concrete:

```{r case-stability, warning=FALSE}
make_case <- function(seed_data, seed_syn, N = 1000) {
  set.seed(seed_data)
  d <- data.frame(
    age_group = factor(sample(age_groups, N, replace = TRUE)),
    sex       = factor(sample(c("M", "F"), N, replace = TRUE)),
    education = factor(sample(edu_levels, N, replace = TRUE,
                              prob = c(0.25, 0.50, 0.25))),
    region    = factor(sample(paste0("R", 1:4), N, replace = TRUE)))
  d$income <- round(exp(10 + age_effect[as.character(d$age_group)] +
                          edu_effect[as.character(d$education)] +
                          rnorm(N, 0, 0.4)))
  set.seed(seed_syn)
  a <- as.data.frame(lapply(d, function(v) sample(v, N, replace = TRUE)))
  b <- d[sample(N, N, replace = TRUE), ]; rownames(b) <- NULL
  b$income <- round(b$income * exp(rnorm(N, 0, 0.15)))
  cc <- d; cc$income <- round(cc$income * exp(rnorm(N, 0, 0.03)))
  list(orig = d, A = a, B = b, C = cc)
}

seed_pairs <- list(c(123, 456), c(1, 2), c(11, 22), c(7, 8), c(99, 100),
                   c(2024, 2025), c(31, 41), c(5, 15), c(77, 88), c(300, 400))
front <- t(vapply(seed_pairs, function(sp) {
  cs <- make_case(sp[1], sp[2])
  set.seed(42)
  r <- rumap(cs$orig, list(A = cs$A, B = cs$B, C = cs$C),
             risk_measures = c("dcap", "tcap", "ims"),
             utility_measures = c("pmse", "wasserstein"),
             key_vars = qi, target_var = sens, seed = 42)
  r$pareto
}, logical(3)))
colnames(front) <- c("A", "B", "C")
round(100 * colMeans(front))
```

Across `r length(seed_pairs)` replicates Method C is Pareto-optimal every
time, while A and B move on and off the front. The dominance of Method B
visible in Figure \ref{fig:rumap-scatter} is therefore a property of this
particular draw, not of the three synthesis strategies. We report it because it
is the honest reading of the tool, and because it identifies what a
composite-based R-U map still needs: a notion of dominance that carries its own
uncertainty, such as a bootstrap Pareto front [@thees2026beyond].

## Step 3: Decision {#sec-case-decision}

The analysis supports a structured decision:

- **Method A** is the appropriate choice for public release. Its attribution
  risk is the lowest of the three on both DCAP and TCAP, and those are the
  measures that bear on the threat that matters for an open release. Its high
  `ims` value should be checked rather than ignored --- but the check is
  cheap, and here it resolves to `r round(1000*ru$risk$ims[1])` coincidental
  matches on a coarse quasi-identifier grid, not memorization.
- **Method C** delivers the best utility and is Pareto-optimal in every
  replicate, but it carries the highest attribution risk and its records stay
  close to the originals. It is defensible only inside a controlled-access
  environment with a data-use agreement.
- **Method B** cannot be separated from Method A on this evidence. It is
  dominated in this replicate and Pareto-optimal in most others, so the
  composite map does not support a decision between them. Choosing between A
  and B requires either more replicates or a measure that targets the specific
  threat of concern.

The wider lesson is the one the R-U literature tends to skip: a point on a
risk-utility map is an estimate, and a decision that turns on a gap of 0.01
between two such points is not supported by the data behind them.

This iterative workflow --- screen with `disclosure_report()`, compare with
`rumap()`, and refine synthesis parameters --- is the core use case that
\pkg{riskutility} is designed to support.


# Summary and Discussion {#sec-discussion}

## Contributions

The \pkg{riskutility} package makes five contributions to the R ecosystem for
statistical disclosure control:

1. **Comprehensive coverage.** To our knowledge, it is the first \proglang{R} package to
   unify all six risk-assessment paradigms --- frequency-based privacy models,
   attribution (CAP), ML-based (\pkg{RAPID}), distance-based, record linkage, and
   membership inference --- under a single API; the most comprehensive
   comparable evaluation suites (the \proglang{Python} libraries \pkg{synthcity} and \pkg{SynthEval})
   sit outside R.

2. **Novel implementations.** To our knowledge, \pkg{riskutility} provides the first
   implementations on CRAN of the singling-out and linkability attacks
   operationalizing two of the Article 29 Working Party anonymization criteria
   [@wp29anonymisation; @giomi2023anonymeter]; of $t$-closeness; of DOMIAS
   density-ratio membership inference [@hu2023domias]; and of eight-method
   record linkage with bijective and optimal-transport matching. \pkg{RAPID} was
   first released in the authors' standalone \pkg{RAPID} package
   [@templ2026rapid]; \pkg{riskutility} supersedes it, adding the permutation test,
   confidence intervals and data-driven threshold selection used in
   Section \ref{sec-rapid}.

3. **Unified API.** The `synth_pair` container and consistent S3 class
   pattern eliminate parameter repetition and ensure practitioners can switch
   between risk measures without learning new interfaces.

4. **Multivariate R-U mapping.** The `rumap()` function implements, as
   software, the multivariate Risk-Utility methodology of @thees2026beyond
   (under review at the *Journal of Official Statistics*) for comparing multiple
   synthesis approaches on many risk and utility dimensions simultaneously. The
   methodological framework and its statistical properties are the subject of
   that companion paper; here `rumap()` is one component of the software.

5. **Ecosystem integration.** The `from_sdcMicro()`, `from_synthpop()`, and
   `from_simPop()` constructors allow practitioners to evaluate data produced
   by any of the three main \proglang{R} packages.

## Partially vs Fully Synthetic Data

The privacy-utility evaluation differs depending on the synthesis approach
[@drechsler2011synthetic]:

- **Fully synthetic data**: All records are synthetic; all risk metrics are
  directly applicable.
- **Partially synthetic data**: Only sensitive values are replaced.
  Attribution metrics (DCAP, \pkg{RAPID}) are particularly relevant because original
  quasi-identifiers provide a direct matching key. Distance-based metrics
  should focus on the synthesized variables.
- **Multiple synthetic datasets**: When $m > 1$ datasets are generated, evaluate
  each separately and report the worst-case risk across all $m$ releases.

## Remediation

If disclosure risk is too high:

1. Add more noise to the synthesis process
2. Reduce granularity of quasi-identifiers (e.g., age groups instead of exact age)
3. Apply additional anonymization techniques (suppression, generalization)
4. Re-synthesize with different model settings
5. Re-evaluate with **\pkg{riskutility}** --- iterate until risk-utility balance is acceptable

If utility is too low:

1. Use a more flexible synthesizer (CART, Bayesian network, GAN)
2. Reduce the amount of perturbation
3. Identify affected variables/subgroups (`subgroup_utility()`, per-variable Hellinger)
4. Consider partially synthetic data (synthesize only sensitive variables)

## Limitations and Recommendations

**No formal privacy guarantees.** All measures provide empirical risk
assessment. A low DCAP score does not prove that no attacker can succeed.
Empirical and formal approaches (differential privacy) are complementary.

**Key variable selection.** Results depend heavily on the choice of
quasi-identifiers. Practitioners should base QI selection on a realistic
threat model, not on convenience.

**Threshold interpretation.** The pass/fail thresholds used by
`disclosure_report()` are pragmatic defaults. Different contexts require
different thresholds.

We recommend the following minimal evaluation protocol:

1. **Always run** `disclosure_report()` with `compute = "all"` as a first
   screening.
2. **For publication-quality assessment**, use `rumap()` to compare multiple
   synthesis approaches and identify Pareto-optimal methods.
3. **For regulatory compliance** (GDPR), include singling out and
   linkability tests.
4. **Interpret distance-based metrics cautiously.** Following @yao2025dcr,
   do not rely on DCR/NNDR alone.

## Future Work

Four extensions are planned: (i) a Shiny dashboard for interactive evaluation;
(ii) integration with differential privacy frameworks;
(iii) computational optimizations for large datasets ($n > 50{,}000$); and
(iv) population-level risk estimation from sample data.


# Computational Details {#sec-computational}

All computations were performed using R `r getRversion()` [@R2025] with the
\pkg{riskutility} package version `r packageVersion("riskutility")`. Core
dependencies include \pkg{data.table} [@pkg_datatable] for efficient data
manipulation, \pkg{ggplot2} [@pkg_ggplot2] for all visualizations, \pkg{VIM}
[@pkg_vim] for Gower distances and \pkg{ranger} [@pkg_ranger] for the default
random-forest backend. The optional backends \pkg{rpart} [@pkg_rpart],
\pkg{xgboost} [@pkg_xgboost], \pkg{caret} [@pkg_caret] and \pkg{clue}
[@pkg_clue] are loaded conditionally via `requireNamespace()`. The package is
available from CRAN at \url{https://CRAN.R-project.org/package=riskutility}.
Table \ref{tab:benchmark} reports measured runtimes for representative metrics.
All code shown in this paper is available as a single replication script,
`riskutility-jss-replication.R`, generated from this manuscript with
`knitr::purl()` and distributed alongside it.

```{r scalability-benchmark, echo=FALSE}
bench_fun <- function(f, p) {
  round(median(replicate(3, system.time(f(p))[["elapsed"]])), 3)
}
mk_pair <- function(n) {
  set.seed(1)
  d1 <- data.frame(
    age = sample(18:85, n, TRUE),
    sex = factor(sample(c("M", "F"), n, TRUE)),
    education = factor(sample(c("Primary", "Secondary", "Tertiary"), n, TRUE)),
    region = factor(sample(paste0("R", 1:5), n, TRUE)),
    income = round(rlnorm(n, log(40000), 0.5)))
  d2 <- d1
  d2$age <- sample(18:85, n, TRUE)
  d2$income <- round(rlnorm(n, log(40000), 0.5))
  synth_pair(d1, d2, key_vars = c("age", "sex", "education", "region"),
             target_var = "income")
}
fns <- list("dcap()" = function(p) dcap(p),
            "tcap()" = function(p) tcap(p),
            "kanonymity()" = function(p) kanonymity(p),
            "ims()" = function(p) ims(p))
sizes <- c(1000, 5000, 20000)
timing <- sapply(sizes, function(n) {
  p <- mk_pair(n)
  vapply(fns, bench_fun, numeric(1), p = p)
})
timing <- data.frame(Metric = names(fns), timing, check.names = FALSE)
colnames(timing)[-1] <- paste0("n = ", format(sizes, big.mark = ","), " (s)")
knitr::kable(timing, row.names = FALSE,
  caption = paste0("Median elapsed time over three runs, measured on ",
                   R.version$platform, " (R ", getRversion(),
                   "). The attribution, frequency, and exact-match metrics scale near-linearly.\\label{tab:benchmark}"))
```

Quadratic-time measures (`dcr()`, `energy_distance()`, and exact `mmd()`) scale
as $O(n^2)$ in memory and time; for large $n$ we recommend the
random-Fourier-feature variant `mmd(method = "rff")` or sub-sampling. The
multivariate `rumap()` runtime is the sum of its component measures.

```{r session-info, echo=FALSE, results="asis"}
si <- sessionInfo()
cat(sprintf(paste0("All results were obtained using %s on %s, with the ",
                   "\\pkg{riskutility} package version %s, ",
                   "\\pkg{data.table} %s, \\pkg{ggplot2} %s, ",
                   "\\pkg{ranger} %s and \\pkg{synthpop} %s. ",
                   "\\proglang{R} and all packages used are available from ",
                   "CRAN at https://CRAN.R-project.org/.\n"),
            si$R.version$version.string, si$running,
            packageVersion("riskutility"), packageVersion("data.table"),
            packageVersion("ggplot2"), packageVersion("ranger"),
            packageVersion("synthpop")))
```


# References {-}
